# Introduction {#intro}

- R Project on local directory. 
- Get a repo at github which is connected to pages, only 1 for free. 
- Only this can turn the branch into docs?
- usethis in R:

- Is it ok to commit them? 1 Absolutely (always different?)

<https://jules32.github.io/bookdown-tutorial/setup.html>

not working

git remote add origin <paste your copied awesome-book github url here>

git push --set-upstream origin master

what was recommended from a new git repo:

git remote add origin https://github.com/MarcoKuehne/awesome-book.git

what I can try

git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.io

already exists


https://marcokuehne.github.io/

next: git branch -M main

next: git push -u origin main

perhaps delete everything in my github pages repo first

https://stackoverflow.com/questions/1221840/remote-origin-already-exists-on-git-push-to-a-new-repository

git remote add origin https://github.com/MarcoKuehne/awesome-book.git

git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.git

git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.io

- look at: `git remote -v show`
- delete repo: `git remote rm origin`
- copy paste into terminal with:  CTRL + SHIFT + V 
- create git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.git 
- push: git push -u origin main


This tutorial worked to push a book to a standard repo

https://jules32.github.io/bookdown-tutorial/setup.html

but it means that i cannot be displayed since there is no doc branch?!!?!? 




You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

  <script type="text/javascript" src="https://ssl.gstatic.com/trends_nrtr/2578_RC01/embed_loader.js"></script>
  <script type="text/javascript">
    trends.embed.renderExploreWidget("TIMESERIES", {"comparisonItem":[{"keyword":"homeoffice","geo":"DE","time":"today 5-y"}],"category":0,"property":""}, {"exploreQuery":"date=today%205-y&geo=DE&q=homeoffice","guestPath":"https://trends.google.de:443/trends/embed/"});
  </script>
  


