--- 
title: "A Minimal Book Example"
author: "Yihui Xie"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
description: This is a minimal example of using the bookdown package to write a book.
  The output format for this example is bookdown::gitbook.
---

# Prerequisites

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Introduction {#intro}

- I need a R project on my local system. 
- Get a repo at github which is connected to pages, only 1 for free. 
- Only this can turn the branch into docs?
- usethis in R:

In R consol `usethis::use_git()` Always gives three different answers. 

Is it ok to commit them? Confirm.

# Git commands

Use the terminal inside RStudio. Copy paste into terminal with:  CTRL + SHIFT + V 

## git init

Use git init to initialize the repository. It is used to create a new empty repository or directory consisting of files' with the hidden directory. '.git' is created at the top level of your project, which places all of the revision information in one place.

Don't know what initiliaze means.

`git init`

I would like to have something like `git remote add ...` and `git push ...`, not working.

Show what is connected with `git remote -v show`

I can add and remove origins, dont know what it means: `git remote rm origin`

Use git add . in your bash to add all the files to the given folder.

Use git status in your bash to view all the files which are going to be staged to the first commit.

## git remote add

Create **git remote add**

git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.git 
git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.io
git remote add origin https://github.com/MarcoKuehne/marcokuehne.github.io.git

git remote add git@github.com:<username>/<repository-name>.git
git remote add git@github.com:MarcoKuehne/marcokuehne.github.git
git remote add git@github.com:MarcoKuehne/marcokuehne.github.io does not appear to be a repository 
git remote add git@github.com:MarcoKuehne/marcokuehne.github.io.git combine both 

## git push 

**Push means upload** something

git push -u origin main
git push -u origin master
git push origin master
git push origin main 
git push --set-upstream origin main
git push -f origin main # worked somehow!!!

If positive, it asks for github credentials. 

# Resources

<https://jules32.github.io/bookdown-tutorial/setup.html>

<https://stackoverflow.com/questions/1221840/remote-origin-already-exists-on-git-push-to-a-new-repository>

<https://jules32.github.io/bookdown-tutorial/setup.html>

This tutorial worked to push a book to a standard repo. But it means that i cannot be displayed since there is no doc branch?!!?!? 

<https://www.datacamp.com/community/tutorials/git-push-pull>


# Something else


You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

  <script type="text/javascript" src="https://ssl.gstatic.com/trends_nrtr/2578_RC01/embed_loader.js"></script>
  <script type="text/javascript">
    trends.embed.renderExploreWidget("TIMESERIES", {"comparisonItem":[{"keyword":"homeoffice","geo":"DE","time":"today 5-y"}],"category":0,"property":""}, {"exploreQuery":"date=today%205-y&geo=DE&q=homeoffice","guestPath":"https://trends.google.de:443/trends/embed/"});
  </script>
  




