<!-- # (PART) Data Analysis {-}  -->

# Intro to R {-}
<!-- Comments start with one or multiple hash tags.  -->
<!-- http://seminars-in-applied-economics.de/wp-content/uploads/2020/08/SOEP_Practice_1.txt -->

An .R file is a text document. Open (and edit) it with a any text editor (or RStudio).^[Here is a minimal R script <https://github.com/MarcoKuehne/marcokuehne.github.io/blob/main/data/intro-r.R>] 

<!-- You can collapse sections in RStudio. -->
<!-- You can navigate between sections in RStudio. -->
<!-- Check out your four panels and what they contain. -->
<!-- You can run/execute code in the console.  -->
<!-- Type or copy commands in the console and hit enter.  -->
<!-- Alternatively click 'run' in RStudio. -->
<!-- Alternatively use shortcuts like STRG + ENTER (run 1 line) -->

## Watching paint dry {-}

<!-- ```{r, include=FALSE} -->
<!-- tutorial::go_interactive(greedy=FALSE) -->
<!-- ``` -->

<!-- By default, `tutorial` will convert all R chunks. -->

R is a calculator. Use this R demo in the browser to explore basic features of R. Commands in the  **script.R** tab are executed by the **Run** bottom. It runs the entire script and prints out results in the **R Console**. This procedure is simplified but similar to a more complex integrated developer environment (IDE) like RStudio.

<!-- <script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script> -->
<script type="text/javascript" src="//cdn.datacamp.com/dcl-react.js.gz"></script>

Copy and paste (or modify) commands from lower sections into the script. 

::: {.challenge}
Try `1:20` and hit Run.
:::

```{r tut=TRUE}
# Comment: R as a calculator is a boring as watching paint dry.
21*2 
```

## When paint becomes art {-}

Copy paste the below code into **script.R** and hit **Run**. It opens a new tab **Plots**. 

```{r tut=FALSE, collapse=TRUE, eval=F} 
plot(x = iris$Sepal.Length, y = iris$Petal.Length,  # x and y variables
     col = iris$Species,                            # color by species! 
     pch = 16, cex = 2,                             # aesthetic parameters 
     xlab = "Sepal Length", ylab = "Petal Length",  # axis labels
     main = "Flower characteristics in Iris")       # title

 # Adding grid: 
grid()

# Adding a legend: 
legend (x = 4.5, y = 7, legend = levels(iris$Species), 
        pch = 16, col = c(1:3))
```

<iframe width='100%' height='300' src='https://rdrr.io/snippets/embed/' frameborder='0'></iframe>

<!-- Help, check the note of the pie chart function ?pie -->

## Define objects {-}

Define R objects for later use. Objects are **case-sensitive** (`X` is different from `x`). Objects can take any name, but its best to use something that makes sense to you, and will likely make sense to others who may read your code.

### Numeric Variables {-}
```{r tut=FALSE, collapse=TRUE} 
a <- 2
b = 3
a + b
```

### Logical Variables {-}

```{r prompt=TRUE}
harvard               <-        TRUE # spacing doesn't matter
yale      <- FALSE
princeton <- F # short for FALSE

# Attention: FALSE=0, TRUE=1
harvard + 1
```

### String Variables {-}

```{r }
emily  <- 'She is a friend.'     # string / character class / plain text
libby  <- "she is a coworker"    # use ' and " interchangeably
other  <- "people"               # prefer "
``` 

### Factor Variables {-}

A factor is an ordered categorical variable. `c()` is a generic function which combines its arguments. 

```{r tut=FALSE}
fruit  <- as.factor(c("banana", "apple")  ) # The default ordering is alphabetic
fruit

dose <- as.factor(c("low", "medium", "high")  ) # The default ordering is alphabetic
dose
```

### Combine objects {-}

```{r tut=FALSE}
# Declare new objects using other variables
c <- a + b + 10

# Open z object or put everything in parentheses
(c <- a + b + 10)
```

### Vectors {-}

Think of a vector as a single column in a spreadsheet.

```{r tut=FALSE}
vectorA <- c(1,2,b)
vectorB <- c(TRUE,TRUE,FALSE)
vectorC <- c(emily, libby, other) 

# Vector Operations
vectorA - vectorB # Vector operation AND auto-change TRUE =1, FALSE=0
```

### Data frame {-}

```{r tut=FALSE}
# think of it conceptually like a spreadsheet
dataDF <- data.frame(numberVec    = vectorA,
                     trueFalseVec = vectorB,
                     stringsVec   = vectorC)

# Examine an entire data frame  
dataDF
```

```{r tut=FALSE}
# Declare a new column
dataDF$NewCol <- c(10,9,8)

# Examine with new column
dataDF

# Examine a single column
dataDF$numberVec # by name
dataDF[,1] # by index...remember ROWS then COLUMNS

# Examine a single row
dataDF[2,] # by index position

# Examine a single value
dataDF$numberVec[2] # column name, then position (2)
dataDF[1,2] #by index row 1, column 2
```


### Plots {-}

To be announced ... 

