
# Intro to R {-}
<!-- Comments start with one or multiple hash tags.  -->
<!-- http://seminars-in-applied-economics.de/wp-content/uploads/2020/08/SOEP_Practice_1.txt -->

An .R file is a **text document**. Open (and edit) it with a *text editor*, a *web browser* or a more sophisticated software like RStudio.^[Download RStudio <https://posit.co/downloads/>.] 

Click [this link](https://raw.githubusercontent.com/MarcoKuehne/marcokuehne.github.io/main/data/intro-r.R){target="_blank"} that opens an R script in your browser. 

Using .R files or scripts offers efficiency, reproducibility, scalability, collaboration, documentation, and flexibility. They allow you to automate tasks, handle large datasets, collaborate with others, document your work, and customize solutions.

<!-- https://raw.githubusercontent.com/MarcoKuehne/marcokuehne.github.io/main/data/intro-r.R -->
<!-- https://github.com/MarcoKuehne/marcokuehne.github.io/blob/main/data/intro-r.R -->
<!-- You can collapse sections in RStudio. -->

## R is a calculator {-}

<!-- By default, `tutorial` will convert all R chunks. -->
<!-- # ```{r, include=FALSE} -->
<!-- # tutorial::go_interactive(greedy=FALSE) -->
<!-- # ``` -->

R is a calculator. Use this R demo in the browser to explore basic features of R. Commands in the  **script.R** tab are executed by the **Run** bottom. It runs the entire script and prints out results in the **R Console**. This setting is simplified but reflects the procedure in a more complex integrated developer environment (IDE) like RStudio. <u id='intro_r'>Test it.</u> 

<iframe src="DCL/playground1.html" frameborder="0" scrolling="no" style="width:100%;height:360px"></iframe>

```{r, echo=FALSE}
tippy::tippy_this(elementId = "intro_r", tooltip = "Delete all code, type 1:20 and hit Run.")
```

:::: {.defbox}
::: {.titeldefbox}
<h2> Definition </h2>
:::
Basic **arithmetic operators** are:

- `+` Addition
- `-` Subtraction 
- `*` Multiplication
- `/` Division
- `^` Exponent 
::::

<!-- https://cdn.datacamp.com/dcl-react-prod/index.html -->
<!-- <script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script> -->
<!-- <script type="text/javascript" src="//cdn.datacamp.com/dcl-react.js.gz"></script> -->
<!-- <iframe src="DCL/intro_1.html" frameborder="0" scrolling="no" style="width:100%;height:360px"></iframe> -->

## R is more than a calculator {-}

:::: {.challenge}
::: {.titelchallenge}
<h2> Your Turn: Adjust the code.</h2>
:::

If you never saw R before, change the `main` title  to what ever suits you or change the color option `col` from `lightblue` to `aliceblue`. 

If you have some experience, order the bars using the `sort()` command.

<iframe src="DCL/playground_graph.html" frameborder="0" scrolling="no" style="width:100%;height:360px"></iframe>
::::

<!-- rdrr.io is working but does not look nice -->
<!-- <iframe width='100%' height='500' src='https://rdrr.io/snippets/embed/' frameborder='1'></iframe> -->

## Define objects {-}

Define R objects for later use. Objects are **case-sensitive** (`X` is different from `x`). Objects can take any name, but its best to use something that makes sense to you, and will likely make sense to others who may read your code.

### Numeric Variables {-}

The standard assignment operator is `<-`, the equal sign `=` works as well. The code assigns the object `a` the value 2 and `b` the value 3. The sum is 5. 

```{r tut=FALSE, collapse=TRUE} 
a <- 2
b = 3
a + b
```

### Logical Variables {-}

Logical values are `TRUE` and `FALSE`. Abbreviations work. You can write `T` instead of `TRUE`. 

```{r prompt=TRUE}
harvard               <-        TRUE # spacing doesn't matter
yale      <- FALSE
princeton <- F # short for FALSE

# Attention: FALSE=0, TRUE=1
harvard + 1
```

Although spacing technically doesn't matter in R, there are some  best practices to consider. 

> “Good coding style is like using correct punctuation. You can manage without it, but it sure makes things easier to read.”</b>
`r tufte::quote_footer('-- Hadley Wickam -- [Style guide](http://adv-r.had.co.nz/Style.html)')`

:::: {.reading}
::: {.titelreading}
<h2> Reading </h2>
:::
Place spaces around all binary operators (=, +, -, <-, etc.). Do not place a space before a comma, but always place one after a comma. 

Read more in [Google's R Style Guide](https://web.stanford.edu/class/cs109l/unrestricted/resources/google-style.html) at Uni Stanford.
::::

### String Variables {-}

Text is stored as *string* or *character*.

```{r }
emily  <- 'She is a friend.'     # string / character class / plain text
libby  <- "she is a coworker"    # use ' and " interchangeably
other  <- "people"               # prefer "
``` 

### Factor Variables {-}

A factor is an ordered categorical variable. `c()` is a generic function which combines its arguments. 

```{r tut=FALSE}
fruit <- factor(c("banana", "apple")  ) # The default ordering is alphabetic
fruit

dose <- factor(c("low", "medium", "high")  ) # The default ordering is alphabetic
dose
```

Factor levels inform about the order of the components, i.e. `apple` comes before `banana` and `high` comes comes before `low`, than comes `medium`. Of course, the apple-banana order does not makes any sense, and the high-low-medium order is just wrong. Software cannot know whether an ordering makes sense, that's job of the data scientist. Use the `levels` option inside the `factor()` function to tell R the ordering.

```{r tut=FALSE}
dose <- factor(c("low", "medium", "high"), levels = c("low", "medium", "high") ) 
dose
```

### Combine objects {-}

```{r tut=FALSE}
# Declare new objects using other variables
c <- a + b + 10

# Open z object or put everything in parentheses
(c <- a + b + 10)
```

### Vectors {-}

Think of a vector as a single column in a spreadsheet.

```{r tut=FALSE}
vectorA <- c(1,2,b)
vectorB <- c(TRUE,TRUE,FALSE)
vectorC <- c(emily, libby, other) 

# Vector Operations
vectorA - vectorB # Vector operation AND auto-change TRUE =1, FALSE=0
```

### Data frame {-}

```{r tut=FALSE}
# think of it conceptually like a spreadsheet
dataDF <- data.frame(numberVec    = vectorA,
                     trueFalseVec = vectorB,
                     stringsVec   = vectorC)

# Examine an entire data frame  
dataDF
```

```{r tut=FALSE}
# Declare a new column
dataDF$NewCol <- c(10,9,8)

# Examine with new column
dataDF

# Examine a single column
dataDF$numberVec # by name
dataDF[,1] # by index...remember ROWS then COLUMNS

# Examine a single row
dataDF[2,] # by index position

# Examine a single value
dataDF$numberVec[2] # column name, then position (2)
dataDF[1,2] #by index row 1, column 2
```

## Plots {-}

There are *base R* graphs. There are `ggplot2` plots. 

```{r tut=FALSE}
# Create some variables
x <- 1:10
y1 <- x*x
y2  <- 2*y1

# Create a first line
plot(x, y1, type = "b", frame = FALSE, pch = 19, 
     col = "red", xlab = "x", ylab = "y")
# Add a second line
lines(x, y2, pch = 18, col = "blue", type = "b", lty = 2)
# Add a legend to the plot
legend("topleft", legend=c("Line 1", "Line 2"),
       col=c("red", "blue"), lty = 1:2, cex=0.8)
```

