# Structured Data {#structured-data}

::: {.analysis-subtitle}
Rows, units, time, and where they live
:::

Some data arrives already rectangular: rows, columns, a value in every cell. The work then is not to build the table but to understand what a row *means* — one person, one person-year, one month, one country. A surprising share of the mistakes in applied research can be traced back to a misunderstanding on exactly that point.

This chapter walks the three shapes a rectangle can take. A cross-section holds many units at one moment. A panel holds many units at many moments. A time series holds one unit across many moments. They look alike on screen and demand entirely different analysis.

A fourth section asks what the first three take for granted: where does the rectangle actually lie? Frequently not in a file on your own machine, but on a server — too large to copy, or shared by too many people. The habit that follows from this is worth learning early: stop moving data around, start sending questions to it.

```{r child-tabular-data, child='L02-01-tabular-data.Rmd'}
```

```{r child-panel-data, child='L02-02-panel-data.Rmd'}
```

```{r child-time-data, child='L02-03-time-data.Rmd'}
```

```{r child-remote-data, child='L02-04-remote-data.Rmd'}
```
