## Common Tests Are Linear Models {#tests-as-models}

::: {.analysis-subtitle}
One framework behind the whole toolbox
:::

<small>

> "Teaching statistics as a zoo of named tests is like teaching zoology as a list of animal names."
>
> `r tufte::quote_footer('The complaint that motivated this section')`

</small>

Look back at what the Compare chapter asked you to learn. A one-sample t-test. A two-sample t-test, in a Student and a Welch flavour. A paired t-test. A correlation test. A one-way ANOVA. An ANCOVA. A Wilcoxon test, a Mann-Whitney test, a Kruskal-Wallis test. Nine names, nine sets of assumptions, nine entries in a decision tree that students memorise for an exam and forget by Christmas.

Now that you know what a linear model is, the honest summary can finally be given: **there are not nine tests. There is one model, fitted with different right-hand sides.** Every test in that list is `lm()` wearing a costume.

This observation is not new, but its clearest modern statement is a cheat sheet by Jonas Kristoffer Lindeløv, *Common statistical tests are linear models*, which lines up each classical test against its equivalent `lm()` formula.[^tm-lindeloev] This section works through that correspondence on our own data, so that you can check every claim rather than take it on faith — and then pushes the list further than the cheat sheet does, into repeated measures, variance tests, model comparison, and the counts and proportions that need one step sideways into `glm()`.

```{r tm-load, echo=TRUE, message=FALSE, warning=FALSE}

coursedata <- read.csv("data/Course/GF_AllTime.csv", sep = ";") %>%
  filter(Age < 100)
```

### The one model {#tm-one-model}

Recall the notation. In R, `y ~ 1` means "predict `y` with an intercept and nothing else", and the intercept of a model with no predictors is simply the mean of `y`. Everything else is built by adding terms to the right of the tilde:

| Right-hand side | What it says | Classical name |
|:---|:---|:---|
| `y ~ 1` | one number predicts `y` | one-sample t-test |
| `(y2 - y1) ~ 1` | one number predicts the within-unit change | paired t-test |
| `y ~ 1 + x` | intercept plus a slope on continuous `x` | correlation / simple regression |
| `y ~ 1 + G` | one intercept per group (2 groups) | two-sample t-test |
| `y ~ 1 + G` | one intercept per group (3+ groups) | one-way ANOVA |
| `y ~ 1 + G + x` | group intercepts plus a slope | ANCOVA |
| `y ~ 1 + G + id` | group effects, holding each subject fixed | repeated-measures ANOVA |
| `y ~ 1 + G * S` | group effects that differ by a second factor | two-way ANOVA |
| `rank(y) ~ ...` | the same models fitted to ranks | the non-parametric family |

Read the fourth and fifth rows together. The two-sample t-test and the one-way ANOVA are *the same formula*; the only difference is how many levels the grouping variable happens to have. The wall between "t-test" and "ANOVA" that every introductory course erects is a wall between two and three, and there is nothing there.

::: {.analysis-note}
One caveat before the demonstrations. The equivalences below are **exact** for the parametric tests and **approximate** for the rank-based ones — good approximations, and better the larger the sample, but approximations. Where the match is exact, R will agree with itself to fifteen decimal places. Where it is approximate, the p-values will be close rather than identical, and we say so.
:::

### One mean is an intercept {#tm-one-sample}

Section \@ref(cm-one-sample) tested whether students' self-rated R background differs from the scale midpoint of 3, and found $t = -20.70$. Here is the same question asked of a regression with no predictors at all. Subtract the benchmark from the outcome, then let the intercept do the work:

```{r tm-one-sample, echo=TRUE}
t.test(coursedata$Background.in.R, mu = 3)$statistic

lm(I(Background.in.R - 3) ~ 1, data = coursedata) %>% summary() %>% coef()
```

Identical to four decimal places, because they are the same computation. The intercept of `y ~ 1` is the mean; its standard error is $sd(y)/\sqrt{n}$; the ratio is Gosset's $t$. What the t-test calls "testing against $\mu_0$" the regression calls "is the intercept zero", and every regression table you will ever read tests exactly that for every coefficient it prints.

### A paired test is a one-sample test on differences {#tm-paired}

Section \@ref(cm-paired) already made this point without the vocabulary: pairing is subtraction. Build the differences and fit the emptiest model there is.

```{r tm-paired, echo=TRUE}
differences <- coursedata$Background.in.Statistics - coursedata$Background.in.R

t.test(coursedata$Background.in.Statistics,
       coursedata$Background.in.R, paired = TRUE)$statistic

lm(differences ~ 1) %>% summary() %>% coef()
```

Both give $t = 14.10$. The paired t-test was never a separate test — it is `y ~ 1` applied to a column you constructed.

### Two groups is a dummy variable {#tm-two-sample}

This is the equivalence that section \@ref(comparing-means) promised and deferred. Male students in this course are older than female students. The classical test and the regression:

```{r tm-two-sample, echo=TRUE}
t.test(Age ~ Gender, data = coursedata, var.equal = TRUE)$statistic

lm(Age ~ Gender, data = coursedata) %>% summary() %>% coef() %>% round(3)
```

Both report $|t| = 2.33$; the sign differs only because R's dummy coding takes `Female` as the reference and measures the *male* effect, whereas the t-test subtracts in the other order.

Read the coefficients as the chapter taught you to. The intercept, 24.500, is the mean age of the reference group — the women. The slope on `GenderMale`, 1.122, is *not* the mean age of men but the **difference** between the groups: men average $24.500 + 1.122 = 25.622$ years. A dummy variable turns a group label into a number, and once it is a number the regression machinery does not care that it only takes two values.

There is a third name for this same number, and it is worth collecting. If you code gender 0/1 and correlate it with age, you get what psychometricians call a **point-biserial correlation** — presented in many textbooks as a separate coefficient with its own formula:

```{r tm-pointbiserial, echo=TRUE}
male <- as.numeric(coursedata$Gender == "Male")

cor(coursedata$Age, male)                      # "point-biserial correlation"
cor.test(coursedata$Age, male)$statistic       # its significance test
```

$r = 0.152$, and $t = 2.329$ — the same $t$ as the two-sample test and the same $t$ as the regression slope. Three names, three formulas in three textbooks, one number. The point-biserial correlation is not a special coefficient for binary variables; it is Pearson's correlation, which never cared how many values your variable took.

::: {.analysis-warning}
Note `var.equal = TRUE`. Ordinary least squares assumes one common error variance across the whole model, so it reproduces **Student's** t-test, not Welch's. If the groups genuinely differ in spread, the regression inherits the same problem the classical test had, and the fix is the same: model the variances (`nlme::gls` with `weights = varIdent(form = ~1|G)`), or use robust standard errors.

This is worth noticing rather than glossing over. "Everything is a linear model" does not mean everything is *safe* — it means the assumptions you were taught to check per test are, underneath, the same three assumptions checked once.
:::

### Correlation is a slope {#tm-correlation}

Section \@ref(relationships) tested whether age and total semesters move together. A correlation test and a regression of one on the other:

```{r tm-correlation, echo=TRUE}
cor.test(coursedata$Age, coursedata$Total.Semesters)$statistic

lm(Total.Semesters ~ Age, data = coursedata) %>% summary() %>% coef() %>% round(4)
```

Again identical: $t = 1.7264$, $p = 0.086$, a positive but unconvincing association. The correlation coefficient and the regression slope are different numbers — $r = 0.113$ is unit-free, the slope is in semesters per year — but they are the same *evidence*, because the slope is just the correlation rescaled by the two standard deviations. Testing "is $r$ zero" and "is the slope zero" cannot give different answers.

And **Spearman's** rank correlation, usually taught as the non-parametric alternative with its own table of critical values, is Pearson's correlation applied to ranks. Not approximately — by definition:

```{r tm-spearman, echo=TRUE}
cor(coursedata$Age, coursedata$Total.Semesters, method = "spearman")

cor(rank(coursedata$Age), rank(coursedata$Total.Semesters))    # the same thing

coef(summary(lm(rank(Total.Semesters) ~ rank(Age), data = coursedata)))[2, ]
```

All three give $\rho = 0.1726$, and the regression on ranks returns $p = 0.0083$ — which is exactly what `cor.test(..., method = "spearman")` reports.

Notice that Spearman finds the association ($p = 0.008$) where Pearson did not ($p = 0.086$). Ranking removed the leverage of a handful of very old students with unusual semester counts. When the two disagree, the disagreement is a fact about outliers, and worth a scatterplot before it is worth a decision.

### Many groups is many dummies {#tm-anova}

The step from two groups to seven changes nothing structurally. Our students arrive across seven terms; do the cohorts differ in age?

```{r tm-anova, echo=TRUE}
summary(aov(Age ~ Term, data = coursedata))[[1]][1, 4]   # F from aov()

summary(lm(Age ~ Term, data = coursedata))$fstatistic[1] # F from lm()
```

$F = 2.15$ on both sides, $p = 0.049$ — a whisker under the conventional threshold, which is a fine reminder of the last box in section \@ref(cm-threshold). Behind the scenes `lm()` created six dummy variables for the seven terms, and the F-statistic asks whether all six slopes are zero at once. That is precisely the question ANOVA's F-test asks. The two functions print different tables because they were written for different audiences, not because they compute different things.

Adding a continuous variable to the same formula turns it into ANCOVA, which is why section \@ref(partitioning-variation) could hand the dog-size effects to a height covariate without changing tools. `aov(y ~ G + x)` and `lm(y ~ G + x)` are one model.

#### Spending degrees of freedom on the right question {#tm-trend}

Once groups are dummies, a new option appears that the ANOVA framing hides. Our seven terms are not just seven categories — they are *ordered in time*. Treating them as an unordered factor spends six degrees of freedom asking "are these cohorts different in any way at all?" Treating the term as a number spends **one**, asking the sharper question: "have students been getting older over the years?"

```{r tm-trend, echo=TRUE}
chrono <- c("SS 2020" = 1, "WS 2020 2021" = 2, "SS 2021" = 3, "SS 2022" = 4,
            "WS 2022 2023" = 5, "SS 2023" = 6, "WS 2023 2024" = 7)
coursedata$term_number <- chrono[coursedata$Term]

coef(summary(lm(Age ~ term_number, data = coursedata)))[2, ]   # 1 df: a trend
anova(lm(Age ~ Term, data = coursedata))[1, ]                  # 6 df: any difference
```

The results point in opposite directions, and the contrast is the lesson. The six-degree-of-freedom test is significant, $F = 2.15$, $p = 0.048$: the cohorts do differ. The one-degree-of-freedom trend test is nowhere near, slope $0.035$ years per term, $p = 0.784$: they do not differ *in a straight line over time*. Cohorts vary, but not by drifting older.

This is what a classical toolbox calls a "test for trend" and hands you as yet another named procedure. In the linear-model framework it is not a new test — it is a decision about how to code a variable, and coding an ordered factor as a number is often the single most powerful move available, because it aims all your evidence at one question instead of scattering it across six.

### Repeated measures: subjects are just another factor {#tm-repeated}

Here the framework earns its keep, because repeated-measures ANOVA is where the classical presentation becomes genuinely forbidding — `Error(id/condition)` strata, sphericity corrections, tables that nobody can read.

Our data has the design built in. Each student rated three things: statistics, R, and academic writing. That is three measurements on 233 people — the paired comparison of section \@ref(cm-paired), extended from two conditions to three.

```{r tm-long, echo=TRUE}
long <- coursedata %>%
  mutate(id = factor(row_number())) %>%
  select(id, Statistics = Background.in.Statistics,
             R          = Background.in.R,
             Writing    = Background.in.Academic.Writing) %>%
  pivot_longer(-id, names_to = "Skill", values_to = "Rating") %>%
  mutate(Skill = factor(Skill))

head(long, 4)
```

Now the two ways of asking whether the three ratings differ:

```{r tm-rm, echo=TRUE}
# the classical incantation
summary(aov(Rating ~ Skill + Error(id / Skill), data = long))

# a linear model with the student as a factor
anova(lm(Rating ~ Skill + id, data = long))["Skill", ]
```

$F = 105.0$ on 2 and 464 degrees of freedom, both ways. The `Error(id/Skill)` stratum that makes repeated-measures ANOVA look like a separate branch of statistics is doing one thing: giving every student their own intercept. Written as `+ id` in a formula, it stops being mysterious.

And it matters enormously, exactly as pairing mattered in section \@ref(cm-paired-vs-unpaired):

```{r tm-rm-naive, echo=TRUE}
anova(lm(Rating ~ Skill, data = long))["Skill", "F value"]   # ignoring the student
```

$F = 73.1$ instead of $105.0$. Throwing away the student identifier throws away the fact that the same person produced three of the numbers, and the F-statistic drops by a third. This is the paired t-test's lesson generalised: **whenever the same unit appears more than once, its identity belongs in the model.** With two conditions we called it pairing; with three we call it repeated measures; with many periods and many units we call it panel data (section \@ref(panel-data)) and the term for `+ id` becomes *fixed effects*. Same move every time.

::: {.analysis-note}
One honest limitation. `lm(y ~ C + id)` reproduces the F-test for the condition, which is what people usually want. It does not reproduce everything a dedicated repeated-measures routine offers — sphericity corrections such as Greenhouse-Geisser, or unbalanced designs with missing cells, which are better served by a mixed model (`lme4::lmer(y ~ C + (1|id))`). The mixed model is the honest generalisation; treating the subject as a fixed factor is the version that shows you what is going on.
:::

### Even the assumption checks are models {#tm-levene}

Section \@ref(partitioning-variation) used tests that compare *spreads* rather than means, and they look like a different family entirely. They are not. **Levene's test** — and its more robust cousin, the Brown-Forsythe test — is defined as an ordinary ANOVA run on the absolute deviations from each group's centre. Not approximately: that is the definition.

```{r tm-levene, echo=TRUE}
spread <- coursedata %>%
  group_by(Term) %>%
  mutate(abs_dev = abs(Age - median(Age))) %>%   # distance from your group's centre
  ungroup()

anova(lm(abs_dev ~ Term, data = spread))[1, ]
```

$F = 0.37$, $p = 0.90$: no evidence that the seven cohorts differ in how spread out their ages are. The trick is worth internalising because it generalises. A question about *variability* becomes a question about *means* the moment you replace each observation by its distance from its group's centre — and questions about means are what linear models answer. If you can define your quantity of interest per observation, you can usually test it with `lm()`.

### One step sideways: counts and proportions {#tm-glm}

Two families genuinely sit outside `lm()`, because their outcomes are not continuous. But they sit only just outside: the linear predictor is the same, and only the link function and the error distribution change. This is the **generalised** linear model of section \@ref(generalised-models), and the classical tests for tables and proportions live there.

Take the chi-square test of independence. Are academic level and gender related in this course?

```{r tm-chisq, echo=TRUE}
tab <- table(coursedata$Gender, coursedata$Academic.level)
tab

chisq.test(tab, correct = FALSE)$statistic     # Pearson's X-squared

counts <- as.data.frame(tab) %>% rename(Gender = Var1, Level = Var2, n = Freq)
anova(glm(n ~ Gender + Level,  data = counts, family = poisson),   # independence
      glm(n ~ Gender * Level,  data = counts, family = poisson),   # association
      test = "Chisq")[2, ]
```

Pearson's statistic is 1.155 and the log-linear model's likelihood-ratio statistic is 1.156, on 1 degree of freedom, $p = 0.28$ either way. No association: Bachelor and Master students are split much the same way across genders.

The two numbers differ in the fourth decimal because Pearson's $X^2$ and the likelihood-ratio $G^2$ are two different approximations to the same quantity, and the Poisson model reproduces $G^2$ **exactly**. Reading the model rather than the test also tells you something the test cannot: the interaction *coefficient* is the log odds ratio, with a standard error and a confidence interval attached.

Proportions work the same way, one link function over:

```{r tm-prop, echo=TRUE}
master <- as.numeric(coursedata$Academic.level == "Master")

prop.test(table(coursedata$Gender, master)[, 2:1], correct = FALSE)$p.value

coef(summary(glm(master ~ Gender, family = binomial, data = coursedata)))[2, 4]
```

$p = 0.2825$ against $p = 0.2830$ — the same conclusion by two routes that make slightly different large-sample approximations. And `glm(y ~ 1, family = binomial)` tests a single proportion against 0.5, which is the classical one-proportion test.

::: {.analysis-warning}
These glm equivalences are **approximate**, and the approximation has a known weak spot: Wald standard errors become unreliable when a proportion approaches 0 or 1. The sign test on our 161 non-tied students, 151 of whom rate statistics above R, is a case in point — `binom.test()` gives $p \approx 10^{-33}$ while the Wald test from `glm()` gives $p \approx 10^{-17}$. Both are overwhelming, so nothing is at stake here, but near the boundary the exact test is the one to trust.
:::

### Comparing two models is also a test {#tm-nested}

One last generalisation, and it is the one that frees you from the list altogether. Every F-test in this section — ANOVA's, the trend test's, Levene's — is a comparison between a **restricted** and an **unrestricted** model: does letting these extra terms into the model explain enough additional variation to justify them?

Written out, that comparison is a named test in econometrics. Does the relationship between age and semesters studied differ between Bachelor and Master students? Fit both models and ask:

```{r tm-chow, echo=TRUE}
pooled <- lm(Total.Semesters ~ Age, data = coursedata)                   # one line
split  <- lm(Total.Semesters ~ Age * Academic.level, data = coursedata)  # two lines

anova(pooled, split)
```

$F = 8.89$ on 2 and 229 degrees of freedom, $p = 0.0002$: the relationship genuinely differs. Pooling Bachelor and Master students into one regression line hides two different lines — an instance of the aggregation warning from section \@ref(relationships).

Economists call this a **Chow test** for a structural break. It has no separate function in R because it does not need one: it is `anova()` on two nested models, which is the same machinery that produced every F-statistic above.

### The extended list {#tm-list}

Here is the full correspondence, ours and Lindeløv's combined. $G$ and $S$ are grouping factors, $x$ is continuous, `id` identifies the unit, `signed_rank(x)` is `sign(x) * rank(abs(x))`, and `within_rank(y)` ranks each unit's own measurements against each other.

| Classical test | Built-in R function | Equivalent model | Exact? |
|:---|:---|:---|:---|
| **One outcome, one sample** | | | |
| One-sample t-test | `t.test(y)` | `lm(y ~ 1)` | exact |
| Paired t-test | `t.test(y1, y2, paired = TRUE)` | `lm(y1 - y2 ~ 1)` | exact |
| Wilcoxon signed-rank | `wilcox.test(y)` | `lm(signed_rank(y) ~ 1)` | approx, $n > 14$ |
| Sign test | `binom.test(sum(y > 0), n)` | `glm(y > 0 ~ 1, binomial)` | approx |
| One proportion | `prop.test(k, n)` | `glm(y ~ 1, binomial)` | approx |
| **Two variables** | | | |
| Pearson correlation | `cor.test(x, y)` | `lm(y ~ x)` | exact |
| Spearman correlation | `cor.test(x, y, method = "spearman")` | `lm(rank(y) ~ rank(x))` | exact |
| Point-biserial correlation | `cor.test(y, G01)` | `lm(y ~ G)` | exact |
| Two-sample t-test | `t.test(y ~ G, var.equal = TRUE)` | `lm(y ~ G)` | exact |
| Welch's t-test | `t.test(y ~ G)` | `gls(y ~ G, weights = varIdent(~1\|G))` | exact |
| Mann-Whitney U | `wilcox.test(y ~ G)` | `lm(rank(y) ~ G)` | approx, $n > 11$ |
| Two proportions | `prop.test(tab)` | `glm(y ~ G, binomial)` | approx |
| **Three or more groups** | | | |
| One-way ANOVA | `aov(y ~ G)` | `lm(y ~ G)` | exact |
| Kruskal-Wallis | `kruskal.test(y ~ G)` | `lm(rank(y) ~ G)` | approx, $n > 11$ |
| ANCOVA | `aov(y ~ G + x)` | `lm(y ~ G + x)` | exact |
| Two-way ANOVA | `aov(y ~ G * S)` | `lm(y ~ G * S)` | exact |
| Test for trend | `prop.trend.test` and relatives | `lm(y ~ as.numeric(G))` | exact, 1 df |
| MANOVA / Hotelling's $T^2$ | `manova(...)` | `lm(cbind(y1, y2) ~ G)` | exact |
| **Repeated measures** | | | |
| Repeated-measures ANOVA | `aov(y ~ C + Error(id/C))` | `lm(y ~ C + id)` | exact for the F on C |
| Friedman test | `friedman.test(y ~ C \| id)` | `lm(within_rank(y) ~ C + id)` | approx |
| **Spread rather than centre** | | | |
| Levene / Brown-Forsythe | `car::leveneTest(y ~ G)` | `lm(abs(y - median_G) ~ G)` | exact, by definition |
| **Counts and tables** | | | |
| Chi-square independence | `chisq.test(tab)` | `glm(n ~ row + col, poisson)` | exact for $G^2$ |
| Chi-square goodness of fit | `chisq.test(y)` | `glm(n ~ 1, poisson)` | exact for $G^2$ |
| Cochran-Armitage trend | `prop.trend.test` | `glm(y ~ dose, binomial)` | approx |
| **Model against model** | | | |
| Chow test / structural break | — | `anova(lm(y ~ x), lm(y ~ x * G))` | exact |
| Nested model F-test | — | `anova(model1, model2)` | exact |

::: {.analysis-takeaway}
Count the right-hand column. Of twenty-eight named procedures, roughly two-thirds are **exactly** a linear model and the rest are close approximations or a single step into `glm()`. The decision tree you were asked to memorise describes one function.
:::

### Why this matters {#tm-why}

Three reasons, in rising order of importance.

**It halves what you have to remember.** A decision tree with dozens of leaves becomes one function and a formula syntax. When you meet a design this book never covered — three groups measured twice, a covariate plus an interaction, an ordinal predictor — you do not go looking for the name of the right test. You write the model.

**It makes the assumptions visible.** Every test in the list above inherits the same handful of assumptions from its parent model: the errors are independent, roughly constant in variance, roughly normal. Presented as twenty-eight separate tests, those become eighty-four assumptions to memorise. Presented as one model, they are three things to check, and `plot(model)` checks them.

**It moves the conversation from testing to estimating.** This is the real prize. A t-test hands you a verdict; a linear model hands you a coefficient, a standard error and a confidence interval — how big, in what units, how precisely known. Section \@ref(cm-threshold) argued that significance is a threshold on evidence rather than a statement about importance. The linear-model framing makes that argument structural: once every test is a model, the *estimate* is what you naturally report, and the p-value becomes one column of the table rather than the point of it.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: What the framework does not swallow</h2>
:::
Honesty requires the boundaries. "Everything is a linear model" is a claim about the tests in an introductory syllabus, not about all of statistics.

**Exact tests** for small samples are not approximations of a linear model at all — they enumerate possibilities. Fisher's exact test, the binomial test, and the permutation logic behind the lady tasting tea in section \@ref(partitioning-variation) count outcomes rather than fit parameters, which is precisely why they remain valid when every large-sample approximation has broken down.

**Distributional tests** have no linear model behind them because they are not about means, slopes, or spreads. The Kolmogorov-Smirnov test asks whether two entire distributions differ in shape; Shapiro-Wilk asks whether one is normal. Section \@ref(partitioning-variation) used both, and neither can be written as a regression.

**Extensions rather than special cases.** Mixed and multilevel models generalise `+ id` into a random effect; survival models handle censored time; the structural equation models of chapter \@ref(reveal) estimate relationships between variables nobody observed. Quantile regression replaces the mean with the median or any other quantile — the same idea with a different loss function. Each of these contains the linear model rather than fitting inside it.

The right summary is therefore narrower than the slogan and more useful: *the classical tests you were taught as a list are special cases of one model, and almost everything that is not a special case is a generalisation of the same idea.* Either way there is one framework, not a zoo.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
1. You want to compare mean income across four regions. Which call does the job? `r mcq(c("aov(income ~ region) - lm() only works for two groups", answer = "lm(income ~ region) - the number of levels does not matter", "lm(income ~ 1) with the regions filtered one at a time"))`

2. `lm(y ~ G)` where `G` has exactly two levels is equivalent to which classical test? `r mcq(c("One-sample t-test", answer = "Two-sample t-test with var.equal = TRUE", "Welch's t-test", "Paired t-test"))`

3. In `lm(Age ~ Gender)` the intercept was 24.500 and the slope 1.122. What is the mean age of men? `r fitb(25.622, num = TRUE, tol = 0.01)`

4. `r torf(FALSE)` Because every test is a linear model, the assumptions of the linear model do not need to be checked for t-tests.

5. Six cohorts are ordered in time. Testing them as an unordered factor costs 5 degrees of freedom; testing them as a number costs 1. `r torf(TRUE)` The second test can find a pattern the first misses, even though it uses less of the data's freedom.

6. Which of these can **not** be rewritten as a linear model? `r mcq(c("Levene's test for equal variances", "Spearman's rank correlation", answer = "The Kolmogorov-Smirnov test", "Repeated-measures ANOVA"))`

7. A colleague reports "we used a Mann-Whitney test because the data were skewed". Explain in one sentence what model they fitted, in the language of this section.

8. You have three measurements on each of 60 patients and analyse them with `lm(y ~ condition)`, forgetting `+ id`. Will your F-statistic be too large or too small, and why? (Compare section \@ref(tm-repeated).)
::::

[^tm-lindeloev]: Lindeløv, J. K., *Common statistical tests are linear models (or: how to teach stats)*, 2019, <https://lindeloev.github.io/tests-as-linear/>. The accompanying one-page cheat sheet tabulates each test, its built-in R function, the equivalent `lm()` call, and whether the correspondence is exact or approximate. The idea itself is older — it is the standard view in the general-linear-model literature — but Lindeløv's presentation is what made it teachable.
