## Items {#items}

::: {.analysis-subtitle}
Item response theory
:::

```{r irt-setup, echo=FALSE}
library("eRm")           # Rasch family, conditional maximum likelihood
library("psychotools")   # real item-response data sets
library("patchwork")     # combining two panels into one figure

theme_set(theme_minimal(base_size = 11))
```

It is a Monday morning in Year 3. Twenty-four children sit down in front of the
same sheet of paper. Fourteen small mathematics tasks: count backwards from 30,
find a number on a number line, read a clock, work out half of 18, pull a value
out of a little table. Some children finish in ten minutes, some are still
staring at task 9 when the bell rings.

By lunchtime the teacher has twenty-four numbers between 0 and 14. And then the
awkward questions begin. Is a 9 good? Marie got 9 and Jonas got 9, but they got
different tasks right — did they learn the same amount? Task 12 was solved by six
children last year and by three children this year; was this year's class weaker,
or was this year's version of task 12 harder? And if the school down the road
uses a different sheet with twelve tasks, can the two classes be compared at all?

Every one of those questions is about the same thing: the sheet of paper and the
children are tangled up in each other. A raw score confounds *how much a child
can do* with *how hard the tasks happened to be*. Factor analysis, in Section
\@ref(latent-variables), took several indicators and asked what latent variable
might explain why they covary. Item response theory turns the problem around and
looks at one response at a time:

> How likely is a person with a particular level of competence to answer this
> particular item correctly?

That single change of viewpoint makes **persons and items equally visible**. A
wrong answer can be wrong because the child has less of the relevant competence,
because the task is difficult, or because the model does not describe what
actually happened in that child's head. The rest of this section is about telling
those three apart — and about how much data you need before you are allowed to
try.

::: {.analysis-note}
This section is written around one small classroom, because that is where the
questions are easiest to feel. Everything scales up unchanged to VERA, TIMSS and
PISA; only the number of children changes, and Section \@ref(irt-how-many) is
about exactly how much that number matters.
:::

### Where IRT Sits among Latent Variable Models {#irt-latent-map}

A first map distinguishes the measurement level of the **observed** indicators
from the form of the **latent** variable behind them.

```{r irt-map-table, echo=FALSE}
knitr::kable(
  data.frame(
    `Observed indicators` = c("Continuous", "Categorical"),
    `Continuous latent variable` = c("factor analysis", "item response theory / categorical factor analysis"),
    `Categorical latent variable` = c("latent profile analysis", "latent class analysis"),
    check.names = FALSE
  ),
  caption = "Four cells, two questions: what kind of data did you observe, and what kind of thing do you believe is behind it? Item response theory lives in the bottom-left cell."
)
```

Item response theory is therefore **not one isolated model** in one cell. It is
the main family of latent-trait models for categorical item responses.
Categorical confirmatory factor analysis (Section \@ref(confirmatory-factors))
occupies almost the same mathematical territory but speaks a different dialect:
loadings and thresholds rather than discrimination and difficulty.

The table is a map, not a taxonomy. Real models mix several continuous and
categorical latent variables, several item formats, repeated measurements,
multilevel structure, testlets, response times, or latent classes of response
process.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
An **item response model** states the probability of each possible answer to each
item as a function of (i) one or more latent person variables and (ii) one or
more item parameters. The **Rasch model** is the member of that family in which
the only item parameter is a difficulty, and in which competence and difficulty
enter only through their difference.
::::

### A Short History of the Idea {#irt-history}

The idea did not arrive fully formed. It was assembled over sixty years, twice
independently, in two intellectual traditions that still argue with each other.

```{r irt-history-table, echo=FALSE}
knitr::kable(
  data.frame(
    Year = c(1925, 1952, 1960, 1968, 1969, 1972, 1973, 1978, 1979, 1981, 1982, 1989, 1991, 2000, 2015),
    Who = c("Thurstone", "Lord", "Rasch", "Birnbaum", "Samejima", "Bock", "Fischer / Andersen",
            "Andrich", "Wright & Stone", "Bock & Aitkin", "Masters", "Warm", "Mislevy",
            "OECD", "OECD"),
    `What arrived` = c(
      "persons and items scaled onto one continuum",
      "the item characteristic curve as the object of a theory of testing",
      "the one-parameter model, and specific objectivity as a requirement",
      "discrimination and guessing parameters: the 2PL and 3PL",
      "the graded response model, for ordered categories",
      "the nominal response model, for unordered ones",
      "item difficulty explained by item features; a conditional test of the model",
      "the rating scale model",
      "the person-item map, and Rasch measurement as a movement",
      "marginal maximum likelihood made practical by an EM algorithm",
      "the partial credit model",
      "weighted likelihood estimation",
      "plausible values, so a survey can report a population correctly",
      "PISA scales its items with a Rasch and partial credit model",
      "PISA switches to a 2PL and generalised partial credit model"
    ),
    check.names = FALSE
  ),
  caption = "Sixty years of item response theory in fifteen lines. The left half of the table is a disagreement about what a measurement model is for; the right half is the machinery that made large-scale assessment possible."
)
```

**Two traditions, two problems.** The American line begins with Louis Leon
Thurstone's absolute scaling of the Binet-Simon tasks[^irt-thurstone] and matures
in Frederic Lord's monograph and Ledyard Tucker's *item characteristic
curve*.[^irt-lord] Its motive is practical: the Educational Testing Service has to
equate test forms that are not identical. Allan Birnbaum then replaces the normal
ogive with the logistic function and adds a discrimination parameter, and after it
a lower asymptote for guessing.[^irt-birnbaum] More parameters, better description.

**Rasch came the other way round.** Georg Rasch was a Danish mathematician who had
spent 1935 in London with R. A. Fisher and absorbed Fisher's ideas about
sufficiency. After the war he was handed two very concrete jobs: analyse the Danish
military's intelligence tests, and evaluate reading tests for children whose
schooling had been interrupted. Those children had been tested at different ages
with different tests, and he had to compare them anyway. Out of that came
*Probabilistic Models for Some Intelligence and Attainment Tests*.[^irt-rasch] His
motive was not fit but a philosophical requirement he later called **specific
objectivity**: a comparison between two persons must not depend on which items were
used, and a comparison between two items must not depend on which persons answered
them.[^irt-berkeley] The whole disagreement in this section grows from that one
sentence.

**Then the machinery.** Erling Andersen showed how to eliminate the person
parameters by conditioning on the raw score, and turned the resulting conditional
likelihood into a formal test of the model; Gerhard Fischer decomposed item
difficulty into item features.[^irt-andersen] [^irt-fischer] Benjamin Wright turned
Rasch measurement into a movement, a software package and a picture — the
person-item map this section draws in Figure
\@ref(fig:irt-wright-map).[^irt-wright] Fumiko Samejima[^irt-samejima] and Darrell
Bock[^irt-bock] handled ordered and unordered categories, and David
Andrich[^irt-andrich] and Geoff Masters[^irt-masters] extended the Rasch model
itself to items worth 0, 1 or 2 points — where Section \@ref(rasch-family) picks it
up. Bock and Aitkin made marginal maximum likelihood practical;[^irt-bockaitkin]
Thomas Warm removed the bias from likelihood scoring; and Robert
Mislevy, working on the American NAEP survey, invented **plausible values**, without
which PISA could not exist.[^irt-mislevy] The models went to school from 2000
onwards[^irt-pisa] — and in Germany, VERA was developed at Koblenz-Landau in 2002,
first written in 2003, and extended to all sixteen *Länder* by 2007/08.[^irt-vera]


:::: {.amazing}
::: {.titelamazing}
<h2>The most demanding model is the oldest and the smallest</h2>
:::
Normally a field starts simple and adds parameters. Here the opposite happened:
Birnbaum's more flexible 2PL and 3PL models came *after* Rasch's one-parameter
model and were seen as progress, while the Rasch community regarded the extra
parameters as a retreat. The argument is not statistical but philosophical. If a
model with a free discrimination per item always fits better, then no item can
ever fail — and a measuring instrument that can never fail a test is not being
tested. Rasch's model is the restrictive one on purpose.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The person who introduced a discrimination parameter for each item, and thereby
created the 2PL model, was `r mcq(c("Georg Rasch", answer = "Allan Birnbaum", "Frederic Lord", "Benjamin Wright"))`.

The trick that lets a survey report a correct population mean even though every
individual score is uncertain is called `r fitb(c("plausible values", "plausible value"), ignore_case = TRUE)`.

PISA changed its scaling model in `r fitb(2015, num = TRUE, tol = 0.5)`, moving
from a one-parameter to a two-parameter model.

True or false: Rasch developed his model as a simplification of Lord's earlier
work. `r torf(FALSE)` `r hide("Why?")` The two traditions grew up independently
and for different reasons. Lord's problem was equating test forms at an American
testing agency; Rasch's problem was comparing Danish children who had been given
different reading tests at different ages. Rasch's route to the model ran through
Fisher's sufficiency, not through Lord's normal ogive. `r unhide()`
::::

### Persons and Items on a Common Scale {#irt-person-item-scale}

For a binary item, the Rasch model is

$$
P(X_{pi} = 1 \mid \theta_p, b_i)
=
\frac{\exp(\theta_p - b_i)}{1 + \exp(\theta_p - b_i)},
$$

where $X_{pi} = 1$ means that person $p$ answered item $i$ correctly, $\theta_p$
is that person's latent competence, and $b_i$ is that item's difficulty. On the
log-odds scale the same statement is a subtraction:

$$
\operatorname{logit}\{P(X_{pi} = 1)\} = \theta_p - b_i.
$$

Everything that makes the model useful is in that subtraction:

- if $\theta_p = b_i$, the predicted probability is exactly $.50$;
- one logit above the item gives about $.73$;
- one logit below gives about $.27$;
- two logits above gives about $.88$.

A higher number means more competence for a person and more difficulty for an
item, and the two live on the same axis. That is why a Rasch difficulty can be
printed next to a child's estimate and compared directly, which a percentage
never can.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
A **logit** is a unit of log-odds. A difference of one logit multiplies the odds
of a correct answer by $e \approx 2.72$; a difference of $0.69$ logits doubles
them. Unlike a percentage, a logit means the same thing everywhere on the scale:
the step from $.10$ to $.21$ and the step from $.50$ to $.67$ are both one logit.
::::

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: probability, odds and log-odds</h2>
:::
The whole of it is one ladder with three rungs, and the misreadings all come from
stepping over a rung.

**Rung one: probability into odds.** A probability is trapped between 0 and 1,
which makes it a bad thing to add to. Odds are not:

$$
\text{odds} = \frac{p}{1-p}.
$$

Odds compare the chance of *right* with the chance of *wrong*. They run from 0 to
infinity and they *multiply* rather than add. A probability of $.5$ is odds of 1
— evens; $.75$ is 3, read "three to one"; $.90$ is 9; $.20$ is $0.25$, read "one to
four". Odds of 2:1 are **not** 200 per cent — they are $2/(2+1) = 66.7$ per cent.

**Rung two: the logarithm turns multiplying into adding.** That is the only thing
a logarithm does, and it is the entire reason it appears here:

$$
\ln(a \cdot b) = \ln a + \ln b.
$$

The natural logarithm is used because its inverse, $e^{x}$, undoes it exactly:
$\ln(e^{x}) = x$. In R, `log()` already means $\ln$.

**Rung three: the logit is the log of the odds** — not the log of the probability.

$$
\operatorname{logit}(p) = \ln\frac{p}{1-p}.
$$

Now the Rasch formula rearranges in four lines. Write $x = \theta_p - b_i$:

$$
\begin{aligned}
p &= \frac{e^{x}}{1 + e^{x}} && \text{the model} \\
p\,(1 + e^{x}) &= e^{x} && \text{clear the denominator} \\
p &= e^{x}(1 - p) && \text{collect the } e^{x} \text{ terms} \\
\frac{p}{1-p} &= e^{x} && \text{divide by } (1-p)
\end{aligned}
$$

The left-hand side is the odds. Take $\ln$ of both sides and $\ln(e^{x}) = x$
leaves

$$
\operatorname{logit}(p) = \theta_p - b_i .
$$

**Person minus item equals the log-odds of a correct answer.** That is the sentence
the whole model reduces to.

```{r irt-logit-ladder, echo=TRUE}
p <- c(0.01, 0.10, 0.20, 0.25, 0.50, 0.667, 0.75, 0.90, 0.99)

knitr::kable(
  data.frame(
    probability = p,
    odds        = round(p / (1 - p), 3),
    logit       = round(qlogis(p), 3),
    back        = round(plogis(qlogis(p)), 3)
  ),
  caption = "The ladder, in both directions. Notice the symmetry the probability scale hides: .25 and .75 sit at -1.10 and +1.10, .10 and .90 at -2.20 and +2.20. In R, qlogis() is the logit and plogis() is its inverse."
)
```

**What one logit is worth.** From $\text{odds} = e^{\,\theta - b}$:

- $\theta - b = 0 \Rightarrow$ odds $1$, $p = 50.0\%$;
- $\theta - b = 1 \Rightarrow$ odds $e = 2.72$, $p = 73.1\%$;
- $\theta - b = -1 \Rightarrow$ odds $0.37$, $p = 26.9\%$;
- $\theta - b = 2 \Rightarrow$ odds $7.39$, $p = 88.1\%$.

So one logit **multiplies the odds by $e \approx 2.72$** — it does not double them.
Doubling the odds is $\ln 2 = 0.693$ logits, which corresponds to $p = 66.7\%$.

**A complete numerical example.** A person at $\theta = 1.2$ meets an item at
$b = 0.7$:

```{r irt-logit-example, echo=TRUE}
theta <- 1.2; b <- 0.7

round(c(logit       = theta - b,
        odds        = exp(theta - b),
        probability = plogis(theta - b)), 3)
```

Half a logit above the item, odds of about 1.65 to 1, a solution probability of
62.2 per cent.

**And the five misreadings**, which are worth more than the algebra:

```{r irt-logit-mistakes, echo=FALSE}
knitr::kable(
  data.frame(
    `Said` = c(
      "\"the logit is just $\\log(p)$\"",
      "\"$\\theta = 2$ is twice as able as $\\theta = 1$\"",
      "\"$\\theta - b = 2$ means odds of 2 to 1\"",
      "\"$p = .75$ means odds of 0.75\"",
      "\"b is the percentage of people who solve the item\""
    ),
    `Should be` = c(
      "the logit is $\\ln\\{p/(1-p)\\}$, the log of the *odds*",
      "the logit scale has no natural zero and is not a ratio scale; only distances mean anything",
      "odds of $e^2 = 7.39$ to 1; odds of 2 to 1 is $\\ln 2 = 0.693$ logits",
      "odds are $.75/.25 = 3$ to 1",
      "b is a position on the shared logit scale; the percentage also depends on who took the test"
    ),
    check.names = FALSE
  ),
  caption = "Five things that get said about logits, and what each of them should be. The second is the one that causes real damage in reporting."
)
```

The second row is worth dwelling on, because the rest of this section depends on
it. In the Rasch model the probability depends only on the **difference**
$\theta - b$, so $\theta = 2$ with $b = 1$ and $\theta = 12$ with $b = 11$ describe
exactly the same situation. The scale therefore needs a convention fixing where
zero is — see [somebody has to choose where zero is](items.html#irt-origin) — and
that zero is a reference point, not an origin of ability. Distances are meaningful;
positions are meaningful only relative to the convention.

Finally, why this is elegant rather than merely clever. On the probability scale
the relation between ability and difficulty is S-shaped and non-linear; after the
logit transformation it is a subtraction. That is what puts persons and items on
one metric, and why a step of one logit means the same thing at the bottom of the
scale as at the top.
::::

#### The model is also a recipe {#irt-build-the-class -}

A response model says how the data would have arisen. That means it can be run
forwards as well as backwards, and the fastest way to understand a model is to use
it to build a world you already know the answer to. So let us hire a teacher, write
fourteen tasks, and enrol twenty-four children.

Here is the test the children will sit. The right-hand column is the only thing the
model needs: where each task sits on the logit scale of the previous page.

```{r irt-task-table, echo=FALSE}
task_sheet <- data.frame(
  task = c("count_back", "number_line", "place_value", "add_2digit", "sub_2digit",
           "times_table", "half_of", "word_problem", "shape_names", "symmetry",
           "clock_read", "measure_cm", "data_table", "estimate_sum"),
  `what the child is asked to do` = c(
    "count backwards from 30 in steps of one",
    "mark 40 on a number line running from 0 to 100",
    "say how many tens are in 342",
    "add 27 and 46 in writing",
    "subtract 38 from 71 in writing",
    "give 7 x 8 from memory",
    "work out half of 18",
    "a two-sentence word problem with one operation",
    "name a rectangle, a triangle and a circle",
    "draw the mirror image of a shape across a line",
    "read 07:45 from an analogue clock",
    "measure a line to the nearest centimetre",
    "pull one value out of a small table",
    "estimate 198 + 403 without calculating it"
  ),
  difficulty = c(-2.4, -1.9, -1.5, -1.1, -0.8, -0.4, -0.1,
                  0.3,  0.6,  1.0,  1.4,  1.8,  2.2,  2.7),
  check.names = FALSE
)

knitr::kable(
  task_sheet,
  caption = "The fourteen tasks, ordered from easy to hard. A child at competence 0.3 - the average of the class we are about to build - solves the first task with probability .94 and the last with probability .08. These numbers are the truth in this section, because the data are about to be invented from them."
)
```

Three lines of R turn that column into a class. There is no library call and no
model fitting here — this is the formula from the top of the section, run forwards.

```{r irt-simulate-class, echo=TRUE}
set.seed(2026)

difficulty <- setNames(task_sheet$difficulty, task_sheet$task)

# Twenty-four children, competence drawn from a normal distribution.
competence <- rnorm(24, mean = 0.3, sd = 1.1)

# The Rasch model itself: one subtraction and one coin flip per cell.
prob      <- plogis(outer(competence, difficulty, "-"))
responses <- matrix(rbinom(length(prob), size = 1, prob = prob), nrow = 24,
                    dimnames = list(sprintf("P%02d", 1:24), names(difficulty)))
```

```{r irt-first-rows, echo=FALSE}
knitr::kable(
  cbind(responses[1:6, 1:8], total = rowSums(responses)[1:6]),
  caption = "The first six children and the first eight of the fourteen tasks, with each child's total out of fourteen in the last column. This is the entire raw material of every analysis in this section: one row per child, one column per task, 1 for correct."
)
```

Two things about this class are worth saying out loud, because the rest of the
section leans on both. First, we know the truth: the fourteen difficulties and the
twenty-four competences are written down above, so every estimate can be checked
against the answer. Second, the data were generated by a *perfect* Rasch model — no
misfitting item, no guessing, no second dimension. Whatever goes wrong from here is
not the model's fault. It is the sample size.

Nobody scored 0 and nobody scored 14, which is lucky and will matter in Section
\@ref(irt-score-options).

```{r irt-icc-figure, echo=FALSE, fig.cap="Three of the fourteen tasks, drawn as item characteristic curves. All Rasch curves have the same shape and differ only in where they sit; each crosses the dotted line at .50 exactly above its own difficulty.", fig.align='center', out.width="90%"}
shown_items <- c("count_back", "times_table", "data_table")

icc_data <- expand.grid(
  theta = seq(-4, 4, length.out = 200),
  item  = shown_items,
  stringsAsFactors = FALSE
)
icc_data$b     <- difficulty[icc_data$item]
icc_data$prob  <- plogis(icc_data$theta - icc_data$b)
icc_data$label <- sprintf("%s (b = %.1f)", icc_data$item, icc_data$b)

ggplot(icc_data, aes(theta, prob, colour = label)) +
  geom_hline(yintercept = 0.5, linetype = 3) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Competence (logits)", y = "Probability of a correct answer", colour = NULL) +
  theme(legend.position = "top")
```

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
A child's competence is exactly equal to an item's difficulty. The probability
that the child answers correctly is `r fitb(0.5, num = TRUE, tol = 0.01)`.

The same child meets a second item that is one logit *easier*. The probability is
now closest to `r mcq(c(".50", ".62", answer = ".73", ".88"))`.

True or false: doubling a child's probability of a correct answer always means
adding the same number of logits, no matter where on the scale you start.
`r torf(FALSE)` `r hide("Why?")` Probabilities are bounded at 1. Going from .10 to
.20 is possible; going from .60 to 1.20 is not. Logits are additive in the
*odds*, not in the probability. `r unhide()`
::::

### Why Begin with Rasch? {#why-rasch}

The Rasch model is attractive because it is small, interpretable and demanding.
Its defining restrictions are:

1. every item has its own difficulty $b_i$;
2. all items discriminate equally strongly;
3. there is no item-specific guessing parameter;
4. given competence, responses are locally independent.

Those four restrictions buy four consequences.

#### One additive comparison {-}

The response probability depends only on $\theta_p - b_i$. A difference of one
logit means the same thing at the bottom of the scale and at the top. This is
what makes it legitimate to say that a child *moved* half a logit between two
measurements.

#### A raw score can be sufficient {-}

For a complete response vector to the same set of binary Rasch items, **the
number of correct answers is a sufficient statistic** for the person parameter.
Two children with the same raw score receive the same Rasch estimate, even if
they solved entirely different items.

This is Fisher's sufficiency, and it is the reason Rasch's model has the shape it
has. It also stops holding the moment children receive different item sets, items
are allowed different discriminations, responses go missing in a way that depends
on ability, or the model gains a second dimension.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: watching the person parameter disappear</h2>
:::
Sufficiency is not a slogan; it is a cancellation you can do on the back of an
envelope. Take two items, `times_table` ($b_1 = -0.4$) and `symmetry`
($b_2 = 1.0$), and a child who got exactly one of them right. Which one?

$$
P\big(X = (1,0) \mid r = 1\big)
= \frac{e^{\theta - b_1}}{e^{\theta - b_1} + e^{\theta - b_2}}
= \frac{e^{-b_1}}{e^{-b_1} + e^{-b_2}}
= \frac{e^{0.4}}{e^{0.4} + e^{-1.0}}
= 0.80 .
$$

The $e^{\theta}$ appears in every term of numerator and denominator and cancels.
Whoever the child is — the weakest in the class or the strongest — given that they
solved exactly one of the two, the odds are four to one that it was the easier
one. The competence has left the equation, and what remains is a likelihood that
contains only item parameters.

That is **conditional maximum likelihood**, the estimator `RM()` uses. It is why
item calibration in a Rasch model does not need to assume anything about the
distribution of competence in the sample — and why it exists only inside the Rasch
family, because no other model lets the person parameter cancel like that.
::::

#### Item and person comparisons can be separated {-}

Under model fit, comparisons among persons do not depend on which items were
used, and comparisons among items do not depend on which persons answered them —
apart from the uncertainty that finite data always introduce. Rasch called this
**specific objectivity**, and it is the property that makes item banking, booklet
designs and trend comparison possible at all.

#### The model can be tested rather than merely fitted {-}

Rasch measurement treats equal discrimination and invariance as substantive
requirements. When an item misbehaves, the first question is not "which model has
the better information criterion" but "what is wrong with this item, this
scoring, this dimension or this population". Section \@ref(irt-checks) is a list
of ways to ask.

This is also the limitation, and it should be said plainly. The Rasch model may
simply be too restrictive for the data at hand. If some items separate weak from
strong children far more sharply than others, a two-parameter model describes
what happened better — and pretending otherwise does not make the measurement
more objective.

::: {.analysis-warning}
"Rasch fits better" and "Rasch is the right model" are different claims. A 2PL
model has more parameters and will essentially always fit at least as well. The
Rasch argument is that a set of items which *needs* free discriminations has not
yet been shown to measure one thing on one scale.
:::

:::: {.amazing}
::: {.titelamazing}
<h2>There is exactly one model in which counting works</h2>
:::
The four restrictions above look like a list of things somebody decided to
assume. They are not. Andersen proved the converse: within a broad class of
item-response models, **the Rasch model is the only one** for which the raw score
is a sufficient statistic for the person parameter.[^irt-sufficiency]

So the choice is not between Rasch and something slightly less tidy. It is
between a model in which "how many did you get right" is a complete summary of
your answers, and every other model, in which it is not. Everyone who has ever
added up ticks on a test has been assuming a Rasch model without knowing it.
::::

:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
The Rasch-versus-2PL argument has been running for sixty years and is worth
watching first-hand rather than being summarised. Andrich's *Controversy and the
Rasch model* sets out the measurement position with unusual clarity, and treats
the disagreement as a genuine paradigm difference rather than a technical
dispute.[^irt-controversy]
::::

### IRT Is a Family of Models {#irt-model-family}

Different item formats and different response processes need different response
functions. The catalogue below is the map most people actually want when they ask
"what is the difference between IRT and the Rasch model".

```{r irt-family-table, echo=FALSE}
irt_family <- data.frame(
  Model = c(
    "Rasch model", "1PL model", "2PL model", "3PL model", "4PL model",
    "Partial credit model (PCM)", "Rating scale model (RSM)",
    "Generalised partial credit model (GPCM)", "Graded response model (GRM)",
    "Nominal response model (NRM)", "Many-facet Rasch model (MFRM)",
    "Linear logistic test model (LLTM)", "Multidimensional IRT (MIRT)",
    "Testlet / bifactor model", "Mixture IRT", "Latent class analysis"
  ),
  Family = c(
    "Rasch", "Rasch", "Birnbaum", "Birnbaum", "Birnbaum",
    "Rasch", "Rasch", "Birnbaum", "Birnbaum", "Birnbaum",
    "Rasch", "Rasch", "either", "either", "either", "categorical latent"
  ),
  Data = c(
    "0/1", "0/1", "0/1", "0/1 multiple choice", "0/1 multiple choice",
    "ordered 0,1,2,...", "ordered, same categories for all items",
    "ordered 0,1,2,...", "ordered 0,1,2,...", "unordered categories",
    "0/1 with raters or occasions", "0/1 with known item features",
    "any", "0/1 nested in shared stimuli", "any", "categorical"
  ),
  `Free per item` = c(
    "difficulty", "difficulty (one shared slope)", "difficulty, discrimination",
    "difficulty, discrimination, guessing", "difficulty, discrimination, guessing, slipping",
    "one step difficulty per category boundary", "location only; steps shared",
    "step difficulties and discrimination", "thresholds and discrimination",
    "one slope and intercept per category", "difficulty plus rater and facet effects",
    "weights of item features", "difficulties on several dimensions",
    "difficulty plus a testlet effect", "parameters within each latent class",
    "response probability within each class"
  ),
  Year = c(1960, 1968, 1968, 1968, 1981, 1982, 1978, 1992, 1969, 1972, 1989, 1973, 1985, 1999, 1990, 1950),
  `R package` = c(
    "eRm, TAM, mirt", "TAM, mirt", "mirt, ltm, TAM", "mirt, ltm", "mirt",
    "eRm, TAM, mirt", "eRm, TAM", "mirt, TAM", "mirt, ltm", "mirt",
    "TAM, immer", "eRm", "mirt, TAM", "mirt", "mirt, psychomix", "poLCA, mclust"
  ),
  check.names = FALSE
)

DT::datatable(
  irt_family,
  rownames = FALSE,
  filter = "top",
  caption = "Sixteen models, one question each. Sort by Family to see the two traditions of the previous section; sort by Year to watch the family grow; filter Data by '0/1' to see how many models compete for the simplest case.",
  options = list(pageLength = 8, dom = "tip")
)
```

Two clarifications that save a lot of confusion later.

**Rasch versus 1PL.** Both impose equal discrimination, so as response models
they are the same thing. They differ in how the scale is fixed: a Rasch
parameterisation sets all slopes to exactly 1 and lets the latent variance be
estimated, while a 1PL parameterisation estimates one common slope and fixes the
latent variance to 1. After rescaling, the fitted probabilities are identical.
The names carry a cultural difference, not a mathematical one.

**Partial credit versus graded response.** These two look interchangeable and are
not, and the difference is worth ten minutes once rather than a decade of vague
unease. Both take an item scored 0, 1, 2. Both produce thresholds. Both usually
rank people almost identically. What differs is the question each threshold
answers.

```{r irt-pcm-grm-table, echo=FALSE}
knitr::kable(
  data.frame(
    ` ` = c("The question a threshold answers", "Kind of comparison",
            "Discrimination", "Is the raw score sufficient?",
            "Can a category be 'squeezed out'?", "Estimated by"),
    `Partial credit model` = c(
      "given that you scored 1 or 2, how likely is 2?",
      "adjacent categories, one step at a time",
      "equal for all items, by construction",
      "yes - it is a Rasch model",
      "yes: disordered thresholds are possible and informative",
      "eRm, TAM, mirt"
    ),
    `Graded response model` = c(
      "how likely are you to score at least 2?",
      "cumulative: everything at or above a category",
      "one per item, freely estimated",
      "no",
      "no: the thresholds are ordered by construction",
      "mirt, ltm"
    ),
    check.names = FALSE
  ),
  caption = "The two standard models for an item scored 0, 1, 2. The first row is the whole difference; everything below it follows from that one choice."
)
```

The last two rows are where it becomes practical. Because the partial credit model
compares *adjacent* categories, it can report a threshold ordering that runs
backwards — the second threshold below the first — and that is not a failure but a
message: hardly anybody ever lands in the middle category, so "partial credit" is
not doing any work on this item. The verbal-aggression data of the next section
contains exactly one such item. A graded response model cannot say this, because its cumulative thresholds
are ordered by construction, so the same problem simply disappears into the
parameters.

::: {.analysis-note}
A useful way to read the catalogue: everything in the *Rasch* family keeps the
raw score (or the weighted raw score) sufficient and therefore keeps specific
objectivity. Everything in the *Birnbaum* family trades that away for a better
description of the data.
:::

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: what a discrimination parameter actually does</h2>
:::
The 2PL model adds one number per item:

$$
P(X_{pi} = 1) = \frac{\exp\{a_i(\theta_p - b_i)\}}{1 + \exp\{a_i(\theta_p - b_i)\}} .
$$

Geometrically $a_i$ is the steepness of the curve at its midpoint. A steep item
separates children just below its difficulty from children just above it very
sharply, and says almost nothing about anyone else. A flat item says a little
about everybody.

Two consequences follow, and they are the whole argument.

First, a steep item deserves more *weight* in the person score than a flat one, so
the sufficient statistic is no longer the raw score but the sum of responses
weighted by $a_i$. Two children with the same number correct can now receive
different competence estimates depending on *which* items they got right.

Second, the ordering of items by difficulty can now depend on which children you
ask, because curves with different slopes cross. Two items may swap places between
a weak and a strong group. Specific objectivity is gone — not damaged, gone —
and that is precisely what the Rasch tradition refuses to give up.
::::

### Models within the Rasch Family {#rasch-family}

The binary Rasch model is only the beginning. Five extensions cover almost
everything a school assessment needs.

#### Dichotomous Rasch model {#irt-rasch-dichotomous -}

Each item is scored 0 or 1, each item has one difficulty. This is the model of
the classroom test in this section.

#### Partial credit model {#irt-pcm -}

An item awards ordered scores such as 0, 1, 2 — no credit, partial credit, full
credit. Masters' insight was that you do not need a new idea, only a repeated
one: apply the Rasch logic to *each boundary between adjacent categories*. An
item with three categories therefore has two **thresholds**, $\tau_{i1}$ and
$\tau_{i2}$:

$$
P(X_{pi} = k) \propto \exp\left( k\,\theta_p - \sum_{j \le k} \tau_{ij} \right),
\qquad k = 0, 1, 2 .
$$

The first threshold is the point on the scale where "no credit" and "partial
credit" are equally likely; the second is where "partial credit" and "full
credit" are equally likely. Item difficulty becomes the average of the
thresholds, and each item is free to have its own spacing between them.

Here is the model on real data. The `VerbalAggression` data set contains 316
students who answered 24 statements about frustrating situations with *no*,
*perhaps* or *yes*.[^irt-verbal]

```{r irt-pcm-fit, echo=TRUE}
data("VerbalAggression", package = "psychotools")

aggression <- as.matrix(VerbalAggression$resp)
storage.mode(aggression) <- "integer"

pcm_fit <- PCM(aggression[, 1:6])
round(thresholds(pcm_fit)$threshtable[[1]], 2)
```

Read the last row: for *"I would shout at the bus driver"* the two thresholds sit
at 0.86 and 1.35 logits, so the middle category *perhaps* occupies a narrow strip
of the scale less than half a logit wide. Figure \@ref(fig:irt-pcm-figure) draws
that strip.

When that strip closes entirely, the thresholds come out **disordered** — the
second below the first — and the model is telling you that the middle category is
never the most likely answer for anybody:

```{r irt-pcm-disordered, echo=TRUE}
all_thresholds <- thresholds(PCM(aggression))$threshtable[[1]]
disordered     <- all_thresholds[, "Threshold 2"] < all_thresholds[, "Threshold 1"]

round(all_thresholds[disordered, ], 2)
```

One item of the twenty-four: *"A clerk gives me faulty information and I miss my
train - I would shout."* It offers three answers and effectively uses two. People
say no or they say yes; *perhaps* is a door nobody walks through. That is a finding about the
response format, not a violation of the model — and it is a finding only a
partial credit model can report.

```{r irt-pcm-figure, echo=FALSE, fig.cap="Category probabilities for the item 'I would shout at the bus driver'. The middle category never becomes the most likely answer by much: between the two dotted thresholds it peaks at only .39, because the two thresholds are close together.", fig.align='center', out.width="90%"}
tau <- as.numeric(thresholds(pcm_fit)$threshtable[[1]]["S1DoShout", c("Threshold 1", "Threshold 2")])
grid_theta <- seq(-4, 4, length.out = 300)
numerator <- cbind(1, exp(grid_theta - tau[1]), exp(2 * grid_theta - tau[1] - tau[2]))
category_prob <- numerator / rowSums(numerator)

data.frame(
  theta   = grid_theta,
  no      = category_prob[, 1],
  perhaps = category_prob[, 2],
  yes     = category_prob[, 3]
) %>%
  tidyr::pivot_longer(-theta, names_to = "category") %>%
  ggplot(aes(theta, value, colour = category)) +
  geom_vline(xintercept = tau, linetype = 3) +
  geom_line(linewidth = 1) +
  labs(x = "Verbal aggressiveness (logits)", y = "Probability of the category", colour = NULL) +
  theme(legend.position = "top")
```

#### Rating scale model {#irt-rsm -}

Andrich's rating scale model is the partial credit model with one restriction
added: the *spacing* of the thresholds is the same for every item, and items
differ only in overall location. That is the natural model for a questionnaire
where every statement offers the same four boxes from "strongly disagree" to
"strongly agree".

The restriction is testable, because the rating scale model is nested inside the
partial credit model.

```{r irt-rsm-compare, echo=TRUE}
rsm_fit <- RSM(aggression[, 1:6])

lr_statistic <- 2 * (pcm_fit$loglik - rsm_fit$loglik)
lr_df        <- pcm_fit$npar - rsm_fit$npar

c(
  PCM_logLik = round(pcm_fit$loglik, 2),
  RSM_logLik = round(rsm_fit$loglik, 2),
  LR         = round(lr_statistic, 2),
  df         = lr_df,
  p_value    = round(pchisq(lr_statistic, lr_df, lower.tail = FALSE), 3)
)
```

For these six items the more restrictive model survives: five extra parameters
buy a likelihood improvement of 5.43, which is exactly what chance would give
($p = 0.37$). The shared category structure is a reasonable description, so the
simpler model can be kept.

#### Many-facet Rasch model {#irt-mfrm -}

When a human being marks the answer, the mark depends on three things, not two:
the child, the task, and the rater. The many-facet model adds them additively,

$$
\operatorname{logit}\{P(\text{correct})\} = \theta_p - b_i - s_r,
$$

with $s_r$ the severity of rater $r$. Further facets — occasion, test mode,
interviewer — enter the same way. This is how essay marking, diving competitions
and clinical assessments are put onto one scale.

#### Multidimensional and explanatory Rasch models {#irt-lltm -}

Two more directions. *Multidimensional* models give the person several
competences at once — reading and listening, say — and let each item load on the
ones it needs. *Explanatory* models go the other way and take the item apart:
Fischer's linear logistic test model writes item difficulty as a weighted sum of
item features, so that "this task is hard because it needs two operations and a
context sentence" becomes a testable claim rather than a hunch.

::: {.analysis-note}
Every extension in this list keeps the additive structure — something minus
something on a logit scale. That is what makes them Rasch models. What they add
is a further term, not a further multiplier.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
A questionnaire uses the same five-point agreement scale for all twenty
statements. The model that assumes the five categories are spaced the same way on
every statement is the `r mcq(c("partial credit model", answer = "rating scale model", "graded response model", "nominal response model"))`.

An oral examination is marked by four different examiners, and you suspect one of
them is systematically harsher. The model that turns that suspicion into a
parameter is the `r mcq(c("linear logistic test model", answer = "many-facet Rasch model", "testlet model", "3PL model"))`.

True or false: the partial credit model allows the gap between "0 and 1" to be
larger than the gap between "1 and 2" for one item and smaller for another.
`r torf(TRUE)`
::::

### Three Choices That Are Often Mixed Together {#irt-three-choices}

Three decisions get compressed into the sentence "we used a Rasch model", and
separating them removes most of the confusion in this field.

#### First: the response model {#irt-choice-model -}

How does the probability of each answer depend on person and item? Rasch, 2PL,
partial credit, graded response, multidimensional — the catalogue in Section
\@ref(irt-model-family).

#### Second: estimation of the parameters {#irt-choice-estimation -}

How are the item parameters calibrated from the data?

```{r irt-estimation-table, echo=FALSE}
knitr::kable(
  data.frame(
    Method = c("Conditional ML (CML)", "Marginal ML (MML)", "Joint ML (JML)", "Bayesian / MCMC"),
    `Basic idea` = c(
      "Condition on the raw score, so the person parameters cancel out of the item likelihood",
      "Integrate over an assumed population distribution of competence",
      "Estimate all person and all item parameters simultaneously",
      "Combine the likelihood with priors and sample the whole posterior"
    ),
    `Needs` = c(
      "sufficiency, so only the Rasch family",
      "an assumption about the population",
      "nothing beyond the model",
      "priors"
    ),
    `Typical home` = c("classical Rasch work, eRm", "large-scale assessment, TAM and mirt",
                       "historically important, biased with short tests", "complex hierarchical models"),
    check.names = FALSE
  ),
  caption = "Four ways to calibrate the same response model. Conditional maximum likelihood is the one that uses sufficiency, which is why it exists only inside the Rasch family."
)
```

#### Third: person scoring after calibration {#irt-choice-scoring -}

Once the items are calibrated, how is a person's uncertain competence summarised?
EAP, MAP, WLE and plausible values are alternatives at this **third** stage. They
are not different Rasch models, and choosing between them is not a modelling
decision. Sections \@ref(irt-person-posterior) and \@ref(irt-score-options) are
about this stage alone.

::: {.analysis-warning}
"PISA uses plausible values" and "PISA uses a 2PL model" are statements about two
different stages. A plausible value can be drawn from any calibrated IRT model,
and a Rasch model can be scored with any of the five rules.
:::

### From a Response Pattern to a Person Distribution {#irt-person-posterior}

After calibration, a child's answers produce a **likelihood** for $\theta_p$:

$$
L(\theta_p \mid \mathbf{x}_p, \hat{\mathbf{b}})
=
\prod_i P(X_{pi} = x_{pi} \mid \theta_p, \hat b_i).
$$

Multiplying that likelihood by a population distribution gives a **posterior**:

$$
p(\theta_p \mid \mathbf{x}_p) \propto L(\theta_p \mid \mathbf{x}_p)\, p(\theta_p).
$$

The object that matters is the whole curve. A person score is a summary of it —
or, in the case of a plausible value, a single draw from it.

Take P16, the strongest child in the class, who answered twelve of the fourteen
tasks correctly. Three curves, one child:

```{r irt-posterior-figure, echo=TRUE, fig.cap="One child, three curves. The likelihood of P16's twelve-out-of-fourteen pattern peaks at 2.63 logits; the population distribution peaks at 0; the posterior lands at 1.68. P16's true competence, which we happen to know, is 1.78 — closer to the shrunken estimate than to the likelihood peak.", fig.align='center', out.width="90%"}
pattern <- responses["P16", ]
grid    <- seq(-5, 6, length.out = 1000)

likelihood <- sapply(grid, function(theta) {
  p <- plogis(theta - difficulty)
  prod(p^pattern * (1 - p)^(1 - pattern))
})

population <- dnorm(grid, mean = 0, sd = 1)
posterior  <- likelihood * population

rescale <- function(x) x / max(x)

data.frame(
  theta      = grid,
  Likelihood = rescale(likelihood),
  Population = rescale(population),
  Posterior  = rescale(posterior)
) %>%
  tidyr::pivot_longer(-theta, names_to = "curve") %>%
  ggplot(aes(theta, value, colour = curve)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Competence (logits)", y = "Height (rescaled to 1)", colour = NULL) +
  theme(legend.position = "top")
```

The pull towards the population mean is called **shrinkage**, and it is not a
flaw. With fourteen items the likelihood is broad, so the population distribution
still carries real information about where this child probably is — in this case
it carries the child *closer* to the truth, not further from it. With eighty items
the likelihood would be narrow and the prior would barely move it.

::: {.analysis-warning}
Shrinkage is fair on average and unfair in individual cases. The strongest child
in a class is systematically pulled down and the weakest is systematically pulled
up, which is exactly what you do not want if the number is going on a report card
next to that child's name. This is the main argument for using WLE rather than
EAP for individual feedback.
:::

### Person-Score Options {#irt-score-options}

```{r irt-score-table, echo=FALSE}
knitr::kable(
  data.frame(
    Result = c("ML", "WLE", "MAP", "EAP", "Plausible value"),
    Definition = c(
      "value maximising the response likelihood",
      "bias-corrected likelihood estimate",
      "mode of the posterior",
      "mean of the posterior",
      "random draw from the posterior"
    ),
    `Uses a prior?` = c("no", "no population prior", "yes", "yes", "yes"),
    `Main strength` = c(
      "depends only on the answers and the calibrated items",
      "removes most of the small-sample bias of ML; usually the choice for an individual report",
      "finite and stable, including at the extremes",
      "minimises expected squared error; comes with a posterior SD",
      "carries the uncertainty into population analyses"
    ),
    `Main caution` = c(
      "undefined for all-wrong and all-correct patterns",
      "still a point estimate; report the standard error with it",
      "shrinks towards the population",
      "shrinks towards the population; depends on the population model",
      "not an individual score and never to be averaged into one"
    ),
    check.names = FALSE
  ),
  caption = "Five summaries of one posterior distribution. Only the last one is not a single number, and that is the point of it."
)
```

#### EAP: the centre of gravity {#irt-eap -}

$$
\hat\theta_p^{\,\text{EAP}} = E(\theta_p \mid \mathbf{x}_p)
$$

The posterior mean. Stable for short tests and extreme patterns, because the
population distribution supplies information the answers do not. The price is
shrinkage.

#### MAP: the highest point {#irt-map -}

$$
\hat\theta_p^{\,\text{MAP}} = \operatorname*{arg\,max}_{\theta_p} p(\theta_p \mid \mathbf{x}_p)
$$

The posterior mode. If the posterior is symmetric, MAP and EAP nearly coincide;
if it is skewed, they separate. MAP is often cheaper to compute in
high-dimensional models because it maximises rather than integrates.

#### WLE: a likelihood score with a bias correction {#irt-wle -}

Warm's weighted likelihood estimate solves a modified likelihood
equation.[^irt-warm] It uses no population prior, so it shrinks far less than EAP
or MAP, and unlike ML it stays finite for a perfect or an empty score. When a
single number has to be handed to a teacher or a parent, this is usually the one.

#### Plausible values: several completed versions of the latent variable {#irt-pv -}

$$
\theta_p^{(m)} \sim p(\theta_p \mid \mathbf{x}_p, \mathbf{z}_p),
$$

where $\mathbf{z}_p$ may contain background variables in a latent regression
model. Several draws create several completed data sets; a population analysis is
run once per draw and the results are pooled with the multiple-imputation rules
of Section \@ref(missing-data).

Plausible values are the right tool for population means, group gaps,
correlations and regression coefficients. They are deliberately unsuitable for
ranking individual children.

::: {.analysis-note}
EAP, MAP and WLE answer: *which single value should stand for this child's
uncertain competence?* Plausible values answer a different question: *how can a
later analysis of the whole population keep that uncertainty instead of pretending
the latent variable was observed?*
:::

### Why the Methods Give Different Scores {#irt-score-differences}

The five rules agree when the likelihood is narrow and disagree when it is broad.
They separate when

- the test is short;
- many responses are missing;
- the pattern is all correct or all incorrect;
- the child sits far from the centre of the item difficulties;
- the posterior is skewed;
- background variables strongly predict the trait;
- the item model is multidimensional.

Section \@ref(irt-r-example) draws the disagreement for the classroom test, and
it is larger than most people expect.

### A Minimal Rasch Application in R {#irt-r-example}

Everything is in place: a class, fourteen tasks, and a model. Four steps follow —
estimate the item difficulties, look at them, put children and tasks on one
picture, and score the children four different ways.

The package is **eRm**.[^irt-erm] It estimates the Rasch family by *conditional*
maximum likelihood, which is the estimator that uses the sufficiency of the raw
score and therefore the one that delivers specific objectivity. Section
\@ref(irt-own-data) shows the same analysis in `TAM` and `mirt`, which is what a
large-scale assessment would use.

#### Step 1: estimate the model {#irt-step-estimate -}

```{r irt-fit, echo=TRUE}
rasch_fit <- RM(responses)

# eRm reports item EASINESS; difficulty is its negative.
item_difficulty <- -coef(rasch_fit)
names(item_difficulty) <- sub("beta ", "", names(coef(rasch_fit)))

# The scale has no natural origin, so centre it on the average task.
item_difficulty <- item_difficulty - mean(item_difficulty)
```

::: {.analysis-warning}
The sign trap. `eRm` returns **easiness** parameters, `TAM` and `mirt` return
**difficulty**. A whole analysis can silently come out upside down if the
conversion is skipped. The check takes two seconds: the item everybody solved
must end up with the *lowest* number.
:::

#### Step 2: read the item table {#irt-step-items -}

```{r irt-item-table, echo=TRUE}
item_table <- data.frame(
  solved     = colMeans(responses),
  truth      = difficulty - mean(difficulty),
  estimate   = item_difficulty,
  std_error  = rasch_fit$se.beta
)

knitr::kable(
  item_table[order(item_table$estimate), ],
  digits  = 2,
  caption = "Fourteen tasks in one class of twenty-four. 'truth' is the value the data were generated from; 'estimate' is what the model recovered. They correlate at .89 and are individually wrong by up to 1.1 logits, and the standard errors say so."
)
```

Three things to notice, in increasing order of importance.

The ordering of `estimate` is exactly the ordering of `solved`. That is not a
coincidence and not a weakness — for a complete dichotomous Rasch test, the item
difficulty estimate is a strictly increasing function of the number of children
who got the item wrong. Rasch does not reorder the items. What it does is put
them on a scale with a defined unit, on which the children also live.

The standard errors run from 0.47 to 0.99 logits. The best-measured task in this
class has a 95% interval about two logits wide.

And `count_back`, which was generated as the second-easiest task, came out third,
while `place_value` came out easiest by a full logit. Nothing is wrong with the
model, the items or the children. There are simply twenty-four of them.

#### What range do these numbers have? {#irt-ranges -}

The first thing anyone wants after a fit is a sense of what counts as a big number,
and logits give no help until you have one. The answer follows from the ladder of
Section \@ref(irt-person-item-scale) and it is worth pinning down once.

```{r irt-range-table, echo=TRUE}
distance <- 0:4

knitr::kable(
  data.frame(
    `theta - b`               = distance,
    `P(correct)`              = round(plogis(distance), 3),
    `information p(1-p)`      = round(plogis(distance) * (1 - plogis(distance)), 3),
    `share of the maximum`    = round(plogis(distance) * (1 - plogis(distance)) / 0.25, 2),
    check.names = FALSE
  ),
  caption = "How fast an item stops telling you anything. An item is most informative about a person standing exactly on it; two logits away it carries 42% of that, three logits away 18%, four logits away 7%."
)
```

That decay is why logits do not run far in practice, and it sets three rules of
thumb that carry through the rest of this section.

**Item difficulties normally sit between about $-3$ and $+3$.** Beyond that an item
is solved by almost everybody or almost nobody, so its difficulty is estimated from
a handful of responses and its standard error explodes. In this class the range is
$-3.70$ to $+3.56$, and the two items at the ends have standard errors of $0.99$
against $0.47$ in the middle — the model is telling you it barely saw them. Real,
carefully assembled tests are narrower: the thirteen Innsbruck exam items of
Section \@ref(irt-real-data) span $-1.27$ to $+2.31$.

**Person estimates normally sit between about $-4$ and $+4$.** A test cannot
distinguish beyond the range its items cover, and above four logits every item is
answered correctly with probability $.98$, so there is nothing left to measure with.
This is also why perfect and empty score patterns are special: maximum likelihood
sends them to $\pm\infty$, and WLE returns a finite value only because the bias
correction pins them somewhere — $-5.16$ and $+4.91$ in
Table \@ref(tab:irt-score-table-display), numbers produced by the *test*, not by the
children.

**A population has a standard deviation of about one logit — by construction.** In
this class the twenty-four children were drawn with $\text{sd} = 1.1$, and a real
reference population is *defined* to have a spread of about one (Section
\@ref(irt-bista)). So a difference of one logit is roughly one standard deviation
of the population: large. Half a logit is a substantial difference; a tenth of a
logit is nothing. On the Bildungsstandards metric those become 90 points, 45 points
and 9 points respectively.

::: {.analysis-note}
Put together, the three rules give a quick reading test for any IRT output. If item
difficulties run from $-8$ to $+8$, something has gone wrong — usually an item that
nobody or everybody solved, or a scale that was not centred. If person estimates
cluster inside $\pm 0.5$, the test was far too easy or too hard for this group. And
if a report describes a difference of $0.05$ logits as meaningful, it is describing
noise.
:::

:::: {.amazing}
::: {.titelamazing}
<h2>Rasch does not tell you which item is harder — you knew that already</h2>
:::
With complete data and dichotomous items, the Rasch ranking of items is identical
to the ranking by percentage correct, and the Rasch ranking of children is
identical to the ranking by raw score. Every bit of the *ordering* was already in
the crosstab.

What the model adds is everything the ordering cannot do: a unit that means the
same thing everywhere, a standard error for each number, the ability to compare
children who sat different tests, and a set of assumptions that the data are
allowed to contradict. Measurement is not about finding out who is first. It is
about knowing how far apart people are, and how sure you are.
::::

#### Step 3: put children and tasks on one picture {#irt-step-wright -}

This is Ben Wright's picture, and it is the single most useful output of a Rasch
analysis.

```{r irt-wright-map, echo=FALSE, fig.cap="A person-item map. Children above, tasks below, one shared logit axis. The class piles up between -1 and +1.5, where five of the fourteen tasks sit; place_value and estimate_sum are so far outside the class that they measure almost nobody.", fig.align='center', out.width="95%"}
person <- person.parameter(rasch_fit)

child_scale <- data.frame(theta = person$theta.table[, 1]) %>%
  filter(!is.na(theta))

task_scale <- data.frame(task = names(item_difficulty), b = as.numeric(item_difficulty)) %>%
  arrange(b) %>%
  mutate(task = factor(task, levels = task))

axis_range <- c(-4.3, 4.3)

children_panel <- ggplot(child_scale, aes(theta)) +
  geom_dotplot(binwidth = 0.3, dotsize = 0.85, fill = "grey55", colour = "white") +
  scale_x_continuous(limits = axis_range) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(x = NULL, title = "Children") +
  theme(plot.title = element_text(size = 10))

tasks_panel <- ggplot(task_scale, aes(b, task)) +
  geom_segment(aes(x = min(axis_range), xend = b, yend = task), colour = "grey88") +
  geom_point(size = 2) +
  scale_x_continuous(limits = axis_range) +
  labs(x = "Common scale (logits)", y = NULL, title = "Tasks") +
  theme(plot.title = element_text(size = 10))

children_panel / tasks_panel + plot_layout(heights = c(1, 2.4))
```

A test is **well targeted** when the tasks sit where the children sit. This one is
not, quite: `estimate_sum` was solved by one child in twenty-four and
`place_value` by twenty-three, so both spend almost all their information on parts
of the scale where nobody is standing. Section \@ref(irt-how-many) puts a number
on what that costs.

#### Step 4: score the children four ways {#irt-step-score -}

The five scoring rules of Section \@ref(irt-score-options) are all summaries of the
same posterior curve, so the interesting question is not how to compute them but
how far apart they end up. Table \@ref(tab:irt-score-table-display) answers that for
every score a child could have got on this test.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: the four scoring rules in twenty lines</h2>
:::
No package is needed for any of this, and writing it out once removes the mystery
for good. The grid is the posterior of Section \@ref(irt-person-posterior),
evaluated at 1,201 points; everything else is a different way of pointing at it.

```{r irt-score-function, echo=TRUE}
score_pattern <- function(pattern, b, prior_sd = 1) {

  grid  <- seq(-6, 6, length.out = 1201)
  p_mat <- plogis(outer(grid, b, "-"))
  lik   <- apply(p_mat, 1, function(p) prod(p^pattern * (1 - p)^(1 - pattern)))

  post <- lik * dnorm(grid, mean = 0, sd = prior_sd)
  post <- post / sum(post)

  # ML: the top of the likelihood, undefined at the two extremes
  ml <- if (sum(pattern) %in% c(0, length(pattern))) NA_real_ else
    optimize(function(t) -sum(pattern * (t - b) - log1p(exp(t - b))), c(-8, 8))$minimum

  # WLE: Warm's bias-corrected likelihood equation, solved for its root
  warm <- function(t) {
    p <- plogis(t - b)
    sum(pattern - p) + sum(p * (1 - p) * (1 - 2 * p)) / (2 * sum(p * (1 - p)))
  }
  wle <- uniroot(warm, c(-10, 10))$root

  eap <- sum(grid * post)                     # posterior mean

  c(raw    = sum(pattern),
    ML     = ml,
    WLE    = wle,
    SE_WLE = 1 / sqrt(sum(plogis(wle - b) * (1 - plogis(wle - b)))),
    MAP    = grid[which.max(post)],           # posterior mode
    EAP    = eap,
    SD_EAP = sqrt(sum((grid - eap)^2 * post)))
}
```

Four lines carry the whole argument of Section \@ref(irt-score-options). `ml`
returns `NA` when the pattern is all right or all wrong, because the likelihood
never turns over. `wle` solves Warm's corrected equation instead and therefore
always returns something. `post` is where the population distribution enters —
delete the `dnorm()` factor and MAP and EAP stop shrinking. And `SE_WLE` is the
test information at that point, which is the only line that depends on how many
items there are.
::::

```{r irt-score-table-class, echo=TRUE}
easiest_first <- order(item_difficulty)

score_table <- t(sapply(0:14, function(r) {
  pattern <- rep(0, 14)
  if (r > 0) pattern[easiest_first[1:r]] <- 1
  score_pattern(pattern, item_difficulty)
}))
```

```{r irt-score-table-display, echo=FALSE}
with_error <- function(value, error) {
  ifelse(is.na(value), "--", sprintf("%.2f (%.2f)", value, error))
}

knitr::kable(
  data.frame(
    `correct out of 14`   = score_table[, "raw"],
    `ML`                  = ifelse(is.na(score_table[, "ML"]), "--",
                                   sprintf("%.2f", score_table[, "ML"])),
    `WLE (standard error)` = with_error(score_table[, "WLE"], score_table[, "SE_WLE"]),
    `MAP`                  = sprintf("%.2f", score_table[, "MAP"]),
    `EAP (posterior SD)`   = with_error(score_table[, "EAP"], score_table[, "SD_EAP"]),
    check.names = FALSE
  ),
  align   = c("r", "r", "r", "r", "r"),
  caption = "Every score a child could have got on this test, summarised five ways, in logits. ML has no answer at all for 0 and 14. WLE always answers, but look at the standard error it answers with. MAP and EAP look confident at the edges because the population distribution, not the child, is doing the talking."
)
```

```{r irt-score-figure, echo=FALSE, fig.cap="The four point estimates across the whole raw-score range. In the middle they are almost the same number. At the edges ML runs off the page, WLE follows it more slowly, and MAP and EAP flatten out because they are being held back by the population distribution.", fig.align='center', out.width="90%"}
as.data.frame(score_table) %>%
  select(raw, ML, WLE, MAP, EAP) %>%
  tidyr::pivot_longer(-raw, names_to = "rule") %>%
  ggplot(aes(raw, value, colour = rule)) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, na.rm = TRUE) +
  labs(x = "Raw score out of 14", y = "Competence estimate (logits)", colour = NULL) +
  theme(legend.position = "top")
```

Between raw scores 5 and 10 all four rules agree to within a tenth of a logit.
Outside that band they diverge by more than two logits. The choice of scoring rule
is irrelevant for the middle of the class and decisive for its edges — which is
precisely where individual decisions about children tend to get made.

#### Step 5: does the model fit? {#irt-step-fit -}

```{r irt-itemfit, echo=TRUE}
itemfit(person)
```

`half_of` is flagged: outfit mean-square 2.04, $p = .001$. And we know for certain
that `half_of` is a perfectly well-behaved Rasch item, because we generated it
that way ourselves.

That is the lesson. Fourteen items were tested, one came out significant at the
1% level, and the data contained no misfit at all. With twenty-four children,
item fit statistics are noise generators. The Andersen test does not even get that
far:

```{r irt-andersen-class, echo=TRUE, warning=TRUE}
LRtest(rasch_fit, splitcr = "median")
```

Seven of the fourteen items had to be dropped because, after splitting
twenty-four children into two groups of twelve, those items were answered
identically by everybody in one group. The test that is supposed to check
specific objectivity ran on half the test.

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
In the score table, the child with 7 correct out of 14 has a WLE standard error of
`r fitb(0.67, num = TRUE, tol = 0.02)` logits — the smallest anywhere in the
table. Multiply it by 1.96 and double it: the 95% interval for the best-measured
child in this class is about `r fitb(2.6, num = TRUE, tol = 0.3)` logits wide.

The child who got everything right has no ML estimate because
`r mcq(c("the likelihood is flat", answer = "the likelihood keeps rising and never turns over", "the raw score is not sufficient", "eRm refuses to compute it"))`.

True or false: because MAP and EAP always produce a number, they are the safer
choice for reporting an individual child's competence. `r torf(FALSE)`
`r hide("Why not?")` They produce a number by borrowing it from the population
distribution. For a child at the edge of the class, a large part of the reported
score is an assumption about children in general rather than evidence about this
child. WLE borrows less, and its standard error tells you honestly how little the
test knew. `r unhide()`
::::

### How Many Children, How Many Items? {#irt-how-many}

Now the question this section was really written for. Would one pupil do? Are
five items enough? What is the minimum?

The answer starts with a distinction that most of the confusion hides behind.
A Rasch analysis does **two different jobs**, and they consume two different
resources.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
**Calibration** estimates the item parameters. Its currency is *persons*: every
additional child sharpens every item difficulty.

**Measurement** estimates one person's competence given calibrated items. Its
currency is *items*: every additional task sharpens that one child's estimate.

Neither resource can be exchanged for the other.
::::

#### Can one pupil be enough? {#irt-one-pupil -}

For calibration, no — and not for a subtle statistical reason. Ask the class data
itself:

```{r irt-minimum-children, echo=TRUE, warning=FALSE}
try_calibration <- function(n) {
  sub      <- responses[1:n, , drop = FALSE]
  constant <- sum(colMeans(sub) %in% c(0, 1))
  outcome  <- tryCatch({
    fit <- suppressWarnings(RM(sub))
    sprintf("%d items used, mean SE %.2f", length(fit$se.beta), mean(fit$se.beta))
  }, error = function(e) "estimation fails")

  data.frame(children = n, `constant items` = constant, outcome = outcome, check.names = FALSE)
}

knitr::kable(
  do.call(rbind, lapply(c(1, 2, 3, 5, 10, 15, 20, 24), try_calibration)),
  caption = "The first n children of the class, and what a Rasch calibration can do with them. An item is 'constant' when everybody in the subsample answered it the same way; such an item carries no information about difficulty and is dropped."
)
```

With **one** child, all fourteen items are constant. Of course they are: that
child either solved a task or did not, so every item is 0 or 1 for the whole
sample. There is nothing left to compare. The difficulty of an item is a
statement about how children *differ* on it, and a sample of one has no
differences in it.

With two children the model still cannot run; with three it runs on nine of the
fourteen items with standard errors of 1.3 logits, which is another way of saying
it has learned nothing. Only at twenty children does every item survive.

::: {.analysis-takeaway}
One pupil can never calibrate an item. One pupil *can* be measured — provided
somebody else already calibrated the items. That is not a technicality; it is the
entire business model of educational assessment, and Section \@ref(irt-anchoring)
shows it in five lines of R.
:::

#### Are five items enough? {#irt-five-items -}

Take the same twenty-four children and keep five tasks.

```{r irt-five-items, echo=TRUE}
short <- responses[, c("number_line", "sub_2digit", "times_table",
                       "word_problem", "symmetry")]

table(rowSums(short))
```

Four children got none of the five right and four got all five. That is **eight
of twenty-four** — a third of the class — for whom maximum likelihood has no
answer at all. And for the rest:

```{r irt-five-items-reliability, echo=TRUE}
short_fit    <- RM(short)
short_person <- person.parameter(short_fit)

c(five_items     = round(SepRel(short_person)$sep.rel, 3),
  fourteen_items = round(SepRel(person)$sep.rel, 3))
```

Person separation reliability collapses from **0.73 to 0.01**. The five-item
test does not rank the children badly; it barely ranks them at all, because the
measurement error is as large as the differences between them.

The arithmetic behind that is short enough to do by hand. For a Rasch model the
standard error of one person's estimate is

$$
\text{se}(\hat\theta_p) = \frac{1}{\sqrt{\sum_i p_i (1 - p_i)}},
$$

and $p(1-p)$ is at most $0.25$, reached when the item is exactly at the child's
level. So even a *perfectly targeted* test cannot beat

$$
\text{se}(\hat\theta_p) \ge \frac{2}{\sqrt{k}} .
$$

```{r irt-items-needed, echo=TRUE}
items_needed <- data.frame(items = c(5, 10, 14, 20, 30, 40, 60))
items_needed$best_case_se  <- 2 / sqrt(items_needed$items)
items_needed$ci95_width    <- 2 * 1.96 * items_needed$best_case_se
items_needed$reliability   <- 1 / (1 + items_needed$best_case_se^2)

knitr::kable(
  items_needed, digits = 2,
  caption = "The best a test of k items can possibly do for one child, assuming every item sits exactly at that child's level. Real tests do worse. Five items cannot produce an interval narrower than three and a half logits, which spans almost the whole range a primary class occupies."
)
```

Reading the table as a set of rules of thumb:

- **fewer than 10 items**: enough to sort a class roughly, not enough to say
  anything about an individual child;
- **20 items**: reliability around .83, the usual floor for classroom feedback;
- **30–40 items**: reliability .88–.91, which is where a test that carries
  consequences for individuals should start;
- **60 items**: reliability .94, and now the booklet is longer than a Year 3
  child's attention span, which is why large-scale assessments give every child a
  *different* subset and let the model link them.

So: five items is not enough for anything you would want to do with a single
child. Fourteen — the classroom test in this section — is enough to see the shape
of a class and not enough to talk about Marie.

#### How many children for the items? {#irt-how-many-children -}

The other resource. This simulation repeats the whole calibration a hundred times
for each sample size and records how precisely the items come out.

```{r irt-calibration-simulation, echo=TRUE, cache=TRUE}
set.seed(451)

calibration_precision <- function(n, reps = 100) {
  true_b <- seq(-2, 2, length.out = 14)
  true_b <- true_b - mean(true_b)

  se <- replicate(reps, {
    theta <- rnorm(n)
    x <- matrix(rbinom(n * 14, 1, plogis(outer(theta, true_b, "-"))), n, 14)
    usable <- rowSums(x) > 0 & rowSums(x) < 14
    x <- x[usable, , drop = FALSE]
    if (nrow(x) < 3 || any(colMeans(x) %in% c(0, 1))) return(NA_real_)
    mean(RM(x)$se.beta)
  })

  data.frame(children = n,
             mean_item_se = mean(se, na.rm = TRUE),
             failed_out_of_100 = sum(is.na(se)))
}

children_needed <- do.call(
  rbind,
  lapply(c(10, 25, 50, 100, 200, 400, 800), calibration_precision)
)

knitr::kable(
  children_needed, digits = 3,
  caption = "A hundred simulated calibrations at each sample size. 'Failed' counts the samples in which the model could not be estimated at all because some item came out constant. With ten children that happened seventy times out of a hundred."
)
```

This reproduces, from scratch, the rule of thumb that Rasch practitioners have
used since Linacre's one-page note in 1994: roughly **30 persons** for a rough
sense of an item's difficulty, **100–150** for $\pm 0.3$ logits, and
**250 or more** when the calibration will be reused as an anchor for other
tests.[^irt-linacre] A single class of twenty-four is a factor of ten short of
the last of those.

#### The experiment that settles it {#irt-two-resources -}

Put both levers in one table. Same simulated world, same model; once with eight
classes instead of one, once with twenty-eight tasks instead of fourteen.

```{r irt-two-resources, echo=TRUE, cache=TRUE}
make_class <- function(n_children, item_difficulty, seed = 2026) {
  set.seed(seed)
  competence <- rnorm(n_children, mean = 0.3, sd = 1.1)
  prob <- plogis(outer(competence, item_difficulty, "-"))
  matrix(rbinom(length(prob), 1, prob), nrow = n_children,
         dimnames = list(sprintf("P%03d", seq_len(n_children)), names(item_difficulty)))
}

report <- function(x, label) {
  fit    <- RM(x)
  people <- person.parameter(fit)
  data.frame(design      = label,
             children    = nrow(x),
             items       = ncol(x),
             item_se     = round(mean(fit$se.beta), 2),
             child_se    = round(mean(people$se.theta[[1]]), 2),
             reliability = round(SepRel(people)$sep.rel, 3))
}

long_test <- c(
  difficulty,
  setNames(seq(-2.3, 2.6, length.out = 14), sprintf("extra_%02d", 1:14))
)

knitr::kable(
  rbind(
    report(make_class(24,  difficulty), "one class, 14 tasks"),
    report(make_class(192, difficulty), "eight classes, 14 tasks"),
    report(make_class(24,  long_test),  "one class, 28 tasks"),
    report(make_class(192, long_test),  "eight classes, 28 tasks")
  ),
  caption = "Two levers, two effects, no overlap. Eight times as many children cut the item standard error from 0.59 to 0.19 and left the children's own precision untouched. Twice as many tasks cut each child's standard error from 0.74 to 0.51 and left the item precision untouched."
)
```

Read the four rows twice, once down each column.

Down `item_se`: 0.59, **0.19**, 0.58, **0.19**. Only the number of children moves
it.

Down `child_se`: 0.74, 0.71, **0.51**, **0.50**. Only the number of items moves
it.

::: {.analysis-takeaway}
More children calibrate items. More items measure children. A teacher who wants
to know more about *her class* needs a longer test; a ministry that wants to
build an item bank needs more schools. Testing more children with the same short
test improves the item parameters beautifully and tells you almost nothing new
about any individual child.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Frau Berger wants a 95% interval no wider than 2 logits for each child, so a
standard error of about 0.5. Using $\text{se} \ge 2/\sqrt{k}$, the smallest number
of perfectly targeted items that could achieve it is `r fitb(16, num = TRUE, tol = 0.5)`.

Her colleague suggests testing all eight Year 3 classes with the same fourteen
tasks instead, "because more data is better". After that, each child's standard
error will be `r mcq(c("about half as large", "about a third as large", answer = "essentially unchanged", "impossible to say"))`.

Now do it properly. Rebuild the class, drop the two badly targeted tasks
(`place_value`, which twenty-three children solved, and `estimate_sum`, which one
child solved), refit, and compare. Report the mean item standard error, the mean
child standard error and the separation reliability.

`r hide("Solution")`

```{r irt-challenge-solution, echo=TRUE}
trimmed <- responses[, !colnames(responses) %in% c("place_value", "estimate_sum")]

trimmed_fit    <- RM(trimmed)
trimmed_person <- person.parameter(trimmed_fit)

c(item_se     = round(mean(trimmed_fit$se.beta), 2),
  child_se    = round(mean(trimmed_person$se.theta[[1]]), 2),
  reliability = round(SepRel(trimmed_person)$sep.rel, 3))
```

The mean item standard error falls from 0.59 to 0.52, because the two dropped
tasks were the two worst-measured ones and they are no longer in the average. The
mean child standard error *rises* slightly, from 0.74 to 0.77, and the
reliability is unchanged at 0.72.

That is the honest answer, and it is more interesting than a clean win. Removing
an off-target item removes an embarrassment from the item table but does not help
the children, because an item nobody or everybody gets right was contributing
almost no information about them in the first place. The way to help the children
is to *add* well-targeted tasks, not to delete bad ones.
`r unhide()`
::::

### VERA as an IRT Application {#irt-vera}

VERA stands for *VERgleichsArbeiten*. The assessments are written in Years 3 and
8, they are not graded, and they are not meant to rank schools. How each of the
sixteen *Länder* actually runs them — who marks, who sees the results, whether
participation is compulsory — differs enough to fill a book chapter of its
own.[^irt-tarkian] Their purpose is
to relate what pupils can do to the national educational standards and the
competence-level models built on them.

The IQB describes competence as a continuous scale divided into a small number of
substantively interpreted levels. Item difficulties and pupil competence live on
the same metric, and the levels are what make that abstract metric usable in a
staff meeting. The whole sequence is the one this section has been building:

$$
\text{educational standards}
\longrightarrow \text{items}
\longrightarrow \text{IRT scale}
\longrightarrow \text{competence levels}
\longrightarrow \text{feedback for teaching}.
$$

A Rasch model is a natural starting point for VERA-style scoring: many responses
are scored right or wrong, comparability across *Länder* and years is the whole
point, and the item bank has to be linked over time. It should not be assumed
automatically, though. Research on VERA data has compared Rasch, 1PL, 2PL and
multidimensional models, and "VERA uses IRT" settles none of the choices in
Section \@ref(irt-three-choices).

VERA and PISA also report for different reasons, and this matters for the scoring
rule. PISA plausible values are built for population estimates and secondary
analysis. VERA feedback may include an individual estimate and a competence-level
classification meant for instruction. The public IQB materials do not name one
universal EAP, MAP or WLE rule for every *Land* and every cycle, so it would be
unsafe to describe all VERA person scores as one particular estimator without the
technical documentation for that implementation.

::: {.analysis-note}
The IQB is explicit that an individual pupil's statistical competence value can be
returned when the difficulties of the administered items are known — the anchoring
of Section \@ref(irt-anchoring) — and equally explicit that such a value needs
other evidence beside it before anything is concluded about one child. Both halves
of that sentence follow directly from the standard errors in Section
\@ref(irt-how-many).
:::

#### The whole process on one page {#irt-vera-overview -}

Everything from here to Section \@ref(irt-unambiguous) is a walk through one
picture. It is worth having in front of you, because the three lanes run on three
completely different clocks, and most of the confusion about VERA comes from
mixing them up.

```{r irt-flow-figure, echo=FALSE, fig.cap="The VERA machinery, from the standards to a competence level on a report. The top lane happens once per generation of educational standards, the middle lane once a year for each subject, and the bottom lane every May in every school. The dashed arrows are the only two things that travel between the lanes: a metric and a table of numbers.", fig.align='center', out.width="100%"}
flow_box <- function(x, y, label, fill) data.frame(x = x, y = y, label = label, fill = fill)
bw <- 1.72; bh <- 0.95

flow_boxes <- rbind(
  flow_box(1.1, 7.4, "KMK\nBildungsstandards",                  "a"),
  flow_box(3.2, 7.4, "Aufgaben-\nentwicklung\n(2-3 years)",      "a"),
  flow_box(5.3, 7.4, "Normierungs-\nstudie\n~12,000 pupils",     "a"),
  flow_box(7.4, 7.4, "national scale\nM = 500, SD = 100",        "a"),
  flow_box(9.5, 7.4, "Kompetenz-\nstufenmodell\n(expert panel)", "a"),
  flow_box(1.1, 4.4, "Pilotierung\n~2,800 pupils\n20 booklets",  "b"),
  flow_box(3.2, 4.4, "free Rasch\ncalibration\n(ConQuest)",      "b"),
  flow_box(5.3, 4.4, "anchor booklet\nmean/mean\nlinking",       "b"),
  flow_box(7.4, 4.4, "shift to\nRP = .625\n(+0.51 logits)",      "b"),
  flow_box(9.5, 4.4, "Itemkennwerte-\ntabelle\nlogits + points", "b"),
  flow_box(1.1, 1.4, "the class\nsits the test\n0/1 per item",   "c"),
  flow_box(3.2, 1.4, "item parameters\nFIXED\n(not estimated)",  "c"),
  flow_box(5.3, 1.4, "WLE per pupil\nin logits\n+ standard error","c"),
  flow_box(7.4, 1.4, "525 + (100/1.105)\nx theta\n= points",     "c"),
  flow_box(9.5, 1.4, "Kompetenzstufe\n+ feedback\nto the school","c")
)

column_x <- c(1.1, 3.2, 5.3, 7.4, 9.5)
flow_across <- do.call(rbind, lapply(c(7.4, 4.4, 1.4), function(y)
  data.frame(x = column_x[-5] + bw/2, xend = column_x[-1] - bw/2, y = y, yend = y)))
flow_down <- data.frame(
  x    = c(7.4, 9.5),          xend = c(5.3, 3.2),
  y    = c(7.4 - bh/2, 4.4 - bh/2), yend = c(4.4 + bh/2, 1.4 + bh/2),
  note = c("the metric travels down", "the item parameters travel down"),
  nx   = c(7.0, 7.2),          ny   = c(5.9, 2.9))

flow_lanes <- data.frame(
  y = c(7.4, 4.4, 1.4),
  label = c("ONCE PER GENERATION OF STANDARDS",
            "EVERY YEAR, FOR EACH SUBJECT",
            "EVERY MAY, IN EVERY SCHOOL"))

ggplot() +
  geom_rect(data = flow_boxes,
            aes(xmin = x - bw/2, xmax = x + bw/2, ymin = y - bh/2, ymax = y + bh/2, fill = fill),
            colour = "grey45", linewidth = 0.35) +
  geom_text(data = flow_boxes, aes(x, y, label = label), size = 2.65, lineheight = 0.95) +
  geom_segment(data = flow_across, aes(x, y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.13, "cm"), type = "closed"),
               linewidth = 0.35, colour = "grey35") +
  geom_curve(data = flow_down, aes(x, y, xend = xend, yend = yend),
             arrow = arrow(length = unit(0.13, "cm"), type = "closed"),
             linewidth = 0.35, colour = "grey35", curvature = 0.22, linetype = 2) +
  geom_text(data = flow_down, aes(x = nx, y = ny, label = note),
            size = 2.6, fontface = "italic", colour = "grey35") +
  geom_text(data = flow_lanes, aes(x = 0.1, y = y + bh/2 + 0.45, label = label),
            hjust = 0, size = 2.9, fontface = "bold", colour = "grey30") +
  scale_fill_manual(values = c(a = "#e8eef6", b = "#eef3e8", c = "#f6efe8"), guide = "none") +
  scale_x_continuous(limits = c(0.05, 10.5)) +
  scale_y_continuous(limits = c(0.7, 8.6)) +
  theme_void()
```

```{r irt-flow-table, echo=FALSE}
knitr::kable(
  data.frame(
    Lane = c("Norming", "Piloting", "Testing"),
    `How often` = c("once per generation of standards, roughly every 15 years",
                    "every year, separately for every subject",
                    "every year, in every participating school"),
    `Who takes part` = c("a sample representative of Germany, all 16 Länder",
                         "randomly drawn classes in several Länder",
                         "whole cohorts, Year 3 and Year 8"),
    `What comes out` = c("the point scale and the competence level model",
                         "item parameters on that scale",
                         "a competence value per pupil"),
    `Where it is described` = c(
      "[the norming study](items.html#irt-norming) and [where zero is](items.html#irt-origin)",
      "Sections \\@ref(irt-calibration) and \\@ref(irt-bista)",
      "Sections \\@ref(irt-anchoring) and \\@ref(irt-unambiguous)"),
    check.names = FALSE
  ),
  caption = "The three clocks. A pupil's competence value in May 2026 rests on item parameters estimated in 2025 and on a scale fixed years earlier on entirely different children."
)
```

::: {.analysis-takeaway}
Two arrows carry everything between the lanes. The **metric** comes down from the
norming study, and it is a definition: it says where zero is and what one point is
worth. The **item parameters** come down from the pilot study, and they are
estimates with standard errors. Everything else stays inside its own lane. In
particular, nothing travels *upwards*: the two million pupils who sit VERA in a
given year do not improve the item parameters, and they do not re-norm the scale.
:::

### Calibration: Where the Difficulties Come From {#irt-calibration}

Section \@ref(irt-how-many) ended with a wall: one class cannot calibrate items.
And yet VERA hands a competence value back for a single child in a single class.
Both are true, and the bridge between them is the word **calibration**.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
**Calibration** is the estimation of item parameters on a defined scale, using a
sample large enough to make them stable. It happens once, in advance, on other
people.

**Scoring** is the estimation of a person's competence *given* those item
parameters. It happens later, for each new person, and needs no sample at all.

A calibrated item is a measuring instrument that has been read against a
standard. Once it has been, using it is a look-up, not a study.
::::

The IQB says it in one sentence in its VERA FAQ: *"Die Schwierigkeit einer
VERA-Aufgabe wird durch eine vorhergehende Erprobung ('Pilotierung') an einer
umfangreichen Stichprobe von Schüler:innen ermittelt."* The task's difficulty is
established beforehand, on a large sample. By the time the sheet of paper reaches
Frau Berger's class, the number attached to each task has already been measured
on thousands of other children.

#### The study that produces the numbers {#irt-norming -}

Two studies stand behind every VERA item, on two different clocks, and it is worth
seeing both with real numbers rather than as words.

Item development itself takes two to three years, from drafts written with teachers
and subject-matter specialists, through pre-pilots in single classes, to the pilot
study proper.[^irt-iqb-development]

**The Pilotierung happens every year.** For VERA-8 mathematics in the 2025/26
cycle, the pilot ran in spring 2025 and looked like this:[^irt-techreport]

```{r irt-pilot-table, echo=FALSE}
knitr::kable(
  data.frame(
    Land = c("Baden-Württemberg", "Bremen", "Hamburg", "Niedersachsen",
             "Nordrhein-Westfalen", "Saarland", "Sachsen-Anhalt", "Thüringen", "Total"),
    `non-grammar classes` = c(15, 3, 7, 13, 22, 4, 6, 4, 74),
    `Gymnasium classes`   = c(13, 3, 7,  8, 20, 3, 4, 4, 62),
    `all classes`         = c(28, 6, 14, 21, 42, 7, 10, 8, 146),
    check.names = FALSE
  ),
  caption = "The realised sample of the VERA-8 mathematics pilot study 2025: 146 classes in eight Länder, 2,827 pupils, average age 14.6 years, 48.9 per cent female. Eight Länder rather than sixteen, because a pilot has to be precise about items, not representative of Germany."
)
```

Those 2,827 pupils saw **219 items** — 160 from 75 newly developed tasks, plus 59
from 35 older tasks brought back for comparison. Nobody saw all of them: the items
were grouped into blocks, the blocks assigned to **twenty booklets** in a
multi-matrix design, and each pupil worked through six blocks in eighty minutes.
A twenty-first booklet contained only already-normed items and did the linking.

Of the 219 items piloted, **27 went into the final test modules**. The rest either
missed the psychometric thresholds of Section \@ref(irt-checks) or were simply not
needed; the modules were filled up with items from the pilots of 2019, 2021, 2022,
2023 and 2024. An operational item pool is a slowly accumulating thing, and any
given year contributes a small fraction of it.

**The Normierungsstudie happens about once per generation of standards.** The
current Sek I metric was fixed in norming studies in 2006 and 2007 for mathematics
and in 2008 for English, and re-normed onto the complete ninth year group in the
IQB-Ländervergleich 2012, and the technical report gives the resulting standard
deviation as 1.105 logits. The primary metric is anchored to the
2011 population of fourth-graders.[^irt-bista-metric] Those are the numbers behind
the $1.105$ and the $500$ in Section \@ref(irt-bista): a scale in use today rests
on children who sat a test between fifteen and twenty years ago.

The KMK revised the Bildungsstandards in 2022, so the whole thing is currently
being rebuilt. New tasks were written from summer 2022, and the
**Normierungsstudie 2026** ran from early March to early May 2026: roughly
**12,000 pupils in Years 3 and 4, in about 455 general and special schools, across
all sixteen Länder**, in German and mathematics. Its data will fix the new point
scales and provide the basis for the new competence level models, which go into
use from 2027.[^irt-norming-2026]

```{r irt-two-studies, echo=FALSE}
knitr::kable(
  data.frame(
    ` `            = c("Purpose", "How often", "Sample", "Länder", "Produces"),
    `Pilotierung`  = c("estimate item parameters",
                       "every year, per subject",
                       "~2,800 pupils, 146 classes",
                       "several (eight in 2025)",
                       "difficulties for the items of the next cycle"),
    `Normierung`   = c("fix the scale and the competence levels",
                       "once per generation of standards, roughly every 15 years",
                       "~12,000 pupils, ~455 schools",
                       "all sixteen",
                       "the metric itself and the Kompetenzstufenmodell"),
    check.names = FALSE
  ),
  caption = "The two studies that stand behind every VERA result. The pilot is precise about items; the norming study is representative of Germany. Confusing them is the commonest way to misread what a VERA number is."
)
```

::: {.analysis-note}
Note what this means for the phrase "500 points". It is not the average of the
children who sat VERA this year, and not the average of Germany today. It is the
average of the norming sample of a particular year, carried forward by anchor items
ever since. A cohort effect, a curriculum change or a pandemic does not move the
scale; it moves where the children sit on it. That is the entire point of a fixed
metric — and the reason a trend can be reported at all.
:::

The statistical core of a norming study is one `RM()` call on a big matrix. Here is
a stand-in with three thousand children:

```{r irt-norming-study, echo=TRUE}
set.seed(1907)

norming_theta <- rnorm(3000, mean = 0, sd = 1)
norming_prob  <- plogis(outer(norming_theta, difficulty, "-"))

norming_data <- matrix(
  rbinom(length(norming_prob), size = 1, prob = norming_prob),
  nrow     = 3000,
  dimnames = list(NULL, names(difficulty))
)

scorable    <- rowSums(norming_data) > 0 & rowSums(norming_data) < 14
norming_fit <- RM(norming_data[scorable, ])

norming_b <- -coef(norming_fit)
names(norming_b) <- sub("beta ", "", names(coef(norming_fit)))
```

```{r irt-norming-table, echo=FALSE}
knitr::kable(
  data.frame(difficulty = norming_b, std_error = norming_fit$se.beta),
  digits  = 3,
  caption = "The same fourteen tasks, calibrated on three thousand children instead of twenty-four. The standard errors are between 0.04 and 0.07 logits, against 0.47 to 0.99 in the classroom of Section \\@ref(irt-r-example). This is what a Pilotierung buys."
)
```

Twenty-four children gave a mean item standard error of 0.59 logits. Three
thousand give 0.047. That is the calibration lever of
Section \@ref(irt-how-many), pulled as far as it goes.

#### Somebody has to choose where zero is {#irt-origin -}

Now the part that trips everyone up, and that no amount of extra data fixes.

The Rasch model says $P = \operatorname{logit}^{-1}(\theta_p - b_i)$. Add three to
every competence and three to every difficulty, and every predicted probability is
unchanged. **The data can never identify the origin of the scale.** They identify
differences and nothing else.

So a convention is imposed. `eRm` uses the sum-zero convention: the difficulties
are shifted so that they average to zero, which puts the origin at "the average of
*these* tasks". That is convenient and arbitrary, and it is the reason two
analyses of the same children can produce different numbers without either of them
being wrong.

A real assessment programme does not leave the origin at the average of a task
set, because that would move whenever the task set moved. It pins the origin to a
**population**: the mean of the norming sample.

```{r irt-official-scale, echo=TRUE}
# Where does the norming population sit on the sum-zero scale?
norming_person <- person.parameter(norming_fit)
origin_shift   <- mean(norming_person$theta.table[, 1], na.rm = TRUE)

# Move the origin so that the average child in Germany scores zero.
official <- norming_b + origin_shift

round(c(origin_shift = origin_shift), 3)
```

The IQB does exactly this and then rescales, so that the norming sample has a mean
of 500 and a standard deviation of 100 on the Bildungsstandards metric. A "500"
in a VERA report is not a raw quantity of mathematics. It is the sentence *"at the
average of the German children who sat the norming study"*, written as a number.

::: {.analysis-takeaway}
Item difficulties are only defined up to a common constant. Everything a
calibration report says about *where* an item or a child sits is a statement
relative to a chosen origin — usually a population, sometimes a task set. Two
numbers can only be compared if they were placed against the same origin, and
that, not the estimation, is what linking is about.
:::

#### Five ways onto an existing scale {#irt-linking-designs -}

Once an official scale exists, every later test has to be attached to it. There
are five standard ways, and they are the "so many variants" that make this
literature hard to enter.

```{r irt-linking-table, echo=FALSE}
knitr::kable(
  data.frame(
    Approach = c("Separate / free calibration",
                 "Fixed-parameter calibration",
                 "Partially fixed (anchored) calibration",
                 "Separate calibration plus a linking transformation",
                 "Concurrent calibration"),
    `What is estimated` = c(
      "all item parameters, from your own data",
      "nothing about the items; only person scores and the population",
      "the new items only; the anchor items are held at their official values",
      "all items in each sample, then a constant is applied to one of them",
      "all items in both samples at once, in one pooled analysis"
    ),
    `Common items needed` = c("none", "none — the items themselves are the link",
                              "yes", "yes", "yes"),
    `Resulting scale` = c("your own, comparable to nothing",
                          "the official one",
                          "the official one",
                          "the official one, after the transformation",
                          "one shared scale for both samples"),
    `Typical use` = c("a first look at your own data",
                      "VERA, TIMSS and PISA scoring of pupils",
                      "adding new items to an existing bank",
                      "linking two studies that were analysed separately",
                      "building a trend across cycles"),
    check.names = FALSE
  ),
  caption = "Five calibration designs. Only the first one produces a scale of your own; the other four are different routes to the same official metric, and which one is available depends on whether your data share items with the reference calibration."
)
```

Two of the five carry almost all the traffic.

**Fixed-parameter calibration** is what a school assessment does with pupils. The
item parameters are taken from the official table and not touched; the only things
estimated are the pupils. This needs no common-item design *because the items
themselves are the common items* — the class sat the calibrated tasks.

**Partially fixed calibration with anchor items** is how an item bank grows. A new
booklet mixes tasks whose difficulties are already known (the **anchors**) with
tasks that are not. The anchors hold the scale in place while the new tasks are
estimated onto it.

#### What it costs to re-estimate {#irt-free-vs-fixed -}

Suppose Frau Berger ignores the official table and calibrates on her own class.
With all fourteen tasks the damage is modest:

```{r irt-own-calibration, echo=TRUE}
own_fit <- RM(responses)
own_b   <- -coef(own_fit)
names(own_b) <- sub("beta ", "", names(coef(own_fit)))
own_b   <- own_b - mean(own_b)

own_scores   <- t(apply(responses, 1, score_pattern, b = own_b))
fixed_scores <- t(apply(responses, 1, score_pattern, b = official))

round(c(own_calibration   = mean(own_scores[, "WLE"]),
        official_scale    = mean(fixed_scores[, "WLE"]),
        difference        = mean(own_scores[, "WLE"]) - mean(fixed_scores[, "WLE"]),
        rank_correlation  = cor(own_scores[, "WLE"], fixed_scores[, "WLE"])), 3)
```

The children are ranked identically ($r = .999$ — the [item table](items.html#irt-step-items)
explained why), and the whole class has drifted 0.22 logits upwards. Not a
disaster, but already enough to turn "slightly below average" into "average".

Now the realistic version. The four hardest tasks discouraged the class, so she
drops them and calibrates on the remaining ten.

```{r irt-own-calibration-subset, echo=TRUE}
ten_easiest <- names(sort(official))[1:10]
short_form  <- responses[, ten_easiest]

subset_fit <- RM(short_form)
subset_b   <- -coef(subset_fit)
names(subset_b) <- sub("beta ", "", names(coef(subset_fit)))
subset_b   <- subset_b - mean(subset_b)

round(c(mean_official_difficulty = mean(official[ten_easiest]),
        own_calibration = mean(apply(short_form, 1, function(r) score_pattern(r, subset_b)["WLE"])),
        official_scale  = mean(apply(short_form, 1, function(r) score_pattern(r, official[ten_easiest])["WLE"]))), 2)
```

On her own scale the class averages **+0.81**. On the official scale, from exactly
the same answers, it averages **−0.27**. The children have gained **1.08 logits**
of imaginary competence, and the reason is written in the first number: the ten
tasks she kept are on average 0.94 logits easier than the average German task, and
free calibration silently redefined "average" to mean "average of these easy
tasks".

::: {.analysis-warning}
This is the single most common way a Rasch analysis goes wrong in practice, and it
never announces itself. Every diagnostic looks fine, the item table is sensible,
the reliability is respectable, and the scale is simply somewhere else. If
official parameters exist, fix them.
:::

#### Anchor items, and how to check them {#irt-anchor-items -}

Now the case where re-estimating is not only allowed but required: the school has
written four new tasks and wants them on the national scale. The design is a
booklet that mixes **six official anchor tasks** with the four new ones, given to
the whole year group.

```{r irt-linking-design, echo=TRUE}
anchor_items <- c("number_line", "sub_2digit", "times_table",
                  "word_problem", "symmetry", "clock_read")

new_items <- c(fractions = -0.6, money = 0.1, area = 0.9, two_step = 1.6)

set.seed(5544)
year_group    <- rnorm(192, mean = 0.3, sd = 1.1)
booklet       <- c(difficulty[anchor_items], new_items)
booklet_prob  <- plogis(outer(year_group, booklet, "-"))

booklet_data <- matrix(
  rbinom(length(booklet_prob), size = 1, prob = booklet_prob),
  nrow = 192, dimnames = list(NULL, names(booklet))
)

usable      <- rowSums(booklet_data) > 0 & rowSums(booklet_data) < ncol(booklet_data)
booklet_fit <- RM(booklet_data[usable, ])

local_b <- -coef(booklet_fit)
names(local_b) <- sub("beta ", "", names(coef(booklet_fit)))
```

`local_b` is on the school's own sum-zero scale. The anchors are the bridge: they
have a known official difficulty and a freshly estimated local one, and in a Rasch
model the two can differ by only one thing — a constant.

```{r irt-linking-constant, echo=TRUE}
linking_constant <- mean(official[anchor_items]) - mean(local_b[anchor_items])
linked           <- local_b + linking_constant

round(c(linking_constant = linking_constant), 3)
```

That is **mean/mean linking**, and in a Rasch model it is the whole method. Before
using it, check that the anchors deserve the name:

```{r irt-anchor-check, echo=TRUE}
knitr::kable(
  data.frame(
    linked   = linked[anchor_items],
    official = official[anchor_items],
    drift    = linked[anchor_items] - official[anchor_items],
    se       = booklet_fit$se.beta[match(anchor_items, names(local_b))]
  ),
  digits  = 3,
  caption = "The anchor check. Each anchor's freshly estimated difficulty, moved onto the official scale, against the value it is supposed to have. The largest discrepancy is 0.15 logits against standard errors of about 0.17, so nothing here is drifting."
)
```

An anchor whose drift is several standard errors has **item parameter drift**: the
task no longer behaves as it did when it was calibrated — a curriculum change, a
translation, a different position in the booklet, a leak. Such an item is dropped
from the anchor set before the constant is computed, never afterwards.

::: {.analysis-warning}
Fixed parameters are only as good as the conditions under which they were fixed.
An item difficulty travels between studies when the item is administered in the
same language, the same format, the same position and to a comparable population —
and stops travelling when any of those change. A task calibrated on paper in 2015
is not automatically the same task on a tablet in 2026, which is precisely what
made the PISA 2015 mode change so hard to absorb: the scaling model changed at the
same time as the medium, and separating the two effects took years of reanalysis.
:::

With the anchors cleared, the four new tasks arrive on the national scale:

```{r irt-new-items-placed, echo=TRUE}
knitr::kable(
  data.frame(
    placed = linked[names(new_items)],
    se     = booklet_fit$se.beta[match(names(new_items), names(local_b))]
  ),
  digits  = 3,
  caption = "Four school-written tasks, calibrated onto the national metric through six anchor items and 184 children. Standard errors around 0.16 logits — good enough to use, not good enough to become anchors themselves."
)
```

:::: {.amazing}
::: {.titelamazing}
<h2>In a Rasch model, linking is one number</h2>
:::
Two calibrations of the same items can differ by a shift, and that is all. So
linking is a subtraction, computed from the anchors, and a Year 3 pupil could
check it.

In a 2PL model the two calibrations can differ by a shift *and* a stretch, because
each item also has a slope that has been estimated on a different scale. Linking
then needs two constants, and there is a small industry devoted to finding them —
mean/sigma, Haebara, Stocking–Lord — each optimising a different notion of
agreement, each giving a slightly different answer.

This is the practical reason item banks and trend studies have loved the Rasch
model for sixty years. It is not that the model fits better. It is that when you
have thousands of items, dozens of booklets and twenty years of cycles to hold
together, the difference between one constant and two is the difference between
arithmetic and a research programme.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
A colleague reports that his class averages 0.8 logits on "the Rasch scale". Before
that number means anything you have to ask
`r mcq(c("which estimator he used", answer = "against which calibration the origin was fixed", "how many items the test had", "whether the model fitted"))`.

Frau Berger's ten easy tasks are on average 0.94 logits easier than the national
average task. Re-estimating on them alone therefore inflated her class by about
`r fitb(1.0, num = TRUE, tol = 0.2)` logits.

An anchor item comes out 0.9 logits away from its official difficulty, with a
standard error of 0.15. The right response is
`r mcq(c("apply the linking constant anyway", "shift the official value to match", answer = "drop it from the anchor set and investigate the item", "increase the sample size"))`.

True or false: fixed-parameter scoring requires the class to be a random sample of
the population the items were calibrated on. `r torf(FALSE)`
`r hide("Why not?")` That is the point of specific objectivity. The item
parameters were established elsewhere; a child's estimate given those parameters
depends only on that child's answers. What *does* depend on the population is the
prior used by EAP and MAP — which is one more reason to report WLE for an
individual child. `r unhide()`
::::

### From Logits to Bildungsstandards Points {#irt-bista}

A fair objection at this point: everything the model has produced is a number like
$-0.26$, and nobody has ever been told anything useful by $-0.26$.

So what exactly *does* come out of a Rasch model, and does the model work at all
without a scale like the Bildungsstandards behind it?

#### What the model hands you {#irt-what-comes-out -}

Two things, and it is worth being precise about both.

**Per item, a difficulty $b_i$.** One number, in logits.

**Per person, a whole posterior distribution** — the curve in Figure
\@ref(fig:irt-posterior-figure). WLE, EAP and MAP are *summaries* of that curve
that you chose; they are not what the model produced. The model produced the
curve, and the standard error is the part of it people throw away first.

Both live in logits, and both are identified only **up to a common additive
constant** — see [somebody has to choose where zero is](items.html#irt-origin) below.
So the honest description of the output is:
*a set of distances along an axis whose zero has not yet been chosen*.

::: {.analysis-note}
Altitude works the same way. "Three hundred metres" is meaningless until somebody
says *above what* — sea level, the valley floor, the car park. The metre is
perfectly well defined; the datum is a convention. A logit is the metre. The
Bildungsstandards metric is the sea level.
:::

#### A difference means something; a value does not {#irt-logit-meaning -}

This part is often skipped, and it matters, because it is not true that logits are
meaningless. A logit *difference* has an exact and universal interpretation that
needs no scale, no population and no committee:

$$
\theta_p - b_i = d \quad\Longrightarrow\quad
\frac{P(\text{correct})}{P(\text{wrong})} = e^{d}.
$$

One logit multiplies the odds by 2.72. Half a logit by 1.65. That statement is
true in mathematics in Year 8, in English reading and in a depression inventory,
and it is true without anybody having normed anything.

What has no meaning is a bare *value*. "Marie is at $0.68$" says nothing until the
sentence is finished — above what? Two ways to finish it exist, and every
reporting system in the world uses one, the other, or both.

```{r irt-reference-table, echo=FALSE}
knitr::kable(
  data.frame(
    Reference = c("Norm-referenced", "Criterion-referenced"),
    `The sentence` = c("above what fraction of a defined population",
                       "which tasks this person can be expected to solve"),
    `What has to be fixed` = c("a reference population and a reporting metric",
                               "a set of described tasks and a probability criterion"),
    `Example` = c("500 points = the average German ninth-grader",
                  "Kompetenzstufe III = can solve these kinds of task"),
    `Fails when` = c("the population changes or is the wrong comparison",
                     "the item pool does not cover that region of the scale"),
    check.names = FALSE
  ),
  caption = "The two ways a scale value acquires meaning. Almost every assessment report uses both at once, which is why the same number gets described as 'average' and as 'level III' in one sentence."
)
```

The second row is where item response theory earns something classical test theory
cannot offer. Because persons and items sit on **one axis**, a person's position
can be described by naming the tasks around it. A percentage-correct score cannot
do that: it is a property of a person *and* a test, and it has no location that a
task could be placed next to.

#### How VERA actually does it {#irt-bista-metric -}

This is not a thought experiment. The IQB publishes a technical report for every
VERA-8 cycle and subject — the same documents behind the pilot figures in
Section \@ref(irt-calibration) — and each one states the whole chain, formulas
included. It is worth walking through, because every step in it
has appeared somewhere in this section.

**1. Pilot the tasks.** For VERA-8 mathematics 2026, 75 newly developed tasks
(160 items) plus 35 older tasks (59 items) were piloted on **2,827 pupils** in 146
classes across eight *Länder*, in a **multi-matrix design**: twenty booklets, each
pupil seeing only a fraction of the pool, which is the
[booklet design](items.html#irt-check-missing) among the checks of Section \@ref(irt-checks).

**2. Include an anchor booklet.** A twenty-first booklet contained *only* items
already normed from the IQB pool. That is the [anchor design](items.html#irt-anchor-items) from the previous
section, used for exactly the reason given there: to attach the new items to an
existing metric and to link the booklets to each other.

**3. Calibrate freely.** One calibration run with the **one-dimensional Rasch
model** in ConQuest, all item parameters estimated, missing and invalid responses
scored as wrong, and the mean person ability fixed to zero — the
[origin convention](items.html#irt-origin), chosen explicitly and stated in the report.

**4. Link with mean/mean equating.** The anchor items' fresh difficulties are
compared with their values from the 2007 norming study, and the constant that
reconciles them is applied to everything else. Mean/mean, one constant, exactly as in
[anchor items and how to check them](items.html#irt-anchor-items).

**5. Shift the items to a response probability of .625.**

$$
b^{*} = b + \ln\!\left(\frac{0.625}{1-0.625}\right) = b + 0.51
$$

**6. Transform to reporting points.**

$$
\text{BiSta} = 525 + \frac{100}{1.105}\left(b^{*} - M_{\text{ref}}\right)
\qquad\text{for items, and}\qquad
\text{BiSta} = 525 + \frac{100}{1.105}\,\hat\theta
\qquad\text{for persons.}
$$

$1.105$ is the standard deviation, in logits, of German ninth-graders in the
2006/2007 norming — the [latent standard deviation](items.html#irt-bista-latent),
cleaned of measurement error. Dividing by it and multiplying by 100 is what makes a
standard deviation worth 100 points. The offset is 525 rather than 500 for a
historical reason the report explains: until 2014 the scale was normed on pupils
aiming for at least a *Mittlerer Schulabschluss*, and re-norming on the complete
ninth year in 2012 moved everything down by 25 points.

English uses the same machinery with its own reference statistics:

$$
\text{BiSta}_{\text{reading}} = 500 + \frac{100}{1.219}\,(\hat\theta + 0.673),
\qquad
\text{BiSta}_{\text{listening}} = 500 + \frac{100}{1.213}\,(\hat\theta + 0.514).
$$

```{r irt-bista-functions, echo=TRUE}
rp_shift <- log(0.625 / (1 - 0.625))          # 0.51 logits

bista_maths   <- function(logit) 525 + (100 / 1.105) *  logit
bista_reading <- function(logit) 500 + (100 / 1.219) * (logit + 0.673)

round(c(rp_shift          = rp_shift,
        one_logit_maths   = 100 / 1.105,
        one_logit_reading = 100 / 1.219), 2)
```

A logit is worth 90 points in mathematics and 82 in English reading. Nothing has
been estimated here; the scale has been relabelled and every distance is preserved.

::: {.analysis-warning}
A linear transformation cannot add information, and it can hide it. A standard
error of 0.47 logits becomes 43 points, which sounds smaller only because the
numbers got bigger. A report that gives points without their standard error has not
become more precise by rescaling; it has stopped saying how imprecise it is.
:::

#### The latent standard deviation {#irt-bista-latent -}

Step 6 needs the spread of the **latent** distribution, not of the estimates.
Estimates are inflated by measurement error, so the variance has to be cleaned
first — the same decomposition that separates true score from error everywhere
else:

```{r irt-latent-sd, echo=TRUE}
norming_wle <- norming_person$theta.table[, 1]
norming_se  <- norming_person$se.theta[[1]]

round(c(observed_variance   = var(norming_wle, na.rm = TRUE),
        mean_error_variance = mean(norming_se^2),
        latent_sd = sqrt(var(norming_wle, na.rm = TRUE) - mean(norming_se^2))), 3)
```

The observed spread of the estimates is 1.53; a third of it is measurement error;
the latent standard deviation is 1.02 — correct, because the simulated norming
population of [the study that produces the numbers](items.html#irt-norming) was
drawn with a standard deviation of
exactly 1. Skip that correction and every reported point value is stretched.

#### Item mapping: what a pupil at 500 can do {#irt-item-mapping -}

Now the criterion-referenced half, and it is the two lines from step 5.

A task sits at $b_i$, which is where it is solved with probability $.50$. Fifty per
cent is a weak notion of "can do", so reporting systems pick a higher **response
probability criterion** — the IQB uses $.625$, TIMSS and the German DESI study
used $.65$ — and locate the task where a person reaches that
probability.[^irt-hartig] Which moves the task up the scale by a constant, and
nothing else changes.

Here is a VERA-8-shaped mathematics module: 24 items, spread to cover the lower
and middle competence levels, in the manner the technical report describes for the
"A" version of the Basismodul.

```{r irt-vera8-maths-module, echo=TRUE}
maths_items <- setNames(seq(-2.2, 0.9, length.out = 24), sprintf("ma_%02d", 1:24))

item_locations <- bista_maths(maths_items + rp_shift)
round(range(item_locations))
```

```{r irt-vera8-item-map, echo=FALSE}
maths_cuts  <- c(355, 435, 515, 595, 675)
stufe_names <- c("I.a", "I.b", "II", "III", "IV", "V")
stufe_of    <- function(points) stufe_names[findInterval(points, maths_cuts) + 1]

knitr::kable(
  as.data.frame(table(Kompetenzstufe = factor(stufe_of(item_locations), levels = stufe_names))),
  col.names = c("Kompetenzstufe", "items in this module"),
  caption = "The twenty-four items of the module, mapped to competence levels at a response probability of .625. The module reaches from 372 to 653 points and spreads its items across four of the six bands - nothing at the very bottom, nothing at the very top, which is what a Basismodul in its 'A' version is for."
)
```

**That mapping is the competence level descriptor.** The prose in a
Kompetenzstufenmodell — *"pupils at this level can carry out multi-step operations
with variables, terms, equations and functions"* — is written by subject
specialists looking at exactly this: which tasks land in which band. The Wright map
of Figure \@ref(fig:irt-wright-map) is not an illustration of the reporting system;
it *is* the reporting system.

:::: {.amazing}
::: {.titelamazing}
<h2>The scale is empty, and the items fill it</h2>
:::
Put the two halves together and something slightly startling appears. The logit
scale has no content whatsoever — no origin, no substantive units, nothing but
distances. And a description of what a pupil can do is produced from it *entirely
by naming the tasks that sit nearby*.

The meaning does not come from the model and it does not come from the metric. It
comes from the item pool. Which is why a scale is only interpretable where it has
items: in a region with no tasks the number is still perfectly estimable and nobody
can say what it means. That is targeting — the [person-item map](items.html#irt-step-wright) — seen from
the other end — and the reason a competence level model is a piece of subject
didactics wearing a psychometric coat.
::::

#### Mathematics: the class report {#irt-ksm-maths -}

The KMK and IQB divide the mathematics scale into six bands, which carry the
standard names twice over — once for the *Erster Schulabschluss* and once for the
*Mittlerer Schulabschluss*, because one scale has to serve both.[^irt-ksm]

```{r irt-ksm-table, echo=FALSE}
knitr::kable(
  data.frame(
    Stufe = c("V", "IV", "III", "II", "I.b", "I.a"),
    Points = c("675 and above", "595-674", "515-594", "435-514", "355-434", "below 355"),
    `for the ESA` = c("Optimalstandard", "Optimalstandard", "Regelstandard plus",
                      "Regelstandard", "Mindeststandard", "below the Mindeststandard"),
    `for the MSA` = c("Optimalstandard", "Regelstandard plus", "Regelstandard",
                      "Mindeststandard", "below the Mindeststandard", "below the Mindeststandard"),
    check.names = FALSE
  ),
  caption = "The integrated competence level model for mathematics, in force since 2011. The bands are 80 points wide, and the boundaries were set by expert panels, not computed."
)
```

Now a Year 8 class sits the module, and is scored with **fixed item parameters**,
exactly as the technical report recommends.

```{r irt-vera8-maths-class, echo=TRUE}
logit_maths <- function(points) (points - 525) * 1.105 / 100

set.seed(826)
maths_class <- rnorm(23, mean = logit_maths(480), sd = 0.75)
names(maths_class) <- sprintf("S%02d", 1:23)

maths_prob <- plogis(outer(maths_class, maths_items, "-"))
maths_resp <- matrix(rbinom(length(maths_prob), 1, maths_prob), nrow = 23,
                     dimnames = list(names(maths_class), names(maths_items)))

maths_scores <- t(apply(maths_resp, 1, score_pattern, b = maths_items))
maths_points <- bista_maths(maths_scores[, "WLE"])
maths_se     <- (100 / 1.105) * maths_scores[, "SE_WLE"]
```

```{r irt-vera8-maths-report, echo=FALSE}
knitr::kable(
  data.frame(
    raw       = maths_scores[, "raw"],
    points    = round(maths_points),
    se        = round(maths_se),
    Stufe     = stufe_of(maths_points),
    truth     = stufe_of(bista_maths(maths_class))
  )[order(-maths_points), ],
  caption = "A Year 8 class on 24 mathematics items, reported the way VERA-8 reports: Bildungsstandards points, a standard error, a Kompetenzstufe. 'truth' is the band each pupil actually occupies, which only a simulation can reveal. Eight of the twenty-three are reported one band away from it."
)
```

```{r irt-vera8-maths-summary, echo=TRUE}
round(c(class_mean   = mean(maths_points),
        true_mean    = mean(bista_maths(maths_class)),
        mean_se      = mean(maths_se),
        band_width   = 80,
        wrong_band   = sum(stufe_of(maths_points) != stufe_of(bista_maths(maths_class)))), 1)
```

A class mean of 447 points against a truth of 442 — the aggregate is fine. The
individual standard error is 43 points against bands 80 points wide, so a pupil's
own interval reaches well into the neighbouring band, and eight of twenty-three end
up in the wrong one.

The IQB says this itself, in the technical report, and it is worth quoting in full
because it is the most honest paragraph in German educational assessment:

> *"Die Zuverlässigkeit bzw. Reliabilität der Messung und der damit verbundenen
> Diagnose ist beschränkt und weist ein breites Vertrauensintervall auf. Dies
> bedeutet, dass die erhaltene Punktschätzung der Personenfähigkeit nicht der
> tatsächlichen Fähigkeit der Schülerin oder des Schülers entsprechen muss."*

Everything in Section \@ref(irt-how-many) is a long version of that sentence.

#### English: the same machinery, European levels {#irt-ksm-english -}

VERA-8 tests English reading and listening, and English shows the criterion half
of this section especially clearly, because its bands were not invented for
Germany. They are **Common European Framework** levels, and the IQB model maps its
point scale onto them.[^irt-ksm-english] The full descriptors come from the
language study whose 2008 norming fixed the metric.[^irt-koeller]

```{r irt-cefr-table, echo=FALSE}
knitr::kable(
  data.frame(
    GER = c("C1", "B2.2", "B2.1", "B1.2", "B1.1", "A2.2", "A2.1", "A1.2", "A1.1"),
    Reading = c("700 and above", "650-699", "600-649", "550-599", "500-549",
                "450-499", "400-449", "350-399", "below 350"),
    Listening = c("740 and above", "675-739", "610-674", "545-609", "480-544",
                  "415-479", "350-414", "285-349", "below 285"),
    `for the MSA` = c("Optimalstandard", "Optimalstandard", "Regelstandard plus",
                      "Regelstandard", "Mindeststandard", "Mindeststandard",
                      "below the Mindeststandard", "below the Mindeststandard",
                      "below the Mindeststandard"),
    check.names = FALSE
  ),
  caption = "The integrated competence level model for English, in force since 2014. Each CEFR level is split in half, so the reading sub-levels are 50 points wide and the listening sub-levels 65 - narrower than the 80-point bands in mathematics, and that turns out to matter."
)
```

```{r irt-vera8-english, echo=TRUE}
set.seed(1808)

reading_items <- setNames(sort(rnorm(20, mean = -0.84, sd = 0.95)), sprintf("lv_%02d", 1:20))
logit_reading <- function(points) (points - 500) * 1.219 / 100 - 0.673

reading_class <- rnorm(26, mean = logit_reading(468), sd = 0.8)
names(reading_class) <- sprintf("S%02d", 1:26)

reading_prob <- plogis(outer(reading_class, reading_items, "-"))
reading_resp <- matrix(rbinom(length(reading_prob), 1, reading_prob), nrow = 26,
                       dimnames = list(names(reading_class), names(reading_items)))

reading_scores <- t(apply(reading_resp, 1, score_pattern, b = reading_items))
reading_points <- bista_reading(reading_scores[, "WLE"])
reading_se     <- (100 / 1.219) * reading_scores[, "SE_WLE"]

cefr_cuts <- c(350, 400, 450, 500, 550, 600, 650, 700)
cefr_name <- c("A1.1", "A1.2", "A2.1", "A2.2", "B1.1", "B1.2", "B2.1", "B2.2", "C1")
cefr_of   <- function(points) cefr_name[findInterval(points, cefr_cuts) + 1]

round(c(class_mean = mean(reading_points),
        true_mean  = mean(bista_reading(reading_class)),
        mean_se    = mean(reading_se),
        band_width = 50,
        wrong_band = sum(cefr_of(reading_points) != cefr_of(bista_reading(reading_class)))), 1)
```

```{r irt-cefr-report, echo=FALSE}
knitr::kable(
  data.frame(
    raw    = reading_scores[, "raw"],
    points = round(reading_points),
    se     = round(reading_se),
    GER    = cefr_of(reading_points),
    truth  = cefr_of(bista_reading(reading_class))
  )[order(-reading_points), ],
  caption = "Twenty English reading items, twenty-six Year 8 pupils, reported as CEFR sub-levels. The standard error is 42 points and the sub-levels are 50 points wide, so eleven of the twenty-six are reported one step away from where they are."
)
```

The English bands are half the width of the mathematics bands, because the CEFR
splits each of A1, A2, B1 and B2 in two. The measurement error did not shrink to
match. **Eleven of twenty-six pupils are placed one sub-level off** — and a
sub-level is precisely the resolution at which a CEFR statement about a pupil would
be made.

This is not an argument against reporting CEFR levels. It is an argument for
reporting them the way the technical report asks: as one indicator among several,
on the global scale, and never as a certificate.

::: {.analysis-takeaway}
So: does the Rasch model work without a scale like the Bildungsstandards? Yes,
completely — it answers every comparative question on its own, because comparisons
are differences and differences are exactly what the model identifies. What it
cannot do alone is say *what this pupil can actually do*. That answer is
manufactured afterwards from three ingredients that are not part of the model: a
reference population that fixes the origin, an item pool whose content is known,
and a panel of people who decide where the boundaries go. The logits are not
meaningless — they are unfinished.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
A pupil is 1.5 logits above an item's difficulty. The odds that she solves it are
about `r fitb(4.5, num = TRUE, tol = 0.4)` to one. `r hide("How?")` $e^{1.5} = 4.48$.
No scale, no population and no norming study was needed — this is the part of the
output that means something on its own. `r unhide()`

On the VERA-8 mathematics metric, one logit is worth
`r fitb(90, num = TRUE, tol = 2)` points, because 100 points is one standard
deviation of the reference population and that population has a spread of 1.105
logits.

Moving an item from a response probability of .50 to .625 shifts it up the scale by
`r fitb(0.51, num = TRUE, tol = 0.03)` logits.

The IQB competence level cut scores are best described as
`r mcq(c("the points where item characteristic curves cross", "quantiles of the population distribution", answer = "boundaries set by expert panels and then kept fixed", "the points where test information is highest"))`.

True or false: rescaling from logits to a 500/100 metric makes an individual
pupil's result more precise. `r torf(FALSE)`
::::

### The VERA Workflow, End to End {#irt-anchoring}

Everything is now in place to do what a German primary school actually does in
May. This is the minimal working example: **binary Rasch, fixed item parameters,
one class**.

#### Step 0: what arrives from the pilot study {#irt-vera-step-arrives -}

Not data. A table. The tasks and their difficulties, established beforehand on a
large sample, exactly as in Section \@ref(irt-calibration):

```{r irt-vera-item-table, echo=TRUE}
round(sort(official), 2)
```

Fourteen numbers on the national metric. Nothing in this section re-estimates
them.

#### Step 1: the class sits the test {#irt-vera-step-test -}

`responses` — twenty-four rows, fourteen columns, one 0 or 1 per cell. That is the
entire input.

#### Step 2: score against the fixed parameters {#irt-vera-step-score -}

```{r irt-vera-scoring, echo=TRUE}
vera_scores <- t(apply(responses, 1, score_pattern, b = official))

class_report <- data.frame(
  raw        = vera_scores[, "raw"],
  competence = vera_scores[, "WLE"],
  std_error  = vera_scores[, "SE_WLE"]
)
class_report$lower <- class_report$competence - 1.96 * class_report$std_error
class_report$upper <- class_report$competence + 1.96 * class_report$std_error

knitr::kable(
  class_report[order(-class_report$competence), ],
  digits  = 2,
  caption = "The class report. Competence is on the national scale, where 0 is the average child in the norming study. Every interval is more than two and a half logits wide, and seventeen of the twenty-four contain 0 — that is, the test cannot distinguish those seventeen children from an average German third-grader."
)
```

That last sentence is the one worth sitting with. The estimates are perfectly
usable, the ordering is right, and the individual intervals are enormous. Nothing
went wrong; this is what fourteen items can do.

#### Step 3: the class as a whole {#irt-vera-step-class -}

```{r irt-vera-class-mean, echo=TRUE}
round(c(class_mean   = mean(vera_scores[, "WLE"]),
        class_sd     = sd(vera_scores[, "WLE"]),
        se_of_mean   = sd(vera_scores[, "WLE"]) / sqrt(24)), 2)
```

A class mean of −0.26 logits with a standard error of 0.25. The class is, as far
as this test can tell, ordinary. And notice what happened: the *aggregate* is
measured far better than any individual in it, because twenty-four noisy estimates
average out. This is why VERA is designed as feedback for schools and teaching and
explicitly not as a ranking of children.

#### Step 4: competence levels {#irt-vera-step-levels -}

VERA does not report logits to teachers. It reports **Kompetenzstufen** — the
continuous scale cut into five or six substantively described bands. Cutting is
easy; the consequence is not.

```{r irt-vera-levels, echo=TRUE}
cut_scores <- c(-1.5, -0.5, 0.5, 1.5)          # five levels

level_of <- function(x) findInterval(x, cut_scores) + 1

# How much of the posterior actually falls in the level we report?
level_probability <- function(pattern, b) {
  grid <- seq(-6, 6, length.out = 1201)
  p    <- plogis(outer(grid, b, "-"))
  lik  <- apply(p, 1, function(q) prod(q^pattern * (1 - q)^(1 - pattern)))
  post <- lik * dnorm(grid)
  post <- post / sum(post)
  tapply(post, factor(level_of(grid), levels = 1:5), sum)
}

posterior_by_level <- t(apply(responses, 1, level_probability, b = official))
posterior_by_level[is.na(posterior_by_level)] <- 0

reported  <- level_of(vera_scores[, "WLE"])
certainty <- posterior_by_level[cbind(seq_len(24), reported)]

round(c(mean_certainty        = mean(certainty),
        below_50_per_cent     = sum(certainty < 0.5),
        wrong_level           = sum(reported != level_of(competence))), 2)
```

Twelve of the twenty-four children are placed in the wrong band, and the average
child's reported level carries only a 52% posterior probability of being the right
one. The cut scores here are a full logit apart, which is *more* generous than the
real Kompetenzstufen, so this is the optimistic version.

::: {.analysis-warning}
A competence level looks like a fact and is a rounded estimate. The rounding is
worst exactly where it matters most — for a child near a cut score, which is the
child a teacher is most likely to act on. The honest report of a level is the
level together with how sure it is, and almost no reporting system provides that.
:::

#### Step 5: one child, in March {#irt-vera-step-one-child -}

A new pupil joins in March and sits the same fourteen tasks. No re-calibration, no
class, no sample:

```{r irt-vera-one-child, echo=TRUE}
new_child <- c(count_back = 1, number_line = 1, place_value = 1, add_2digit = 1,
               sub_2digit = 1, times_table = 1, half_of = 0, word_problem = 1,
               shape_names = 0, symmetry = 0, clock_read = 0, measure_cm = 0,
               data_table = 0, estimate_sum = 0)

round(score_pattern(new_child, official), 2)
```

Seven of fourteen, a weighted-likelihood estimate of −0.18 logits on the national
scale, standard error 0.66. **One row of data, a number on the national metric** —
because three thousand other children were measured first.

That is the full answer to "would one pupil be enough?". One pupil cannot
calibrate anything, as [the minimum-children table](items.html#irt-one-pupil)
showed, and one pupil can be measured
perfectly well against a calibration somebody else paid for.

#### Step 6: and if the booklet were longer? {#irt-vera-step-longer -}

The lever from [the two-resources experiment](items.html#irt-two-resources),
applied to this class. The same
twenty-four children, the same competences, a thirty-task booklet on the same
official scale:

```{r irt-vera-longer, echo=TRUE, cache=TRUE}
set.seed(77)

long_booklet <- setNames(seq(-2.7, 2.4, length.out = 30), sprintf("task_%02d", 1:30))
long_prob    <- plogis(outer(competence, long_booklet, "-"))
long_data    <- matrix(rbinom(length(long_prob), 1, long_prob), nrow = 24,
                       dimnames = list(rownames(responses), names(long_booklet)))

long_scores <- t(apply(long_data, 1, score_pattern, b = long_booklet))

long_by_level <- t(apply(long_data, 1, level_probability, b = long_booklet))
long_by_level[is.na(long_by_level)] <- 0
long_reported  <- level_of(long_scores[, "WLE"])
long_certainty <- long_by_level[cbind(seq_len(24), long_reported)]

knitr::kable(
  data.frame(
    booklet          = c("14 tasks", "30 tasks"),
    mean_se          = c(mean(vera_scores[, "SE_WLE"]), mean(long_scores[, "SE_WLE"])),
    mean_ci_width    = c(mean(2 * 1.96 * vera_scores[, "SE_WLE"]),
                         mean(2 * 1.96 * long_scores[, "SE_WLE"])),
    mean_certainty   = c(mean(certainty), mean(long_certainty)),
    levels_under_50  = c(sum(certainty < 0.5), sum(long_certainty < 0.5)),
    wrong_level      = c(sum(reported != level_of(competence)),
                         sum(long_reported != level_of(competence)))
  ),
  digits  = 2,
  caption = "Same twenty-four children, same competences, same official scale, twice the booklet. The standard error falls from 0.70 to 0.49, the number of children whose reported competence level is less than half certain falls from twelve to three, and the number placed in the wrong band falls from twelve to eight."
)
```

Doubling the booklet does not make the class cleverer and does not improve a single
item parameter. It makes the school's knowledge about each individual child
roughly a third sharper — and cuts the number of children whose reported level is
essentially a guess from twelve to three.

```{r irt-vera-length-figure, echo=FALSE, fig.cap="The same twenty-four children measured twice: once with fourteen tasks, once with thirty. The thirty-task intervals are visibly shorter, and that is the whole benefit of a longer booklet. The second thing the picture shows is less comfortable — the two estimates for the same child sometimes differ by more than a logit, because both are measurements of an unchanged competence and both contain error.", fig.align='center', out.width="95%"}
ranking <- order(vera_scores[, "WLE"])

interval_frame <- function(scores, label) {
  data.frame(
    child   = factor(rownames(responses), levels = rownames(responses)[ranking]),
    est     = scores[, "WLE"],
    se      = scores[, "SE_WLE"],
    booklet = label
  )
}

rbind(interval_frame(vera_scores, "14 tasks"),
      interval_frame(long_scores, "30 tasks")) %>%
  ggplot(aes(est, child, colour = booklet)) +
  geom_vline(xintercept = 0, linetype = 3) +
  geom_errorbarh(aes(xmin = est - 1.96 * se, xmax = est + 1.96 * se),
                 height = 0, linewidth = 0.6, position = position_dodge(width = 0.6)) +
  geom_point(size = 1.6, position = position_dodge(width = 0.6)) +
  labs(x = "Competence on the national scale (logits), 95% interval",
       y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

::: {.analysis-note}
The second reading of that figure is the one to keep. Every child was measured
twice, with the same competence and two different booklets, and several of them
moved by more than a logit. Nothing changed about the children between the two
measurements. That is what a standard error of 0.7 looks like when you can
actually see it.
:::

::: {.analysis-takeaway}
This is the whole section in one workflow. The item difficulties came from a
Pilotierung on thousands of children, because that is the only way to get them
(Section \@ref(irt-how-many)). They were fixed, not re-estimated, because
re-estimating would move the origin and quietly change what the numbers mean
(see [what it costs to re-estimate](items.html#irt-free-vs-fixed)). A single
child could then be placed on the
national scale from one row of data. And how precisely that child was placed
depended on one thing only: how many tasks were on the sheet.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The new pupil in March got seven of fourteen and scored −0.18 logits. A second
pupil gets seven of fourteen but solves a *different* seven. Her estimate will be
`r mcq(c("higher, if she solved harder tasks", "lower, if she solved easier tasks", answer = "exactly −0.18", "impossible to say without the pattern"))`.
`r hide("Why?")` The raw score is sufficient (Section \@ref(why-rasch)). Under a
complete dichotomous Rasch test with fixed parameters, the pattern adds nothing to
the point estimate — though it does tell you something about *fit*, which is a
different question. `r unhide()`

Going from fourteen to thirty tasks cut the mean standard error from 0.70 to
about `r fitb(0.49, num = TRUE, tol = 0.04)`. That is close to the theoretical
factor of $\sqrt{14/30} =$ `r fitb(0.68, num = TRUE, tol = 0.03)`.

Now do it yourself. Take the same class, keep the official parameters fixed, and
score the children on only the six tasks whose official difficulty lies between
−1.5 and +1.5 — a short but *well-targeted* form. Compare the mean standard error
with the six *easiest* tasks.

`r hide("Solution")`

```{r irt-vera-targeting, echo=TRUE}
targeted <- names(official)[official > -1.5 & official < 1.5][1:6]
easiest  <- names(sort(official))[1:6]

se_of <- function(items) {
  sub <- responses[, items, drop = FALSE]
  mean(apply(sub, 1, function(r) score_pattern(r, official[items])["SE_WLE"]))
}

round(c(targeted = se_of(targeted), easiest = se_of(easiest)), 2)
```

Six well-targeted tasks give a mean standard error of 0.99 logits, six easy ones
1.16 — a fifteen per cent gain for no extra testing time at all. Targeting is worth
items: a short form that sits where the children sit is worth more than a longer
one that does not. This is the same lesson as the
person–item map in Figure \@ref(fig:irt-wright-map), now with a price attached.
`r unhide()`
::::

### Is the Result Unambiguous? {#irt-unambiguous}

A teacher, a parent and a ministry official ask the same question in three
different registers. *Is this number right?* *Does it mean my child is bad at
maths?* *Can we compare our Land to the neighbours?*

The answer is not yes or no. It is: **which number are you asking about**. Some of
what comes out of a VERA-style analysis is exact, some is stable, some is
reproducible less than half the time, and the differences are not intuitive.
This section takes them apart one at a time, because the intuitions people bring
to it are almost all wrong — including the two most common ones, that a bigger
data set makes an individual result safer, and that the class a pupil sits in has
nothing to do with it.

#### Does the class have to be pooled with anybody? {#irt-pooling -}

A real *Lerngruppe* can be one pupil. So: does scoring work only when many classes
are analysed together?

With fixed item parameters, no — and not approximately, exactly. Take one response
pattern and score it inside samples of wildly different size:

```{r irt-pooling-demo, echo=TRUE}
one_pupil <- rep(0, 24)
one_pupil[c(1:11, 14, 17)] <- 1
names(one_pupil) <- names(maths_items)

score_inside_sample <- function(n) {
  set.seed(500 + n)
  classmates <- rnorm(max(n - 1, 0), logit_maths(480), 0.75)
  if (n > 1) {
    p <- plogis(outer(classmates, maths_items, "-"))
    data <- rbind(one_pupil, matrix(rbinom(length(p), 1, p), nrow = n - 1))
  } else {
    data <- matrix(one_pupil, nrow = 1)
  }
  s <- score_pattern(data[1, ], maths_items)          # item parameters fixed
  c(sample_size = n, WLE = s[["WLE"]], points = bista_maths(s[["WLE"]]))
}

knitr::kable(
  t(sapply(c(1, 24, 100, 1000, 10000), score_inside_sample)),
  digits  = 6,
  caption = "The same pupil, scored alone and scored inside samples of up to ten thousand. Not approximately equal - equal to six decimal places, because the surrounding sample never enters the calculation."
)
```

The reason is in Section \@ref(why-rasch) and it is worth restating in one line:
with the item parameters fixed, the weighted-likelihood estimate is a function of
that pupil's response pattern and nothing else — and, by sufficiency, really only
of the raw score. The ten thousand other pupils in the *Land* contribute exactly
nothing to this pupil's number.

:::: {.amazing}
::: {.titelamazing}
<h2>Ten thousand observations that change nothing</h2>
:::
Almost every other statistical procedure gets better with more data, so the
instinct that a state-wide analysis produces a safer individual value is a
reasonable one. It is also completely wrong here, and the reason is structural
rather than a matter of degree: individual scoring is a *look-up against a fixed
table*, not an estimation problem with a sample size.

The sample already did its work — years earlier, in the Pilotierung, when the table
was made. Everything after that is arithmetic. What still limits the individual
number is the number of tasks the pupil answered, and nothing else can be
substituted for it.
::::

#### Where the group does enter {#irt-prior-dependence -}

That was WLE. Change the estimator and the answer changes, because EAP and MAP
multiply the pupil's likelihood by a population distribution — and *somebody has to
choose which population*.

```{r irt-prior-demo, echo=TRUE}
eap_under_prior <- function(mu, sd) {
  grid <- seq(-6, 6, length.out = 2001)
  p    <- plogis(outer(grid, maths_items, "-"))
  lik  <- apply(p, 1, function(q) prod(q^one_pupil * (1 - q)^(1 - one_pupil)))
  post <- lik * dnorm(grid, mean = mu, sd = sd)
  bista_maths(sum(grid * post / sum(post)))
}

prior_effect <- data.frame(
  assumed_population = c("the whole country", "a weak group", "a strong group"),
  points = c(eap_under_prior(0, 1), eap_under_prior(-0.7, 0.7), eap_under_prior(0.7, 0.7))
)
prior_effect$Stufe <- stufe_of(prior_effect$points)

knitr::kable(prior_effect, digits = 0,
  caption = "Identical answers, three assumptions about the population the pupil belongs to. The reported value moves by 38 points and the competence level changes from II to III - and not one of the pupil's responses was different.")
```

The pupil answered thirteen of twenty-four tasks in all three rows. Assume she sits
in a strong group and she is reported at Kompetenzstufe III; assume a weak one and
she is at II. WLE, which uses no population at all, gives 484 points regardless.

::: {.analysis-warning}
This is the concrete reason individual feedback should use WLE, and the reason
plausible values must never be attached to a named pupil. A plausible value is
drawn from a posterior that deliberately contains background variables — the whole
point is to carry population information *into* the individual draw. Excellent for
estimating a national gap; indefensible on a report that a family reads.
:::

#### The class is not in the model {#irt-clustering -}

Now the structural question. A Rasch model has persons and items. It has no
classes, no schools, no *Länder*. Pupils are treated as exchangeable draws, and
nothing in `RM()` or in `score_pattern()` knows that twenty-four of them sat in the
same room with the same teacher.

For an individual score that is harmless, for the reason just given: the individual
score does not use the other pupils. For **anything aggregated it is not
harmless**, because pupils in a class resemble each other, so a class of
twenty-four carries much less information than twenty-four unrelated pupils.

```{r irt-clustering-demo, echo=TRUE, cache=TRUE}
set.seed(303)
n_classes <- 40; per_class <- 24; icc <- 0.35; sd_total <- 0.85

class_id    <- rep(1:n_classes, each = per_class)
class_level <- rep(rnorm(n_classes, logit_maths(490), sd_total * sqrt(icc)), each = per_class)
pupil_level <- rnorm(n_classes * per_class, 0, sd_total * sqrt(1 - icc))

p   <- plogis(outer(class_level + pupil_level, maths_items, "-"))
raw <- rowSums(matrix(rbinom(length(p), 1, p), nrow = length(class_id)))

wle_lookup <- t(sapply(0:24, function(r) {
  pat <- rep(0, 24); if (r > 0) pat[order(maths_items)[1:r]] <- 1
  score_pattern(pat, maths_items)[c("WLE", "SE_WLE")]
}))
points <- bista_maths(wle_lookup[raw + 1, "WLE"])

naive_se   <- sd(points) / sqrt(length(points))
cluster_se <- sd(tapply(points, class_id, mean)) / sqrt(n_classes)

round(c(mean            = mean(points),
        naive_se        = naive_se,
        cluster_se      = cluster_se,
        design_effect   = (cluster_se / naive_se)^2), 2)
```

Nine hundred and sixty pupils, mean 494 points. Treat them as independent and the
standard error is 2.65 points, giving a confidence interval of 489 to 500. Treat
the forty **classes** as the independent units, which is what they are, and the
standard error is 6.22 — an interval of 482 to 507. The design effect is 5.5: those
960 pupils carry about as much information as 175 unrelated ones.

::: {.analysis-note}
The intraclass correlation used here, 0.35, is not pessimistic for Germany. A
school system that sorts pupils into different school types at the end of Year 4
produces classes that differ from each other far more than classes in a
comprehensive system, so between-class variance is large by construction.
:::

Large-scale assessments handle this outside the item response model, in three
places: the **sampling design** is a stratified cluster sample and is documented as
such; standard errors are computed with **replicate weights** or jackknife rather
than the naive formula; and plausible values are drawn from a **background model**
that can be multilevel. The VERA-8 technical report's warning that module
combinations must be entered as dummy variables in the background model of an MML
scaling is the same problem in miniature: whenever a grouping is real and the model
does not know about it, something gets underestimated.

#### Does the answer change with 100 or 10,000? {#irt-sample-stability -}

Two very different things can change, and conflating them is the source of most of
the confusion.

**Random error shrinks with sample size.** If item parameters are re-estimated, a
bigger sample estimates them more precisely, and that noise propagates into person
scores.

**The choice of origin does not shrink at all.** It is not noise. It is a decision,
and a hundred thousand pupils will reproduce a wrong decision with magnificent
precision.

```{r irt-stability-demo, echo=TRUE, cache=TRUE}
set.seed(1)

wobble <- function(n, reps = 60) {
  out <- replicate(reps, {
    theta <- rnorm(n, logit_maths(480), 0.85)
    p <- plogis(outer(theta, maths_items, "-"))
    x <- rbind(one_pupil,
               matrix(rbinom(length(p), 1, p), n, dimnames = list(NULL, names(maths_items))))
    usable <- rowSums(x) > 0 & rowSums(x) < 24
    x <- x[usable, , drop = FALSE]
    if (nrow(x) < 5 || any(colMeans(x) %in% c(0, 1))) return(c(NA, NA))

    fit <- suppressWarnings(RM(x))
    b   <- -coef(fit); names(b) <- sub("beta ", "", names(coef(fit)))
    own <- b - mean(b)                               # eRm's own origin
    c(bista_maths(score_pattern(one_pupil, own)[["WLE"]]),
      bista_maths(score_pattern(one_pupil, own + mean(maths_items))[["WLE"]]))
  })
  c(n = n,
    own_origin = mean(out[1, ], na.rm = TRUE), own_sd = sd(out[1, ], na.rm = TRUE),
    anchored   = mean(out[2, ], na.rm = TRUE), anchored_sd = sd(out[2, ], na.rm = TRUE))
}

knitr::kable(
  t(sapply(c(50, 200, 1000), wobble)), digits = 2,
  caption = "The same pupil, the same answers, sixty re-estimations at each sample size. 'own_origin' re-estimates the items and keeps the software's default origin; 'anchored' re-estimates and then puts the scale back where the official calibration had it. Fixed parameters give 483.9 points."
)
```

Read the two `sd` columns and the two mean columns separately, because they say
opposite things.

The standard deviations behave exactly as expected: 1.44 points at $n = 50$, 0.30
at $n = 1000$. So yes, the results are random-stable, and they become more so — but
notice the scale. Even at $n = 50$, the wobble from re-estimating the items is
**1.4 points**, against a measurement error for this pupil of **43 points**. The
sampling noise in the item parameters is a rounding error next to the pupil's own
uncertainty.

The mean columns are the finding. Re-estimating on your own sample puts this pupil
at **543 points** whether you have fifty pupils or a thousand. The official
calibration puts her at **484**. The gap is 59 points — three quarters of a
Kompetenzstufe — and it is exactly as large at $n = 1000$ as at $n = 50$, because
it is not an error at all. It is the [origin](items.html#irt-origin), silently redefined as "the average of the tasks in this booklet".

::: {.analysis-takeaway}
More data makes a wrong scale more precise, not more correct. Sample size fixes
noise and cannot touch a definition. This is why [what it costs to re-estimate](items.html#irt-free-vs-fixed)
insisted on fixing the published parameters: it is the one error in the whole
pipeline that no amount of testing will reveal or repair.
:::

And one more asymmetry, which is the reason VERA is a school-development instrument
rather than a diagnostic one:

```{r irt-aggregate-vs-individual, echo=TRUE, cache=TRUE}
precision <- function(n) {
  set.seed(9000 + n)
  theta <- rnorm(n, logit_maths(480), 0.85)
  p     <- plogis(outer(theta, maths_items, "-"))
  raw   <- rowSums(matrix(rbinom(length(p), 1, p), n))
  c(pupils             = n,
    se_of_the_mean     = sd(bista_maths(wle_lookup[raw + 1, "WLE"])) / sqrt(n),
    mean_individual_se = mean((100 / 1.105) * wle_lookup[raw + 1, "SE_WLE"]))
}

knitr::kable(
  t(sapply(c(24, 100, 1000, 10000, 100000), precision)), digits = 2,
  caption = "As the sample grows, the mean becomes almost perfectly known and each individual stays exactly as uncertain as before. A Land can report its average to within a third of a point and still not be able to place one of its pupils in a competence level."
)
```

#### The same pupil, tested twice {#irt-retest -}

Here is the number to give a parent, and it is the most honest summary of the
whole section. Suppose a pupil sat an equivalent test a second time, on a different
Tuesday, with nothing about her competence changed. How often does she land in the
same Kompetenzstufe?

```{r irt-retest, echo=TRUE, cache=TRUE}
retest <- function(n_items, n = 5000, seed = 42) {
  items <- setNames(seq(-2.2, 0.9, length.out = n_items), sprintf("t%02d", 1:n_items))
  table <- t(sapply(0:n_items, function(r) {
    pat <- rep(0, n_items); if (r > 0) pat[order(items)[1:r]] <- 1
    score_pattern(pat, items)[c("WLE", "SE_WLE")]
  }))

  set.seed(seed)
  theta <- rnorm(n, logit_maths(490), 0.9)
  sit   <- function() rowSums(matrix(rbinom(n * n_items, 1,
                                            plogis(outer(theta, items, "-"))), n))

  first  <- bista_maths(table[sit() + 1, "WLE"])
  second <- bista_maths(table[sit() + 1, "WLE"])

  c(items          = n_items,
    mean_se        = mean((100 / 1.105) * table[sit() + 1, "SE_WLE"]),
    same_level     = mean(stufe_of(first) == stufe_of(second)),
    correct_level  = mean(stufe_of(first) == stufe_of(bista_maths(theta))),
    two_apart      = mean(abs(match(stufe_of(first),  stufe_names) -
                              match(stufe_of(second), stufe_names)) >= 2))
}

knitr::kable(
  t(sapply(c(12, 24, 48, 96), retest)), digits = 3,
  caption = "Five thousand pupils, each sitting an equivalent test twice with unchanged competence. 'same_level' is how often the two administrations agree on the Kompetenzstufe; 'correct_level' is how often the first one matches the truth."
)
```

With a twenty-four item module — a realistic VERA-8 Basismodul — **the same pupil
gets the same competence level less than half the time**. Not because the model is
broken and not because the items are bad: because a 44-point standard error is
being pushed through 80-point bands, twice.

The consolations are real but limited. Agreement of 47% is far above the 17% that
random assignment to six levels would give, so the level does carry information.
Landing two or more levels apart happens 7% of the time, so the result is rarely
absurd. And doubling the test to 48 items lifts agreement to 61%.

```{r irt-bands-figure, echo=FALSE, fig.cap="The Year 8 class from [the class report](items.html#irt-ksm-maths), each pupil with a 95% interval, drawn against the competence bands. Twenty-two of the twenty-three intervals stretch across three bands. The point estimates are perfectly usable and the ordering is right; what no single pupil has is a level.", fig.align='center', out.width="95%"}
band_frame <- data.frame(
  lo  = c(200, maths_cuts),
  hi  = c(maths_cuts, 760),
  lab = stufe_names
)

pupil_frame <- data.frame(
  rank = rank(maths_points, ties.method = "first"),
  name = names(maths_points),
  est  = maths_points,
  se   = maths_se
)

ggplot() +
  geom_rect(data = band_frame,
            aes(xmin = lo, xmax = hi, ymin = -Inf, ymax = Inf, fill = lab), alpha = 0.18) +
  geom_text(data = band_frame, aes(x = (lo + hi) / 2, y = 24.4, label = lab),
            size = 3.2, colour = "grey25") +
  geom_errorbarh(data = pupil_frame,
                 aes(y = rank, xmin = est - 1.96 * se, xmax = est + 1.96 * se),
                 height = 0, linewidth = 0.6) +
  geom_point(data = pupil_frame, aes(y = rank, x = est), size = 1.8) +
  scale_y_continuous(breaks = pupil_frame$rank, labels = pupil_frame$name,
                     limits = c(0.5, 25)) +
  scale_x_continuous(limits = c(200, 760), breaks = maths_cuts) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  labs(x = "Bildungsstandards points", y = NULL)
```

#### So: is it unambiguous? {#irt-verdict -}

```{r irt-verdict-table, echo=FALSE}
knitr::kable(
  data.frame(
    Claim = c("she solved 13 of 24 tasks",
              "she is at 484 points",
              "she is above the Mindeststandard",
              "she is at Kompetenzstufe II",
              "this class averages 447 points",
              "this class is below the national average",
              "this Land averages 494 points",
              "she improved since last year"),
    Status = c("exact",
               "stable, given fixed parameters",
               "usually safe",
               "reproducible under half the time",
               "stable to about 20 points",
               "depends how far below",
               "stable, if clustering is in the standard error",
               "the weakest claim of all"),
    `Depends on` = c("nothing - it is a count",
                     "the official calibration being used, not re-estimated",
                     "how far she is from the cut score",
                     "test length, and her distance from the nearest cut",
                     "class size and test length",
                     "the standard error of the class mean, about 20 points",
                     "the sampling design, not the item model",
                     "two standard errors, so about 1.4 times a single one"),
    check.names = FALSE
  ),
  caption = "Eight sentences a VERA report could contain, ordered from unarguable to fragile. The model is the same in every row; what differs is how much of it is being asked to carry."
)
```

Four things decide where any given sentence lands in that table.

**How many tasks the pupil answered.** The only lever on individual precision, and
the one nobody outside the technical reports talks about. Twelve items, twenty-four
items and ninety-six items are three different instruments.

**Whether the published item parameters were used.** Not a matter of precision but
of meaning, and the one mistake that grows more convincing with more data.

**Which estimator produced the number.** WLE ignores the group; EAP and MAP import
it; plausible values are drawn from it deliberately.

**How far the pupil sits from the nearest cut score.** A pupil in the middle of a
band is placed reliably. A pupil near a boundary is being sorted by a coin whose
bias is small. Every aggregate statistic hides this, and every individual report
depends on it.

::: {.analysis-takeaway}
So: does something unambiguous come out of a VERA-style test? At the level of a
school or a *Land*, yes — the aggregates are precise, and that is what the
instrument was built for. At the level of one child, a point estimate and its
standard error are honest, and a competence level is a rounded version of them that
reproduces itself less than half the time. Both statements are true at once, and
the IQB says so itself in the technical report. A result is not made unambiguous by
being reported without its uncertainty; it is only made to look that way.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
A colleague argues that the *Land*-wide file with 10,000 pupils gives every
individual pupil a more trustworthy value than a single class would. He is
`r mcq(c("right, because more data reduces error", "right, but only for the item parameters", answer = "wrong: with fixed parameters the individual value does not depend on the sample at all", "impossible to judge without the standard errors"))`.

The same 960 pupils in 40 classes gave a naive standard error of 2.65 points and a
cluster standard error of 6.22. The design effect is therefore about
`r fitb(5.5, num = TRUE, tol = 0.4)`.

Re-estimating the items on your own sample moved this pupil from 484 to 543 points.
Going from 50 to 1000 pupils changes that gap by
`r mcq(c("about half", "about a tenth", answer = "essentially nothing", "it reverses"))`.
`r hide("Why?")` The gap is not sampling error but a different origin. It is a
systematic shift of the whole scale, and sample size has no effect on a definition.
`r unhide()`

A parent asks whether the reported competence level is her daughter's "real" level.
On a twenty-four item module, the probability that a second, equivalent test would
put her in the same level is about `r fitb(0.47, num = TRUE, tol = 0.05)`.
::::

### The Same Model on Real Exam Data {#irt-real-data}

Simulated classes are honest teachers because the truth is known, but they never
misbehave. Now a real one: 729 first-year business and economics students at the
University of Innsbruck sat a thirteen-item written mathematics exam, and the
whole response matrix ships with the `psychotools` package.[^irt-mathexam]

```{r irt-mathexam-fit, echo=TRUE}
data("MathExam14W", package = "psychotools")

exam <- as.matrix(MathExam14W$solved)
storage.mode(exam) <- "integer"

# Conditional maximum likelihood cannot use all-wrong or all-correct students.
scorable <- rowSums(exam) > 0 & rowSums(exam) < ncol(exam)
exam_fit <- RM(exam[scorable, ])

exam_difficulty <- -coef(exam_fit)
names(exam_difficulty) <- sub("beta ", "", names(coef(exam_fit)))

knitr::kable(
  data.frame(
    solved     = colMeans(exam[scorable, ]),
    difficulty = exam_difficulty - mean(exam_difficulty),
    std_error  = exam_fit$se.beta
  )[order(exam_difficulty), ],
  digits  = 3,
  caption = "Thirteen exam items calibrated on 688 students. Compare the standard errors with the classroom table: 0.08 instead of 0.59, which is what twenty-eight times as many people buy."
)
```

The mean item standard error is 0.087 logits, against 0.59 in the class. That is
the calibration lever of Section \@ref(irt-how-many), pulled all the way.

And now the model earns its keep. The exam was written in two sittings, and to
stop the second group from being told the answers, eight of the thirteen items
were given in different versions. The two versions were meant to be equivalent.
Andersen's test asks whether they were:

```{r irt-mathexam-andersen, echo=TRUE}
LRtest(exam_fit, splitcr = as.numeric(MathExam14W$group[scorable]))
```

They were not: $LR = 265$ on 12 degrees of freedom. Split the same students by
gender instead and the model is content ($LR = 18.1$, $p = .11$). The Rasch model
was never told that two versions existed. It found them, because a difficulty
that depends on which group you are in is exactly what specific objectivity
forbids.

```{r irt-mathexam-gender, echo=TRUE}
LRtest(exam_fit, splitcr = as.numeric(MathExam14W$gender[scorable]))
```

::: {.analysis-note}
This is what a *testable* model buys you, and it is the concrete answer to the
"Rasch versus 2PL" argument of Section \@ref(why-rasch). A 2PL model would have
absorbed part of that group difference into item discriminations and reported a
better fit. The Rasch model refused, and the refusal was the finding.
:::

```{r irt-mathexam-itemfit, echo=TRUE}
exam_person <- person.parameter(exam_fit)
itemfit(exam_person)
```

Three items misfit here for real reasons, not for lack of data. `payflow` has an
outfit mean-square of 1.57: it is by far the hardest item (13.8% correct) and
strong students get it wrong more often than the model expects — the signature of
an item that is hard for a reason unrelated to the rest of the exam. `matrix` and
`hesse` sit at 0.79 and 0.82, *too* predictable, which usually means they
duplicate information the other items already carry.

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Of the 729 students who sat the exam, `r fitb(41, num = TRUE, tol = 0.5)` had to
be excluded before conditional maximum likelihood could run. `r hide("Where does that number come from?")`
`729 - sum(scorable)`. They scored either 0 or 13, and a raw score at the boundary
carries no information about where on the scale the student is — only that they
are below or above everything the test can see. `r unhide()`

The Andersen test on the group split gave $LR = 265$ with 12 degrees of freedom.
The right conclusion is that `r mcq(c("the students in group 2 were better", answer = "at least one item behaved differently in the two versions", "the Rasch model is wrong for exam data", "the exam was too hard"))`.

An item with an outfit mean-square of 0.79 is `r mcq(c("misfitting because it is too hard", answer = "more predictable than the model expects", "showing differential item functioning", "fitting perfectly"))`.
::::

### What Must Be Checked before Scores Are Trusted? {#irt-checks}

A fitted model is not automatically a valid measuring instrument. Seven questions,
in the order they are usually worth asking.

#### Dimensionality {#irt-check-dimensionality -}

Do the items mainly measure one competence, or are reading, listening, vocabulary
and language form empirically distinct? A single total score can hide real
multidimensionality, and the confirmatory tools of Section
\@ref(confirmatory-factors) are the natural way to look.

#### Local independence {#irt-check-independence -}

After conditioning on competence, responses should be unrelated. Five questions
about the same reading passage usually are not: they form a **testlet**, and
treating them as five independent items overstates how much the test knows.

#### Item fit {#irt-check-itemfit -}

Do the observed patterns follow the curves the model implies? Outfit is sensitive
to surprising responses far from an item's difficulty, infit to surprises near it.
As the [fit statistics in the classroom](items.html#irt-step-fit) showed, both
need a substantial sample before they
mean anything: a mean-square of 2.0 on twenty-four children is a coin flip, the
same value on seven hundred is a finding.

It is worth seeing what thresholds a real programme uses, because they are less
dramatic than most textbook advice. The VERA-8 technical reports state
theirs: the point-biserial discrimination should exceed $.25$ and
must not fall below $.10$; the **infit should not exceed 1.10**, with values up to
1.25 tolerated in exceptional cases; and no lower bound is set at all, because an
item with an infit below 1 fits the Rasch model imperfectly but discriminates
*better* than the model expects, which is not a problem worth solving. Items that
trip a threshold are not dropped automatically — a summary index flags them and a
person decides, weighing didactic and motivational reasons against the statistics.
Of the 219 mathematics items piloted in 2025, 27 went into the final modules.

#### Differential item functioning {#irt-check-dif -}

Does an item have the same difficulty for different groups *at the same competence
level*? Gender, home language, school type, test mode, federal state. The Andersen
test of Section \@ref(irt-real-data) is the global version; a Wald test per item
says which item is responsible.

#### Targeting and information {#irt-check-targeting -}

Are the item difficulties spread across the range the persons occupy? Figure
\@ref(fig:irt-wright-map) is this check. A test made mostly of easy items measures
strong children badly no matter how many items it has.

#### Missingness and booklet design {#irt-check-missing -}

*Not administered* is not the same as *wrong*. IRT handles missing-by-design
elegantly — that is one of its main reasons for existing — but only if the
booklets are linked by common anchor items and the missingness really is by
design. Missing because a child ran out of time is a different animal, and
Section \@ref(missing-data) has the vocabulary for it.

#### Consequences of reporting {#irt-check-reporting -}

An estimate with a standard error and a competence *level* are different objects.
Cutting a continuous uncertain estimate into levels adds classification error, and
adds it most for exactly the children who sit near a cut score. With the standard
errors of Section \@ref(irt-how-many) in mind, a child reported as "level 3" on a
fourteen-item test could comfortably be level 2 or level 4.

::: {.analysis-warning}
Every one of these seven checks needs more data than a single class can provide.
That is not an argument against a teacher looking at her own item statistics — it
is an argument against a teacher drawing conclusions about individual children
from them.
:::

### Are VERA Items and IQB Data Public? {#irt-vera-data}

It depends what is meant by *the data*.

#### Published example tasks: yes {#irt-vera-tasks -}

The IQB task browser holds many VERA examples from earlier cycles with solutions
and didactic commentary, filterable by subject, competence domain and level, and
much of it released under a Creative Commons licence. Excellent for understanding
task design; not an item-response data set. There is no row per pupil.

#### The operational item pool: not fully open {#irt-vera-pool -}

VERA develops new tasks and reuses tasks from earlier cycles, so the live pool
cannot be an open bank. The released examples are the public face of a larger
system.

#### IQB research data: available on application {#irt-vera-fdz -}

The Research Data Centre at the IQB distributes anonymised Scientific Use Files
and Campus Use Files for non-commercial research and teaching. Access is normally
free but requires an application. Documentation, empty data structures and scaling
manuals are often available without one, and reading a scaling manual is a better
introduction to real practice than any textbook chapter.

No nationwide VERA pupil-by-item file appears to be offered as an unrestricted
download. The closest German alternative is an IQB education-monitoring data set
through the FDZ.

#### Directly downloadable alternatives {#irt-vera-alternatives -}

```{r irt-data-access, echo=FALSE}
knitr::kable(
  data.frame(
    Source = c("IQB VERA task browser", "VERA-8 technical reports", "IQB Kompetenzstufenmodelle",
               "IQB FDZ Scientific Use File",
               "OECD PISA public-use files", "psychotools::MathExam14W",
               "psychotools::VerbalAggression", "sirt::data.timss"),
    `Item texts` = c("yes, selected past tasks", "task names only", "example tasks per level",
                     "depends on the study", "coded item information",
                     "described in the help page", "all 24 statements printed", "identifiers and metadata"),
    `Response matrix` = c("no", "no", "no", "yes", "yes", "yes", "yes", "yes"),
    `Immediate access` = c("yes", "yes", "yes", "application required", "yes", "yes, with the package",
                           "yes, with the package", "yes, with the package"),
    `Best use here` = c("understand items and competence levels",
                        "the scaling procedure, formulas and cut scores",
                        "what each band means in words",
                        "German secondary analysis",
                        "a full large-scale-assessment workflow", "a real dichotomous Rasch analysis",
                        "a real partial credit analysis", "an educational booklet design"),
    check.names = FALSE
  ),
  caption = "Where item-response data and scaling documentation can actually be obtained. No pupil-by-item file is public, but the entire method is: the technical reports give the model, the estimation software, the equating procedure and the transformation formulas."
)
```

:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
The single most useful document for anyone who wants to understand a real
assessment rather than a textbook one is a **VERA-8 technical report**. Twenty
pages per subject per cycle, freely available, and they contain what papers
normally leave out: the booklet design, the sample, the coder agreement, the
estimation software, the item selection thresholds, the equating procedure, the
transformation formulas, and a frank paragraph about how imprecise an individual
result is. Sections \@ref(irt-calibration) and \@ref(irt-bista) are, to a large
extent, a walk through one of them. They are not secret — they are working
documents written for the *Länder*, and the numbers quoted in this section come
from the 2026 mathematics and English reports. What is *not* public is the
Itemkennwertetabelle that travels with them: the actual difficulty of each live
item, which would compromise the test if it circulated. The method is open; the
key is not.
::::

### What You Need for Your Own VERA-Style Analysis {#irt-own-data}

Item difficulties alone are not enough. At minimum you need

- one row per pupil;
- one column per item or sub-item;
- scored responses — `0`, `1`, and possibly ordered partial-credit categories;
- `NA` for genuinely unadministered items, distinguished from wrong answers;
- stable item identifiers that match the difficulty table;
- which booklet or module each pupil received;
- ideally item metadata and grouping variables, so that Section \@ref(irt-checks)
  can be worked through.

The minimal layout is unglamorous:

```{r irt-layout, echo=FALSE}
knitr::kable(
  data.frame(pupil_id = c(1001, 1002, 1003),
             item_01 = c(1, 1, 0), item_02 = c(0, 1, 0),
             item_03 = c(1, 1, NA), item_04 = c(NA, 0, 0)),
  caption = "One row per pupil, one column per item, NA for items the pupil never saw. Everything in this section runs on a matrix of this shape."
)
```

#### The same analysis in TAM and mirt {#irt-tam-mirt -}

`eRm` was the right tool for this section because conditional maximum likelihood
is the estimator that makes the Rasch logic visible. Operational large-scale
assessment uses marginal maximum likelihood instead, because it handles booklet
designs, background models and plausible values. In R that means
`TAM`[^irt-tam] or `mirt`[^irt-mirt]. The chunks below are not evaluated when the
book builds; remove `eval=FALSE` to run them.

The choice is not cosmetic, and the VERA-8 technical reports spell out why in a
passage that reads like a footnote and is not one. If several
module combinations of different difficulty are scaled together, MML software such
as ConQuest or `BILOG-MG` needs the combinations entered as dummy variables in the
background model — otherwise the differences between the difficulty levels are
*underestimated*, because marginal maximum likelihood assumes one population
distribution and quietly averages the groups together. In CML or JML software such
as `eRm` or Winsteps, no such specification is necessary "und auch nicht möglich",
because conditioning on the raw score removed the population from the item
likelihood in the first place. Section \@ref(why-rasch) called that specific
objectivity; here it is as a line in an operations manual.

```{r irt-tam, echo=TRUE, eval=FALSE}
library("TAM")

# Marginal maximum likelihood; the default is the Rasch / 1PL model.
tam_fit <- tam.mml(resp = responses)
tam_fit$xsi                                  # item difficulties and SEs

# Person scores: WLE for individual reporting ...
tam_wle <- tam.wle(tam_fit)

# ... plausible values for population statistics.
tam_pv <- tam.pv(tam_fit, nplausible = 10)

# A latent regression: background variables enter the population model itself.
tam_background <- tam.mml(resp = responses, Y = background_variables)

# Partial credit instead of dichotomous:
tam_pcm <- tam.mml(resp = aggression, irtmodel = "PCM2")

# Anchoring: hold the item parameters fixed at an official calibration.
anchors  <- cbind(seq_along(difficulty), difficulty)
tam_anch <- tam.mml(resp = new_responses, xsi.fixed = anchors)
```

```{r irt-mirt, echo=TRUE, eval=FALSE}
library("mirt")

mirt_fit <- mirt(responses, model = 1, itemtype = "Rasch", verbose = FALSE)
coef(mirt_fit, IRTpars = TRUE, simplify = TRUE)$items

fscores(mirt_fit, method = "WLE", full.scores.SE = TRUE)
fscores(mirt_fit, plausible.draws = 5)

# The comparison that the "Why begin with Rasch?" section is an argument about:
two_pl <- mirt(responses, model = 1, itemtype = "2PL", verbose = FALSE)
anova(mirt_fit, two_pl)
```

::: {.analysis-warning}
If the item difficulties are already calibrated on an official scale, fix them as
anchors rather than re-estimating them from one class. Re-estimating produces a
scale that is internally consistent and comparable to nothing — the numbers will
look fine and mean something else.
:::

### The Main Connections {#irt-connections}

The chain is complete:

$$
\text{categorical indicators}
\longrightarrow \text{latent response model}
\longrightarrow \text{item calibration}
\longrightarrow \text{person posterior}
\longrightarrow
\begin{cases}
\text{EAP, MAP or WLE point estimate},\\
\text{plausible values for population analysis}.
\end{cases}
$$

Rasch is valuable because it makes every link in that chain visible. It is not the
only IRT model, and IRT is not the only occupant of the
categorical-indicator cell in Section \@ref(irt-latent-map). But it is the
clearest place to see how a test can put a task and a child on the same axis — and
why the number that comes out is always an estimate rather than an observation.

Four sentences are worth carrying out of this section, and none of them is about
software.

A raw score confounds the child with the sheet of paper; a Rasch scale separates
them, at the price of assumptions that can fail. More children buy better items,
more items buy better children, and no amount of one substitutes for the other.
The scale has no origin of its own, so every number is relative to a calibration
somebody chose — which is why fixing the official item parameters matters more
than any modelling decision downstream of it. And a single pupil can be placed on
a national scale from one row of data, but only because several thousand other
pupils were measured first — while a competence level for that same pupil would
reproduce itself, on an equivalent test, less than half the time
(Section \@ref(irt-unambiguous)).

::: {.analysis-takeaway}
Item response theory is the branch of latent-variable modelling for categorical
responses (Section \@ref(irt-latent-map)), and the Rasch model is its most
restrictive and most testable member (Section \@ref(why-rasch)). Its restrictions
buy sufficiency, specific objectivity and a scale on which persons and items can
be compared directly. What it cannot buy is data: five items cannot measure a
child, one child cannot calibrate an item, and one class cannot do either job
properly (Section \@ref(irt-how-many)). Which is why real assessment splits the
work in two — a calibration on thousands of children, once, and a scoring against
fixed parameters for everybody afterwards (Sections \@ref(irt-calibration) and
\@ref(irt-anchoring)). The numbers that come out are then exact, stable, fragile or
close to a coin flip depending entirely on which sentence you try to build with
them (Section \@ref(irt-unambiguous)). The next chapter, on identification, asks
a different question of the same kind of data — not *how much does this person
have*, but *what would have happened otherwise*.
:::

### Readings {#irt-readings -}

- Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment Tests*. Copenhagen: Danmarks Paedagogiske Institut. The original, and far more readable than its reputation.
- Wright, B. D., & Stone, M. H. (1979). *Best Test Design*. Chicago: MESA Press. Where the person–item map comes from.
- Bond, T. G., Yan, Z., & Heene, M. (2020). *Applying the Rasch Model* (4th ed.). New York: Routledge. The standard applied introduction.
- De Boeck, P., & Wilson, M. (eds) (2004). *Explanatory Item Response Models*. New York: Springer. IRT as generalised mixed modelling; the source of the `VerbalAggression` data.
- Mair, P. (2018). *Modern Psychometrics with R*. Cham: Springer. Chapters 4 and 5 are the `eRm` and `mirt` companions to this section.
- Paul, S., Krüger, S., & Sachse, K. A. (2025). *Vergleichsarbeiten 2026, 8. Jahrgangsstufe, Mathematik: Aufgaben der Einzelmodule. Technischer Bericht.* Berlin: IQB. With the parallel reports for English (Niemietz, Paul, Jungermann & Sachse, 2025) and German (Wagner, Wäckerle & Sachse, 2025). Twenty pages per subject per cycle, and the only place the scaling model, the estimator, the equating method and the transformation formulas are written down together.
- Tarkian, J., Maritzen, N., Eckert, M., & Thiel, F. (2019). Vergleichsarbeiten (VERA) — Konzeption und Implementation in den 16 Ländern. In F. Thiel et al. (eds), *Datenbasierte Qualitätssicherung und -entwicklung in Schulen* (pp. 41–103). Wiesbaden: Springer VS. [DOI](https://doi.org/10.1007/978-3-658-23240-5_4)
- Stanat, P., Schipolowski, S., Schneider, R., Sachse, K. A., Weirich, S., & Henschel, S. (eds) (2022). *IQB-Bildungstrend 2021*. Münster: Waxmann. Open access; Chapter 10 is the methods chapter that names the scaling model, the estimator and the plausible values. [DOI](https://doi.org/10.31244/9783830996064)
- Hartig, J. (2007). Skalierung und Definition von Kompetenzniveaus. In B. Beck & E. Klieme (eds), *Sprachliche Kompetenzen*. Weinheim: Beltz. [pedocs](https://www.pedocs.de/volltexte/2010/3143/)
- [IQB: VERA overview](https://www.iqb.hu-berlin.de/de/bista/implementation-und-ueberpruefung/vera/)
- [IQB: Normierungsstudie 2026](https://www.iqb.hu-berlin.de/de/schule/primarstufe/bistatest/normierungsstudie-2026/) · [BiStaTest](https://www.iqb.hu-berlin.de/bista/bistatest)
- [VERA information site (FAQ and feedback)](https://iqb-berlin.github.io/vera-info/faq.html)
- [IQB: competence-level models](https://www.iqb.hu-berlin.de/de/bista/entwicklung/kompetenzstufenmodelle/)
- [IQB: VERA example tasks and materials](https://www.iqb.hu-berlin.de/de/schule/aufgaben/)
- [IQB VERA information: evaluation and interpretation](https://iqb-berlin.github.io/vera-info/intro/feedback.html)
- [FDZ at IQB: using research data](https://fdz.iqb.hu-berlin.de/de/daten-nutzen/)
- [OECD: PISA data and methodology](https://www.oecd.org/en/about/programmes/pisa/pisa-data.html)
- [CRAN: `eRm`](https://cran.r-project.org/package=eRm) · [`TAM`](https://cran.r-project.org/package=TAM) · [`mirt`](https://cran.r-project.org/package=mirt) · [`psychotools`](https://cran.r-project.org/package=psychotools)

[^irt-thurstone]: Thurstone, L. L. (1925). A method of scaling psychological and educational tests. *Journal of Educational Psychology*, 16(7), 433–451. https://doi.org/10.1037/h0073357

[^irt-lord]: Lawley, D. N. (1943). On problems connected with item selection and test construction. *Proceedings of the Royal Society of Edinburgh, Section A*, 61(3), 273–287. https://doi.org/10.1017/S0080454100006282 · Lord, F. M. (1952). *A theory of test scores*. Psychometric Monograph No. 7. Richmond: Psychometric Society.

[^irt-rasch]: Rasch, G. (1960). *Probabilistic Models for Some Intelligence and Attainment Tests*. Copenhagen: Danmarks Paedagogiske Institut. Expanded edition 1980, Chicago: University of Chicago Press.

[^irt-berkeley]: Rasch, G. (1961). On general laws and the meaning of measurement in psychology. In *Proceedings of the Fourth Berkeley Symposium on Mathematical Statistics and Probability*, Vol. 4, 321–333. Berkeley: University of California Press.

[^irt-birnbaum]: Birnbaum, A. (1968). Some latent trait models and their use in inferring an examinee's ability. In F. M. Lord & M. R. Novick, *Statistical Theories of Mental Test Scores* (pp. 397–479). Reading, MA: Addison-Wesley.

[^irt-andersen]: Andersen, E. B. (1973). A goodness of fit test for the Rasch model. *Psychometrika*, 38(1), 123–140. https://doi.org/10.1007/BF02291180

[^irt-sufficiency]: Andersen, E. B. (1977). Sufficient statistics and latent trait models. *Psychometrika*, 42(1), 69–81. https://doi.org/10.1007/BF02293746 The result holds for models with independent responses, a monotone response function and one latent dimension.

[^irt-controversy]: Andrich, D. (2004). Controversy and the Rasch model: A characteristic of incompatible paradigms? *Medical Care*, 42(1 Suppl), I7–I16. https://doi.org/10.1097/01.mlr.0000103528.48582.7c

[^irt-fischer]: Fischer, G. H. (1973). The linear logistic test model as an instrument in educational research. *Acta Psychologica*, 37(6), 359–374. https://doi.org/10.1016/0001-6918(73)90003-6

[^irt-wright]: Wright, B. D., & Stone, M. H. (1979). *Best Test Design: Rasch Measurement*. Chicago: MESA Press.

[^irt-samejima]: Samejima, F. (1969). Estimation of latent ability using a response pattern of graded scores. *Psychometrika Monograph Supplement*, 34(4, Pt. 2), 1–97. https://doi.org/10.1007/BF03372160

[^irt-bock]: Bock, R. D. (1972). Estimating item parameters and latent ability when responses are scored in two or more nominal categories. *Psychometrika*, 37(1), 29–51. https://doi.org/10.1007/BF02291411

[^irt-andrich]: Andrich, D. (1978). A rating formulation for ordered response categories. *Psychometrika*, 43(4), 561–573. https://doi.org/10.1007/BF02293814

[^irt-masters]: Masters, G. N. (1982). A Rasch model for partial credit scoring. *Psychometrika*, 47(2), 149–174. https://doi.org/10.1007/BF02296272

[^irt-bockaitkin]: Bock, R. D., & Aitkin, M. (1981). Marginal maximum likelihood estimation of item parameters: Application of an EM algorithm. *Psychometrika*, 46(4), 443–459. https://doi.org/10.1007/BF02293801

[^irt-mislevy]: Mislevy, R. J. (1991). Randomization-based inference about latent variables from complex samples. *Psychometrika*, 56(2), 177–196. https://doi.org/10.1007/BF02294457

[^irt-pisa]: PISA scaled its cognitive items with a one-parameter logistic model, and the partial credit model for polytomous items, from 2000 to 2012, and moved to a two-parameter logistic and generalised partial credit model in 2015. See Robitzsch, A., Lüdtke, O., Goldhammer, F., Kroehne, U., & Köller, O. (2020). Reanalysis of the German PISA data: A comparison of different approaches for trend estimation with a particular emphasis on mode effects. *Frontiers in Psychology*, 11, 884. https://doi.org/10.3389/fpsyg.2020.00884

[^irt-vera]: VERA was developed from 2002 at the Universität Koblenz-Landau together with the Rhineland-Palatinate education ministry and first written in 2003; seven *Länder* joined in 2004 and all sixteen from the 2007/08 school year, when the primary assessment also moved from Year 4 to Year 3. The IQB was founded in 2004 and has produced the material since. See the IQB VERA pages listed under Readings.

[^irt-verbal]: Vansteelandt, K. (2000), collected for and distributed with De Boeck, P., & Wilson, M. (eds) (2004). *Explanatory Item Response Models: A Generalized Linear and Nonlinear Approach*. New York: Springer. Available as `VerbalAggression` in `psychotools`.

[^irt-erm]: Mair, P., & Hatzinger, R. (2007). Extended Rasch modeling: The `eRm` package for the application of IRT models in R. *Journal of Statistical Software*, 20(9), 1–20. https://doi.org/10.18637/jss.v020.i09

[^irt-warm]: Warm, T. A. (1989). Weighted likelihood estimation of ability in item response theory. *Psychometrika*, 54(3), 427–450. https://doi.org/10.1007/BF02294627

[^irt-iqb-development]: Item development, piloting and norming are described on the IQB's test development page, <https://www.iqb.hu-berlin.de/de/bista/entwicklung/testentwicklung-zu-den-standards/>, and the quoted sentence on task difficulty comes from the VERA FAQ, <https://iqb-berlin.github.io/vera-info/faq.html>. The FAQ also states that the continuous competence scale is divided into five or six substantively described levels, and that the task pool deliberately spans solution rates from about 5–10 per cent to about 90 per cent.

[^irt-techreport]: Paul, S., Krüger, S., & Sachse, K. A. (2025). *Vergleichsarbeiten 2026, 8. Jahrgangsstufe, Mathematik: Aufgaben der Einzelmodule. Technischer Bericht.* Berlin: Institut zur Qualitätsentwicklung im Bildungswesen. The parallel reports for English (Niemietz, J., Paul, S., Jungermann, D., & Sachse, K. A., 2025) and German (Wagner, K., Wäckerle, M., & Sachse, K. A., 2025) follow the same structure, and each cycle publishes a new one. The equating method is mean/mean equating after Loyd, B. H., & Hoover, H. D. (1980). Vertical equating using the Rasch model. *Journal of Educational Measurement*, 17(3), 179-193. https://doi.org/10.1111/j.1745-3984.1980.tb00825.x These reports are official documents produced for each cycle and passed to the Länder together with the item parameter tables; several Länder republish them or the accompanying didactic handbooks on their own education servers. The item parameter tables themselves are not public, but every method described in this section is.

[^irt-hartig]: Hartig, J. (2007). Skalierung und Definition von Kompetenzniveaus. In B. Beck & E. Klieme (Eds.), *Sprachliche Kompetenzen: Konzepte und Messung* (pp. 83-99). Weinheim: Beltz. https://www.pedocs.de/volltexte/2010/3143/

[^irt-ksm]: Integriertes Kompetenzstufenmodell Mathematik, adopted by the KMK in November 2011. The metric was re-normed in the 2012 IQB-Ländervergleich onto the complete ninth year, which shifted the scale down by 25 points relative to the earlier norming on pupils aiming for at least a Mittlerer Schulabschluss; all level boundaries moved with it, so the meaning of the levels did not change. Cut scores and abbreviated level descriptions are reproduced in Table 1 of the VERA-8 2026 mathematics technical report; the full descriptions are at <https://www.iqb.hu-berlin.de/de/bista/entwicklung/kompetenzstufenmodelle/>.

[^irt-ksm-english]: Integriertes Kompetenzstufenmodell Englisch (IQB, 2014), adopted by the KMK in November 2014, referring to the KMK Bildungsstandards for the first foreign language (2003, 2004) and to the Common European Framework of Reference (Council of Europe, 2001). The sub-levels follow the equivalent differentiation made in the HarmoS project of the Swiss EDK (2009) and are placed at the midpoint of each CEFR level. Cut scores from Tables 1.1 and 1.2 of the VERA-8 2026 English technical report.

[^irt-bista-metric]: Stanat, P., Schipolowski, S., Schneider, R., Sachse, K. A., Weirich, S., & Henschel, S. (Eds.) (2022). *IQB-Bildungstrend 2021. Kompetenzen in den Fächern Deutsch und Mathematik am Ende der 4. Jahrgangsstufe im dritten Ländervergleich.* Münster: Waxmann. https://doi.org/10.31244/9783830996064 Open access under CC BY-NC-ND 4.0; the quoted sentence and the cut scores are in Chapter 1, the scaling procedure in Chapter 10.

[^irt-norming-2026]: The Normierungsstudie 2026 ran from early March to early May 2026 with roughly 12,000 pupils in Years 3 and 4 at about 455 general and special schools in all sixteen Länder, in German and mathematics. Its purpose, in the IQB's own words, is that "die Normierungsdaten werden zum einen zur Festlegung nationaler Kompetenzskalen (Punkteskalen) genutzt und bilden zum anderen die Grundlage zur Entwicklung von Kompetenzstufenmodellen". See <https://www.iqb.hu-berlin.de/de/schule/primarstufe/bistatest/normierungsstudie-2026/> and, for the wider rebuilding programme, <https://www.iqb.hu-berlin.de/bista/bistatest>.

[^irt-tarkian]: Tarkian, J., Maritzen, N., Eckert, M., & Thiel, F. (2019). Vergleichsarbeiten (VERA) - Konzeption und Implementation in den 16 Ländern. In F. Thiel, J. Tarkian, E.-M. Lankes, N. Maritzen, T. Riecke-Baulecke & A. Kroupa (Eds.), *Datenbasierte Qualitätssicherung und -entwicklung in Schulen* (pp. 41-103). Wiesbaden: Springer VS. https://doi.org/10.1007/978-3-658-23240-5_4 Sixty pages on how sixteen Länder implement the same instrument differently, which is the part no technical report covers.

[^irt-koeller]: Köller, O., Knigge, M., & Tesch, B. (Eds.) (2010). *Sprachliche Kompetenzen im Ländervergleich.* Münster: Waxmann. The report of the language study whose 2008 norming fixed the English metric and whose descriptors the competence level model uses; cited as the source of the full level descriptions in the IQB's own Kompetenzstufenmodell.

[^irt-linacre]: Linacre, J. M. (1994). Sample size and item calibration stability. *Rasch Measurement Transactions*, 7(4), 328. The often-quoted table gives 16–36 persons for ±1 logit at 99% confidence, 64–144 for ±0.5 logit at 95%, and 108–243 for ±0.3 logit at 95% with a well-targeted test.

[^irt-mathexam]: Zeileis, A., Umlauf, N., & Leisch, F. (2014). Flexible generation of e-learning exams in R: Moodle quizzes, OLAT assessments, and beyond. *Journal of Statistical Software*, 58(1), 1–36. https://doi.org/10.18637/jss.v058.i01 The data are distributed as `MathExam14W` in `psychotools`.

[^irt-tam]: Robitzsch, A., Kiefer, T., & Wu, M. (2024). *TAM: Test Analysis Modules*. R package. https://CRAN.R-project.org/package=TAM

[^irt-mirt]: Chalmers, R. P. (2012). mirt: A multidimensional item response theory package for the R environment. *Journal of Statistical Software*, 48(6), 1–29. https://doi.org/10.18637/jss.v048.i06
