# Identify {#identify}

::: {.analysis-subtitle}
Causal effects
:::

Everything up to here has been an answer to the question *what goes with what*. Chapter \@ref(compare) compared groups, Chapter \@ref(model) fitted lines through clouds, Chapter \@ref(reveal) looked for structure behind the measurements. Each of them describes the data honestly, and none of them can tell you what would happen if you changed something.

That gap is not closed by a better estimator. A regression coefficient is a conditional mean, no more and no less, and whether it deserves to be read as *the effect of* rather than *the difference between* is settled outside the model — by where the data came from and by how the treatment was assigned. This chapter is about that question, which economists call **identification**.

There are only two ways to earn a causal claim. Run the experiment, and let a coin sever the link between the treatment and everything else about the people who receive it. Or find a situation in which something else already did the severing — a lottery, a policy date, an administrative threshold, a rule nobody designed with your research question in mind — and then argue the case. Section \@ref(experiments) is about both: the logic of randomisation, why economics rarely gets to use it, and the vocabulary of research design that the rest of the chapter takes for granted.

The four sections that follow are the standard catalogue of designs, ordered by how much they ask you to believe about what you cannot see. Section \@ref(matching-and-balancing) builds comparability out of measured covariates. Section \@ref(difference-in-differences) builds it out of timing, Section \@ref(discontinuities) out of a threshold, and Section \@ref(instruments) out of a variable that moves the treatment and nothing else.

```{r child-experiments, child='L09-01-experiments.Rmd'}
```

```{r child-matching, child='L09-02-matching.Rmd'}
```

```{r child-difference-in-differences, child='L09-03-difference-in-differences.Rmd'}
```

```{r child-discontinuities, child='L09-04-discontinuities.Rmd'}
```

```{r child-instruments, child='L09-05-instruments.Rmd'}
```
