## Instruments {#instruments}

::: {.analysis-subtitle}
Instrumental-variable methods
:::

```{r iv-setup, echo=FALSE}
library("AER")        # ivreg(), and the diagnostics that come with summary()
library("gridExtra")  # two panels side by side
```

### A Letter That Nobody Has to Open {#iv-a-letter}

The Bundesagentur für Arbeit hands out something called a *Bildungsgutschein*.
It is a voucher for vocational further training, issued under §§ 81 ff. SGB III,
and the arithmetic of it is simple: the agency pays the course fee, the person
decides whether to sit in the classroom. Nobody is compelled. A voucher is an
offer, and an offer can be declined.

Suppose the agency wants to know what the training is worth. It has, in its own
files, every person who took a course and every person who did not, and it has
their earnings two years later. The comparison is one line of R and it will be
wrong, because the people who walk into a classroom on a Monday morning are not
the people who do not. They are more ambitious, better organised, closer to a
labour market that rewards the effort. Section \@ref(matching-and-balancing)
would try to measure that difference and adjust for it. Section
\@ref(difference-in-differences) would try to difference it away. Both need an
assumption about the part of ambition that no administrative file records.

Now change one thing. Suppose the agency, running a pilot, draws a random half
of the eligible register and sends them a letter — *you are entitled to a
voucher, here is how to claim it* — and sends nothing to the other half. The
letter does not put anyone in a classroom. Most people who receive one do
nothing at all. But some people who would not otherwise have gone now go, and
because the letter went out by lottery, the group that received it is otherwise
identical to the group that did not.

That is the entire idea of this section. There is a randomised thing (the
letter) that is not the treatment (the course). It is only a nudge, it works on
a minority, and it is worthless as a policy in its own right. What it is worth
is an argument: whatever difference in earnings appears between the letter group
and the no-letter group *must* have come through the courses, because nothing
else about the two groups differs. Divide the earnings gap by the participation
gap and the result is a causal effect — for the people the letter moved, and for
nobody else.

::: {.analysis-takeaway}
An instrument is a variable that shoves the treatment around without touching
the outcome by any other route. It buys identification the way randomisation
does, except that the randomisation happened to something adjacent to the thing
you care about.
:::

The pilot above is invented; the voucher is not. The design is not invented at
all. It is the oldest identification strategy in econometrics, it is older than
the phrase *causal inference*, and it was first written down by a man studying
the price of butter.

### What Philip Wright Could Not See {#iv-what-philip-wright-could-not-see}

In 1928 the Institute of Economics published a book by Philip Green Wright
called *The Tariff on Animal and Vegetable Oils*. It is a policy monograph about
whether a duty on imported oils would raise the domestic price, and to answer
that it needs to know how much buyers and sellers respond to price. Wright could
not get the number out of the data, and the reason he could not is the reason
this section exists.

The problem was already famous. E. J. Working had set it out the year before in
the *Quarterly Journal of Economics* under a title that is itself the question:
*What Do Statistical "Demand Curves" Show?* Suppose you collect, year after
year, the price of butter and the quantity sold. You plot the pairs. You fit a
line. What have you drawn?

Nothing, is the honest answer. Each point is where a demand curve met a supply
curve in that year, and both curves move between years. If demand jumps around
while supply sits still, the points trace out the supply curve. If supply jumps
around while demand sits still, they trace out demand. If both move — which is
what actually happens — the fitted line is a blend of the two, and it is not an
estimate of anything.

```{r iv-supply-demand, echo=FALSE, fig.align='center', fig.width=7.4, fig.height=3.6, fig.cap="Two markets, one estimator, two opposite answers. The dashed lines are the true demand and supply curves, identical in both panels; the solid line is a regression of price on quantity through the observed pairs. Where demand does the moving the regression recovers supply, and where supply does the moving it recovers demand. Nothing in the scatter says which world you are in."}
set.seed(4)
markt <- function(sd_demand, sd_supply, n = 70) {
  a <- 10 + rnorm(n, 0, sd_demand)       # demand intercept, shifting
  b <-  2 + rnorm(n, 0, sd_supply)       # supply intercept, shifting
  p <- (a - b) / 2                       # market clearing: 10-1.2p = 2+0.8p
  q <- b + 0.8 * p
  tibble(p = p, q = q)
}

kurven <- tibble(q = seq(2.5, 7.5, length.out = 50)) %>%
  mutate(supply = (q - 2) / 0.8, demand = (10 - q) / 1.2) %>%
  tidyr::pivot_longer(c(supply, demand), names_to = "curve", values_to = "p")

bind_rows(markt(1.6, 0.3) %>% mutate(welt = "only demand shifts"),
          markt(0.3, 1.6) %>% mutate(welt = "only supply shifts")) %>%
  ggplot(aes(q, p)) +
  geom_line(data = kurven, aes(group = curve), linetype = "dashed",
            colour = "grey65", linewidth = .4) +
  geom_point(colour = "#2c6e8e", alpha = .7, size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "grey25", linewidth = .6) +
  coord_cartesian(xlim = c(2.5, 7.5), ylim = c(0, 7)) +
  facet_wrap(~ welt) +
  labs(x = "Quantity", y = "Price") +
  theme_minimal(base_size = 10)
```

Wright's answer is in Appendix B, pages 286 to 319.^[In the book's own table of
contents the appendix is "Effects of a Duty on Price, Output, and Imports", in
three parts: formulae for estimating the effect of a duty, explanations and
qualifications, and the handling of data. It was reprinted almost a century
later, as "Effects of a duty on price and output with special reference to butter
and flaxseed", in *The Econometrics Journal* 28(1), 2025, pages 21–40, with an
editorial introduction — a rare honour for an appendix.] He calls the
method *introducing external factors*, and the sentence in which he states it is
worth reading slowly:

> Such additional factors may be factors which (A) affect demand conditions
> without affecting cost conditions or which (B) affect cost conditions without
> affecting demand conditions.
>
> --- Philip G. Wright (1928), quoted in Stock and Trebbi (2003, p. 180)

That is the definition of an instrumental variable, seventeen years before
anybody called it that. A factor of type (B) — something that shifts the supply
curve and nothing else — pins supply down and lets the demand curve be traced.
For flaxseed the natural candidate is *yield per acre*, which is mostly weather:
a good harvest pushes the supply curve out and has no business in anyone's
demand for linseed oil.^[Wright does not actually name his instruments in the
appendix. Stock and Trebbi (2003) reconstruct them from context as the price of a
substitute, which shifts demand, and yield per acre, which shifts supply. The
story you sometimes read, that Wright used regional rainfall for butter, is not
in the text.]

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
A variable $z$ is an **instrument** for the treatment $d$ in the equation

$$y = \beta_0 + \beta_1 d + \epsilon$$

if it satisfies three conditions:

1. **Relevance.** $\text{cov}(z, d) \neq 0$ — the instrument actually moves the
   treatment. *Testable.*
2. **Independence.** $z$ is as good as randomly assigned with respect to
   everything that determines $y$. *Testable only for what you measured.*
3. **Exclusion.** $z$ reaches $y$ through $d$ and through no other channel, so
   that $\text{cov}(z, \epsilon) = 0$. **Not testable, ever.**

Under these conditions

$$\beta_1 = \frac{\text{cov}(z, y)}{\text{cov}(z, d)}$$

which is the ratio Wright wrote down. Everything below is a way of computing it.
::::

Notice what the third condition costs. Relevance you can check in the data.
Independence you can probe, imperfectly, by looking for imbalance in measured
covariates. Exclusion is a claim about a quantity — the error term — that you
have never seen and never will. It is defended by argument, by institutional
detail, by knowing where the variation came from. It is never defended by a
p-value. Almost everything that goes wrong with instrumental variables in
practice goes wrong there.

:::: {.amazing}
::: {.titelamazing}
<h2>Amazing Fact</h2>
:::
For decades the credit for Appendix B went to the wrong Wright.

Philip's son was **Sewall Wright**, the geneticist who invented path analysis
and one of the three founders of population genetics. Appendix B contains a
path-analytic derivation in exactly his style, and from Goldberger (1972) onwards
a series of distinguished economists concluded that the son must have written it
— perhaps the appendix, perhaps the whole idea. Angrist and Krueger, writing in
2001, still put it in a footnote: "It is quite likely that Sewall Wright deserves
much of the credit for his father's use of instrumental variables."

Two years later James Stock and Francesco Trebbi settled the question, and they
settled it with a method borrowed from literary scholarship. They took 69
function-word frequencies and 18 grammatical statistics — the fingerprints of
prose style, the same technique Mosteller and Wallace had used on the disputed
*Federalist Papers* — reduced them by principal components, and ran a linear
discriminant against known writings of father and son. Their verdict, in the
*Journal of Economic Perspectives* 17(3), pages 177 to 194: "The stylometric
evidence clearly points to Philip G. Wright."

Principal components, from Section \@ref(components), used to decide who invented
instrumental variables.
::::

### From Measurement Error to Simultaneity {#iv-from-measurement-error-to-simultaneity}

Wright's appendix sank without trace. What happened next happened somewhere
else, for a different reason, and only later turned out to be the same thing.

The Oslo school of the 1930s was worried about **errors in variables**. Ragnar
Frisch's *Statistical Confluence Analysis by Means of Complete Regression
Systems* (1934) asks what happens to a regression when the regressors are
measured with noise, which in economic time series they always are. A mismeasured
regressor is correlated with the equation's error by construction, and its
coefficient is pulled towards zero. Frisch's own remedy, the bunch map, was a
diagnostic rather than an estimator.

Two answers came out of that tradition. Abraham Wald, in the *Annals of
Mathematical Statistics* in 1940, proposed splitting the sample into two groups
and taking the ratio of the difference in mean $y$ to the difference in mean $x$.
That is algebraically identical to instrumental variables with a binary
instrument, which is why the ratio in the definition box above is called the
**Wald estimator**. And Olav Reiersøl, in a 1945 dissertation titled *Confluence
Analysis by Means of Instrumental Sets of Variables*, gave the thing its
name.^[Reiersøl had the idea earlier — see his *Econometrica* paper of 1941 on
lag moments — and R. C. Geary arrived at something similar independently in the
1940s. Aldrich (1993) argues that Geary's conception is closer to the modern one
than Reiersøl's is.]

Meanwhile Trygve Haavelmo was making Wright's problem respectable. His 1943
*Econometrica* paper *The Statistical Implications of a System of Simultaneous
Equations*, and the enormous 1944 supplement *The Probability Approach in
Econometrics*, established that when an equation sits inside a system — when
price is determined jointly with quantity — ordinary least squares on that
equation is not merely imprecise but *inconsistent*, and no quantity of data will
save it. He received the Nobel Memorial Prize in 1989 "for his clarification of
the probability theory foundations of econometrics and his analyses of
simultaneous economic structures".

The Cowles Commission then spent a decade building the machinery. Anderson and
Rubin's *Estimation of the Parameters of a Single Equation in a Complete System
of Stochastic Equations* (1949) gave the first workable estimator, limited
information maximum likelihood, and — as Section \@ref(iv-inference-that-survives)
will show — a test that has aged rather better than the estimator has. Two-stage
least squares arrived twice: in two unpublished memoranda by Henri Theil at the
Dutch Central Planning Bureau in 1953, and in Robert Basmann's *Econometrica*
paper of 1957. James Durbin (1954) noticed that the errors-in-variables people
and the simultaneous-equations people had built the same estimator, and J. D.
Sargan (1958) supplied the overidentification test that carries his name.

By 1960 instrumental variables was a technique for fitting systems of
macroeconomic equations, taught as an answer to simultaneity and mostly forgotten
as an answer to anything else. What changed was not the algebra.

:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
James H. Stock and Francesco Trebbi, "Retrospectives: Who Invented Instrumental
Variable Regression?", *Journal of Economic Perspectives* 17(3), 2003, 177–194 —
the detective story, and a clear account of what Appendix B actually does.

Joshua D. Angrist and Alan B. Krueger, "Instrumental Variables and the Search for
Identification: From Supply and Demand to Natural Experiments", *Journal of
Economic Perspectives* 15(4), 2001, 69–85 — the same history told by the people
who changed what the method was for.
::::

### What Changed Was the Instrument {#iv-what-changed-was-the-instrument}

In 1990 Joshua Angrist published a paper in the *American Economic Review* about
the Vietnam draft lottery. Between 1970 and 1973 the United States assigned a
random sequence number, 1 to 365, to every birth date; men whose number fell
below a ceiling were draft-eligible. Eligibility did not determine service — some
eligible men were never drafted, some ineligible men volunteered — but it shifted
the probability sharply, and it was drawn from a drum on television. Angrist
matched the lottery to Social Security earnings records and divided. White
veterans, he found, were earning about 15% less in the early 1980s than
comparable non-veterans.^[Angrist (1990), *AER* 80(3), 313–336. The paper has an
erratum: two figures were transposed, and a sampling-variance formula in footnote
7 was wrong in a way that raises the standard errors in Table 3 by about ten per
cent. The corrected headline is a loss of \$2,384 with a standard error of \$778,
in 1978 dollars. Angrist and Chen (2011) return to the same lottery with data
through 2000 and find the penalty has gone to roughly zero.]

Nothing about that is a system of equations. The instrument is not a supply
shifter; it is a *lottery*, and the identification comes from the institution
that ran it. Within a decade the method had been turned inside out. Instead of
asking what estimator fits a model, economists began asking where in the world
somebody had accidentally run an experiment. That question is Section
\@ref(exp-natural); instrumental variables is the machinery that answers it.

The studies below are the catalogue. Read the middle column first — the
instrument is the paper.

```{r iv-catalogue, echo=FALSE}
katalog <- tibble::tribble(
  ~Study, ~Instrument, ~Treatment, ~Finding,
  "Angrist (1990), *AER*",
  "draft lottery number", "military service in Vietnam",
  "white veterans earned about 15% less in the early 1980s; by 2000 the gap has gone",
  "Angrist & Krueger (1991), *QJE*",
  "quarter of birth", "years of schooling",
  "compulsory-attendance laws keep roughly a quarter of would-be dropouts in school; the return is 8.9-9.3%, slightly above OLS",
  "Card (1995)",
  "grew up near a four-year college", "years of schooling",
  "the return to schooling is 13.2%, nearly twice the OLS figure of 7.5% - and the gap between them is the finding",
  "Angrist & Evans (1998), *AER*",
  "first two children of the same sex", "having a third child",
  "a third child cuts the probability that the mother works by roughly eight percentage points; fathers are unaffected",
  "Acemoglu, Johnson & Robinson (2001), *AER*",
  "settler mortality in the colonial period", "quality of institutions today",
  "the institutional gap between Nigeria and Chile is worth a sevenfold difference in income",
  "Miguel, Satyanath & Sergenti (2004), *JPE*",
  "rainfall growth", "GDP growth in Africa",
  "a negative growth shock of five percentage points raises the chance of civil conflict by one half",
  "Kling (2006), *AER*",
  "which judge you drew", "length of prison sentence",
  "no consistent evidence that longer sentences damage later employment or earnings",
  "Maestas, Mullen & Strand (2013), *AER*",
  "which disability examiner you drew", "receiving disability insurance",
  "for the 23% of applicants on the margin of entry, employment would have been 28 points higher without benefits",
  "Davey Smith & Ebrahim (2003), *IJE*",
  "a genetic variant", "the exposure that variant governs",
  "Mendelian randomisation: a variant such as ALDH2, which makes alcohol unpleasant, is dealt out at conception and so instruments for the exposure it governs",
  "Goldsmith-Pinkham, Sorkin & Swift (2020), *AER*",
  "local industry shares x national shocks", "local labour demand",
  "the Bartik instrument, and a proof that its identification lives in the shares rather than the shocks")

knitr::kable(katalog,
  caption = "Ten instruments. In every row the middle column is a variable that moves the treatment and has no plausible business in the outcome equation - and in every row that second claim is an argument rather than a test.") %>%
  kableExtra::kable_styling(font_size = 11)
```

Two things about that table. The first is that no two instruments are alike: a
lottery, a birthday, a map, a coin toss in the genome, a bureaucrat's caseload.
There is no method for finding one. The second is that the strongest entries —
the lottery, the judge, the examiner, the genetic variant — are all things that
were *randomised by somebody*, whether a government or meiosis. That is not a
coincidence, and Section \@ref(iv-an-instrument-designed-on-purpose) comes back
to it.

### The Idea in Four Numbers {#iv-the-idea-in-four-numbers}

Strip the algebra away and an instrumental-variables study is four numbers.

Take the letter from the opening. Compute mean earnings among the people who got
a letter and mean earnings among the people who did not; the difference is the
**intention-to-treat** effect, and it is a perfectly honest causal quantity — the
effect of *sending letters*, which is what a minister could actually decide to
do. Then compute the participation rate in each group; that difference is the
**first stage**, the amount of treatment the letter bought. The
intention-to-treat effect is small because the letter moved few people. Dividing
one by the other scales the small honest number up by the fraction of people it
worked on.

$$\hat{\beta}_{\text{IV}} = \frac{\bar{y}_{z=1} - \bar{y}_{z=0}}{\bar{d}_{z=1} - \bar{d}_{z=0}} = \frac{\text{intention to treat}}{\text{first stage}}$$

That is the Wald estimator. Everything else in this section is either a way of
computing it with covariates in the way (Section \@ref(iv-two-stages)), a warning
about what happens when the denominator is small (Section
\@ref(iv-when-the-instrument-is-weak)), or a question about whose effect the
answer is (Section \@ref(iv-who-is-the-answer-about)).

::: {.analysis-warning}
The denominator is the whole danger. If the letter moves participation by thirty
percentage points, you are dividing by 0.3 and multiplying every error in the
numerator — including every violation of the exclusion restriction — by 3.3. If
it moves participation by one percentage point, you are multiplying them by 100.
A weak instrument is not merely imprecise. It is *fragile*, in a way the printed
standard error does not reveal.
:::

### A Programme We Can Check {#iv-a-programme-we-can-check}

As in Sections \@ref(mat-data) and \@ref(did-data), the honest way to see what an
estimator does is to build a world in which the answer is already known. The
function below is that world. Twelve thousand people on the register; a coin
decides who gets a letter; each person has an unobserved `ehrgeiz` — drive,
ambition, whatever you want to call it — which raises their wage, raises their
chance of taking a course, *and* raises how much the course is worth to them.

That last clause is what makes the example hard. If the returns were the same for
everybody, selection would be a nuisance. Because the people most likely to enrol
are also the people who gain most from enrolling, there is no single number
called "the effect of the course", and the choice of estimator becomes a choice
of *whose* effect to estimate.

```{r iv-simulate, echo=TRUE}
simulate_weiterbildung <- function(n = 12000, seed = 1848, arme = 2,
                                   staerke = 1, verletzung = 0) {
  set.seed(seed)

  ehrgeiz <- rnorm(n)                                  # unobserved, and decisive
  alter   <- round(runif(n, 25, 58))
  frau    <- rbinom(n, 1, 0.48)
  schule  <- pmin(pmax(round(11 + 1.8 * rnorm(n) + 0.7 * ehrgeiz), 8), 18)
  lohn0   <- 2.55 + 0.045 * schule + 0.004 * (alter - 40) -
             0.09 * frau + 0.07 * ehrgeiz + rnorm(n, 0, 0.16)

  arm       <- sample(0:(arme - 1), n, replace = TRUE)  # the lottery
  einladung <- as.integer(arm >= 1)                     # letter, or no letter
  beratung  <- as.integer(arm == 2)                     # letter plus a phone call

  # what each person would do under each arm, monotone in one index
  index <- 0.80 * ehrgeiz + 0.10 * (schule - 11) - 0.20 * frau
  d0 <- as.integer(index > 1.00)                        # goes with no letter
  d1 <- as.integer(index > 1.00 - 0.90 * staerke)       # goes with a letter
  d2 <- as.integer(index > 1.00 - 1.30 * staerke)       # goes with letter and call
  teilnahme <- ifelse(arm == 0, d0, ifelse(arm == 1, d1, d2))

  typ <- ifelse(d0 == 1, "always-taker",
         ifelse(d1 == 1, "complier",
         ifelse(d2 == 1, "late complier", "never-taker")))

  # the return to the course, rising in the same drive that drives take-up
  tau <- 0.030 + 0.075 * pnorm(ehrgeiz) + 0.010 * (schule - 11) / 4

  lohn <- lohn0 + tau * teilnahme + 0.055 * ehrgeiz +
          verletzung * einladung + rnorm(n, 0, 0.13)

  tibble(id = seq_len(n), arm, einladung, beratung, teilnahme, typ,
         tau, ehrgeiz, alter, frau, schule, lohn0, lohn)
}

kurs <- simulate_weiterbildung()
```

`lohn` is a log hourly wage, so a coefficient of 0.09 means roughly nine per
cent. `staerke` is a dial for how hard the letter pushes, `verletzung` a dial for
the letter having a direct effect on wages; both are switched off for now and
both come back later. The `arme` argument adds a third treatment arm, used once,
in Section \@ref(iv-testing-what-can-be-tested).

Here is the world in two pictures. The left panel is the design: a threshold on
an index nobody observes, and a letter that moves the threshold. The right panel
is why the design matters.

```{r iv-world, echo=FALSE, fig.align='center', fig.width=8, fig.height=3.6, fig.cap="Left: the unobserved index that decides participation, with the two thresholds the letter moves between. Everyone to the right of the solid line goes anyway, everyone between the lines goes only if the letter arrives, everyone to the left never goes. Right: the true return to the course by type. The three groups are not interchangeable, which is the whole difficulty of this section."}
kurs2 <- kurs %>%
  mutate(index = 0.80 * ehrgeiz + 0.10 * (schule - 11) - 0.20 * frau)

p1 <- ggplot(kurs2, aes(index)) +
  geom_density(fill = "grey85", colour = "grey50") +
  geom_vline(xintercept = 1.00, linewidth = .6) +
  geom_vline(xintercept = 0.10, linewidth = .6, linetype = "dashed") +
  annotate("text", x = -1.5, y = .05, label = "never-takers",
           size = 3, colour = "grey25") +
  annotate("text", x = 0.55, y = .05, label = "compliers",
           size = 3, colour = "#2c6e8e") +
  annotate("text", x = 1.8, y = .05, label = "always-\ntakers",
           size = 3, colour = "grey25") +
  labs(x = "unobserved index of drive, schooling and constraints",
       y = "density", title = "The letter moves one threshold") +
  theme_minimal(base_size = 10)

p2 <- kurs %>%
  filter(typ != "late complier") %>%
  group_by(typ) %>%
  summarise(tau = mean(tau)) %>%
  mutate(typ = factor(typ,
                      levels = c("never-taker", "complier", "always-taker"))) %>%
  ggplot(aes(typ, tau)) +
  geom_col(fill = "#2c6e8e", width = .55) +
  geom_text(aes(label = sprintf("%.3f", tau)), vjust = -0.4,
            size = 3, colour = "grey25") +
  scale_y_continuous(limits = c(0, 0.13)) +
  labs(x = NULL, y = "true return to the course",
       title = "And they are not the same people") +
  theme_minimal(base_size = 10)

grid.arrange(p1, p2, nrow = 1)
```

Because we built the world, we can write down every quantity an economist might
mean by "the effect of the course" and watch them come out as four different
numbers.

```{r iv-truths, echo=FALSE}
tibble(
  Quantity = c("ATE --- everyone on the register",
               "ATT --- those who actually took a course",
               "ATU --- those who did not",
               "LATE --- those the letter moved (compliers)"),
  Truth = c(mean(kurs$tau),
            mean(kurs$tau[kurs$teilnahme == 1]),
            mean(kurs$tau[kurs$teilnahme == 0]),
            mean(kurs$tau[kurs$typ == "complier"]))) %>%
  knitr::kable(digits = 4,
    caption = "Four honest answers to the question 'what is the effect of the course', in a world where we are allowed to look. Nothing in the data will tell you which one an estimator has produced; only the design will.")
```

::: {.analysis-note}
The register splits into `r sprintf("%.1f%%", 100 * mean(kurs$typ == "always-taker"))`
always-takers, `r sprintf("%.1f%%", 100 * mean(kurs$typ == "complier"))` compliers
and `r sprintf("%.1f%%", 100 * mean(kurs$typ == "never-taker"))` never-takers.
You will never see this table in a real study. Which type a given person belongs
to is unknowable — we observe what they did under the arm they were given, never
under the other one — and Section \@ref(iv-who-is-the-answer-about) shows how
much of it can be recovered anyway.
:::

### What the Naive Comparison Says {#iv-what-the-naive-comparison-says}

Start where a competent analyst without an identification strategy would start.

```{r iv-naive, echo=TRUE}
lm(lohn ~ teilnahme, data = kurs) %>% summary() %>% coef() %>% round(4)
```

Thirty-five log points. Now add everything the agency's files contain — age, sex,
years of schooling — and then, going well beyond what any real file offers, the
person's own baseline wage.

```{r iv-naive-controls, echo=TRUE}
kontrollen <- list(
  "raw difference"        = lm(lohn ~ teilnahme, data = kurs),
  "+ age, sex, schooling" = lm(lohn ~ teilnahme + alter + frau + schule,
                               data = kurs),
  "+ baseline wage"       = lm(lohn ~ teilnahme + alter + frau + schule + lohn0,
                               data = kurs))

tibble(Specification = names(kontrollen),
       Estimate     = sapply(kontrollen, function(m) coef(m)["teilnahme"]),
       `Std. error` = sapply(kontrollen,
                             function(m) sqrt(diag(vcov(m)))["teilnahme"]),
       `True LATE`  = mean(kurs$tau[kurs$typ == "complier"])) %>%
  knitr::kable(digits = 4, row.names = FALSE,
    caption = "Three regressions, each better controlled than the last, converging on the wrong answer. The truth for the group an instrument will identify is in the final column; every specification sits above it, and the standard errors are so small that no interval comes close.")
```

The controls help, a great deal in fact, because we handed the model the baseline
wage, which absorbs most of what drive does to earnings. And the best-controlled
specification is still nearly double the true return for the group we will
actually be able to learn about, with a standard error that misses the truth by a
factor of twenty.

::: {.analysis-warning}
This is the pattern from Section \@ref(matching-and-balancing) in a different
costume: adding controls moves the estimate a long way and never tells you
whether it has moved far enough. Between the second row and the third the
estimate falls by ten log points. There is no diagnostic in the output that says
"and it should fall by another seven".
:::

### The Wald Ratio by Hand {#iv-the-wald-ratio-by-hand}

Now use the lottery. Four numbers, computed with `group_by()`.

```{r iv-four-numbers, echo=TRUE}
vier <- kurs %>%
  group_by(einladung) %>%
  summarise(n = n(), participation = mean(teilnahme), wage = mean(lohn))
vier
```

```{r iv-wald, echo=TRUE}
itt         <- diff(vier$wage)             # intention to treat
first_stage <- diff(vier$participation)    # the first stage

c(itt = itt, first_stage = first_stage, wald = itt / first_stage)
```

The letter raised participation by thirty percentage points and raised average
log wages by not quite three. Neither number is the effect of a course. Their
ratio is — for the people the letter moved.

::: {.analysis-takeaway}
The intention-to-treat effect is the effect of the *policy*. The Wald ratio is
the effect of the *treatment*, for compliers. A minister deciding whether to send
letters wants the first; a person deciding whether to enrol wants the second.
Reporting only one of them is a decision about whose question you are answering.
:::

### Two Stages, and Why the Second One Lies {#iv-two-stages}

The Wald ratio works only when the instrument is binary and nothing else is in
the model. Two-stage least squares is the generalisation. Regress the treatment
on the instrument and keep the fitted values — the part of participation the
lottery explains, and therefore the part that cannot be contaminated by drive —
then put those fitted values into the outcome equation in place of the treatment.

```{r iv-manual-2sls, echo=TRUE}
stufe1 <- lm(teilnahme ~ einladung, data = kurs)
stufe2 <- lm(kurs$lohn ~ fitted(stufe1))
coef(summary(stufe2)) %>% round(4)
```

The point estimate is exactly the Wald ratio. The standard error is wrong.

It is wrong for a reason worth understanding rather than memorising. `lm()`
computes the second-stage standard error from that regression's own residuals,
which are $y - \hat{\beta} \hat{d}$. The residual you actually want is
$y - \hat{\beta} d$, built from the real treatment rather than its fitted shadow.
The two differ, and there is no rule about which direction the error goes. Do it
properly and the software takes care of it:

```{r iv-ivreg, echo=TRUE}
iv <- ivreg(lohn ~ teilnahme | einladung, data = kurs)
summary(iv, diagnostics = TRUE)
```

The formula syntax is `outcome ~ regressors | instruments`. Everything left of
the bar is the model you want; everything right of it is the list of variables
allowed to be exogenous. Exogenous controls therefore appear on *both* sides,
which trips up everybody exactly once:

```{r iv-ivreg-controls, echo=TRUE}
ivreg(lohn ~ teilnahme + alter + frau + schule |
             einladung + alter + frau + schule, data = kurs) %>%
  summary() %>% coef() %>% head(2) %>% round(4)
```

::: {.analysis-note}
`summary(iv, diagnostics = TRUE)` prints three tests that no
instrumental-variables table should be without. *Weak instruments* is the
first-stage F on the excluded instruments — Section
\@ref(iv-when-the-instrument-is-weak). *Wu-Hausman* asks whether the IV and OLS
estimates differ by more than sampling noise, which is a test of whether the
treatment was endogenous at all. *Sargan* is the overidentification test, blank
here because with one instrument and one endogenous regressor there is nothing to
overidentify — Section \@ref(iv-testing-what-can-be-tested).
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The letter raised participation from 11.3% to 41.4% and average log wages by
0.027.

The Wald ratio, to two decimal places, is `r fitb(0.09, tol = 0.011)`.

Now suppose only half as many people had responded to the letter — a first stage
of fifteen points instead of thirty — while the intention-to-treat effect stayed
the same. The Wald ratio would be
`r mcq(c("unchanged", answer = "twice as large", "half as large"))`.

And the harder one. In that world, is the estimate more or less sensitive to a
small direct effect of the letter on wages?
`r mcq(c(answer = "more sensitive, because the same numerator error is divided by a smaller denominator", "less sensitive, because fewer people are affected", "equally sensitive, since the letter was randomised either way"))`
::::

### Who Is the Answer About? {#iv-who-is-the-answer-about}

The estimate is `r sprintf("%.4f", coef(iv)["teilnahme"])`. The average treatment
effect in this population is `r sprintf("%.4f", mean(kurs$tau))`. These are not
the same number, and the estimator is not making a mistake.

Imbens and Angrist worked out what the ratio identifies in a 1994 *Econometrica*
note, and Angrist, Imbens and Rubin gave the pieces their names in the *Journal
of the American Statistical Association* two years later. Divide the population
by what each person *would* do under each value of the instrument:

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
With a binary instrument $z$ and a binary treatment $d$, write $d(1)$ and $d(0)$
for what a person would do with and without the instrument. Four types exhaust
the population:

|  | $d(0) = 0$ | $d(0) = 1$ |
|---|---|---|
| **$d(1) = 0$** | never-taker | defier |
| **$d(1) = 1$** | **complier** | always-taker |

**Monotonicity** is the assumption that there are no defiers: the instrument
pushes everybody in the same direction, or leaves them alone. Under relevance,
independence, exclusion and monotonicity the Wald ratio identifies

$$\text{LATE} = E[\tau_i \mid d_i(1) > d_i(0)]$$

the **local average treatment effect** — the average effect among compliers, and
among nobody else.
::::

Compliers are not a group you can point at. Anybody who took a course after
receiving a letter is either a complier or an always-taker, and no data will
separate them. What *can* be recovered is the size of the group and its average
characteristics, and both fall out of the same trick: the first stage is the
complier share.

```{r iv-complier-share, echo=TRUE}
c(estimated = unname(first_stage),
  truth     = mean(kurs$typ == "complier"))
```

Averages of covariates among compliers come from a difference of the same shape.
Take the mean of $x \cdot d$ in the letter group, subtract it in the no-letter
group, divide by the first stage.

```{r iv-complier-profile, echo=TRUE}
complier_mean <- function(x) {
  z <- kurs$einladung
  d <- kurs$teilnahme
  (mean(x * d * z) / mean(z) -
   mean(x * d * (1 - z)) / mean(1 - z)) / first_stage
}

tibble(Variable = c("years of schooling", "age", "share women"),
       `Compliers (estimated)` = c(complier_mean(kurs$schule),
                                   complier_mean(kurs$alter),
                                   complier_mean(kurs$frau)),
       `Compliers (truth)` = c(mean(kurs$schule[kurs$typ == "complier"]),
                               mean(kurs$alter[kurs$typ == "complier"]),
                               mean(kurs$frau[kurs$typ == "complier"])),
       `Whole register` = c(mean(kurs$schule), mean(kurs$alter),
                            mean(kurs$frau))) %>%
  knitr::kable(digits = 3,
    caption = "Who the compliers are, estimated from observable data and checked against the truth. They have more schooling than the register as a whole, which is exactly the group a letter about a training voucher would be expected to move.")
```

That table is the most under-used diagnostic in applied work. It costs three
lines, and it turns an abstract complaint — "LATE is only for compliers" — into a
concrete description of a group the reader can picture.

```{r iv-estimands, echo=FALSE, fig.align='center', fig.width=7.2, fig.height=3.2, fig.cap="Every estimand in this world, with the instrumental-variables estimate and its interval laid over them. The estimator lands on the local average treatment effect and misses the population average by a third; the fully controlled regression misses everything."}
ci <- coef(iv)["teilnahme"] + c(-1.96, 1.96) * sqrt(diag(vcov(iv)))["teilnahme"]
ols_voll <- coef(lm(lohn ~ teilnahme + alter + frau + schule + lohn0,
                    data = kurs))["teilnahme"]

tibble(was = factor(c("ATU", "ATE", "LATE (compliers)", "ATT"),
                    levels = c("ATU", "ATE", "LATE (compliers)", "ATT")),
       wert = c(mean(kurs$tau[kurs$teilnahme == 0]), mean(kurs$tau),
                mean(kurs$tau[kurs$typ == "complier"]),
                mean(kurs$tau[kurs$teilnahme == 1]))) %>%
  ggplot(aes(wert, was)) +
  annotate("rect", xmin = ci[1], xmax = ci[2], ymin = -Inf, ymax = Inf,
           fill = "#2c6e8e", alpha = .15) +
  geom_vline(xintercept = coef(iv)["teilnahme"], colour = "#2c6e8e",
             linewidth = .7) +
  geom_vline(xintercept = ols_voll, colour = "grey40",
             linetype = "dashed", linewidth = .6) +
  geom_point(size = 3, colour = "grey20") +
  annotate("text", x = coef(iv)["teilnahme"], y = 4.4, label = "2SLS",
           size = 3, colour = "#2c6e8e", hjust = -0.15) +
  annotate("text", x = ols_voll, y = 4.4, label = "OLS, fully controlled",
           size = 3, colour = "grey40", hjust = 1.05) +
  labs(x = "effect on log hourly wage", y = NULL) +
  theme_minimal(base_size = 10)
```

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: the objection Deaton makes</h2>
:::
Not everybody finds the local average treatment effect a satisfying answer. Angus
Deaton, in the *Journal of Economic Literature* in 2010, put the objection about
as sharply as it can be put:

> Instead, we have a procedure in which the choice of the instrument, which is
> guided by criteria designed for a situation in which there is no heterogeneity,
> is implicitly allowed to determine the parameter of interest. This goes beyond
> the old story of looking for an object where the light is strong enough to see;
> rather, we have at least some control over the light, but choose to let it fall
> where it may, and then proclaim that whatever it illuminates is what we were
> looking for all along.

James Heckman and Sergio Urzúa make a related argument in the *Journal of
Econometrics* the same year: an instrument answers the question it happens to
pose, and the question a policy-maker asked is generally a different one.

The reply, which is Angrist and Imbens's, is that a well-defined local answer is
worth more than an ill-defined global one, and that the complier population can
be described — see the table above — rather than left as a mystery. Both sides
are right about something. Neither side disputes the empirical fact: change the
instrument and the number changes, because the compliers changed.
::::

### When the Instrument Is Weak {#iv-when-the-instrument-is-weak}

Everything above assumed a letter that moved thirty per cent of the register.
Most instruments in the literature move far less. Turn the `staerke` dial down
and watch.

```{r iv-weak-study, echo=TRUE, cache=TRUE}
weak_study <- function(staerke, draws = 400) {
  out <- lapply(seq_len(draws), function(i) {
    d  <- simulate_weiterbildung(n = 4000, seed = 20000 + i, staerke = staerke)
    fs <- lm(teilnahme ~ einladung, data = d)
    m  <- ivreg(lohn ~ teilnahme | einladung, data = d)
    tibble(F    = unname(summary(fs)$fstatistic[1]),
           iv   = unname(coef(m)["teilnahme"]),
           se   = unname(sqrt(diag(vcov(m)))["teilnahme"]),
           late = mean(d$tau[d$typ == "complier"]))
  })
  bind_rows(out) %>% mutate(staerke = staerke)
}

schwach <- bind_rows(lapply(c(1, 0.35, 0.15, 0.07), weak_study))
```

```{r iv-weak-table, echo=FALSE}
schwach %>%
  group_by(staerke) %>%
  summarise(`median F`        = median(F),
            `true LATE`       = mean(late),
            `median estimate` = median(iv),
            `10th pct`        = quantile(iv, .10),
            `90th pct`        = quantile(iv, .90),
            `share absurd`    = mean(iv < 0 | iv > 0.3),
            coverage          = mean(abs(iv - late) < 1.96 * se)) %>%
  arrange(desc(staerke)) %>%
  rename(`letter strength` = staerke) %>%
  knitr::kable(digits = 3,
    caption = "Four hundred studies at each of four instrument strengths, strongest first. The median estimate stays near the truth almost all the way down; the tenth and ninetieth percentiles do not, and by the last row five studies in six report a negative or implausibly large return. Coverage looks fine throughout, which is the trap.")
```

Read the coverage column last, and read it carefully. It sits at the nominal 95%
in the first row and *rises* from there. A reader checking only whether the
confidence interval covers the truth would conclude that nothing is wrong. What
has happened is that the intervals have grown so wide that they cover everything
— zero, the OLS answer, numbers no economist would defend. The estimator has not
become biased so much as *uninformative*, and an uninformative answer that
arrives with stars beside it is worse than no answer at all.

```{r iv-weak-picture, echo=FALSE, fig.align='center', fig.width=7.6, fig.height=3.2, fig.cap="The sampling distribution of the two-stage least squares estimate at four instrument strengths, truncated to a readable range. The vertical line is the truth. As the first stage weakens the distribution does not shift so much as spread, growing tails far beyond anything the reported standard error suggests."}
schwach %>%
  group_by(staerke) %>%
  mutate(label = sprintf("median F = %.0f", median(F))) %>%
  ungroup() %>%
  mutate(label = factor(label,
                        levels = unique(label[order(-staerke)]))) %>%
  filter(iv > -0.3, iv < 0.5) %>%
  ggplot(aes(iv)) +
  geom_histogram(bins = 40, fill = "#2c6e8e", colour = "white",
                 linewidth = .15) +
  geom_vline(xintercept = mean(schwach$late), colour = "grey20",
             linewidth = .5) +
  facet_wrap(~ label, nrow = 1, scales = "free_y") +
  labs(x = "2SLS estimate", y = NULL) +
  theme_minimal(base_size = 10)
```

### The F Statistic and Its Thresholds {#iv-the-f-statistic}

The standard summary of instrument strength is the F statistic on the excluded
instruments in the first stage. Its folklore is worth unpicking, because most of
what gets said about it in seminars is a little bit wrong.

The number everybody quotes is ten, and it comes from a remark rather than a
theorem. Douglas Staiger and James Stock, in *Econometrica* in 1997, wrote that
"it is our impression that, in applications of two-stage least squares, it is
common for the first stage F statistic … to take on a value less than 10." The
profession turned an observation about what is common into a threshold for what
is acceptable.

Stock and Yogo (2005) did the work properly and produced tables of critical
values, and those tables say two different things depending on what you are
afraid of. If you are afraid of *bias* — of two-stage least squares drifting back
towards ordinary least squares — the critical values sit in the low teens and ten
is a fair approximation. If you are afraid of *size* — of a nominal 5% test
rejecting far more often than five per cent of the time — the numbers are much
larger. With one endogenous regressor and one instrument, the value that keeps
the true size of a 5% test below 10% is 16.38.

Then Lee, McCrary, Moreira and Porter, in the *American Economic Review* in 2022,
showed that even that is optimistic. In the just-identified case, the conventional
$t$-test with the usual 1.96 cut-off attains correct size only if the first-stage
F exceeds **104.7**.

```{r iv-thresholds, echo=FALSE}
tibble(
  Threshold = c("10", "16.38", "104.7"),
  Source = c("Staiger & Stock (1997), as folklore",
             "Stock & Yogo (2005), one instrument",
             "Lee, McCrary, Moreira & Porter (2022)"),
  `What it promises` = c(
    "roughly, that 2SLS bias stays under a tenth of the OLS bias --- a statement about the point estimate only",
    "that a nominal 5% Wald test really rejects at most 10% of the time --- a statement about size",
    "that the conventional t-test with the 1.96 cut-off is valid at the 5% level in the just-identified case")) %>%
  knitr::kable(
    caption = "Three thresholds for the same statistic, promising three different things: a rule of thumb about bias, a bound on test size, and the requirement for the t-statistic you actually printed to mean what it says.")
```

The authors of the third paper went and looked. Across 1,311 single-instrument
specifications in 61 *AER* papers, roughly a quarter of the results significant at
the 5% level with an F above ten stop being significant once the correction is
applied.

::: {.analysis-warning}
Card's college-proximity instrument in Section
\@ref(iv-returns-to-schooling-with-real-data) has a first-stage F of about
thirteen. That clears the folklore threshold, sits below the Stock-Yogo size
threshold, and is an order of magnitude below what Lee and co-authors require. It
is one of the most cited instruments in economics.
:::

### Inference That Survives a Weak Instrument {#iv-inference-that-survives}

If the F statistic is a diagnostic that arrives too late, the alternative is not
to diagnose better but to use a procedure that does not care. The Anderson-Rubin
test — the same Anderson and Rubin, the same 1949 paper — is that procedure, and
it works by refusing to estimate anything.

To test whether the true effect equals some particular value $b$, subtract it
out. Under that null the quantity $y - b\,d$ contains no treatment effect at all,
so if the instrument is valid it should be unrelated to the instrument. Regress it
on $z$ and test the coefficient. Repeat over a grid of $b$ and keep every value
the data cannot reject.

```{r iv-ar, echo=TRUE}
ar_statistic <- function(data, b) {
  rest <- data$lohn - b * data$teilnahme
  summary(lm(rest ~ data$einladung))$coefficients[2, 3]^2   # squared t
}

gitter   <- seq(-0.2, 0.5, by = 0.0005)
stark    <- vapply(gitter, function(b) ar_statistic(kurs, b), numeric(1))
behalten <- gitter[stark < qchisq(0.95, df = 1)]
range(behalten)
```

Compare that with the interval `ivreg()` implies:

```{r iv-ar-compare, echo=TRUE}
coef(iv)["teilnahme"] + c(-1.96, 1.96) * sqrt(diag(vcov(iv)))["teilnahme"]
```

Practically identical, and they should be: with a first stage this strong the
conventional approximation is excellent. Now run the same two calculations in a
world where the letter barely works.

```{r iv-ar-weak, echo=TRUE}
kurs_schwach <- simulate_weiterbildung(staerke = 0.05)
iv_schwach   <- ivreg(lohn ~ teilnahme | einladung, data = kurs_schwach)

schwach_stat <- vapply(gitter,
                       function(b) ar_statistic(kurs_schwach, b), numeric(1))

c(anteil_des_gitters  = mean(schwach_stat < qchisq(0.95, df = 1)),
  konventionell_unten = unname(coef(iv_schwach)["teilnahme"] -
                          1.96 * sqrt(diag(vcov(iv_schwach)))["teilnahme"]),
  konventionell_oben  = unname(coef(iv_schwach)["teilnahme"] +
                          1.96 * sqrt(diag(vcov(iv_schwach)))["teilnahme"]))
```

The Anderson-Rubin set is the entire grid. Every value of the treatment effect we
tried — every one — is consistent with the data at the 5% level, which is the
correct answer, because a letter that moves almost nobody tells you almost
nothing. The conventional interval, meanwhile, is finite and prints a
respectable-looking pair of numbers.

```{r iv-ar-picture, echo=FALSE, fig.align='center', fig.width=7.4, fig.height=3.2, fig.cap="The Anderson-Rubin statistic against the hypothesised treatment effect, for a strong and a weak instrument. Where the curve dips below the dashed critical value the effect cannot be rejected. On the left the curve rises steeply on both sides and the set is a short interval; on the right it never rises at all, and the honest confidence set is unbounded."}
tibble(b    = rep(gitter, 2),
       stat = pmin(c(stark, schwach_stat), 40),
       welt = rep(c("strong first stage (30 points)",
                    "weak first stage (about one point)"),
                  each = length(gitter))) %>%
  ggplot(aes(b, stat)) +
  geom_line(colour = "#2c6e8e", linewidth = .6) +
  geom_hline(yintercept = qchisq(0.95, 1), linetype = "dashed",
             colour = "grey40") +
  facet_wrap(~ welt) +
  labs(x = "hypothesised effect on log wage",
       y = "Anderson-Rubin statistic") +
  theme_minimal(base_size = 10)
```

::: {.analysis-takeaway}
An unbounded confidence set is not a failure of the method. It is the method
telling you the truth about how much the data know. A procedure that always
returns a tidy interval is not being more informative; it is being less honest.
:::

Andrews, Stock and Sun surveyed the practice in the *Annual Review of Economics*
in 2019, and their recommendation for the single-instrument case is exactly this:
report the identification-robust interval. They also counted. Of seventeen *AER*
papers between 2014 and 2018 with "instrument" in the abstract, fifteen reported
at least one first-stage F — and many of those were below ten.

### Testing What Can Be Tested {#iv-testing-what-can-be-tested}

With two instruments and one endogenous regressor there is a degree of freedom
left over, and it can be spent on a test. Give the pilot a third arm: some people
get the letter, some get the letter and a phone call from an adviser, some get
nothing.

```{r iv-three-arms, echo=TRUE}
kurs3 <- simulate_weiterbildung(arme = 3)

kurs3 %>%
  group_by(arm) %>%
  summarise(n = n(), participation = mean(teilnahme), wage = mean(lohn))
```

Each arm gives its own Wald ratio against the control group:

```{r iv-arm-walds, echo=TRUE}
a <- kurs3 %>% group_by(arm) %>% summarise(w = mean(lohn), d = mean(teilnahme))

c(`letter only`      = (a$w[2] - a$w[1]) / (a$d[2] - a$d[1]),
  `letter plus call` = (a$w[3] - a$w[1]) / (a$d[3] - a$d[1]))
```

Two-stage least squares with both instruments combines them, and Sargan's test
asks whether the combination was reasonable:

```{r iv-overid, echo=TRUE}
iv3 <- ivreg(lohn ~ teilnahme | einladung + beratung, data = kurs3)
summary(iv3, diagnostics = TRUE)$diagnostics %>% round(4)
```

Sargan is comfortable. Here is what that comfort is worth. In this world the two
instruments identify **different** local effects — the phone call reaches a group
with lower returns, and the true LATEs are
`r sprintf("%.4f", mean(kurs3$tau[kurs3$typ == "complier"]))` and
`r sprintf("%.4f", mean(kurs3$tau[kurs3$typ %in% c("complier", "late complier")]))`
respectively — and the test cannot see it, because sampling noise in the two
ratios is larger than the gap between them.

::: {.analysis-warning}
An overidentification test asks whether your instruments agree with each other.
It does not ask whether they are valid. Two instruments that violate the exclusion
restriction in the *same* direction will pass it comfortably, and two perfectly
valid instruments identifying two genuinely different local effects will fail it.
A passed Sargan test is a weak piece of evidence dressed as a strong one.
:::

### The Exclusion Restriction Is an Argument {#iv-the-exclusion-restriction}

In our world the letter has no effect on wages except through the course. Turn
the `verletzung` dial and it does — imagine the letter also carries a page of
job-search advice, or simply reminds the recipient that the agency is watching.

```{r iv-exclusion, echo=TRUE, cache=TRUE}
verletzt <- lapply(c(0, 0.002, 0.005, 0.010, 0.020), function(v) {
  stark   <- simulate_weiterbildung(verletzung = v)
  schwach <- simulate_weiterbildung(verletzung = v, staerke = 0.15)
  tibble(`direct effect`   = v,
         `first stage (strong)` = unname(coef(lm(teilnahme ~ einladung,
                                                 stark))[2]),
         `2SLS (strong)` = unname(coef(ivreg(lohn ~ teilnahme | einladung,
                                             data = stark))[2]),
         `first stage (weak)`   = unname(coef(lm(teilnahme ~ einladung,
                                                 schwach))[2]),
         `2SLS (weak)`   = unname(coef(ivreg(lohn ~ teilnahme | einladung,
                                             data = schwach))[2]))
}) %>% bind_rows()
```

```{r iv-exclusion-table, echo=FALSE}
knitr::kable(verletzt, digits = 4, row.names = FALSE,
  caption = "The same small violation of the exclusion restriction, run through a strong instrument and a weak one. The true local effect is 0.086 in every row. A direct effect of two log points --- a rounding error in any real setting --- moves the strong estimate by seven log points and the weak one by sixty-three.")
```

The arithmetic is one line. A direct effect $\gamma$ of the instrument on the
outcome adds $\gamma / \pi$ to the estimate, where $\pi$ is the first stage. With
$\pi = 0.30$ the violation is multiplied by 3.3; with $\pi = 0.03$ it is
multiplied by more than thirty. Note also the first row, where the letter is
perfectly valid: the weak instrument is already half again too high before any
violation at all, because a small denominator makes an estimator noisy as well as
fragile.

::: {.analysis-warning}
This is why "my instrument is weak but at least it is valid" is not a defence.
Weakness and invalidity multiply. An instrument that is exactly valid can afford
to be weak; an instrument that is *almost* valid cannot.
:::

Three cautionary tales, all of them instruments that once looked obvious.

**Quarter of birth.** Angrist and Krueger's compulsory-schooling instrument
assumes that when in the year you were born affects your earnings only through
how long you stayed at school. Buckles and Hungerman, in the *Review of Economics
and Statistics* in 2013, documented that mothers who give birth in winter differ
systematically from mothers who give birth in summer — they are
disproportionately teenagers and disproportionately unmarried — and that family
background can explain up to half of the relationship between season of birth and
adult outcomes.

**Rainfall.** Rain shocks have been used as instruments for income, for conflict,
for agricultural output, for migration. Heather Sarsons studied Indian districts
downstream of large irrigation dams, where income is much less sensitive to rain,
and found that rain shocks predicted riots there just as strongly — direct
evidence that rain reaches conflict by some route other than income. Jonathan
Mellon has since catalogued 194 separate variables that the literature links to
weather, each of them a potential exclusion violation, across a review of 289
studies.

**Streams.** Caroline Hoxby instrumented the amount of school choice in an
American metropolitan area with the number of streams running through it, on the
theory that rivers historically fragmented school districts. Jesse Rothstein
could not reproduce the sample or the hand-counted stream variable, and found the
estimates sensitive to both the instrument list and the sample definition. The
exchange ran in the *AER* in 2007 and is worth reading in full.

:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
Jonathan Mellon, "Rain, rain, go away: 194 potential exclusion-restriction
violations for studies using weather as an instrumental variable", *American
Journal of Political Science*, 2025 — a list, and an education.

Isaiah Andrews, James H. Stock and Liyang Sun, "Weak Instruments in Instrumental
Variables Regression: Theory and Practice", *Annual Review of Economics* 11, 2019,
727–753 — what to report, and why.
::::

### Returns to Schooling with Real Data {#iv-returns-to-schooling-with-real-data}

Enough simulation. The most instructive instrumental-variables data set in
economics is David Card's, from a 1993 working paper published in 1995 as *Using
Geographic Variation in College Proximity to Estimate the Return to Schooling*.
The data are the National Longitudinal Survey of Young Men — men interviewed from
1966, with wages measured in 1976 — and the instrument is whether the respondent
grew up in a local labour market that contained a four-year college.

The argument is that living near a college lowers the cost of attending one
without otherwise determining what you earn twenty years later.

```{r iv-card-load, echo=TRUE}
card <- read.csv(file.path("data", "Causal", "card.csv"))
c(observations = nrow(card), near_a_college = mean(card$nearc4))
```

Before any estimator, the same four numbers as before.

```{r iv-card-wald, echo=TRUE}
roh <- card %>%
  group_by(nearc4) %>%
  summarise(n = n(), schooling = mean(educ), log_wage = mean(lwage))
roh

diff(roh$log_wage) / diff(roh$schooling)
```

Nineteen per cent per year of schooling, which is implausible, and it is
implausible for a reason the next table makes obvious. Men who grew up near a
college did not grow up in comparable places.

```{r iv-card-balance, echo=TRUE}
card %>%
  group_by(nearc4) %>%
  summarise(n = n(),
            urban_1966  = mean(smsa66),
            south_1966  = mean(south66),
            black       = mean(black),
            father_educ = mean(fatheduc, na.rm = TRUE)) %>%
  knitr::kable(digits = 3,
    caption = "Balance on the college-proximity instrument, and there is none. Four-year colleges are in cities, so the instrument is partly a measure of having grown up urban --- which is why every specification below carries a long list of controls.")
```

So the instrument is defensible only conditional on controls, and the controls do
a great deal of work. This is Card's specification: experience and its square,
race, region, urban status now and in 1966.

```{r iv-card-models, echo=TRUE}
kontr <- paste("exper + expersq + black + south + smsa +",
               "reg661 + reg662 + reg663 + reg664 + reg665 +",
               "reg666 + reg667 + reg668 + smsa66")

card_ols <- lm(as.formula(paste("lwage ~ educ +", kontr)), data = card)
card_fs  <- lm(as.formula(paste("educ  ~ nearc4 +", kontr)), data = card)
card_iv  <- ivreg(as.formula(paste("lwage ~ educ +", kontr,
                                   "| nearc4 +", kontr)), data = card)
```

```{r iv-card-table, echo=FALSE}
modelsummary(list("OLS" = card_ols,
                  "First stage" = card_fs,
                  "2SLS" = card_iv),
  coef_map = c("educ" = "Years of schooling",
               "nearc4" = "Grew up near a college"),
  gof_map = "nobs",
  title = "Card's returns to schooling: least squares, the first stage, and two-stage least squares. Proximity buys about a third of a year of schooling; instrumenting with it nearly doubles the estimated return and multiplies its standard error by fifteen.",
  notes = "All specifications include experience and its square, race, region and urban status. Standard errors in parentheses.")
```

Ordinary least squares says 7.5% per year. Two-stage least squares says 13.2%,
with a standard error of 5.5 points. The first stage is
`r sprintf("%.3f", coef(card_fs)["nearc4"])` years of schooling, which is real
but modest, and its F statistic is:

```{r iv-card-f, echo=TRUE}
summary(card_iv, diagnostics = TRUE)$diagnostics %>% round(3)
```

Thirteen. Above the folklore threshold of ten, below the Stock-Yogo threshold of
16.38, and far below the 104.7 that Lee and co-authors require for the
$t$-statistic in the table above to mean what it appears to mean. So compute the
interval that does not depend on the first stage being strong.

```{r iv-card-ar, echo=TRUE, cache=TRUE}
ar_card <- function(b) {
  d <- transform(card, rest = card$lwage - b * card$educ)
  voll <- lm(as.formula(paste("rest ~ nearc4 +", kontr)), data = d)
  ohne <- lm(as.formula(paste("rest ~", kontr)), data = d)
  anova(ohne, voll)$F[2]
}

gitter_card   <- seq(-0.2, 0.6, by = 0.002)
stat_card     <- vapply(gitter_card, ar_card, numeric(1))
behalten_card <- gitter_card[stat_card < qf(0.95, 1, nrow(card) - 17)]

c(ar_lower = min(behalten_card), ar_upper = max(behalten_card),
  conventional_lower = unname(coef(card_iv)["educ"] -
    1.96 * sqrt(diag(vcov(card_iv)))["educ"]),
  conventional_upper = unname(coef(card_iv)["educ"] +
    1.96 * sqrt(diag(vcov(card_iv)))["educ"]))
```

The identification-robust interval is wider and shifted upwards, and it is the
one to quote. The estimate remains distinguishable from zero and remains almost
entirely uninformative about whether the true return to a year of schooling is
three per cent or twenty-eight.

Now the part that made the paper famous. Two-stage least squares is *larger* than
ordinary least squares, and the textbook prediction runs the other way: if
schooling is correlated with unmeasured ability, and ability raises wages, then
OLS should be biased upwards and the instrument should push it down. Card's own
explanation is the one this section has been building towards. Proximity to a
college does not move the sons of professors, who were going to university
regardless. It moves men whose parents did not go, for whom the cost of leaving
home was the binding constraint — and those men, precisely because they were
constrained, have unusually high returns to the schooling they were missing. The
complier population is not the population, and Card says so in the abstract: the
gains "are concentrated among men with poorly-educated parents".

```{r iv-card-sensitivity, echo=TRUE}
varianten <- c("Card's controls"           = "",
               "+ parental education"      = "+ fatheduc + motheduc",
               "+ parental education, IQ"  = "+ fatheduc + motheduc + IQ")

lapply(names(varianten), function(nm) {
  zusatz <- varianten[[nm]]
  d <- card
  if (grepl("fatheduc", zusatz)) d <- d[!is.na(d$fatheduc) & !is.na(d$motheduc), ]
  if (grepl("IQ", zusatz))       d <- d[!is.na(d$IQ), ]
  cc <- paste(kontr, zusatz)
  m  <- ivreg(as.formula(paste("lwage ~ educ +", cc, "| nearc4 +", cc)),
              data = d)
  fs <- lm(as.formula(paste("educ ~ nearc4 +", cc)), data = d)
  tibble(Specification   = nm,
         n               = nrow(d),
         OLS             = unname(coef(lm(as.formula(
                             paste("lwage ~ educ +", cc)), d))["educ"]),
         `2SLS`          = unname(coef(m)["educ"]),
         `Std. error`    = unname(sqrt(diag(vcov(m)))["educ"]),
         `First-stage F` = unname(summary(fs)$coefficients["nearc4",
                                                           "t value"]^2))
}) %>% bind_rows() %>%
  knitr::kable(digits = 4, row.names = FALSE,
    caption = "The same instrument under three control sets. Adding the parental education that Card's own story says matters cuts the estimate almost in half, halves the first-stage F and drops eight hundred observations --- and every one of these specifications has appeared in print somewhere.")
```

::: {.analysis-note}
That last table is not an argument that Card was wrong. It is an argument about
what a single instrumental-variables estimate can carry. The point estimate moves
from 0.13 to 0.08 to 0.04 as controls are added, and none of the three intervals
excludes any of the three point estimates. The data are consistent with a great
many stories, and the instrument is not strong enough to choose between them.
:::

### An Instrument Somebody Designed on Purpose {#iv-an-instrument-designed-on-purpose}

Return to the Malawi experiment of Section \@ref(exp-thornton). Rebecca Thornton
gave every person tested for HIV a randomly drawn voucher, worth between nothing
and about three dollars, redeemable only on collecting their result. She then
asked what *learning your status* does to behaviour.

That second question cannot be answered by comparing people who collected their
result with people who did not, for the same reason the training example could
not be answered by comparing participants with non-participants. But the voucher
was randomised, it moved collection enormously, and there is no obvious route
from a one-dollar voucher to a person's later behaviour except through the
knowledge it bought. That is an instrument, and it is an instrument by design
rather than by discovery.

```{r iv-thornton, echo=TRUE}
hiv <- read.csv(file.path("data", "Causal", "thornton_hiv.csv")) %>%
  rename(learned = got, incentive = any, distance = distvct) %>%
  filter(!is.na(learned), !is.na(incentive))

erste_stufe <- lm(learned ~ incentive, data = hiv)

c(first_stage = unname(coef(erste_stufe)["incentive"]),
  F           = unname(summary(erste_stufe)$fstatistic[1]))
```

Forty-five percentage points, and an F statistic of about 550. Set that beside
Card's thirteen. Both are published, both are in the canon, and one of them is
forty times the strength of the other.

::: {.analysis-takeaway}
The difference between the two is not statistical skill. It is that Thornton
*built* her instrument and Card *found* his. A designed instrument can be made
strong, can be documented, and can have its exclusion restriction argued from the
protocol. A found instrument takes the strength the world happens to give it.
:::

The teaching extract distributed with this book carries the first stage but not
the behavioural outcome, so the second stage cannot be run here.^[In the paper it
can. Thornton reports that among sexually active HIV-positive individuals,
learning one's result raised condom purchases by 1.69 condoms, with a standard
error of 0.784 — an effect that is real, statistically detectable and small in
absolute terms. There was no detectable effect among HIV-negative individuals.]
What the extract does show is the shape of a well-built encouragement design: a
strong first stage, a randomised nudge, and a treatment nobody was forced into.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: every experiment with imperfect compliance is an IV study</h2>
:::
This is the connection worth carrying away from the whole chapter. Randomise a
treatment; some of the treatment group refuse, some of the control group find a
way in. Compare the groups as *assigned* and you have the intention-to-treat
effect, which is unbiased and answers a different question from the one you
asked. Compare the groups as *treated* and you have thrown the randomisation
away. Divide the first by the compliance rate and you have a local average
treatment effect, estimated by instrumental variables, with assignment as the
instrument.

Fuzzy regression discontinuity, Section \@ref(rdd-sharp-and-fuzzy), is the same
algebra with the threshold as the instrument. The Wald ratio there is this Wald
ratio. The 2021 Nobel committee's *Scientific Background* makes the same point in
its own way: the framework Angrist and Imbens built is what Hahn, Todd and van
der Klaauw (2001) used to work out the conditions under which the discontinuity
design identifies a causal effect at all.
::::

### Opening the Envelope {#iv-opening-the-envelope}

Back to the world where the answer is known, and this time not for one data set
but for three hundred.

```{r iv-envelope, echo=TRUE, cache=TRUE}
ziehungen <- lapply(1:300, function(i) {
  d <- simulate_weiterbildung(n = 4000, seed = 30000 + i)
  m <- ivreg(lohn ~ teilnahme | einladung, data = d)
  o <- lm(lohn ~ teilnahme + alter + frau + schule + lohn0, data = d)
  tibble(late   = mean(d$tau[d$typ == "complier"]),
         iv     = unname(coef(m)["teilnahme"]),
         iv_se  = unname(sqrt(diag(vcov(m)))["teilnahme"]),
         ols    = unname(coef(o)["teilnahme"]),
         ols_se = unname(sqrt(diag(vcov(o)))["teilnahme"]))
}) %>% bind_rows()
```

```{r iv-envelope-table, echo=FALSE}
tibble(
  Estimator = c("OLS, fully controlled", "2SLS with the letter"),
  `Mean estimate`  = c(mean(ziehungen$ols), mean(ziehungen$iv)),
  `Std. deviation` = c(sd(ziehungen$ols), sd(ziehungen$iv)),
  `Bias vs LATE`   = c(mean(ziehungen$ols - ziehungen$late),
                       mean(ziehungen$iv  - ziehungen$late)),
  `RMSE vs LATE`   = c(sqrt(mean((ziehungen$ols - ziehungen$late)^2)),
                       sqrt(mean((ziehungen$iv  - ziehungen$late)^2))),
  `Coverage`       = c(mean(abs(ziehungen$ols - ziehungen$late) <
                              1.96 * ziehungen$ols_se),
                       mean(abs(ziehungen$iv - ziehungen$late) <
                              1.96 * ziehungen$iv_se))) %>%
  knitr::kable(digits = 4,
    caption = "Three hundred studies of the same programme. The controlled regression is about six times more precise than two-stage least squares and covers the truth in none of the three hundred; the instrument is noisy and covers it in ninety-four per cent of them. Precision about the wrong quantity is not an asset.")
```

Two things to take from that table. The instrument is *expensive*: its standard
deviation across draws is about six times the regression's, and every one of
those extra units of imprecision buys nothing except the right to be unbiased.
And the regression's coverage is zero — not low, zero — because a tight interval
around a wrong number never contains the right one.

```{r iv-envelope-picture, echo=FALSE, fig.align='center', fig.width=7.4, fig.height=3.2, fig.cap="Three hundred estimates from each of two estimators, against the local average treatment effect both are aimed at. The regression is a narrow spike in the wrong place; the instrumental-variables estimator is a wide cloud in the right one."}
ziehungen %>%
  select(iv, ols) %>%
  tidyr::pivot_longer(everything(), names_to = "estimator",
                      values_to = "estimate") %>%
  mutate(estimator = factor(estimator, levels = c("ols", "iv"),
                            labels = c("OLS, fully controlled",
                                       "2SLS with the letter"))) %>%
  ggplot(aes(estimate, fill = estimator)) +
  geom_histogram(bins = 40, colour = "white", linewidth = .15, alpha = .9) +
  geom_vline(xintercept = mean(ziehungen$late), colour = "grey20",
             linewidth = .6) +
  scale_fill_manual(values = c("grey55", "#2c6e8e"), guide = "none") +
  facet_wrap(~ estimator, scales = "free_y") +
  labs(x = "estimate of the effect on log hourly wage", y = NULL) +
  theme_minimal(base_size = 10)
```

### What Instrumental Variables Cannot Do {#iv-what-instrumental-variables-cannot-do}

#### The central assumption is not testable

Everything in this section rests on the claim that the letter reaches wages only
through the course, and no quantity of data can establish that claim. The
overidentification test compares instruments with each other. The balance table
compares what you measured. The first-stage F measures strength, not validity.
Where the exclusion restriction is concerned statistics has nothing to offer, and
the argument has to be made from institutional knowledge: who wrote the rule,
what else the rule does, what the affected people would say if you asked them.

#### It answers a question the instrument chose

The estimate is a local average treatment effect, and *local* means local to the
compliers. Change the instrument and you change the population, and therefore the
number, with no contradiction and no error anywhere. This is why two
well-executed studies of the same treatment can report different effects and both
be right, and it is why reporting the complier profile — the three-line table of
Section \@ref(iv-who-is-the-answer-about) — should be routine and is not.

#### It is expensive

The envelope table above is the general result. Instrumental variables throws away
all the variation in the treatment that the instrument does not explain, which is
usually most of it, and pays for the discarding in standard errors. Alwyn Young
re-ran 1,309 instrumental-variables regressions from thirty published papers and
found that two-stage least squares "rarely rejects the OLS point estimate", that
weak-instrument pre-tests were "largely uninformative of both size and bias", and
that a substantial share of significant results do not survive a bootstrap.

#### It cannot be rescued with more data

Bias from a violated exclusion restriction does not shrink as $n$ grows; it is
inconsistency, not imprecision. The estimate converges, beautifully and with
ever-tighter intervals, on the wrong number. This is the failure mode no
robustness table catches, and it is why the honest sequence is always the same:
first ask where the variation came from, then choose the estimator.

### Readings {#iv-readings -}

Joshua D. Angrist and Jörn-Steffen Pischke, *Mostly Harmless Econometrics*
(Princeton, 2009), chapter 4 — still the clearest exposition of the local average
treatment effect, and the source of most of the vocabulary used above.

Guido W. Imbens and Joshua D. Angrist, "Identification and Estimation of Local
Average Treatment Effects", *Econometrica* 62(2), 1994, 467–475 — nine pages that
changed what the estimator was understood to be estimating.

Joshua D. Angrist, Guido W. Imbens and Donald B. Rubin, "Identification of Causal
Effects Using Instrumental Variables", *Journal of the American Statistical
Association* 91(434), 1996, 444–455, with discussion — where compliers,
always-takers, never-takers and defiers get their names, and where the discussion
by Robins, Greenland, Rosenbaum and others is worth as much as the paper.

John Bound, David A. Jaeger and Regina M. Baker, "Problems with Instrumental
Variables Estimation When the Correlation Between the Instruments and the
Endogenous Explanatory Variable Is Weak", *JASA* 90(430), 1995, 443–450 — the
paper that started the weak-instrument literature by replacing a real instrument
with random numbers and getting results that looked perfectly reasonable.

Scott Cunningham, *Causal Inference: The Mixtape* (Yale, 2021), chapter 7
[@cunningham2021causal] — the same material with more code and a lighter touch.

This section closes the catalogue of designs that began in Section
\@ref(experiments). Randomisation severs the link between treatment and
confounders directly; matching rebuilds comparability out of covariates; timing
and thresholds borrow it from institutions; an instrument borrows it from a
variable with no business in the outcome equation. Chapter \@ref(extract) turns
from what the data can identify to what can be pulled out of them.
