## Discontinuities {#discontinuities}

::: {.analysis-subtitle}
Regression discontinuity designs
:::

```{r rdd-setup, echo=FALSE}
library(rdrobust)
library(rddensity)
library(rdlocrand)
library(estimatr)
```

### A Line Drawn Through a List of Schools {#rdd-a-line}

In the spring of 2024 the school ministry of North Rhine-Westphalia sorted every
one of its schools by a single number. The number is the *Schulsozialindex*,
first built at the Ruhr-Universität Bochum in 2020 and revised in 2023, and it
is a factor score over four administrative indicators: the share of
under-eighteens in the school's catchment area living in households on basic
income support, the share of pupils whose family language is not German, the
share who moved to Germany themselves, and the share with a diagnosed need for
support in learning, language or emotional and social development. The ministry
is explicit that the last of these is not counted on its own — schools carrying
*both* a high poverty rate and a high share of such pupils are weighted more
heavily, which is an interaction in everything but name. The score is then
pressed into nine steps. Since the 2023 revision, step 9 holds the most
disadvantaged five per cent of schools and steps 1 to 8 split the remaining
ninety-five at equal intervals.[^rdd-sozialindex]

Then the ministry drew a line. Primary schools in steps 6 to 9 and secondary
schools in steps 7 to 9 were invited to join the *Startchancen* programme —
twenty billion euros over ten years, which the federal ministry calls the
largest and longest-running education programme in the history of the Federal
Republic. Schools below the line were not invited. North Rhine-Westphalia took
its schools in two tranches: 400 started in the 2024/25 school year and 523
followed in 2025/26, 923 in all.

Consider two primary schools in the same city. One has an index score that puts
it barely inside step 6. The other is a fraction of a point below and lands in
step 5. Everything we can measure about them is nearly identical: the same kind
of neighbourhood, the same share of families on benefits, the same staffroom
problems. One gets a decade of money, staff and support. The other gets a letter
saying it was not selected.

In 2030 somebody will have to say whether the money worked. Section
\@ref(difference-in-differences) showed one way to answer that: compare the
change over time in funded schools with the change over time in unfunded ones,
and defend the parallel-trends assumption. This section shows the other way, and
it is a way that needs no assumption about trends at all. It needs only the two
schools above — and the observation that whatever separates them, it cannot be
much.

::: {.analysis-takeaway}
When a programme is handed out by a rule, and the rule has a threshold, the
threshold itself is the experiment. The people on either side of it are as good
as randomised, and the rule tells you so.
:::

### What Thistlethwaite and Campbell Were Trying to Fix {#rdd-what-thistlethwaite-and-campbell-were-trying-to-fix}

The method was invented in 1958 and published in 1960, by Donald L.
Thistlethwaite and Donald T. Campbell, in the *Journal of Educational
Psychology*. The title tells you what problem they had in mind: *Regression-Discontinuity
Analysis: An Alternative to the Ex Post Facto Experiment*.

The ex post facto experiment was the dominant design in mid-century American
sociology, associated above all with F. Stuart Chapin's *Experimental Designs in
Sociological Research* (1947) and Ernest Greenwood's *Experimental Sociology*
(1945). The recipe: wait until a programme has run, then find, for every person
who received it, a person who did not but who looks the same on a list of
measured background variables, and compare the pairs. You will recognise it. It
is matching, done before anyone had worked out what matching can and cannot
promise — the subject of Section \@ref(matching-and-balancing). Campbell's
objection was not that matching is useless. It was that matching on a handful of
fallible measurements buys the appearance of comparability without the substance,
and that regression to the mean will then manufacture an effect out of nothing.

Thistlethwaite and Campbell had in front of them a setting where the selection
rule was not a mystery to be reconstructed but a published number. The National
Merit Scholarship programme gave two grades of public recognition to American
high-school students on the strength of a qualifying test: a *Certificate of
Merit* to those above a state-specific cut-off, a *Letter of Commendation* to
those below it. The two groups were treated differently by their schools, their
teachers and their own self-image — and the only thing that separated a student
at the bottom of the certificate group from one at the top of the commendation
group was a point or two on a test.

So they plotted the outcomes against the test score and looked for a step in the
line at the cut-off. That is the whole idea. There is no matching, no covariate
list, no assumption that treated and untreated units are alike overall. There is
only the claim that a curve which is smooth everywhere else has no business
jumping at exactly the point where the rule changes.

Campbell went on to make this the centrepiece of a much larger argument. In the
chapter he wrote with Julian Stanley for the *Handbook of Research on Teaching*
(1963), later reprinted as *Experimental and Quasi-Experimental Designs for
Research*, the method appears as Design 16. In *Reforms as Experiments*
(*American Psychologist*, 1969) it becomes a proposal for how governments should
work: administer scarce benefits by an explicit rule, record the rule, and the
evaluation writes itself.

> If there is no significant difference for these at the decision-point
> intercept, then the tie-breaking experiment should show no difference. In
> cases where the tie breakers would show an effect, there should be an abrupt
> discontinuity in the regression line.
>
> --- Campbell (1969, p. 420)

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
A **regression discontinuity design** exploits a treatment whose assignment — or
whose *eligibility* — is a known function of an observed, continuous variable — the **running variable** or
**score** — crossing a fixed **cut-off**. The causal quantity of interest is the
vertical gap between the two regression functions at the cut-off:

$$\tau = \lim_{x \downarrow c} E[y \mid x] - \lim_{x \uparrow c} E[y \mid x]$$

Everything in the design turns on that limit being taken from data near $c$, and
on there being no reason other than the treatment for the two limits to differ.
::::

### Waiting for Life to Arrive {#rdd-waiting-for-life-to-arrive}

And then, for thirty-five years, almost nothing happened.

Thomas Cook told this story in a paper whose title is a nod to Beckett:
*Waiting for Life to Arrive: A history of the
regression-discontinuity design in Psychology, Statistics and Economics*
(*Journal of Econometrics*, 2008). Cook counts the citations and finds an
invention that nobody used.

Part of the reason is that the inventor talked it down. Campbell and Stanley
described the design as "very limited in its range of applications", and Cook
notes drily that this is "not a ringing endorsement, particularly when coming
from the method's own developer". Statisticians found the problem uninteresting
for a different reason: once you know the assignment rule completely, where is
the difficulty? And economists, Cook writes, had the design "off the radarscope"
by 1980, "not to re-emerge for another 15 years or so".

The most interesting part of Cook's history is that the design kept being
invented from scratch, in fields that did not read each other. Psychology got it
in 1960. Economics got it in 1972, from Arthur Goldberger. Statistics got it
again in 1977, from Donald Rubin. And public health rediscovered it in 1996
under the name *risk-based allocation*. Goldberger, the economist,
did not cite Campbell and did not think he was writing about regression
discontinuity at all. Goldberger was asking a different question: what happens
to a treatment-effect estimate when people are selected on a *measured*
pre-test rather than on their true, unmeasured ability? His answer was that
selection on the observed score produces no bias — the distinction that we now
call selection on observables versus selection on unobservables. Regression
discontinuity is the limiting case of his argument.

Goldberger also computed the price. For a cut-off at the midpoint of a normally
distributed score, he found the design roughly 2.75 times less efficient than an
experiment with the same number of units.[^rdd-goldberger] Keep that number. It is the single most useful fact about regression
discontinuity in practice, and it explains most of the disappointments.

::: {.analysis-note}
Two and three-quarter times less efficient means you need nearly three times the
sample to reach the same precision. A threshold study with 500 units on each
side — a thousand in all — has roughly the power of a randomised trial with 360
units in total. Regression
discontinuity is not a cheap experiment; it is an expensive one that somebody
else already paid for.
:::

Economics came back to the design in the late 1990s, and it came back fast:
Wilbert van der Klaauw on financial aid offers (circulated from the
mid-1990s, published in 2002), Joshua Angrist and Victor Lavy
on Israeli class sizes, Sandra Black on school-attendance boundaries. The
theoretical account arrived in 2001, when Jinyong Hahn, Petra Todd and van der
Klaauw published nine pages in *Econometrica* that set out what the design
actually assumes. Their three results organise everything that follows in this
section: identification rests on *continuity*, not on unconfoundedness; the
estimand is a local effect at the cut-off and nowhere else; and when the rule is
obeyed only partly, the estimator is a ratio — an instrumental-variables
estimator in disguise.

:::: {.amazing}
::: {.titelamazing}
<h2>The prize that was not for this design</h2>
:::
The 2021 Nobel Memorial Prize went to David Card, Joshua Angrist and Guido
Imbens, and it is often reported as a prize for regression discontinuity. It was
not. The citations read "for his empirical contributions to labour economics"
and "for their methodological contributions to the analysis of causal
relationships", and the words *regression discontinuity* do not appear in the
press release at all.

They do appear twice in the committee's *Scientific Background*, and both times
the credit goes elsewhere. Once to name Donald Campbell as the man who
"developed one of the main empirical methods --- the regression discontinuity
(RD) design" for natural experiments. And once to note that the framework
Angrist and Imbens built was later used to work out "the conditions under which
a causal effect can be identified when using other methods for causal inference,
including the regression discontinuity design (Hahn, Todd, and van der Klaauw,
2001)".

That second sentence is the real connection, and it is worth more than the
misattribution. The language in which Hahn, Todd and van der Klaauw stated what
a fuzzy discontinuity identifies --- compliers, monotonicity, a local average
treatment effect --- is exactly the language Angrist and Imbens had invented for
instrumental variables a decade earlier. A prize for a vocabulary is still a
prize for the sentences you can write in it.
::::

### A Catalogue of Thresholds {#rdd-a-catalogue-of-thresholds}

Before the machinery, the payoff. The following studies all answer a question
that had resisted every other design, and they do it because somebody once wrote
a rule with a number in it.

```{r rdd-catalogue, echo=FALSE}
catalogue <- tibble::tribble(
  ~Study, ~Year, ~Field, ~`Running variable`, ~`Cut-off`, ~Finding,
  "Ludwig & Miller, QJE", 2007L, "Education",
  "county poverty rate, 1960 census", "59.20%",
  "Head Start grant-writing help cut child mortality from Head-Start-treatable causes by a third to a half of the control mean of 3.2 per 100,000",
  "Carpenter & Dobkin, AEJ: Applied", 2009L, "Public health",
  "age in days", "21st birthday",
  "all-cause mortality jumps 9%, about 10 extra deaths per 100,000 person-years; alcohol-related deaths +35%, motor vehicle +15%, suicide +16%",
  "Card, Dobkin & Maestas, QJE", 2009L, "Health economics",
  "age in days at admission", "65th birthday",
  "Medicare eligibility cuts 7-day mortality of emergency patients by nearly one percentage point on a base of about 5%, a 20% reduction",
  "Almond, Doyle, Kowalski & Williams, QJE", 2010L, "Health economics",
  "birth weight in grams", "1,500 g",
  "the 'very low birth weight' label buys about 4,000 USD more care and cuts one-year mortality by roughly one percentage point",
  "Angrist & Lavy, QJE", 1999L, "Education",
  "grade enrolment", "40, 80, 120 ...",
  "Maimonides' rule splits classes; ten fewer pupils raise fifth-grade reading by about 0.25 to 0.3 standard deviations",
  "Lee, J. Econometrics", 2008L, "Political economy",
  "margin of victory", "0",
  "winning a US House seat by a hair is worth 7.7-8.1 percentage points at the next election",
  "Hoekstra, REStat", 2009L, "Education",
  "SAT score minus the GPA-specific admission cut-off", "0",
  "attending the flagship state university raises later earnings of white men by about 20%",
  "Dell, Econometrica", 2010L, "Development",
  "location relative to a colonial boundary", "the mita border",
  "districts inside Peru's colonial mining draft have 25% lower household consumption today",
  "Bor, Moscoe, Mutevedzi, Newell & Barnighausen, Epidemiology", 2014L, "Public health",
  "first CD4 count", "200 cells/ul",
  "eligibility for antiretroviral therapy cut mortality, hazard ratio 0.65 (95% CI 0.45-0.94)"
)

DT::datatable(
  catalogue,
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 5, scrollX = TRUE, dom = "ftip",
                  columnDefs = list(list(width = "40%", targets = 5))),
  caption  = "Nine thresholds and what they bought. The table is sortable: click a column heading to reorder it, and use the boxes at the top to search. Read the cut-off column first - every one of these numbers is an administrative rule that somebody wrote down, and every finding exists because of it.")
```

Look at what these have in common. Not one of them is a variable a researcher
chose. A county poverty rate taken from the 1960 census and used once, in 1965;
a birthday; a gram on a hospital scale; a rabbinical ruling from the twelfth century about how many children one
teacher may instruct. The design does not find natural experiments; it finds
*bureaucracies*, and bureaucracies leave thresholds behind them the way glaciers
leave moraines.

::: {.analysis-note}
Two of the entries deserve a second look because of what the naive comparison
says. Sicker newborns get more intensive care, so the raw association between
treatment intensity and infant mortality is *positive*: more care, more death.
And in Israel, small classes are concentrated in schools that need them, so the
raw association between class size and achievement is also positive: bigger
classes, better results. In both cases the threshold reverses the sign.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The table above is sortable, and reading a table is a skill. Click the **Year**
heading to reorder it.

The oldest study in the catalogue was published in `r fitb(1999, num = TRUE)`.

Now sort by **Field**. Four of the nine come from public health or health
economics. Of those four, how many use a *date* — an age or a birthday — as the
running variable? `r fitb(2, num = TRUE)`

Type `birthday` into the search box at the top. Two studies come back. In one of
them, crossing the threshold makes people worse off.
`r mcq(c("Card, Dobkin and Maestas, at the 65th birthday", answer = "Carpenter and Dobkin, at the 21st birthday", "both of them", "neither of them"))`

And one question the table can answer only if you read the third column
properly. Eight of these nine cut-offs were written down by an administrator.
Which one was not?
`r mcq(c("59.20%, the county poverty rate", "1,500 g, the very-low-birth-weight label", answer = "0, the margin of victory in an election", "200 cells/µl, the therapy eligibility threshold"))`
::::


:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
The two Cambridge Elements volumes by Matias Cattaneo, Nicolás Idrobo and Rocío
Titiunik — *A Practical Introduction to Regression Discontinuity Designs:
Foundations* (2020) and *Extensions* (2024) — are the current standard, they are
written around R and Stata code, and the working-paper versions are free on
arXiv. For the older canon, Imbens and Lemieux (2008) and Lee and Lemieux (2010)
are still the papers everyone cites. Cattaneo and Titiunik (2022) in the *Annual
Review of Economics* is the best short overview.
::::

### Sharp and Fuzzy {#rdd-sharp-and-fuzzy}

The design comes in two versions, and the difference is whether the rule is
obeyed.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
In a **sharp** design, treatment is a deterministic step function of the score:
everyone above the cut-off is treated, everyone below is not.

$$d_i = \mathbb{1}\{x_i \geq c\}$$

In a **fuzzy** design, crossing the cut-off changes the *probability* of
treatment but does not settle it:

$$\lim_{x \downarrow c} P(d = 1 \mid x) \neq \lim_{x \uparrow c} P(d = 1 \mid x)$$

The sharp estimand is the jump in the outcome. The fuzzy estimand is the ratio
of the jump in the outcome to the jump in the treatment probability:

$$\tau_{\text{fuzzy}} = \frac{\lim_{x \downarrow c} E[y \mid x] - \lim_{x \uparrow c} E[y \mid x]}{\lim_{x \downarrow c} E[d \mid x] - \lim_{x \uparrow c} E[d \mid x]}$$
::::

The Startchancen case is fuzzy, and for a mundane reason. In North
Rhine-Westphalia the index governed who was *eligible*; it did not settle who
ended up in. The ministry reports both a pre-selection step by the school
inspectorate within the eligible steps, and that "the decision to take part lay
with the schools themselves, in consultation with their local authority". So the
score determines eligibility sharply, and eligibility determines participation
only partly.[^rdd-takeup]

That is exactly the structure of an instrumental variable, and the fuzzy
estimator is a Wald ratio. If this feels familiar, it should: it is the
same algebra as Section \@ref(instruments), applied inside a narrow window
around a number. The assumptions travel too. There must be no *defiers* — no
school that would have joined had it not been invited and refused once it was —
and under that monotonicity condition the ratio identifies the average effect
for **compliers at the cut-off**: schools that took part because they were
invited, and that sit right on the line. Continuity identifies the sharp
estimand; continuity *plus* monotonicity identifies the fuzzy one.

::: {.analysis-warning}
Not "the effect of the programme". Not even "the effect of the programme at the
cut-off". The effect *for the schools that respond to the invitation*, *at the
cut-off*. We will measure the distance between those three quantities later in
this section, once we have a world where all three are known.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
In North Rhine-Westphalia the index decided who was *eligible*; each school then
decided for itself whether to join.

The design is therefore `r mcq(c("sharp", answer = "fuzzy"))`.

Had every invited school joined, and no uninvited school found a way in, the
design would have been
`r mcq(c(answer = "sharp", "still fuzzy", "impossible to estimate at all"))`.

A school that would have joined had it *not* been invited, and refused once it
was, is called a `r fitb("defier", ignore_case = TRUE)`.

True or false: monotonicity can be checked by comparing take-up just above and
just below the cut-off. `r torf(FALSE)`

`r hide("Why false?")`

Take-up above and below tells you the *size* of the jump — the first stage.
Monotonicity is about individual schools moving in the same direction, and a
school's behaviour under the arm it did not get is never observed. Every jump in
take-up is consistent with some compliers and some defiers cancelling out.

`r unhide()`
::::


### The Identifying Assumption {#rdd-the-identifying-assumption}

Section \@ref(experiments) decomposed every naive comparison into a treatment
effect plus a selection bias term, and every design in this chapter is an
argument for why the second term vanishes. Matching argues that it vanishes once
you condition on enough covariates. Difference-in-differences argues that it is
constant over time and cancels. Regression discontinuity argues something
narrower and, on the face of it, much weaker.

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
**Continuity.** The conditional expectations of both potential outcomes are
continuous in the running variable at the cut-off:

$$E[y_i(0) \mid x_i = x] \quad \text{and} \quad E[y_i(1) \mid x_i = x]
  \quad \text{are continuous at } x = c$$

Under continuity, the sharp design identifies

$$\tau_{\text{SRD}} = E[y_i(1) - y_i(0) \mid x_i = c]$$

the average treatment effect **for units exactly at the cut-off** (Hahn, Todd
and van der Klaauw, 2001).
::::

Read the assumption carefully, because two things about it are unusual.

First, it says nothing about units away from the cut-off, and it therefore
permits arbitrarily severe confounding everywhere else. Schools with a high
social index really are different from schools with a low one, in every way you
can imagine and several you cannot. Continuity does not deny that. It only says
that the difference does not arrive all at once, at 60.

Second — and this is what makes the design worth having — continuity is a
statement about a *limit*, and limits are things the data can speak to. You
cannot test whether treated and untreated units are comparable after
conditioning on covariates; that is the whole trouble with matching. But you
*can* look at whether things which should be smooth actually are: the density of
the score, the pre-programme outcomes, covariates that no programme could have
touched. None of those checks proves continuity. Together they make a case that
a reader can inspect, which is more than most designs offer.

```{r rdd-q-continuity, echo=FALSE}
opts_continuity <- c(
  "Schools above the threshold are far poorer than schools below it, right across the range of the index.",
  answer = "Crossing the same threshold also releases 1,110 additional teaching posts, quite apart from the programme.",
  "The share of pupils below the minimum standard rises steeply with the index on both sides of the threshold.",
  "The index is measured with error, and that error is unrelated to everything else in the model."
)
```


:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Continuity is a statement about two conditional expectations at one point. Three
of the four situations below leave it intact. Which one breaks it?

`r longmcq(opts_continuity)`

And the sentence that separates this design from Section
\@ref(matching-and-balancing). Continuity permits arbitrarily severe confounding
`r mcq(c("nowhere", answer = "everywhere except at the cut-off", "only below the cut-off"))`.
::::


### A Programme We Can Check {#rdd-a-programme-we-can-check}

The Startchancen evaluation reports for the first time at the end of 2026 or the
beginning of 2027 and concludes in 2030, so nobody yet knows what the answer is. We will therefore
build a world in which we do.

As in Sections \@ref(mat-data), \@ref(did-data) and
\@ref(iv-a-programme-we-can-check), the way to find out what an estimator does is
to build a world in which the answer is already written down. The simulation
below has 4,200 schools in one federal state. Each has a social
index score between 2 and 98, known to the ministry. Schools at 60 or above are
invited. Invited schools join with a probability that depends on something we
will never observe — call it the leadership's capacity to take on a
ten-year project — and a handful of schools below the line join through a
hardship route — about one school in twenty below the line. The outcome is the
share of Year-4 pupils below the minimum
standard in mathematics, measured in 2029. Nationally that share was about 22
per cent in the 2021 IQB survey; in the schools this programme addresses it runs
far higher.

The effect of participation is negative — fewer children below the standard —
and it is deliberately heterogeneous: larger for schools further above the
threshold, and larger for schools with more capacity. That last choice is what
will separate the three estimands we warned about.

```{r rdd-dgp, echo=TRUE}
simulate_startchancen <- function(n = 4200, seed = 4, world = "clean",
                                  strength = 2.5) {
  set.seed(seed)

  # The published social index: 0-100, higher means more disadvantage.
  index <- pmin(pmax(rnorm(n, mean = 50, sd = 15), 2), 98)

  # Unobserved: how much room to manoeuvre the school leadership has.
  capacity <- rnorm(n)

  # A world in which schools just below the line get themselves moved above it.
  if (world == "manipulated") {
    pushed <- (index >= 58 & index < 60) & (capacity > 0.5)
    index[pushed] <- 60 + runif(sum(pushed), 0, 1)
  }

  invited <- as.integer(index >= 60)

  # Take-up. One latent draw, two thresholds: nobody is a defier.
  u <- runif(n)
  participates <- as.integer(ifelse(invited == 1,
                                 u < plogis( 0.7 + 0.9 * capacity),
                                 u < plogis(-3.2 + 0.9 * capacity)))

  # Share of Year-4 pupils below the minimum standard, as a percentage.
  mean_curve <- function(x) 5 + 50 * plogis((x - 58) / 14)
  y0  <- mean_curve(index) - strength * capacity + rnorm(n, 0, 1.5)
  tau <- -3.2 - 0.11 * (index - 60) - 1.3 * capacity

  tibble(
    school     = seq_len(n),
    index      = index,
    step      = as.integer(as.character(cut(index,
                   breaks = c(-Inf, 20, 30, 40, 50, 60, 70, 80, 90, Inf),
                   labels = 1:9, right = FALSE))),
    invited  = invited,
    participates  = participates,
    benefits       = 4 + 0.55 * index + rnorm(n, 0, 6),
    pre_score     = mean_curve(index) - 1.0 - 0.8 * strength * capacity + rnorm(n, 0, 1.5),
    capacity = capacity,
    tau        = tau,
    below_standard    = y0 + participates * tau
  )
}

sc <- simulate_startchancen()
```

Three quantities are true in this world at once, and they are not the same
number.

```{r rdd-truth, echo=TRUE}
set.seed(99)
cap_draw    <- rnorm(8e6)
complier_weight <- plogis(0.7 + 0.9 * cap_draw) - plogis(-3.2 + 0.9 * cap_draw)  # complier weight

truth <- tibble(
  Quantity = c("effect at the cut-off, all schools",
              "effect at the cut-off, compliers only",
              "intention to treat at the cut-off",
              "effect on all participating schools",
              "effect on every school in the state"),
  Value = c(-3.2,
           -3.2 - 1.3 * sum(cap_draw * complier_weight) / sum(complier_weight),
           mean(complier_weight) * (-3.2 - 1.3 * sum(cap_draw * complier_weight) / sum(complier_weight)),
           mean(sc$tau[sc$participates == 1]),
           mean(sc$tau)))

knitr::kable(truth, digits = 3,
  caption = "Five true answers to five different questions, in percentage points of pupils below the minimum standard. Note that the second and the fourth differ by more than half a point, and the fifth is only three fifths of the second.")
```

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Five numbers, one programme, and only one of them is what a discontinuity design
estimates.

Which? `r mcq(c("effect at the cut-off, all schools", answer = "effect at the cut-off, compliers only", "effect on all participating schools", "effect on every school in the state"))`

The effect on every school in the state is much the smallest of the five, in
absolute value. In this world, why?
`r mcq(c("because most schools never receive any money", answer = "because the effect was built to grow with disadvantage, and the average school in the state is far less disadvantaged than a funded one", "because the index is measured with error"))`

A minister reads the estimate as "what we would get by doubling the programme".
That reading is wrong by roughly what factor? Divide the fifth quantity by the
second: `r fitb(0.6, tol = 0.08)`
::::


```{r rdd-first-look, fig.align='center', fig.cap="Binned means of the outcome and of the participation rate against the social index, with local linear fits on each side. The upper panel drops by about two points at the threshold; the lower panel shows why the design is fuzzy — participation jumps from near zero to about 0.6, not to one.", echo=TRUE}
bin_means <- function(d, yvar, width = 2, lo = 30, hi = 90) {
  d %>%
    filter(index >= lo, index <= hi) %>%
    mutate(centre = floor((index - 60) / width) * width + width / 2 + 60) %>%
    group_by(centre) %>%
    summarise(y = mean(.data[[yvar]]), .groups = "drop") %>%
    mutate(side = ifelse(centre < 60, "below", "above"))
}

local_fit <- function(d, yvar, h = 7) {
  b <- d %>%
    filter(abs(index - 60) <= h) %>%
    mutate(z = index - 60, g = 1 - abs(z) / h, above = as.integer(index >= 60))
  m <- lm(reformulate("z * above", yvar), data = b, weights = b$g)
  tibble(z = c(seq(-h, 0, length.out = 50), seq(0, h, length.out = 50)),
         above = rep(0:1, each = 50)) %>%
    mutate(fit = predict(m, newdata = .),
           index = z + 60,
           side = ifelse(above == 1, "above", "below"))
}

colours <- c(below = "grey35", above = "#C1440E")

points <- bind_rows(
  bin_means(sc, "below_standard")   %>% mutate(panel = "below minimum standard (%), 2029"),
  bin_means(sc, "participates") %>% mutate(panel = "participation rate"))

lines <- bind_rows(
  local_fit(sc, "below_standard")   %>% mutate(panel = "below minimum standard (%), 2029"),
  local_fit(sc, "participates") %>% mutate(panel = "participation rate"))

ggplot(points, aes(centre, y)) +
  geom_vline(xintercept = 60, linetype = "dashed", colour = "grey40") +
  geom_point(aes(colour = side), size = 1.6) +
  geom_line(data = lines, aes(index, fit, colour = side), linewidth = 0.9) +
  scale_colour_manual(values = colours, guide = "none") +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "social index score", y = NULL) +
  theme_minimal(base_size = 11)
```

### What the Naive Comparison Says {#rdd-what-the-naive-comparison-says}

Start with the answer a competent analyst would produce if the threshold had
never occurred to anyone.

```{r rdd-naive, echo=TRUE}
sc %>%
  group_by(participates) %>%
  summarise(n = n(),
            `below standard` = mean(below_standard),
            `social index`   = mean(index), .groups = "drop") %>%
  knitr::kable(digits = 2,
    caption = "Participating and non-participating schools, compared directly. The programme schools look about eight points worse.")

models <- list(
  "raw"                = lm(below_standard ~ participates, data = sc),
  "+ index"            = lm(below_standard ~ participates + index, data = sc),
  "+ index, covariates" = lm(below_standard ~ participates + index + benefits + pre_score, data = sc))

purrr::imap_dfr(models, ~tibble(
  Model = .y,
  Estimate = coef(.x)["participates"],
  SE = summary(.x)$coefficients["participates", 2],
  `CI low`  = confint(.x)["participates", 1],
  `CI high` = confint(.x)["participates", 2])) %>%
  knitr::kable(digits = 3,
    caption = "Three regressions, three stories. The raw comparison has the wrong sign; adding the index flips it; adding covariates shrinks the standard error to a tenth of a point.")
```

The raw comparison says the programme made things markedly worse. Nobody would
publish that, so the analyst controls for the index, and the sign flips to about
$-3.7$. Add the covariates and the estimate is $-3.9$ with a standard error of
one tenth of a percentage point. The true effect for compliers at the cut-off is
$-3.49$. So the controlled regression is *nearly right* — and that is the most
dangerous outcome of all, because there is no way to know it from the output.

Here is the demonstration. Everything in the world stays fixed except how much
the unobserved leadership capacity matters for the outcome. That parameter
changes no treatment effect at all: the effect for compliers at the cut-off is
$-3.49$ in every row below. It changes only how much of the difference between
participants and non-participants is confounding.

```{r rdd-drift, cache=TRUE, echo=TRUE}
drift <- purrr::map_dfr(c(0, 1, 2.5, 5), function(st) {
  v <- vapply(1:100, function(i) {
    d <- simulate_startchancen(seed = 2000 + i, strength = st)
    c(coef(lm(below_standard ~ participates + index + benefits + pre_score, data = d))["participates"],
      rdrobust(d$below_standard, d$index, c = 60, fuzzy = d$participates)$coef[1, 1])
  }, numeric(2))
  tibble(`unobserved factor` = st,
         OLS = mean(v[1, ]), `OLS sd` = sd(v[1, ]),
         RD = mean(v[2, ]), `RD sd` = sd(v[2, ]))
})

knitr::kable(drift, digits = 3,
  caption = "One hundred simulated studies for each row. The regression estimate walks from -3.08 to -4.31 as the unobserved factor grows; the discontinuity estimate stays put and only loses precision. The truth is -3.49 in every row.")
```

Even in the first row, where capacity does not touch the outcome at all, the
regression sits at $-3.08$ rather than $-3.49$. That last gap is not
confounding: with heterogeneous effects, a regression on everybody is aiming at
a different average than the compliers at the threshold. It never was going to
land on $-3.49$.

::: {.analysis-takeaway}
The controlled regression is not wrong by a lot. It is wrong by an amount that
depends on something you cannot see, and its standard error does not know that.
The discontinuity estimate is noisier and honest about it: it drifts by a sixth
of a point across the four rows and carries a small smoothing bias of its own,
but nothing like the regression's march.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The covariate-adjusted regression gave $-3.9$ with a standard error of one tenth
of a point. The truth for compliers at the cut-off is $-3.49$.

So the regression is wrong by about `r fitb(0.4, tol = 0.06)` points, which is
about `r fitb(4, tol = 0.6)` standard errors.

Across the four rows of the table, the regression walked from $-3.08$ to
$-4.31$. What was changed between the rows?
`r mcq(c("the sample size", "the true treatment effect", answer = "how strongly an unobserved variable affects the outcome", "the bandwidth"))`

True or false: a reader who saw only the third row of that table could have told
from the output that the estimate was too large. `r torf(FALSE)`
::::


### The Jump, by Hand {#rdd-the-jump-by-hand}

Before reaching for a package, do it with `lm()`. Three decisions have to be
made, and the code makes all three visible.

*How wide a window?* Take schools within seven index points of the threshold.
*What shape inside it?* A straight line on each side. *Which observations
count most?* Those closest to the cut-off, through a triangular weight that
falls linearly to zero at the edge of the window.

```{r rdd-by-hand, echo=TRUE}
h <- 7

window <- sc %>%
  filter(abs(index - 60) <= h) %>%
  mutate(z = index - 60, weight = 1 - abs(z) / h)

nrow(window)

# The wrong way first: just compare the two group means inside the window.
mean(window$below_standard[window$invited == 1]) -
  mean(window$below_standard[window$invited == 0])

# The right way: let each side have its own intercept and its own slope.
local_linear <- lm(below_standard ~ z * invited, data = window, weights = weight)
summary(local_linear)$coefficients
```

The difference of means inside the window is $+3.55$: still the wrong sign, even
after throwing away seven of every ten schools. The window is narrow, but the
outcome climbs steeply across it, and the invited schools sit systematically on
the upper half. The `invited` coefficient — the vertical gap between two
fitted lines evaluated at $z = 0$ — is $-1.99$.

That coefficient estimates the intention-to-treat effect: the consequence of being
*invited*, not of taking part. It is a sharp regression discontinuity on
eligibility, and it is the numerator of everything that follows.

::: {.analysis-note}
Why a straight line rather than a curve? Because at a cut-off you are always
extrapolating to a boundary, and boundary estimates from a local *constant* fit
carry a bias of order $h$, while a local *linear* fit reduces it to order $h^2$.
Fitting the mean of a window is the one thing you should never do — as the
$+3.55$ above demonstrates.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Inside a window seven index points wide the difference of group means was
$+3.55$; the local linear estimate on the same schools was $-1.99$.

The two disagree by more than five points on identical data. Why?
`r mcq(c("the local linear fit quietly drops the schools far from the cut-off", answer = "the outcome climbs steeply across the window, and the invited schools all sit on its upper half", "the triangular weights are miscalculated"))`

True or false: shrinking the window far enough would eventually bring the
difference of means into agreement with the local linear estimate.
`r torf(TRUE)`

Fill in the two orders of bias. A local *constant* fit at a boundary carries a
bias of order `r mcq(c(answer = "h", "h²", "1/h", "h³"))`; a local *linear* fit
reduces it to order `r mcq(c("h", answer = "h²", "1/h", "h³"))`.
::::


### The Picture Is a Choice Too {#rdd-the-picture-is-a-choice-too}

Before the estimate, a word about the figure, because in practice the figure is
what readers judge. Every binned scatter plot in this section was built by hand,
and the bin width was a number we made up: two index points above, one and a
half in the polynomial comparison. Widen the bins and any jump looks smoother;
narrow them and noise starts to look like structure. That is exactly the kind of
invisible researcher choice the rest of this section is about.

There is a selector for it. Calonico, Cattaneo and Titiunik (2015) worked out
how many bins to use — either to mimic the underlying variability of the data or
to give an integrated-mean-squared-error-optimal picture — and `rdplot()`
implements it.

```{r rdd-rdplot, fig.align='center', fig.cap="The same discontinuity, binned by rdplot's own selector rather than by hand. The number of bins on each side is chosen to trace the variability of the data instead of being picked by the analyst; the smooth line through them is a global fourth-order polynomial, drawn for the eye.", echo=TRUE}
rdplot(y = sc$below_standard, x = sc$index, c = 60,
       x.label = "social index score",
       y.label = "below minimum standard (%)",
       title = "")
```

One warning about that picture, which we will earn in Section
\@ref(rdd-why-not-just-fit-a-polynomial): the smooth curve `rdplot()` draws
through the bins is a global fourth-order polynomial. It is there to show the
shape of the data, not to produce the estimate, and the jump you think you see
in it is not the number anyone should report.

### Doing It Properly: rdrobust {#rdd-doing-it-properly}

Every choice we made by hand has a literature behind it. `rdrobust`, from
Sebastián Calonico, Matias Cattaneo, Max Farrell and Rocío Titiunik, makes those
choices for you and reports what it chose.

```{r rdd-rdrobust-sharp, echo=TRUE}
itt <- rdrobust(y = sc$below_standard, x = sc$index, c = 60)
summary(itt, all = TRUE)
```

Three lines of output deserve attention.

The **bandwidth**, `h`, was chosen to minimise mean squared error, following
Imbens and Kalyanaraman (2012) as refined by Calonico, Cattaneo and Titiunik
(2014). It came out at 6.6 index points, which leaves 687 schools below the line
and 530 above.

The **three estimates**, printed because we asked for `all = TRUE`; the default
output shows only the last of them. `Conventional` is the local linear fit — the
same estimator we computed by hand, at the bandwidth the package chose. `Bias-Corrected` subtracts an estimate of the
smoothing bias, using a second, wider bandwidth `b` and a quadratic fit.
`Robust` reports the bias-corrected point estimate with a standard error that
accounts for the fact that the bias correction was itself estimated. This last
point is the contribution of Calonico, Cattaneo and Titiunik (2014), and it is
why the robust interval is wider than the conventional one rather than narrower.

::: {.analysis-warning}
The robust confidence interval is centred on the bias-corrected estimate, not on
the conventional one. In awkward samples it can therefore fail to contain the
conventional point estimate. That is not a bug; it means the two are answering
slightly different questions. Report the conventional estimate with the robust
interval, and say that you are doing so. That is the convention the `rdrobust`
output itself invites: the point estimate stays on the MSE-optimal bandwidth,
and the interval carries the bias correction.
:::

### How Wide Should the Window Be? {#rdd-how-wide-should-the-window-be}

The bandwidth is the one choice in a regression discontinuity that a sceptical
reader will always attack, and rightly: it trades bias against variance, and the
optimum depends on curvature you cannot see.

```{r rdd-bandwidth, cache=TRUE, fig.align='center', fig.cap="The intention-to-treat estimate and its robust interval as a function of the bandwidth. The estimate is flat between about six and twenty index points; below six the interval explodes, above twenty it drifts back towards zero as the window swallows schools that resemble nothing near the threshold.", echo=TRUE}
bw_grid <- purrr::map_dfr(seq(2, 30, by = 0.5), function(b) {
  f <- rdrobust(sc$below_standard, sc$index, c = 60, h = b)
  tibble(h = b, est = f$coef[1, 1], lo = f$ci[3, 1], hi = f$ci[3, 2])
})

h_opt <- rdrobust(sc$below_standard, sc$index, c = 60)$bws[1, 1]

ggplot(bw_grid, aes(h, est)) +
  geom_hline(yintercept = -2.064, linetype = "dotted",
             colour = "#1B7837", linewidth = 0.8) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80", alpha = 0.6) +
  geom_line(colour = "#C1440E", linewidth = 0.9) +
  geom_vline(xintercept = h_opt, linetype = "dashed", colour = "grey30") +
  annotate("text", x = h_opt + 0.6, y = 4.5, hjust = 0, size = 3,
           label = paste0("MSE-optimal: h = ", round(h_opt, 1))) +
  annotate("text", x = 29, y = -2.064, vjust = -0.7, hjust = 1, size = 3,
           colour = "#1B7837", label = "truth") +
  coord_cartesian(ylim = c(-6, 6)) +
  labs(x = "bandwidth h (index points)", y = "sharp RD estimate (ITT)") +
  theme_minimal(base_size = 11)
```

`rdrobust` offers several selectors, and they do not agree.

```{r rdd-bwselect, echo=TRUE}
purrr::map_dfr(c("mserd", "msetwo", "cerrd"), function(b) {
  rr <- rdrobust(sc$below_standard, sc$index, c = 60, bwselect = b)
  tibble(selector = b, `h left` = rr$bws[1, 1], `h right` = rr$bws[1, 2],
         estimate = rr$coef[1, 1])
}) %>%
  knitr::kable(digits = 3,
    caption = "Three bandwidth selectors on the same data. `mserd` uses one bandwidth for both sides, `msetwo` allows them to differ, `cerrd` optimises coverage error rather than mean squared error and is therefore deliberately narrower.")
```

The honest way to present this is not to pick one. It is to show the picture
above, state which selector produced the headline number, and let the reader see
that nothing important turns on it.

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
First, without touching R. Three selectors were run on the same data and
disagreed. `cerrd` is deliberately the narrowest of the three because it
optimises `r mcq(c("mean squared error", answer = "coverage error", "the number of observations retained"))`
rather than mean squared error.

The MSE-optimal bandwidth came out at 6.6 index points. From the figure, the
estimate is roughly flat between about six and
`r fitb(20, tol = 2)` index points.

Now re-run the bandwidth grid with `p = 2` inside `rdrobust()`, so that each side
is fitted with a local *quadratic* rather than a local line. Does the MSE-optimal
bandwidth get wider or narrower? Answer before you look:
`r mcq(c(answer = "wider", "narrower", "unchanged"))`

`r hide("The reasoning")`

A quadratic can follow curvature that a line has to treat as bias. The
MSE-optimal bandwidth trades that bias against variance, so once the bias term
shrinks the optimiser is willing to buy more data — the window widens. The price
is that a quadratic fitted at a boundary is estimated less precisely at any given
width, which is why the wider window does not automatically buy a tighter
interval.

```{r rdd-bw-quadratic, eval=FALSE, echo=TRUE}
c(linear    = rdrobust(sc$below_standard, sc$index, c = 60, p = 1)$bws[1, 1],
  quadratic = rdrobust(sc$below_standard, sc$index, c = 60, p = 2)$bws[1, 1])
```

`r unhide()`
::::

### Why Not Just Fit a Polynomial? {#rdd-why-not-just-fit-a-polynomial}

There is an older way to do this, and you will still meet it. Fit one regression
to all the data, with a dummy for being above the cut-off and a polynomial in
the score, and read off the dummy.

```{r rdd-polynomials, echo=TRUE}
sc$z <- sc$index - 60

purrr::map_dfr(1:8, function(p) {
  m <- lm(below_standard ~ invited * poly(z, p, raw = TRUE), data = sc)
  tibble(Order = p,
         Estimate = coef(m)["invited"],
         SE = summary(m)$coefficients["invited", 2])
}) %>%
  knitr::kable(digits = 3,
    caption = "The same data, eight polynomial orders. The estimate wanders from +1.17 to -2.63 and back to -0.99, and the standard error triples. The truth is -2.06.")
```

Order one gets the sign wrong. Orders two to four land within six tenths of a
point of the truth. Orders five upwards drift back towards zero while the standard error
triples.
Nothing in the output tells you which row to believe.

```{r rdd-polynomial-figure, fig.align='center', fig.cap="The same fitted values around the threshold under three global polynomials and under the local linear fit. Watch the gap at 60: the straight line puts a small step upwards where the data have a step downwards, because it is spending its slope on the curvature of the whole range.", echo=TRUE}
curves <- purrr::map_dfr(c(1, 2, 6), function(p) {
  m <- lm(below_standard ~ invited * poly(z, p, raw = TRUE), data = sc)
  tibble(z = c(seq(-58, -0.01, length.out = 300), seq(0, 38, length.out = 300)),
         invited = rep(0:1, each = 300)) %>%
    mutate(fit = predict(m, newdata = .), index = z + 60,
           order = paste("global polynomial, order", p), side = invited)
}) %>%
  bind_rows(local_fit(sc, "below_standard") %>%
              transmute(index, fit, order = "local linear, h = 7", side = above))

curves$order <- factor(curves$order,
  levels = c("global polynomial, order 1", "global polynomial, order 2",
             "global polynomial, order 6", "local linear, h = 7"))

ggplot(bin_means(sc, "below_standard", width = 1.5, lo = 42, hi = 78), aes(centre, y)) +
  geom_vline(xintercept = 60, linetype = "dashed", colour = "grey50") +
  geom_point(colour = "grey65", size = 1.2) +
  geom_line(data = curves, aes(index, fit, group = side),
            colour = "#C1440E", linewidth = 0.9) +
  facet_wrap(~order, nrow = 1) +
  coord_cartesian(xlim = c(45, 75), ylim = c(20, 45)) +
  labs(x = "social index score", y = "below minimum standard (%)") +
  theme_minimal(base_size = 10)
```

One draw proves nothing, so run three hundred. Each draw is a fresh cohort of
4,200 schools from the same world; each estimator gets the same data; and we
record whether its 95 per cent interval contains the true intention-to-treat
effect of $-2.06$.

```{r rdd-polynomial-study, cache=TRUE, echo=TRUE}
itt_true <- mean(complier_weight) * (-3.2 - 1.3 * sum(cap_draw * complier_weight) / sum(complier_weight))

study <- purrr::map_dfr(1:300, function(i) {
  d <- simulate_startchancen(seed = 1000 + i)
  d$z <- d$index - 60

  global <- purrr::map_dfr(c(1, 2, 4, 6), function(p) {
    m <- lm(below_standard ~ invited * poly(z, p, raw = TRUE), data = d)
    e <- coef(m)["invited"]
    s <- summary(m)$coefficients["invited", 2]
    tibble(estimator = paste("global, order", p), est = e,
           covers = abs(e - itt_true) <= 1.96 * s)
  })

  rr <- rdrobust(d$below_standard, d$index, c = 60)
  bind_rows(global, tibble(
    estimator = "local linear (rdrobust)", est = rr$coef[1, 1],
    covers = itt_true >= rr$ci[3, 1] & itt_true <= rr$ci[3, 2]))
})

study %>%
  group_by(estimator) %>%
  summarise(mean = mean(est),
            bias = mean(est) - itt_true,
            sd   = sd(est),
            rmse = sqrt(mean((est - itt_true)^2)),
            coverage = mean(covers), .groups = "drop") %>%
  knitr::kable(digits = 3,
    caption = "Three hundred simulated studies. The global linear specification never once covers the truth; the local linear estimator covers it 95 times in a hundred, which is what a 95 per cent interval is supposed to do.")
```

The global linear model has the smallest standard deviation of any estimator in
the table and a coverage rate of exactly zero. It is precise, stable, replicable
and wrong — a combination that no diagnostic in a regression table will ever
flag. This is the third time the chapter has met that combination and it will not
be the last: Section \@ref(exp-lalonde) found it with 16,177 observations and a
t-statistic of $-12$, and Section \@ref(iv-opening-the-envelope) will find it
again with a coverage rate of zero out of three hundred. **Precision is not
credibility** is the one sentence this chapter would keep if it could keep only
one.

This is the argument Andrew Gelman and Guido Imbens made in *Why High-Order
Polynomials Should Not Be Used in Regression Discontinuity Designs* (*JBES*,
2019): a global polynomial gives large and arbitrary weights to observations far
from the cut-off, and the resulting estimate is noisy in a way that its own
standard error understates.[^rdd-albada]

::: {.analysis-takeaway}
Fit locally, and let the bandwidth do the work of the polynomial order. If you
must show a global fit, show it as a picture, not as a coefficient.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Read the coverage column of the simulation table once more.

The global linear specification covered the truth in `r fitb(0, num = TRUE)` of
three hundred studies.

It also had the *smallest* standard deviation of any estimator in the table.
Which one-line description fits it?
`r mcq(c("imprecise and biased", answer = "precise, stable, replicable and wrong", "unbiased but too noisy to be useful"))`

In the polynomial table, orders two to four all landed within half a point of the
truth. Does that mean a quadratic is a safe choice?
`r mcq(c("yes, orders two to four are the recommended range", answer = "no — nothing in the output identifies which row to believe, and the safe range is a property of this data set", "yes, provided the sample is large"))`

True or false: a global polynomial gives every observation the same weight in
determining the jump at the cut-off. `r torf(FALSE)`
::::


### The First Stage and the Fuzzy Estimate {#rdd-the-first-stage-and-the-fuzzy-estimate}

Everything so far has estimated the effect of the *invitation*. To get the
effect of *participation* we need the denominator: how much the invitation
actually changed participation.

```{r rdd-first-stage, echo=TRUE}
first_stage <- rdrobust(y = sc$participates, x = sc$index, c = 60)

c(jump = first_stage$coef[1, 1],
  se   = first_stage$se[3, 1],
  lo   = first_stage$ci[3, 1],
  hi   = first_stage$ci[3, 2])
```

Crossing the threshold raises the participation probability by 0.55. It is a
strong first stage — nothing like the weak instruments of Section
\@ref(instruments) — but it is not a jump all the way to one, and that gap is
the whole reason the design is fuzzy.[^rdd-two-bandwidths]

```{r rdd-fuzzy, echo=TRUE}
fuzzy <- rdrobust(y = sc$below_standard, x = sc$index, c = 60, fuzzy = sc$participates)
summary(fuzzy, all = TRUE)
```

And by hand, to show that there is no magic in it: the fuzzy estimate is one
local linear regression divided by another.

```{r rdd-fuzzy-by-hand, echo=TRUE}
hh <- fuzzy$bws[1, 1]

f2 <- sc %>%
  filter(abs(index - 60) <= hh) %>%
  mutate(z = index - 60, weight = 1 - abs(z) / hh)

numerator   <- coef(lm(below_standard   ~ z * invited, data = f2, weights = weight))["invited"]
denominator <- coef(lm(participates ~ z * invited, data = f2, weights = weight))["invited"]

c(numerator = numerator, denominator = denominator,
  wald = numerator / denominator)

# The same number as two-stage least squares, with eligibility as the instrument.
iv <- iv_robust(below_standard ~ participates + z + z:invited |
                  invited + z + z:invited,
                data = f2, weights = weight)

c(coef(iv)["participates"], se = iv$std.error[["participates"]])
```

Now open the envelope — the same move Sections \@ref(mat-truth),
\@ref(did-truth) and \@ref(iv-opening-the-envelope) make, and the only reason
any of these four sections can say whether an estimator worked.

```{r rdd-envelope, echo=TRUE}
tibble(
  Quantity = c("effect at the cut-off, compliers (the estimand)",
               "fuzzy RD, conventional",
               "fuzzy RD, bias-corrected",
               "Wald ratio by hand",
               "two-stage least squares",
               "OLS with index and covariates"),
  Value = c(-3.2 - 1.3 * sum(cap_draw * complier_weight) / sum(complier_weight),
            fuzzy$coef[1, 1],
            fuzzy$coef[2, 1],
            numerator / denominator,
            coef(iv)["participates"],
            coef(lm(below_standard ~ participates + index + benefits + pre_score, data = sc))["participates"])) %>%
  knitr::kable(digits = 3,
    caption = "The fuzzy estimate against the truth. Three different routes to the same number, and one confident alternative that is off by four tenths of a point in a direction no output could have revealed.")
```

::: {.analysis-note}
The fuzzy estimator inherits every warning from the instruments section. A weak
first stage makes the ratio explode; the monotonicity assumption is untestable;
and the estimand is defined by who complies, so two studies of the same
programme with different take-up patterns are estimating different things even
if both are correct.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The intention-to-treat estimate was $-2.06$ and the jump in participation
$0.553$. Divide one by the other:
`r fitb(-3.73, tol = 0.15)`

The fuzzy call reports $-3.44$ instead. What accounts for the gap?
`r mcq(c("rounding", answer = "the fuzzy call picks the bandwidth that is optimal for the ratio, not for either piece separately", "the by-hand calculation forgot the triangular weights"))`

Now suppose the invitation had raised participation by only five percentage
points instead of fifty-five, with the same numerator. The estimate would be
`r mcq(c("about a tenth as large", answer = "about eleven times as large", "unchanged"))`,
and its sensitivity to any violation of continuity would be
`r mcq(c("unchanged", "smaller", answer = "eleven times larger"))`.

That last pair is the whole of Section \@ref(iv-when-the-instrument-is-weak),
arriving early.
::::


### Can the Score Be Gamed? {#rdd-can-the-score-be-gamed}

Continuity is an assumption about potential outcomes, and we cannot see those.
But there is one thing we can see, and it is the single most informative check
in the whole design.

If people can move themselves across the threshold — a headteacher who knows the
formula and gets a pupil reclassified, a doctor who records 1,499 grams rather
than 1,501, an official who rounds a poverty rate the helpful way — then the
units just above the line will not be like the units just below. And there will
be a tell: too many of them.

```{r rdd-manipulated, echo=TRUE}
scm <- simulate_startchancen(world = "manipulated")

# In this world, schools just under the line with capable leadership
# get themselves moved just over it.
c(below_60 = sum(sc$index >= 57 & sc$index < 60),
  above_60 = sum(sc$index >= 60 & sc$index < 63),
  below_60_manipulated = sum(scm$index >= 57 & scm$index < 60),
  above_60_manipulated = sum(scm$index >= 60 & scm$index < 63))
```

```{r rdd-density-figure, fig.align='center', fig.cap="The distribution of the social index around the threshold in the two worlds. On the left the histogram runs smoothly across 60; on the right there is a hole immediately below the line and a spike immediately above it, which is what manipulation looks like.", echo=TRUE}
bind_rows(sc  %>% transmute(index, world_label = "as recorded"),
          scm %>% transmute(index, world_label = "with renegotiation")) %>%
  mutate(world_label = factor(world_label, levels = c("as recorded", "with renegotiation"))) %>%
  filter(index >= 45, index <= 75) %>%
  ggplot(aes(index)) +
  geom_histogram(binwidth = 1, fill = "grey60", colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 60, linetype = "dashed", colour = "#C1440E",
             linewidth = 0.8) +
  facet_wrap(~world_label) +
  labs(x = "social index score", y = "number of schools") +
  theme_minimal(base_size = 11)
```

The formal version is a test of whether the density of the running variable is
continuous at the cut-off. Justin McCrary proposed it in 2008; the current
implementation, `rddensity`, uses the local polynomial density estimator of
Cattaneo, Jansson and Ma (2020) and needs no pre-binning.

```{r rdd-density-test, echo=TRUE}
rddensity(X = sc$index,  c = 60)$test$p_jk
rddensity(X = scm$index, c = 60)$test$p_jk
```

A p-value of 0.49 in the clean world, and $2.8 \times 10^{-8}$ in the
manipulated one. Sixty-one schools moved, out of 4,200, and the test found what
the histogram only hints at.

The second check is continuity of things the programme cannot have caused:
covariates fixed before assignment, and outcomes measured before the programme
started. If those jump at the cut-off, the design is telling you that the two
sides of the line differ for reasons other than treatment.

```{r rdd-balance, echo=TRUE}
check_jumps <- function(dat, world_label) {
  purrr::map_dfr(c("benefits", "pre_score", "capacity"), function(v) {
    r <- rdrobust(dat[[v]], dat$index, c = 60)
    tibble(World = world_label, Variable = v, Jump = r$coef[1, 1],
           `CI low` = r$ci[3, 1], `CI high` = r$ci[3, 2], p = r$pv[3, 1])
  })
}

bind_rows(check_jumps(sc, "as recorded"), check_jumps(scm, "with renegotiation")) %>%
  knitr::kable(digits = 3,
    caption = "Continuity checks in both worlds. In the clean world nothing jumps. In the manipulated world the pre-programme outcome and the unobserved capacity both do, and the capacity jump is the one no real analyst would ever see.")
```

Notice which line matters most. In the manipulated world, the share of pupils in
benefit households — the covariate an analyst would naturally check — passes the
test. The pre-programme outcome fails it. And the variable that actually drives
the manipulation is the one we included in this table only because we built the
world and can therefore look at it.

::: {.analysis-warning}
Balance tests can only fail informatively. Passing them is not evidence that the
design is sound; it is evidence that the particular sorting you tested for is
not the sorting that happened.
:::

The standard repair is the **donut**: drop observations in a small band on both
sides of the cut-off, on the theory that manipulation is local, and re-estimate
from what is left. It was popularised by Barreca, Guldi, Lindo and Waddell
(2011), whose comment on Almond and co-authors showed that recorded birth
weights heap at round numbers throughout — and that the babies recorded at
exactly 1,500 grams have markedly higher mortality than those on either side,
unlike the heaps at 1,400 and 1,600.

```{r rdd-donut, cache=TRUE, echo=TRUE}
purrr::map_dfr(0:2, function(d) {
  dat <- scm %>% filter(abs(index - 60) >= d)
  rr <- rdrobust(dat$below_standard, dat$index, c = 60, fuzzy = dat$participates)
  pv <- rdrobust(dat$pre_score,  dat$index, c = 60)
  tibble(Donut = d, `fuzzy estimate` = rr$coef[1, 1],
         `CI low` = rr$ci[3, 1], `CI high` = rr$ci[3, 2],
         `placebo p` = pv$pv[3, 1],
         `density p` = rddensity(X = dat$index, c = 60)$test$p_jk)
}) %>%
  knitr::kable(digits = 3,
    caption = "Widening the donut in the manipulated world. At a hole of two index points the density test and the placebo outcome both stop rejecting, and the estimate moves from -5.97 to -2.82 — past the true -3.49, and into an interval nearly four times as wide that now includes zero.")
```

The manipulation is gone. So is the precision, and the point estimate has
overshot. Nothing here says the repaired number is the right one; it says the
unrepaired number was the wrong one.

::: {.analysis-takeaway}
The donut is a diagnostic dressed as an estimator. If the answer moves a lot as
the hole widens, the design has a problem; if it does not, you have learned
something. Either way, report the whole sequence, not the row you liked.

Every design in this chapter has one of these — a way of turning an untestable
assumption into a quantity a reader can dispute. Matching has the E-value and
Rosenbaum bounds (Section \@ref(mat-truth)); difference-in-differences has the
Rambachan-Roth bound on how large a violation could be (Section
\@ref(did-pretesting)); instrumental variables has the identification-robust
interval (Section \@ref(iv-inference-that-survives)). None of them is a test.
All of them replace a yes-or-no claim with a magnitude, which is the only honest
move available.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Sixty-one schools out of 4,200 moved themselves across the line — that is
`r fitb(1.5, tol = 0.3)` per cent of them — and the density test caught it at
$p = 2.8 \times 10^{-8}$.

A density test that **rejects** tells you
`r mcq(c("that the estimate is biased upwards", answer = "that units could move across the line, so the two sides are no longer comparable", "that the bandwidth is too wide"))`

A density test that **does not reject** tells you
`r mcq(c("that the design is sound", answer = "that the particular sorting you tested for is not the sorting that happened", "nothing whatever"))`

In the manipulated world, which check failed and which passed? Look at the
continuity table again. The covariate an analyst would naturally reach for — the
share of pupils in benefit households —
`r mcq(c("failed", answer = "passed", "was not testable"))`,
while the pre-programme outcome
`r mcq(c(answer = "failed", "passed", "was not testable"))`.

True or false: widening the donut until the tests stop rejecting produces the
right answer. `r torf(FALSE)`
::::


### Placebo Cut-offs {#rdd-placebo-cut-offs}

The last check is the cheapest. Pretend the threshold was somewhere else, and
estimate again. Fit the placebo cut-offs using only data from one side of the
real threshold, so that no genuine treatment effect can contaminate them.

Be clear about what this tests. A density test looks for manipulation; a placebo
cut-off looks for **functional form**. If a local linear fit produces jumps at 45
and at 70 as readily as at 60, then what you are reading at 60 is curvature that
the straight lines could not absorb, and the design has told you nothing. A
clean row of nulls does not make the real estimate causal — it only removes one
way of being fooled.

```{r rdd-placebo, cache=TRUE, echo=TRUE}
purrr::map_dfr(c(40, 45, 50, 55, 60, 65, 70, 75), function(cc) {
  dat <- if (cc < 60) filter(sc, index < 60)
         else if (cc > 60) filter(sc, index >= 60)
         else sc
  rr <- rdrobust(dat$below_standard, dat$index, c = cc)
  tibble(`pretend cut-off` = cc, estimate = rr$coef[1, 1],
         `CI low` = rr$ci[3, 1], `CI high` = rr$ci[3, 2], p = rr$pv[3, 1])
}) %>%
  knitr::kable(digits = 3,
    caption = "Seven false thresholds and the real one. Only 60 produces a discontinuity, which is what a genuine design looks like and what a mis-specified functional form does not.")
```

### When the Score Is Published in Nine Steps {#rdd-when-the-score-is-published-in-nine-steps}

Now the bad news, and it is the reason the section opened with a real programme
rather than an invented one.

The North Rhine-Westphalian social index exists as a continuous factor score
inside the ministry. What is published is the *step*: an integer from one to
nine. Our simulated world has the same feature, so we can find out what the
design is worth when the running variable arrives with nine values instead of
four thousand.

```{r rdd-discrete, echo=TRUE}
sc %>%
  group_by(step) %>%
  summarise(schools = n(),
            `below standard` = mean(below_standard),
            `participation` = mean(participates), .groups = "drop") %>%
  knitr::kable(digits = 3,
    caption = "The same schools, grouped into the nine published steps. Participation jumps between step 5 and step 6 exactly as it should — and the outcome climbs monotonically across every step, which is the problem.")
```

(Our nine steps are equal-width bins rather than the ministry's quantiles, which
makes no difference to the point.) There is no window left to narrow. The nearest thing to a local comparison is
step 5 against step 6, and those two groups differ by ten index points on
average, not by a fraction of one. So the estimate becomes entirely a matter of
what functional form you assume for the trend across steps.

```{r rdd-discrete-estimates, echo=TRUE}
d <- sc %>% mutate(above = as.integer(step >= 6), s = step - 6)

purrr::map_dfr(0:3, function(p) {
  trend_formula <- switch(as.character(p),
    "0" = below_standard ~ above,
    "1" = below_standard ~ above + s,
    "2" = below_standard ~ above + s + I(s^2),
    "3" = below_standard ~ above + s + I(s^2) + I(s^3))
  my <- lm(trend_formula, data = d)
  mt <- lm(update(trend_formula, participates ~ .), data = d)
  tibble(`trend order` = p,
         ITT = coef(my)["above"],
         `first stage` = coef(mt)["above"],
         Wald = coef(my)["above"] / coef(mt)["above"])
}) %>%
  knitr::kable(digits = 3,
    caption = "The nine-step index, four assumptions about the trend, four answers. The truth is -2.06 for the intention to treat and -3.49 for the Wald ratio; the four rows span +27.0 to -2.8.")
```

Four defensible specifications, four answers ranging from $+27$ to $-2.8$. The
first stage is essentially unchanged from row to row — it really does jump by
about 0.59 — but the numerator is a guess. That asymmetry is a property of this
programme rather than a general law: take-up here is flat within each side of the
line, so any trend specification recovers it. Where take-up trends across the
steps, the denominator is guessed too.

There are two honest responses. One is the inference procedure of Kolesár and
Rothe (2018), which builds confidence intervals that are valid under an explicit
bound on how much the trend can curve, rather than pretending the bound is
zero.[^rdd-kolesar] The other is to change
framework entirely.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: when there is no window left to shrink into</h2>
:::
Everything in this section rests on the *continuity* framework: smooth
conditional expectations, a limit at the cut-off, a bandwidth that shrinks. There
is a second framework, developed by Cattaneo, Frandsen and Titiunik (2015), that
assumes something stronger and more concrete — that within some window around
the cut-off, treatment is *as if randomly assigned*, so that the analysis can be
a randomisation test of the kind Fisher ran at Rothamsted (Section
\@ref(experiments)).

The `rdlocrand` package selects that window by widening it until a covariate
balance test fails. On our data it selects a window of a tenth of an index
point.

```{r rdd-locrand, echo=TRUE}
window_lr <- rdwinselect(R = sc$index - 60,
                          X = as.matrix(sc[, c("benefits", "pre_score")]),
                          seed = 11, reps = 1000, quietly = TRUE)

c(left = window_lr$w_left, right = window_lr$w_right)
```

That answer is not a failure of the package; it is the package being right. Our
outcome climbs by nearly a point per index point, so a window of even one point
already contains a real trend, and "as if randomised" is simply false there. The
local randomisation framework earns its keep where the score has few distinct
values and the continuity framework has no window to shrink into — which is
precisely the nine-step case above. The two frameworks are not rivals; they
answer different failures of the same design.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Four defensible specifications on the nine-step index gave answers from $+27.0$
to $-2.8$, while the first stage barely moved. Which half of the ratio was the
guess? `r mcq(c("the denominator, because take-up is hard to measure", answer = "the numerator, because the outcome trends across the steps and any trend specification is an assumption", "both equally"))`

Would that asymmetry hold in general?
`r mcq(c("yes, first stages are always robust", answer = "no — it holds here because take-up is flat within each side of the line; where take-up trends across the steps, the denominator is guessed too", "no, it is an artefact of the simulation seed"))`

Now to R. Run `simulate_startchancen(strength = 8)` and estimate the effect
twice: once with the covariate-adjusted regression and once with
`rdrobust(..., fuzzy = ...)`. Which of the two gives up first, and in which
direction? Then move the cut-off in the simulation from 60 to 50 and estimate
again. The effect parameters have not changed at all — so why has the answer?

`r hide("What you should see")`

The regression marches: `strength = 8` makes the unobserved capacity dominate
the outcome, and the coefficient on participation drifts well past $-5$ while its
standard error barely notices. The discontinuity estimate stays near the truth
and loses precision, which is the honest response.

Moving the cut-off to 50 changes the *estimand*, not the estimator. The true
effect is $\tau = -3.2 - 0.11\,(\text{index} - 60) - 1.3\,\text{capacity}$, so
schools sitting at 50 have an effect about 1.1 points *smaller* in absolute value
than schools sitting at 60 — and the complier mix at the new line is different
too. A local estimate belongs to the line it was estimated at, and nowhere else.

`r unhide()`
::::

::: {.analysis-takeaway}
A published threshold is not enough. You need the score the threshold was
applied to. If a ministry publishes the step and keeps the number, the
discontinuity design goes with the number.
:::

### The Design in the Wild {#rdd-the-design-in-the-wild}

Enough of a world we invented. The canonical teaching data set for this design
is real, small enough to ship, and comes from Cattaneo, Frandsen and Titiunik
(2015): every US Senate race from 1914 to 2010, with the Democratic margin of
victory in *that* race as the running variable and the Democratic vote share in
the next election for the same seat, six years later, as the outcome.

```{r rdd-senate, echo=TRUE}
senate <- readr::read_csv(file.path("data", "Causal", "rdrobust_senate.csv"),
                         show_col_types = FALSE)

nrow(senate)
range(senate$year)
```

The cut-off is zero: a margin of $+0.1$ means the Democrats took the seat by a
whisker, a margin of $-0.1$ means they lost it by the same whisker. Nobody can
decide which side of zero they land on, and that is the point Lee (2008) made in
the paper that turned close elections into the standard illustration of the
design.

```{r rdd-senate-fit, echo=TRUE}
senate_rd <- rdrobust(y = senate$vote, x = senate$margin, c = 0)
summary(senate_rd)
```

Winning a seat is worth about 7.4 percentage points to the party at the next
election for that seat, with a robust interval from 4.1 to 10.9. Lee's House
data gave 7.7 to 8.1.

```{r rdd-senate-figure, fig.align='center', fig.cap="US Senate elections, 1914 to 2010: binned means of the Democratic vote share six years later against the party's margin in the current race for the same seat, with local linear fits. The step at zero is the incumbency advantage; point size is the number of races in each bin.", echo=TRUE}
senate_bins <- senate %>%
  filter(!is.na(margin), !is.na(vote)) %>%
  mutate(centre = floor(margin / 4) * 4 + 2) %>%
  group_by(centre) %>%
  summarise(y = mean(vote), n = n(), .groups = "drop") %>%
  filter(abs(centre) <= 60)

senate_fit <- purrr::map_dfr(c(-1, 1), function(s) {
  dd <- senate %>%
    filter(sign(margin) == s, abs(margin) <= 17.75) %>%
    mutate(g = 1 - abs(margin) / 17.75)
  m <- lm(vote ~ margin, data = dd, weights = dd$g)
  tibble(margin = seq(min(dd$margin), max(dd$margin), length.out = 50)) %>%
    mutate(fit = predict(m, newdata = .), side = s)
})

ggplot(senate_bins, aes(centre, y)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(aes(size = n), colour = "grey45", alpha = 0.8) +
  geom_line(data = senate_fit, aes(margin, fit, group = side),
            colour = "#C1440E", linewidth = 1) +
  scale_size_area(max_size = 3.5, guide = "none") +
  labs(x = "Democratic margin in this election (percentage points)",
       y = "Democratic vote share six years later (%)") +
  theme_minimal(base_size = 11)
```

Run the same checks we ran on the simulation.

```{r rdd-senate-checks, echo=TRUE}
# Is the density of the margin continuous at zero?
rddensity(X = senate$margin[!is.na(senate$margin)], c = 0)$test$p_jk

# Do variables that cannot have been caused by the win jump at zero?
purrr::map_dfr(c("termshouse", "termssenate", "population", "class"), function(v) {
  rr <- suppressWarnings(rdrobust(senate[[v]], senate$margin, c = 0))
  tibble(Variable = v, Jump = rr$coef[1, 1], p = rr$pv[3, 1])
}) %>%
  knitr::kable(digits = 3,
    caption = "Placebo outcomes in the Senate data. None of the four jumps at zero, which is the pattern a credible design produces.")
```

```{r rdd-senate-robustness, echo=TRUE}
purrr::map_dfr(c(5, 10, 20, 50), function(b) {
  rr <- rdrobust(senate$vote, senate$margin, c = 0, h = b)
  tibble(h = b, estimate = rr$coef[1, 1],
         `CI low` = rr$ci[3, 1], `CI high` = rr$ci[3, 2], n = sum(rr$N_h))
}) %>%
  knitr::kable(digits = 3,
    caption = "The Senate estimate at four fixed bandwidths. Above about ten percentage points of margin the answer settles between six and eight; at five it doubles, on a fifth of the data.")

# The global alternative, for comparison.
global4 <- lm(vote ~ I(margin >= 0) * poly(margin, 4, raw = TRUE), data = senate)
coef(summary(global4))[2, 1:2]
```

The density test passes, all four placebo outcomes pass, and the estimate is
stable for bandwidths above about ten. Fit a global fourth-order polynomial
instead and you get 9.4 — two points higher, from the same data, for no reason a
reader could check.

:::: {.amazing}
::: {.titelamazing}
<h2>The design that let itself be audited</h2>
:::
The close-election design has itself been the subject of one of the most useful
methodological arguments in the field. If elections that are nearly tied are
essentially coin flips, each side should win half of them. Caughey and Sekhon
(2011) showed that in US House races between 1942 and 2008 the bare winners and
the bare losers differ markedly on things measured *before* the vote — money,
experience, forecasts — and that the imbalance gets worse in the closest races.

Two replies followed. Eggers and co-authors (2015) looked at more than forty
thousand close races across other US offices and nine other countries and found
the imbalance essentially nowhere else, concluding that the postwar House is the
exception rather than the rule. De la Cuesta and Imai (2016) argued that much of
the dispute rested on a confusion: the design requires continuity, not as-if
randomisation inside a window, and once the multiple-testing problem is handled
the evidence for sorting is considerably weaker than it looked.

What matters here is not who was right. It is that the argument was *possible*.
The design publishes a running variable, and a published running variable can be
audited by anyone with the data. No matching study has ever been contested at
that level of detail, because there is nothing comparable to point at.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The Senate estimate is 7.4 percentage points with a robust interval from 4.1 to
10.9. Lee's House estimate was 7.7 to 8.1. A global fourth-order polynomial on
the same Senate data returns `r fitb(9.4, tol = 0.2)`.

Three things were checked on these data. Which of them, if it had failed, would
have been the most damaging?
`r mcq(c("the estimate moving between bandwidths of 20 and 50", answer = "the density of the margin jumping at zero", "the global polynomial disagreeing with the local fit"))`

True or false: the four placebo outcomes passing their tests establishes that the
close-election design is valid. `r torf(FALSE)`

And a question about the argument in the box above. Caughey and Sekhon found
imbalance; Eggers and co-authors did not; De la Cuesta and Imai said the dispute
partly rested on a confusion. What was the confusion?
`r mcq(c("that a placebo test is a test of continuity", answer = "that the design requires continuity at the cut-off, not as-if randomisation inside a window", "that the running variable was measured with error"))`
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn: one hour with the Senate file</h2>
:::
This one is the size of a seminar exercise rather than a quiz. Everything you
need is in `data/Causal/rdrobust_senate.csv` and in the four packages already
loaded at the top of this section.

**Step 1 — split the century.** The file runs from 1914 to 2010. Estimate the
incumbency advantage separately for races up to and including 1960 and for races
after 1960. Report the conventional point estimate and the robust interval for
each half.

**Step 2 — read what you got.** One of the two halves has an interval that
includes zero. Which?
`r mcq(c(answer = "1914 to 1960", "1962 to 2010", "both", "neither"))`

Round the two point estimates to one decimal place. The early half gives
`r fitb(4.3, tol = 0.35)` and the late half `r fitb(9.9, tol = 0.35)`.

**Step 3 — resist the obvious reading.** The two halves differ by more than five
points. Before concluding that the incumbency advantage grew, ask what else
differs between them.
`r mcq(c("nothing — the design controls for everything", answer = "the early half has fewer races, so its interval is wider, and the two intervals overlap", "the running variable is defined differently before 1960"))`

**Step 4 — a donut on real data.** Re-estimate the full sample dropping races
decided by less than half a point, then less than one point, then less than two.
Does the estimate move? Report the three point estimates and say what the
sequence tells you.

**Step 5 — write two sentences.** Not code: prose. What would you report as the
headline number, and what would you have to say about it in the same breath?

`r hide("The code, and what it gives")`

```{r rdd-senate-exercise, eval=FALSE, echo=TRUE}
# Step 1 --------------------------------------------------------------------
era <- function(rows, label) {
  r <- rdrobust(rows$vote, rows$margin, c = 0)
  tibble(era = label, n = nrow(rows), estimate = r$coef[1, 1],
         `CI low` = r$ci[3, 1], `CI high` = r$ci[3, 2],
         h = r$bws[1, 1], `n in window` = sum(r$N_h))
}

bind_rows(
  era(filter(senate, year <= 1960), "1914-1960"),
  era(filter(senate, year >  1960), "1962-2010"))

# Step 4 --------------------------------------------------------------------
purrr::map_dfr(c(0, 0.5, 1, 2), function(d) {
  keep <- senate %>% filter(abs(margin) >= d, !is.na(margin), !is.na(vote))
  r <- rdrobust(keep$vote, keep$margin, c = 0)
  tibble(donut = d, estimate = r$coef[1, 1],
         `CI low` = r$ci[3, 1], `CI high` = r$ci[3, 2])
})
```

**Step 1.** 1914–1960: 611 races, estimate **4.35**, robust interval
$[-0.20,\ 8.93]$, bandwidth 19.1. 1962–2010: 779 races, estimate **9.94**,
robust interval $[5.44,\ 14.89]$, bandwidth 17.9.

**Step 3.** The intervals overlap between 5.4 and 8.9, so the data are perfectly
consistent with one constant advantage of about seven points across the whole
century. The story "the incumbency advantage grew after the war" is a real one in
political science, and these two numbers are consistent with it — but on their
own they do not establish it. Splitting a sample in two and finding that one half
is significant and the other is not is the most common way of manufacturing a
difference that is not there. The test you want compares the two estimates, not
their p-values.

**Step 4.** 7.41, then 7.09, 6.58, 4.70. That is a drift of nearly three points
by the time two-point margins are gone, and each step also drops observations, so
the intervals widen. Two readings are open: either something is going on very
close to zero, or the estimator is simply losing the observations that identify
it. The density test does not reject ($p = 0.38$), which argues for the second
reading — but the sequence is exactly what you should print, rather than the row
you like.

**Step 5.** Something in the spirit of: *winning a Senate seat by a hair is worth
about seven percentage points to the party at the next election for that seat,
with a robust interval of roughly four to eleven; the estimate is stable across
bandwidths above ten points of margin, and it drifts downwards under a widening
donut, which is worth reporting.*

`r unhide()`
::::

### What Regression Discontinuity Cannot Do {#rdd-what-regression-discontinuity-cannot-do}

Four limits, in increasing order of how often they are ignored.

#### The estimate is local, and "local" means local

Our fuzzy estimate was $-3.44$ against a truth of $-3.49$ — a good answer to the
question it asks. But look back at the table of true values. The effect on all
participating schools is $-4.08$, and the effect if every school in the state
were funded would be $-2.07$. The state-wide figure is barely three fifths of
what we estimated, and the effect on the schools that actually took part is half
a point larger again.

That is not an error. It is the estimand. Schools at the threshold are, by
construction, the least disadvantaged of the funded schools, and in this world
the effect grows with disadvantage. A minister who reads $-3.4$ as "what we
would get by doubling the programme" is reading a number that was never about
that.

#### The threshold may switch on more than one thing

In North Rhine-Westphalia the same social index that decides Startchancen
eligibility also governs the distribution of some seven thousand teaching posts:
roughly 1,645 for language support, 1,250 for social pedagogues in the
early-years phase, 818 for school social work, and over a thousand allocated
school by school to steps 6 to 9. A school that
crosses from step 5 to step 6 does not only receive an invitation to one
programme; it moves into a different administrative category.

Everything estimated at that threshold is then the effect of the *bundle*.
Nothing in the output will say so, and no balance test will catch it, because
the second treatment is as discontinuous as the first. This is the reason to
read the administrative rules before the data, and it is the most common way a
discontinuity study goes quietly wrong.

#### It is expensive

Recall Goldberger's factor of 2.75. Our simulation had 4,200 schools, of which
886 lie within five index points of the cut-off, and it still produced a robust
confidence interval on the fuzzy estimate running from $-5.0$ to $-1.7$. Take a
quarter of the state and that interval covers zero.

The official Startchancen evaluation plans repeated competence measurements at
300 randomly selected programme schools *nationally* — a baseline in early 2026
and a repeat in early 2029, in years 4 and 9 — spread across sixteen states with
sixteen different selection rules.[^rdd-evaluation] Whatever else that design can
support, it cannot support a discontinuity estimate at any one state's
threshold.

There is one legitimate way to buy some of that precision back. Calonico,
Cattaneo, Farrell and Titiunik (2019) show that pre-treatment covariates may be
added to the local polynomial — `rdrobust(..., covs = ...)` — without disturbing
identification, provided they enter additively and are themselves continuous at
the cut-off. Note what has changed since the naive comparison at the start of
this section: there, the covariates were carrying the identification and
failing; here they are carrying nothing but variance reduction.

#### The rule has to be a rule

The design needs assignment to be a function of a score, and many programmes are
not like that. Where a committee has discretion, where quotas by region or school
type cut across the ranking, where a waiting list reshuffles the order — the
threshold stops being a threshold. Of the sixteen German states, some rank
schools on one index and cut. Saxony layers a second, purely pedagogical index
on top of the social one, so that two schools with identical social conditions
can end up on opposite sides of the line for reasons that have nothing to do
with the score you can see.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: kinks, borders and several cut-offs at once</h2>
:::
Three extensions, each a full literature.

**Regression kink.** Sometimes a rule changes a *slope* rather than a level —
unemployment benefit that replaces 60 per cent of earnings up to a ceiling and
nothing above it. The identifying object is then a kink in the outcome, not a
jump, and the estimator is a ratio of changes in derivatives. It is strictly
harder: derivatives are estimated less precisely than levels, so kink designs
need more data than jump designs, which already need 2.75 times an experiment.

**Geographic discontinuities.** When the running variable is a location and the
cut-off is a border, the score is two-dimensional. Black (1999) compared houses
on either side of school-attendance boundaries; Dell (2010) compared Peruvian
districts on either side of a colonial mining boundary drawn in 1573. The
difficulty is that a border separates units in every dimension at once, so
"everything else is continuous" is a much stronger claim than it is for a test
score.

**Multiple cut-offs.** Angrist and Lavy's Maimonides rule bites at 40, 80 and
120 pupils; the German states each set their own threshold on their own index.
Pooling several cut-offs buys precision, but the pooled estimate is a weighted
average of several local effects at several different points of the score
distribution, and the weights are rarely the ones you would have chosen.
::::

::: {.analysis-takeaway}
A threshold is the cheapest experiment a society ever runs, and it runs one every
time it writes a rule with a number in it. The price is that the answer belongs
to the people standing on the line, and to nobody else.

That is why this section sits where it does. Section
\@ref(difference-in-differences) buys a wider population with an assumption about
time; Section \@ref(instruments) buys it with an assumption about exclusion; the
threshold buys the narrowest population of all and asks for the least in return.
The fuzzy estimator of Section \@ref(rdd-the-first-stage-and-the-fuzzy-estimate)
is already an instrumental-variables estimator in disguise, which is where the
next section starts.
:::

### Readings {#rdd-readings -}

- Cattaneo, M. D., Idrobo, N., & Titiunik, R. *A Practical Introduction to Regression Discontinuity Designs: Foundations* (2020) and *... : Extensions* (2024). Cambridge Elements in Quantitative and Computational Methods for the Social Sciences. The current standard, written around R and Stata code; working-paper versions are on arXiv.
- Cattaneo, M. D., & Titiunik, R. (2022). Regression Discontinuity Designs. *Annual Review of Economics* 14, 821-851. The best short overview.
- Lee, D. S., & Lemieux, T. (2010). Regression Discontinuity Designs in Economics. *Journal of Economic Literature* 48(2), 281-355; and Imbens, G. W., & Lemieux, T. (2008). Regression Discontinuity Designs: A Guide to Practice. *Journal of Econometrics* 142(2), 615-635. The older canon, still cited by everyone.
- Cook, T. D. (2008). "Waiting for Life to Arrive": A History of the Regression-Discontinuity Design in Psychology, Statistics and Economics. *Journal of Econometrics* 142(2), 636-654. The history this section leans on.
- Cattaneo, M. D., Frandsen, B. R., & Titiunik, R. (2015). Randomization Inference in the Regression Discontinuity Design: An Application to Party Advantages in the U.S. Senate. *Journal of Causal Inference* 3(1), 1-24. The source of the Senate data used here, and of the local randomisation framework.

[^rdd-sozialindex]: The index was first built for the state by the Ruhr-Universität Bochum in 2020 and evaluated and updated in 2023; the ministry names the four indicators as child and youth poverty, the share of pupils whose family language is predominantly not German, the share who migrated themselves, and the share with support needs in learning, emotional and social development, or language — adding that schools with high values on *both* poverty and support needs are weighted more heavily, which is an interaction in substance if not in name. The hybrid scaling, in which the top five per cent form step 9 and the rest are cut into eight equal intervals, dates from the 2023 revision. Construction and revision are documented in Schräpler, J.-P., & Jeworutzki, S. (2021), *Konstruktion des Sozialindex für Schulen in Nordrhein-Westfalen*, ZEFIR-Materialien Bd. 14, <https://doi.org/10.46586/rub.zefir.213.186>, and Schräpler, J.-P., & Jeworutzki, S. (2025), *Sozialindex für Schulen in NRW — Zielsetzung, Umsetzung und Erfahrungen*, in Sendzik, N., et al. (Eds.), *Ungleich fördern — gerecht steuern*, DDS-Beiheft 20, Waxmann, 55-68, <https://doi.org/10.31244/9783830999812.05>. The four indicators are combined by a confirmatory factor analysis of the kind met in Section \@ref(confirmatory-factors). <https://www.schulministerium.nrw/schulsozialindex>

[^rdd-takeup]: The 923 schools arrived in two tranches — 400 in the 2024/25 school year and 523 in 2025/26 — so a "first cohort" figure is a statement about the rollout, not about how many invitations were declined. Whether a design is sharp or fuzzy is a fact about an administration rather than a property of the method, and here the administration recorded eligibility but not refusals. <https://www.schulministerium.nrw/startchancen>

[^rdd-goldberger]: Goldberger, A. S. (1972). *Selection Bias in Evaluating Treatment Effects: Some Formal Illustrations*. Discussion Paper 123-72, Institute for Research on Poverty, University of Wisconsin-Madison; reprinted in *Advances in Econometrics* 21 (2008), 1-31. The efficiency factor follows from bivariate normality with a cut-off at the mean of the score: $1/(1 - 2/\pi) = 2.752$. Goldberger's own phrasing is that a random sample of 100 is as good as a selected-on-pretest sample of 275.

[^rdd-albada]: The Gelman-Imbens recommendation is not uncontested. Albada, M. R. (2025), *Why High-Order Polynomials Should Not Be Used in Regression Discontinuity Designs. A Comment on Gelman and Imbens*, *Journal of Comments and Replications in Economics* 4, 2025-11, 1-16, <https://doi.org/10.18718/81781.50>, argues that the pathologies blamed on high-order polynomials also appear with local linear and local quadratic fits. Gelman, A., & Imbens, G. (2025), *Stable Adjustments in Regression Discontinuity. A Reply to Albada*, same volume, 2025-12, <https://doi.org/10.18718/81781.51>, reply that this misses the point: adjusting for the running variable alone, whatever the functional form, does not in general correct for imbalance in other pre-treatment variables.

[^rdd-two-bandwidths]: The 0.553 comes from a sharp regression on `participates` with its own MSE-optimal bandwidth of 5.8. The fuzzy call below reports 0.563, because it uses the bandwidth that is optimal for the ratio. Neither is more correct than the other, and the gap between them is a fair measure of how much a bandwidth choice is worth.

[^rdd-kolesar]: Kolesár, M., & Rothe, C. (2018). Inference in Regression Discontinuity Designs with a Discrete Running Variable. *American Economic Review* 108(8), 2277-2304, <https://doi.org/10.1257/aer.20160945>. The paper also shows that the common fix of clustering standard errors on the values of the running variable does not work: the resulting intervals "do not guard against model misspecification" and "have poor coverage properties".

[^rdd-evaluation]: The evaluation's published design is a repeated cross-section at 300 randomly selected programme schools, in years 4 and 9, with a baseline in early 2026 and a repeat in early 2029, alongside monitoring surveys of all programme schools roughly every eighteen months; the project runs to 2030. The consortium is led by infas with IEA Hamburg running the achievement testing. No identification strategy is named in the public documentation — the reading of it as a difference-in-differences with a threshold design alongside is this book's, not the consortium's. <https://evaluation-startchancen.de>
