## Difference-in-Differences {#difference-in-differences}

::: {.analysis-subtitle}
Comparing changes across groups and time
:::

In August 2024 Germany began spending twenty billion euros on about four thousand schools. The **Startchancen-Programm** runs for ten years, splits the money evenly between the federal government and the Länder, and is aimed squarely at schools in the hardest circumstances: money for buildings, a discretionary *Chancenbudget* for each school, and staff for multi-professional teams.[^did-startchancen] The stated target is arithmetic rather than rhetorical — halve, *at the programme's own schools*, the share of pupils who fail to reach the minimum standard in German and mathematics. Nationally that share was 18.8% in reading and 21.8% in mathematics at the end of primary school. At the end of year nine the number depends on which school-leaving standard you measure against, and the two are not comparable: 15.2% missed the minimum standard for the *erster Schulabschluss* in reading in 2022, while roughly a third missed the more demanding *mittlerer Schulabschluss* standard in mathematics in 2024.[^did-iqb]

Now the question this section is about. In 2032, someone will have to say whether it worked. How?

You cannot compare Startchancen schools with other schools, because they were selected for being different — that is the entire point of the programme. You cannot compare the schools with their own past, because German competence scores have been falling for a decade and a school that merely stops falling has done something. And you certainly cannot randomise: no ministry is going to draw lots for a billion euros a year.

What the evaluation consortium — led by infas, with the Leibniz Institute for Educational Trajectories, IEA Hamburg, Tübingen, Potsdam and others — has set up is the raw material for the design in this section.[^did-infas] Measure the treated schools in 2026 and again in 2029. Build a comparison group out of the national assessments. Then compare not the *levels* but the *changes*. Whatever made Startchancen schools different in 2026 is allowed to keep making them different, as long as it does not change over the six years in between.

That single move — subtract twice instead of once — is the most-used identification strategy in policy evaluation, and it is old. John Snow used it in 1855 on the London water supply.[^did-snow] Its modern fame dates from 1994, when David Card and Alan Krueger asked what happened to fast-food employment when New Jersey raised its minimum wage and Pennsylvania did not: employment per store went from 20.4 to 21.0 in New Jersey and from 23.3 to 21.2 across the river, a relative gain of 2.76 full-time-equivalent employees, or 13%.[^did-cardkrueger] Germany ran the same experiment on itself in January 2015, when a national minimum wage of €8.50 arrived in a country that had never had one, and the design has been carrying that literature ever since.[^did-milog] It reaches well beyond labour economics: the staggered arrival of high-yielding wheat and rice varieties across countries is what lets economists put a number on the Green Revolution, and by extension on the work for which Norman Borlaug was given the Nobel Peace Prize in 1970.[^did-greenrev]

This section builds the design from four numbers in a table, and then spends most of its length on the two things that have happened to it recently. The first is that its central assumption turned out to be less testable than everyone believed. The second is that the standard way of estimating it turned out to be wrong whenever a programme is rolled out in waves — which is exactly how Startchancen was rolled out.[^did-rollout]

### The Idea in One Table {#did-2x2}

Two groups, two periods, four averages.

|  | before | after |
|:---|:---:|:---:|
| **treated group** | $\bar{Y}_{T,0}$ | $\bar{Y}_{T,1}$ |
| **comparison group** | $\bar{Y}_{C,0}$ | $\bar{Y}_{C,1}$ |

The treated group changed by $\bar{Y}_{T,1} - \bar{Y}_{T,0}$. Some of that is the programme and some is everything else that happened between the two dates. The comparison group tells you what "everything else" looks like, because it experienced everything else and not the programme. Subtract:

$$\widehat{\text{DiD}} = \underbrace{\left(\bar{Y}_{T,1} - \bar{Y}_{T,0}\right)}_{\text{treated change}} - \underbrace{\left(\bar{Y}_{C,1} - \bar{Y}_{C,0}\right)}_{\text{everything else}}$$

The first difference removes everything that is fixed about a unit — a school's catchment area, its building, its decades-old reputation. The second removes everything that is common to a period — a curriculum reform, a pandemic, a change in how the test is scored. What survives is the change in the treated group that is not shared by anyone else.

::: {.analysis-takeaway}
This is why level differences are not a problem. Startchancen schools score lower than other schools and always did; the design does not care. What it cannot survive is a difference in *trajectories*. Section \@ref(matching-and-balancing) needed the two groups to be comparable; this design needs only their trends to be.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Two groups, two periods, four averages, and the whole design is one subtraction.

Suppose the treated group goes from 40 to 46 and the comparison group from 50 to
53. The difference in differences is `r fitb(3, num = TRUE)`.

Now the same treated numbers, but the comparison group goes from 50 to 58. The
difference in differences is `r fitb(-2, num = TRUE)`.

In the second case the treated group still improved by six points. Would you
report that the programme worked?
`r mcq(c("yes, six points is six points", answer = "no - everyone improved by more, so the programme is associated with falling behind", "impossible to say without a standard error"))`

Which of the two groups' *levels* matters to the estimate?
`r mcq(c("the treated group's", "the comparison group's", answer = "neither - only the two changes enter"))`
::::


The same estimate comes out of a regression, and it is worth seeing why, because every extension in this section is an extension of the regression rather than of the table.

$$Y_{it} = \alpha + \beta\, \text{Treated}_i + \gamma\, \text{After}_t + \delta\, (\text{Treated}_i \times \text{After}_t) + \varepsilon_{it}$$

$\beta$ is the permanent gap between the groups, $\gamma$ is what happened to everyone, and $\hat\delta$ is the difference in differences, numerically identical to the table.

### Parallel Trends {#did-parallel}

:::: {.defbox}
::: {.titeldefbox}
<h2>Definition</h2>
:::
**Parallel trends**: in the absence of the treatment, the average outcome of the treated group would have changed by the same amount as the average outcome of the comparison group,

$$E\!\left[Y_{i1}(0) - Y_{i0}(0) \mid \text{Treated}\right] \;=\; E\!\left[Y_{i1}(0) - Y_{i0}(0) \mid \text{Comparison}\right]$$

Under it, the difference in differences identifies the average treatment effect on the treated.

Two further assumptions travel with it and are usually left implicit. **No anticipation**: the outcome in the pre-period is not already responding to a treatment that has been announced. **SUTVA**: the comparison units are genuinely untreated — no spillovers.
::::

Read the left-hand side again. It contains $Y_{i1}(0)$ for the treated group: what the treated units *would have done* without the programme. That quantity is never observed, in any data set, ever. Parallel trends is therefore not a statement about your data. It is a statement about a counterfactual, and it can be argued for, made plausible, probed — but not confirmed.

::: {.analysis-warning}
Parallel trends is not scale-free. If two groups have parallel trends in levels they generally do *not* have parallel trends in logs, and vice versa. Taking a log of the outcome is not a neutral presentational choice here; it changes the assumption you are making. Susan Athey and Guido Imbens made this the motivation for an entire alternative model, *changes-in-changes*, which asks for something invariant to monotone transformations instead.[^did-athey]
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Two towns. Town A's wage goes from 10 to 12, town B's from 20 to 24.

In levels the changes are `r fitb(2, num = TRUE)` and `r fitb(4, num = TRUE)` —
not parallel.

In logs both changed by the same proportion, so in logs they are
`r mcq(c("still not parallel", answer = "parallel", "impossible to compare"))`.

Which of the two is the right specification?
`r mcq(c("logs, because wages are always modelled in logs", "levels, because the outcome is measured in euros", answer = "neither is right in general - the choice is an assumption about the counterfactual and has to be argued"))`

True or false: taking a log of the outcome is a presentational choice with no
consequences for identification. `r torf(FALSE)`
::::


### A Programme We Can Check {#did-data}

Section \@ref(matching-and-balancing) built its own data so that the right answer would be available at the end, and Sections \@ref(rdd-a-programme-we-can-check) and \@ref(iv-a-programme-we-can-check) will do it again. The trick is worth more here than anywhere, because this section has to show two estimators disagreeing about a number, and there is no way to referee that without knowing the number.

```{r did-packages, echo=TRUE}
library("fixest")   # fast fixed-effects regression, event studies, Sun-Abraham
library("did")      # Callaway and Sant'Anna group-time treatment effects
```

The fictional programme below is Startchancen-shaped: schools are selected because they are struggling, the money arrives in three waves, and the effect builds up the longer a school has been in the programme.

```{r did-dgp, echo=TRUE}
simulate_schools <- function(n_cohort = 150, n_never = 75, years = 12,
                             seed = 7, divergent = 0) {

  set.seed(seed)

  # --- three funding waves plus schools that never get in ---------------
  cohort <- c(rep(c(4, 6, 8), each = n_cohort), rep(0, n_never))
  n      <- length(cohort)

  # funded schools start lower - that is why they were chosen
  start  <- rnorm(n, 50, 6) - 4 * (cohort > 0)

  # effects build with exposure, and the early waves gain fastest
  gain   <- c("4" = 1.10, "6" = 0.45, "8" = 0.08)[as.character(cohort)]
  gain[is.na(gain)] <- 0

  d <- expand.grid(school = seq_len(n), year = seq_len(years))
  d$cohort   <- cohort[d$school]
  d$exposure <- ifelse(d$cohort == 0, 0, pmax(0, d$year - d$cohort + 1))
  d$effect   <- gain[d$school] * d$exposure

  # what every school would have scored without the programme:
  # a common national decline, plus (optionally) a divergent drift
  d$y0 <- start[d$school] - 0.30 * (d$year - 1) +
    divergent * (-0.35) * (d$year - 1) * (d$cohort > 0) +
    rnorm(nrow(d), 0, 2)
  d$y1 <- d$y0 + d$effect

  d$score  <- ifelse(d$exposure > 0, d$y1, d$y0)
  d$funded <- as.integer(d$exposure > 0)

  d[order(d$school, d$year),
    c("school", "year", "cohort", "exposure", "funded", "score",
      "y0", "y1", "effect")]
}

schools <- simulate_schools()

schools %>%
  group_by(cohort) %>%
  summarise(schools = n_distinct(school), first_funded_year = first(cohort))
```

One row is one school in one year. `score` is the school's average competence score on the familiar test metric — a scale with a national mean near 50 and a standard deviation near 10 — and because it is a school average rather than a pupil score, its year-to-year noise is small. `cohort` is the year the school entered the programme, with **0** meaning never. `y0` and `y1` are the two potential outcomes for every school-year; nothing before Section \@ref(did-truth) touches them.

Four features are deliberate.

**Selection on levels, not on trends.** Funded schools start four points below the rest. That is confounding of exactly the kind matching had to work for, and here it is free.

**A national decline.** Everyone loses 0.30 points a year. A funded school that merely holds still is doing well, and only the comparison group can tell you that.

**Effects that grow.** A school in its sixth funded year is doing better than one in its first. This is realistic, and it is the ingredient that will break the standard estimator later.

**Effects that differ by wave.** The first wave gains 1.10 points per year of exposure, the second 0.45, the third 0.08. Real programmes are not equally effective everywhere, and the early adopters are rarely a random sample.

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Four design decisions went into that generator. Match each to what it is there to
break.

Funded schools start four points below the rest. That is
`r mcq(c("a violation of parallel trends", answer = "selection on levels, which the design tolerates", "a violation of SUTVA"))`.

Everyone loses 0.30 points a year. That is there so that
`r mcq(c("the effects grow", answer = "a funded school which merely holds still is already doing well, and only a comparison group can reveal it", "the standard errors are correct"))`.

Effects grow with exposure. That is the ingredient which will later break
`r mcq(c("the parallel-trends assumption", answer = "the two-way fixed effects estimator", "the clustering"))`.

And the `divergent` argument, used later, switches on
`r mcq(c("a treatment effect", answer = "a violation of parallel trends", "a spillover"))`.
::::


```{r did-raw, echo=FALSE, fig.align='center', fig.width=7, fig.height=4, fig.cap="Average score by funding wave. The dotted vertical lines mark when each wave starts receiving money. Everyone drifts downwards; the funded schools bend away from that drift at their own start year, and the first wave bends hardest. This one figure contains everything the rest of the section argues about."}
schools %>%
  mutate(wave = factor(cohort, c(0, 4, 6, 8),
                       c("never funded", "wave 1 (year 4)",
                         "wave 2 (year 6)", "wave 3 (year 8)"))) %>%
  group_by(wave, year) %>%
  summarise(score = mean(score), .groups = "drop") %>%
  ggplot(aes(year, score, colour = wave)) +
  geom_vline(xintercept = c(4, 6, 8) - 0.5, linetype = "dotted",
             colour = "grey60") +
  geom_line(linewidth = .8) +
  scale_x_continuous(breaks = 1:12) +
  scale_colour_manual(values = c("#5a5a5a", "#9b3f2d", "#c8925f", "#2c6e8e")) +
  labs(x = "Year", y = "Mean competence score", colour = NULL,
       title = "Three waves, one comparison group") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

### The Canonical Comparison {#did-canonical}

Start with the textbook case: one treated wave, one comparison group, one year before and one year after. Wave 1 enters in year 4, so year 3 is *before* and year 5 is *after*.

```{r did-cells, echo=TRUE}
two_by_two <- schools %>%
  filter(cohort %in% c(0, 4), year %in% c(3, 5)) %>%
  mutate(funded_school = as.integer(cohort == 4),
         after         = as.integer(year == 5))

cells <- two_by_two %>%
  group_by(funded_school, after) %>%
  summarise(mean_score = mean(score), .groups = "drop")

cells
```

Funded schools went from 46.46 to 47.96, a gain of 1.50. Never-funded schools went from 49.03 to 48.73, a loss of 0.30. The difference in differences is **1.80**.

```{r did-cells-hand, echo=TRUE}
m <- with(cells, tapply(mean_score, list(funded_school, after), identity))

(m["1", "1"] - m["1", "0"]) - (m["0", "1"] - m["0", "0"])
```

The regression gives the same number and adds an inferential apparatus. Standard errors are clustered on the school, for reasons the section returns to.

```{r did-reg, echo=TRUE}
fit_2x2 <- feols(score ~ funded_school * after, data = two_by_two,
                 cluster = ~ school)
fit_2x2
```

Read the three coefficients as the table. `funded_school` is $-2.57$: the permanent gap, and not an effect of anything. `after` is $-0.30$: the national decline. `funded_school:after` is $+1.80$ with a standard error of 0.39 — the estimate.

::: {.analysis-note}
The comparison group earns its keep in the middle coefficient. Look at the funded schools alone and you see a gain of 1.50 points; a before-and-after study would report that. The comparison group says the year was worth $-0.30$ to everybody, so the programme is credited with 1.80 rather than 1.50. Here the correction is small and upward. In a year with a curriculum reform or a pandemic it can be larger than the effect itself, and of either sign.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The regression printed three coefficients: $-2.57$, $-0.30$ and $+1.80$.

Which one is the estimate? `r fitb(1.8, tol = 0.05)`

The $-2.57$ on `funded_school` is `r mcq(c("the effect of the programme before it started", answer = "the permanent gap between the two groups, and not an effect of anything", "a sign that the randomisation failed"))`.

A colleague reports the $-0.30$ on `after` as "the effect of the year on the
comparison group". Is that right?
`r mcq(c(answer = "yes - it is what happened to everyone, and it is the whole reason the design works", "no, it is the effect on the treated group", "no, it is a nuisance parameter with no interpretation"))`

And a small arithmetic check on the note above. The funded schools gained 1.50
and were credited with 1.80. If the year had been worth $+0.60$ to everybody
instead of $-0.30$, the programme would have been credited with
`r fitb(0.9, tol = 0.05)`.
::::


### Event Studies {#did-event-study}

Two periods waste the data. With twelve years there is a coefficient for every year relative to the start of funding, and that object — the **event study** — is where most of the credibility of a modern DiD paper lives.

$$Y_{it} = \alpha_i + \lambda_t + \sum_{k \neq -1} \delta_k \, D_{it}^{k} + \varepsilon_{it}$$

$\alpha_i$ is a school fixed effect, $\lambda_t$ a year fixed effect, and $D_{it}^k$ indicates that school $i$ is $k$ years from its funding start in year $t$. One relative period has to be dropped as the reference; convention is $k = -1$, the year before the money arrives.

```{r did-event, echo=TRUE}
# a coefficient label from i() ends in its relative period: "rel::-3", "rel::4"
relative_period <- function(labels) {
  as.numeric(sub(".*?(-?[0-9]+)$", "\\1", labels, perl = TRUE))
}

event_study <- function(data, first_year = 4) {
  d <- data %>%
    filter(cohort %in% c(0, first_year)) %>%
    mutate(rel = ifelse(cohort == 0, -1, year - first_year))
  fit <- feols(score ~ i(rel, ref = -1) | school + year,
               data = d, cluster = ~ school)
  list(fit = fit,
       tidy = data.frame(rel      = relative_period(names(coef(fit))),
                         estimate = as.numeric(coef(fit)),
                         se       = sqrt(diag(vcov(fit)))))
}

es <- event_study(schools)
round(es$tidy, 3)
```

The two pre-period coefficients are $-0.09$ and $-0.34$, both smaller than their own standard errors. The post-period coefficients climb from 1.21 in the first funded year to 9.53 in the ninth. The true effects, which we are not supposed to know yet, are 0 and 0 before and $1.1, 2.2, \dots, 9.9$ after.

```{r did-event-fig, echo=FALSE, fig.align='center', fig.width=7, fig.height=4.4, fig.cap="Event-study coefficients with 95% confidence intervals, in the world where trends really are parallel (left) and in a world where funded schools were already drifting downwards before the money arrived (right). The dashed grey line is the truth. On the right the two pre-period estimates are positive but individually insignificant - the violation is real, the diagnostic is quiet, and every post-period estimate is too low."}
truth_line <- data.frame(rel = -3:8, true = c(0, 0, 0, 1.1 * (1:9)))

bind_rows(
  event_study(schools)$tidy %>% mutate(world = "Parallel trends hold"),
  event_study(simulate_schools(divergent = 1))$tidy %>%
    mutate(world = "Funded schools were already drifting")
) %>%
  rename(est = estimate) %>%
  bind_rows(data.frame(rel = -1, est = 0, se = 0,
                       world = c("Parallel trends hold",
                                 "Funded schools were already drifting"))) %>%
  mutate(world = factor(world, c("Parallel trends hold",
                                 "Funded schools were already drifting"))) %>%
  ggplot(aes(rel, est)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = -0.5, linetype = "dotted", colour = "grey60") +
  geom_line(data = truth_line, aes(rel, true), linetype = "dashed",
            colour = "grey55") +
  geom_pointrange(aes(ymin = est - 1.96 * se, ymax = est + 1.96 * se),
                  colour = "#2c6e8e", size = .3) +
  facet_wrap(~ world) +
  scale_x_continuous(breaks = -3:8) +
  labs(x = "Years since funding started", y = "Effect on mean score",
       title = "The same programme, two assumptions about the counterfactual") +
  theme_minimal(base_size = 10)
```

An event study does three jobs at once, and they are worth separating. It shows whether the effect grows, plateaus or fades. It shows whether anything happened *before* the treatment, which would suggest anticipation or a pre-existing divergence. And it shows the reader the whole shape rather than one summary number, which makes the paper harder to fool anybody with.

::: {.analysis-warning}
A flat pre-period is evidence, not proof, and it is evidence about the pre-period only. Parallel trends is a claim about what the treated group would have done *after* the treatment. Nothing in the data speaks to that directly. Two groups can move together for five years and part company in the sixth for reasons that have nothing to do with the programme — and then the design attributes the parting to the programme.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The event study dropped one relative period as the reference. Which, and why?

The dropped period is $k =$ `r fitb(-1, num = TRUE)`, the year before the money
arrives, because
`r mcq(c("it has the most observations", answer = "the coefficients are only ever differences from some baseline, and one has to be fixed at zero", "it is the least noisy"))`.

An event study does three jobs at once. Which of these is **not** one of them?
`r mcq(c("it shows whether the effect grows, plateaus or fades", "it shows whether anything happened before the treatment", answer = "it tests whether parallel trends holds after the treatment", "it shows the reader the whole shape rather than one number"))`

In the divergent world the first post-period estimate was 0.86 against a truth of
1.1, and the ninth 6.38 against 9.9. So the bias
`r mcq(c("is constant", answer = "grows the further from the start you look", "shrinks"))`,
which is exactly what an ongoing divergence does when you keep subtracting it.
::::


### When Trends Are Not Parallel {#did-violations}

The right-hand panel of the figure above comes from the same generator with one argument changed. `divergent = 1` makes funded schools lose an additional 0.35 points a year *throughout*, before and after the money — the deprived schools were falling behind anyway. The programme's true effect is unchanged.

```{r did-divergent, echo=TRUE}
drifting <- simulate_schools(divergent = 1)

event_study(drifting)$tidy %>%
  mutate(true = c(0, 0, 1.1 * (1:9))) %>%
  round(3)
```

Every post-period estimate is too low, and the gap widens the further from the start you look: 0.86 against a true 1.1 in the first year, 6.38 against 9.9 in the ninth. The design has silently subtracted the ongoing divergence from the programme's effect.

The historically famous version of this runs the other way and has a name.

:::: {.amazing}
::: {.titelamazing}
<h2>Ashenfelter's dip</h2>
:::
In 1978 Orley Ashenfelter looked at the earnings of people who enrolled in American job-training programmes and found something that has shaped programme evaluation ever since: their earnings fell sharply in the year or two *before* they enrolled.[^did-ashenfelter]

It is not mysterious. People sign up for retraining after they lose a job, or after a bad year. The dip is the reason they are in the programme at all.

Now put that into the two-period table. Pick the pre-period at the bottom of the dip and the treated group will bounce back on its own — mean reversion — and the design will credit the bounce to the training. Pick it earlier and you get a different answer. The estimate depends on the choice of "before", which is a devastating thing to have to admit about a research design.

The dip is why event studies became compulsory rather than optional, and why the phrase *selection into treatment on the basis of transitory shocks* appears in referee reports. It also generalises well beyond training: firms adopt a new technology after a bad quarter, schools get an intervention after a bad cohort, patients start a drug after a flare-up.
::::

### Pre-Trend Tests Are Weaker Than They Look {#did-pretesting}

The standard defence is to test the pre-period coefficients jointly and report that the test does not reject. Since we can generate as many studies as we like, we can ask how well that defence actually works.

```{r did-pretest, echo=TRUE}
# joint chi-square test that every pre-period coefficient is zero
pretrend_p <- function(fit, rel) {
  lead <- which(rel < 0)
  b <- coef(fit)[lead]
  V <- vcov(fit)[lead, lead, drop = FALSE]
  stat <- as.numeric(t(b) %*% solve(V) %*% b)
  pchisq(stat, df = length(lead), lower.tail = FALSE)
}

pretrend_study <- function(divergent, reps = 200) {
  out <- t(replicate(reps, {
    e <- event_study(simulate_schools(seed = sample(1e6, 1),
                                      divergent = divergent))
    c(reject = as.numeric(pretrend_p(e$fit, e$tidy$rel) < 0.05),
      est    = e$tidy$estimate[e$tidy$rel == 4])   # effect five years in
  }))
  c(rejection_rate      = mean(out[, "reject"]),
    mean_estimate       = mean(out[, "est"]),
    mean_if_test_passed = mean(out[out[, "reject"] == 0, "est"]))
}

set.seed(99)
round(pretrend_study(divergent = 0), 3)   # trends really are parallel
round(pretrend_study(divergent = 1), 3)   # trends really are not
```

In the world where parallel trends holds, the test rejects about 5% of the time. That is exactly what a 5% test should do, and it is reassuring about the machinery: the estimate averages 5.50 against a true 5.50.

In the world where parallel trends fails, the test rejects roughly a third of the time. Two studies in three are told that their trends look parallel when they are not. And here is the part that turns a nuisance into a bias: among the studies that *passed* the test, the average estimate is about **3.6** against a true effect of **5.5**. Passing the pre-trend test does not mark out the studies that got the right answer. If anything it selects the samples in which the pre-period noise happened to hide the drift — which are disproportionately the samples in which the post-period estimate is off.

::: {.analysis-takeaway}
This is Jonathan Roth's result, and it is the most useful thing to have learned about DiD in the last decade.[^did-roth] A pre-trend test is underpowered against exactly the violations that matter, and conditioning the analysis on having passed one distorts the estimate that follows. "We checked for parallel trends and found none" is a much weaker sentence than it sounds, and the honest version of it reports the power of the check, not just its p-value.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Read the two simulation results together. In the world where trends really are
parallel, the test rejected about `r fitb(0.06, tol = 0.02)` of the time — which
is what a 5% test should do.

In the world where trends are not parallel, it rejected about
`r mcq(c("nineteen times in twenty", answer = "a third of the time", "never"))`.

Now the part that matters. Among the studies that *passed* the test, the average
estimate was 3.6 against a true 5.5. Why is that worse than simply being
underpowered?
`r mcq(c("it is not worse, it is the same thing", answer = "passing selects the samples in which pre-period noise happened to hide the drift, and those are disproportionately the samples whose post-period estimate is off", "the test statistic is computed incorrectly"))`

True or false: reporting the p-value of a pre-trend test is enough.
`r torf(FALSE)`

`r hide("What to report instead")`

Report the power. A pre-trend test that rejects a third of the time against the
violation you are worried about is telling you almost nothing when it fails to
reject, and the honest write-up says how large a violation it *could* have
caught. Rambachan and Roth's approach goes one step further and drops the
hypothesis test altogether: assume the post-period violation is no more than
$\bar{M}$ times what is visible in the pre-period, and report the range of
effects that survives. Same move as the E-value in Section
\@ref(matching-and-balancing) — turn an untestable assumption into a magnitude a
reader can argue with.

`r unhide()`
::::


The constructive response is not to abandon the design. It is to stop treating parallel trends as a binary that data can settle, and start treating the possible violation as a quantity with a size. Ashesh Rambachan and Roth's approach asks: if the post-period violation were no larger than what is visible in the pre-period — or at most $\bar{M}$ times as large — what range of effects would remain consistent with the data?[^did-rambachan] That produces a set rather than a point, and it makes the argument about magnitudes, which is where it belongs. It is the same move as the E-value in Section \@ref(matching-and-balancing): turn an untestable assumption into a quantity a reader can dispute.

### Standard Errors {#did-standard-errors}

Every regression in this section clusters on the school, and the reason is one of the most cited papers in applied economics.

Marianne Bertrand, Esther Duflo and Sendhil Mullainathan took real data, invented placebo laws that never happened, and ran the standard DiD regression on them. With conventional standard errors, the effect of a law that did not exist was significant at the 5% level in up to 45% of the placebo runs.[^did-bdm] The cause is serial correlation: outcomes within a unit are correlated over time, so a panel with twelve years of data does not contain twelve independent observations per school. Treating it as though it does inflates the apparent precision by a factor of several.

Clustering on the unit of treatment fixes it, provided there are enough clusters. With few — a treatment assigned at the level of sixteen Länder, say — cluster-robust errors are themselves unreliable, and the standard remedy is the wild cluster bootstrap.[^did-cgm]

::: {.analysis-warning}
Cluster at the level at which the treatment varies, not at the level at which the data were collected. If a policy switches on for a whole Bundesland, the effective number of independent observations is the number of Länder, however many pupils are in the file. This is the mistake that most often turns a null result into a published one.
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Bertrand, Duflo and Mullainathan invented laws that never happened and found an
"effect" significant at the 5% level in up to
`r fitb(45, num = TRUE)` per cent of their placebo runs.

The cause was `r mcq(c("a coding error", answer = "serial correlation - outcomes within a unit are correlated over time, so twelve years is not twelve independent observations", "too small a sample"))`.

A free-childcare reform switches on for one Bundesland. Your file has 40,000
pupils. Cluster on
`r mcq(c("the pupil", "the school", answer = "the Bundesland", "the year"))`,
which leaves you with `r fitb(16, num = TRUE)` clusters.

With that few clusters, cluster-robust errors are themselves unreliable. The
standard remedy is the `r fitb("wild cluster bootstrap", ignore_case = TRUE, width = "24")`.
::::


### Everyone Starts at a Different Time {#did-staggered}

Real programmes do not switch on everywhere at once. Startchancen began with 2,125 schools in the 2024/25 school year and had passed four thousand by 2026; the Länder legislate at different times; firms adopt technologies in waves.[^did-rollout] The natural generalisation of the regression is to replace *Treated × After* with a single indicator that is one whenever a unit is currently treated, and to absorb everything else in unit and year fixed effects:

$$Y_{it} = \alpha_i + \lambda_t + \delta\, D_{it} + \varepsilon_{it}$$

This is the **two-way fixed effects** (TWFE) estimator. For twenty years it was what "doing difference-in-differences" meant. It takes one line.

```{r did-twfe, echo=TRUE}
feols(score ~ funded | school + year, data = schools, cluster = ~ school)
```

The estimate is $-0.006$, with a standard error of 0.146 and a p-value of 0.96.

A programme that lifted every wave it touched, in a data set we built ourselves, comes back as a precisely estimated zero. Not a wide confidence interval — a tight one, centred on nothing. Any evaluator would report that the programme had no effect, and the reporting would be careful and the referee would nod.

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Before reading on, commit to an explanation.

The estimate is $-0.006$ with a standard error of 0.146. Which of these is it?
`r mcq(c("the programme really had no effect in this data set", "the sample is too small to detect the effect", answer = "the estimator is averaging over comparisons that should never have been made"))`

If you had seen only this output, what would have told you something was wrong?
`r mcq(c("the standard error", "the p-value", "the R-squared", answer = "nothing in the output at all"))`

True or false: adding more years of data would have revealed the problem.
`r torf(FALSE)`
::::


### The Forbidden Comparison {#did-forbidden}

The reason is a comparison the regression makes without being asked.

With one treated group and one comparison group there is one 2×2 table. With three waves and a never-funded group there are many, and the TWFE coefficient is a weighted average of all of them. Andrew Goodman-Bacon worked out exactly which ones.[^did-bacon] Some are the comparisons you meant to make: each wave against the schools that never get funded. Some compare an early wave against a later wave *before the later one starts*, which is also fine — the later wave is untreated then, so it is a legitimate control.

And some compare a **later wave against an earlier wave that is already being treated**.

```{r did-forbidden, echo=TRUE}
compare_2x2 <- function(d, group_a, group_b, year_before, year_after) {
  w <- d %>%
    filter(cohort %in% c(group_a, group_b),
           year %in% c(year_before, year_after)) %>%
    mutate(b = as.integer(cohort == group_b),
           a = as.integer(year == year_after))
  cell <- with(w %>% group_by(b, a) %>%
                 summarise(m = mean(score), .groups = "drop"),
               tapply(m, list(b, a), identity))
  (cell["1", "1"] - cell["1", "0"]) - (cell["0", "1"] - cell["0", "0"])
}

c(clean     = compare_2x2(schools, group_a = 0, group_b = 4, 6, 10),
  forbidden = compare_2x2(schools, group_a = 4, group_b = 8, 6, 10)) %>%
  round(3)
```

The same programme, the same two years. Compared against schools that were never funded, wave 1 gained **+4.21** points. Compared against wave 1 — which by year 6 had been funded for three years and was still improving — wave 3 *lost* **4.39** points.

Nothing is wrong with the arithmetic. Wave 3's score rose by 0.24 points between year 6 and year 10. Wave 1's rose by 4.4 over the same stretch, because its effect was still growing. Subtract, and a programme that helped both groups reports a large negative number. The already-treated control is not a control at all: it is a treated unit whose treatment is still doing work, and that work gets subtracted from the effect you are trying to measure.

::: {.analysis-note}
A common misreading is that the Goodman-Bacon weights are negative. They are not — they are products of variances and sample shares, and they are non-negative. What is negative is the *estimand* of the forbidden 2×2 itself, because it differences out a growing treatment effect. Negative weights on the underlying treatment effects are a related but distinct result, due to Clément de Chaisemartin and Xavier D'Haultfœuille, and they can make TWFE report a negative number when every single unit-level effect is positive.[^did-chaisemartin]
:::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
The same programme, the same two years: $+4.21$ against the never-funded schools,
$-4.39$ against wave 1.

Wave 3's score rose by 0.24 points between year 6 and year 10; wave 1's rose by
`r fitb(4.4, tol = 0.3)`. Subtract, and you get the second number.

What is wrong with using wave 1 as a control for wave 3?
`r mcq(c("wave 1 is too small", answer = "wave 1 is being treated over exactly the period in question, and its ongoing gains are subtracted from wave 3's", "wave 1 started from a different level"))`

Now the sentence that saves twenty years of applied work. Two-way fixed effects
is unbiased when treatment effects are
`r mcq(c("large", "positive", answer = "the same for every unit and at every duration"))`.

And the precision. Which of the two is the Goodman-Bacon result and which the de
Chaisemartin-D'Haultfœuille one? Negative *weights on treatment effects* are
`r mcq(c("Goodman-Bacon", answer = "de Chaisemartin and D'Haultfœuille"))`;
a decomposition into 2×2 comparisons with non-negative weights is
`r mcq(c(answer = "Goodman-Bacon", "de Chaisemartin and D'Haultfœuille"))`.
::::


The condition under which none of this matters is worth stating plainly, because it is the condition under which twenty years of applied work was fine: **TWFE is unbiased when treatment effects are constant** — the same for every unit and, crucially, the same at every duration. The moment effects grow with exposure, which is the normal case for anything educational or organisational, the forbidden comparisons start pulling the average towards zero and past it.

### Estimators That Survive Staggering {#did-modern}

The fix, in the form given by Brantly Callaway and Pedro Sant'Anna, is to refuse to average anything until the end.[^did-cs] Estimate a separate effect for every combination of funding wave $g$ and year $t$, each one from a clean 2×2: the wave's change from the year before it was funded, minus the change over the same years in a group that is genuinely untreated. Only then aggregate, deliberately, with weights you choose.

It is short enough to write by hand, which is the best way to see that nothing clever is happening.

```{r did-cs-hand, echo=TRUE}
att_gt_hand <- function(d, g, t) {
  w <- d %>%
    filter(year %in% c(g - 1, t), cohort %in% c(0, g)) %>%
    mutate(treated = as.integer(cohort == g),
           post    = as.integer(year == t))
  cell <- with(w %>% group_by(treated, post) %>%
                 summarise(m = mean(score), .groups = "drop"),
               tapply(m, list(treated, post), identity))
  (cell["1", "1"] - cell["1", "0"]) - (cell["0", "1"] - cell["0", "0"])
}

group_time <- expand.grid(g = c(4, 6, 8), t = 1:12) %>%
  filter(t >= g) %>%
  mutate(att = mapply(att_gt_hand, list(schools), g, t))

group_time %>%
  group_by(wave = g) %>%
  summarise(periods = n(), mean_att = mean(att)) %>%
  mutate(mean_att = round(mean_att, 3))
```

Wave 1 averages **+5.25** over its nine funded years, wave 2 **+1.83** over seven, wave 3 **+0.21** over five. Averaged over all treated school-years, the estimate is **+2.91**.

The package does the same thing with proper standard errors, and it wants the never-treated group coded as `0` — which is why the generator codes it that way.

```{r did-cs-package, echo=TRUE}
cs <- att_gt(yname = "score", tname = "year", idname = "school",
             gname = "cohort", data = schools,
             control_group = "nevertreated", bstrap = FALSE, cband = FALSE)

aggte(cs, type = "simple", bstrap = FALSE, cband = FALSE)
aggte(cs, type = "group",  bstrap = FALSE, cband = FALSE)
```

Liyang Sun and Sarah Abraham attack the same problem from the event-study side: in the ordinary specification each lead and lag coefficient is contaminated by effects from *other* relative periods, so an event study can show a spurious pre-trend that is nothing but treatment-effect heterogeneity.[^did-sunab] Their interaction-weighted estimator is one function call inside `feols` — with the opposite coding convention, which is a genuine trap.

```{r did-sunab, echo=TRUE}
# fixest reads any cohort value that does not appear among the periods as
# never-treated. Its own examples use 1000; `did` insists on 0.
staggered <- schools %>%
  mutate(cohort_sa = ifelse(cohort == 0, 1000L, cohort))

feols(score ~ sunab(cohort_sa, year) | school + year,
      data = staggered, cluster = ~ school) %>%
  summary(agg = "att")
```

::: {.analysis-warning}
Two packages, two conventions, and the trap is not the one you would guess. `did::att_gt()` requires never-treated units to carry a group value of **0**. `fixest::sunab()` asks instead for a cohort value that lies **outside the range of the periods** — its own examples use 1000 — and it identifies the never-treated as *any cohort that does not appear among the period values*.

Read that rule twice, because it cuts both ways. In this data set the periods run from 1 to 12, so a value of 0 is outside the range too, and coding the never-treated as 0 happens to give exactly the same 6,300 observations and the same estimate. What the rule cannot survive is a cohort that is genuinely **always** treated — a school funded before the panel begins, coded 1990 in a panel that starts in 1995. That cohort also fails to appear among the periods, so `sunab()` files it under *never treated* and uses it as a control. The package's own documentation says so, and nothing in the output warns you. Check `nobs()` whenever you change the coding, and check the cohort values against the period range whenever you do not.
:::

### Opening the Envelope {#did-truth}

The data frame has carried `y0` and `y1` all along.

```{r did-truth, echo=TRUE}
funded <- schools$funded == 1

truth <- c(overall = mean(schools$effect[funded]),
           wave1   = mean(schools$effect[funded & schools$cohort == 4]),
           wave2   = mean(schools$effect[funded & schools$cohort == 6]),
           wave3   = mean(schools$effect[funded & schools$cohort == 8]))

round(truth, 3)
```

```{r did-truth-table, echo=FALSE}
tibble(
  Estimator = c("Before-and-after, funded schools only",
                "Two-way fixed effects",
                "Callaway-Sant'Anna, aggregated",
                "Callaway-Sant'Anna, wave 1",
                "Callaway-Sant'Anna, wave 2",
                "Callaway-Sant'Anna, wave 3"),
  Estimate = c(
    with(schools, mean(score[funded == 1]) - mean(score[cohort > 0 & funded == 0])),
    coef(feols(score ~ funded | school + year, data = schools))[["funded"]],
    mean(group_time$att),
    mean(group_time$att[group_time$g == 4]),
    mean(group_time$att[group_time$g == 6]),
    mean(group_time$att[group_time$g == 8])),
  Truth = c(NA, truth[["overall"]], truth[["overall"]],
            truth[["wave1"]], truth[["wave2"]], truth[["wave3"]])) %>%
  mutate(Error = Estimate - Truth) %>%
  kable(digits = 2, row.names = FALSE,
        caption = "What each estimator recovers, against the effect that was written into the data. The before-and-after row has no target because it estimates nothing in particular. Two-way fixed effects misses a three-point effect entirely; the group-time estimator finds it, and finds the differences between waves as well.")
```

Two-way fixed effects reports $-0.01$ against a true $+3.01$. The group-time estimator reports $+2.91$, and recovers the three waves separately at $5.25$, $1.83$ and $0.21$ against true values of $5.50$, $1.80$ and $0.24$.

```{r did-gt-fig, echo=FALSE, fig.align='center', fig.width=7, fig.height=4.2, fig.cap="Group-time treatment effects: one estimate for every funding wave in every year it was funded, with the truth as hollow points and the two-way fixed effects estimate as a single flat line. The line is a weighted average of everything above it and lands on zero - which is what averaging becomes when some of the weight sits on comparisons that subtract one wave's ongoing progress from another's."}
group_time %>%
  mutate(truth = mapply(function(g, t)
           mean(schools$effect[schools$cohort == g & schools$year == t]), g, t),
         wave = factor(g, c(4, 6, 8),
                       c("wave 1 (year 4)", "wave 2 (year 6)",
                         "wave 3 (year 8)"))) %>%
  ggplot(aes(t, att, colour = wave)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_hline(yintercept = coef(feols(score ~ funded | school + year,
                                     data = schools))[["funded"]],
             linetype = "dashed", colour = "#5a5a5a") +
  geom_line(linewidth = .7) +
  geom_point(size = 1.8) +
  geom_point(aes(y = truth), shape = 21, fill = NA, size = 2.6) +
  annotate("text", x = 11.4, y = 0.75, label = "two-way fixed effects",
           size = 3, colour = "#5a5a5a", hjust = 1) +
  scale_x_continuous(breaks = 4:12) +
  scale_colour_manual(values = c("#9b3f2d", "#c8925f", "#2c6e8e")) +
  labs(x = "Year", y = "Effect on mean score", colour = NULL,
       title = "Every wave gained. The standard estimator says none did.") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn: five variations on one generator</h2>
:::
This is the section's longer exercise, and it is worth an hour. Everything runs
on `simulate_schools()`, which is already defined, and on `feols()`, which is
already loaded. The point is not to learn the syntax; it is to find out *when*
the broken estimator is broken and when it is not.

Define one helper and reuse it:

```{r did-exercise-helper, eval=FALSE, echo=TRUE}
twfe <- function(d) coef(feols(score ~ funded | school + year, data = d))[["funded"]]
truth <- function(d) mean(d$effect[d$funded == 1])
```

**Variation 1 — one wave only.** Keep the never-funded schools and wave 1, drop
the other two. Run `twfe()`. Is it still wrong?

The estimate comes out at about `r fitb(5.4, tol = 0.3)` against a truth of 5.5,
so with one wave two-way fixed effects is
`r mcq(c("still badly biased", answer = "essentially right", "impossible to compute"))`.

**Variation 2 — a constant effect.** Modify the generator so that the effect is a
flat 3 points from the first funded year onwards, instead of growing with
exposure, and rerun on all three waves.

The estimate is about `r fitb(2.9, tol = 0.25)` against a truth of 3.0. So the
staggering is not the problem on its own; the problem is
`r mcq(c("staggering", "the number of waves", answer = "staggering combined with effects that change with duration"))`.

**Variation 3 — starve it of controls.** Set `n_never = 0`, so that the only
available comparisons are between waves.

The estimate moves to about `r fitb(-1.5, tol = 0.4)` — *further* from the truth
than the baseline $-0.006$, because now every comparison is a forbidden one.

**Variation 4 — feed it.** Set `n_never` to 300 and then 900 and watch. The
estimates go to roughly 1.3 and 1.8. Does two-way fixed effects ever reach the
truth of 3.01?
`r mcq(c("yes, at n_never = 900", answer = "no - it converges towards a variance-weighted blend that is not the quantity you asked for, however many controls you add", "yes, in the limit"))`

**Variation 5 — write the sentence.** In two lines, and without using the word
"bias", say what two-way fixed effects estimates in this data set. Then check it
against the takeaway below.

`r hide("The code")`

```{r did-exercise-solution, eval=FALSE, echo=TRUE}
# 1 -- one wave
one_wave <- schools %>% filter(cohort %in% c(0, 4))
c(twfe = twfe(one_wave), truth = truth(one_wave))

# 2 -- a constant effect: edit the generator's `effect` line to
#      d$effect <- 3 * (d$exposure > 0)
#      and rerun on all three waves.

# 3 and 4 -- vary the size of the never-funded group
purrr::map_dfr(c(0, 25, 75, 300, 900), function(nn) {
  d <- simulate_schools(n_never = nn)
  tibble(n_never = nn, twfe = twfe(d), truth = truth(d))
})
```

The five rows come out at $-1.46$, $-0.55$, $-0.01$, $+1.28$, $+1.81$ against a
truth of 3.01 throughout. The never-funded schools are the only clean controls in
the data set, and the more weight they carry the closer the blend gets — but the
forbidden comparisons never disappear, so the sequence approaches something that
is not the answer.

`r unhide()`
::::

::: {.analysis-takeaway}
The lesson is not that two-way fixed effects is broken. It is that it answers a question nobody asked: a variance-weighted blend of every available comparison, including comparisons that should never have been made. The modern estimators are not more sophisticated. They are more *literal* — they compute the thing you meant, one wave and one year at a time, and let you choose the average afterwards.

If you read one empirical paper published before about 2019 that uses staggered adoption, this is the first thing to check.
:::

### What Difference-in-Differences Cannot Do {#did-limits}

The design is cheap, transparent and widely applicable, which is precisely why it is worth listing what it does not survive.

**Anticipation.** If schools knew in 2023 that money was coming in 2024 and started hiring, the pre-period is already contaminated and the reference year is not a clean baseline. The usual response is to move the reference earlier and show that the result does not move with it.

**Spillovers.** Comparison schools in the same city may hire the teachers that funded schools trained, or lose pupils to them. SUTVA fails, the comparison group is treated, and the estimate shrinks towards zero — or grows, if the spillover is negative.

**Composition change.** This one is specific to the Startchancen design and is easy to miss. The evaluation compares repeated cross-sections: year 4 in 2026, a *different* year 4 in 2029. If the programme changes who attends those schools — and a visibly improving school in a deprived area very plausibly does — then the two cohorts are not drawn from the same population, and part of the measured gain is a change in the intake rather than in the teaching. Panel data on the same pupils would not have this problem, and would have attrition instead.

**Scale.** Parallel in levels is not parallel in logs, as noted above. Report which one you assumed and why.

**A comparison group at all.** When a policy hits everybody at once there is nothing to difference against. Section \@ref(experiments) named the usual escape — build a comparison group out of weighted others, which is the synthetic control method — and noted that it buys identification with a stronger assumption than anything here.

:::: {.dedicated}
::: {.titeldedicated}
<h2>Truly Dedicated: what happens when parallel trends only holds conditionally</h2>
:::
A frequent situation: unconditionally the two groups were on different tracks, but *given covariates* — school size, share of pupils from low-income households, prior achievement — the trends look parallel. That is a weaker and often more defensible assumption, and it is where this section shakes hands with the last one.

Alberto Abadie's semiparametric estimator does exactly this: weight the comparison units by their propensity score so that their covariate distribution matches the treated group's, then take the difference in differences of the reweighted data.[^did-abadie] The propensity score is the same object as in Section \@ref(matching-and-balancing), and it earns its place for the same reason — one number standing in for many covariates.

In `did`, it is an argument: `att_gt(..., xformla = ~ prior_score + size)`, with `est_method` choosing between outcome regression, inverse probability weighting and the doubly robust combination of the two. Callaway and Sant'Anna's default is the doubly robust one, for the reason Section \@ref(matching-and-balancing) gave: two chances to be right.

The catch is the same as always. Conditioning on covariates buys conditional parallel trends, and conditional parallel trends is still untestable. It replaces one assumption with a slightly weaker one; it does not replace assumptions with evidence.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2>Your Turn</h2>
:::
Funded schools went from 46.46 to 47.96 and never-funded schools from 49.03 to 48.73.

The difference in differences is `r fitb(1.8, tol = 0.05)`.

A before-and-after study of the funded schools alone would have reported a gain of 1.50 rather than 1.80. Why is the true answer *larger* than the naive one here? `r mcq(c("because the funded schools were selected", answer = "because scores were falling everywhere, so standing still was already an improvement", "because the sample is too small"))`

An event study shows two pre-period coefficients that are not statistically significant. What does that establish? `r mcq(c("that parallel trends holds", "that the treatment had no anticipation effect", answer = "very little - the test is underpowered against exactly the drifts that matter"))`

Three waves of schools are funded in different years and the effect grows with exposure. What does two-way fixed effects do wrong? `r mcq(c("it clusters the standard errors at the wrong level", answer = "it uses already-funded schools as controls for later waves, subtracting their ongoing gains", "it drops the never-funded schools"))`

True or false: if the treatment effect were identical for every unit and every duration, two-way fixed effects would be fine. `r torf(TRUE)`

And the coding trap, which is subtler than it looks. `fixest::sunab()` treats any cohort value that does not appear among the periods as never-treated. Which of these is the dangerous case? `r mcq(c("never-treated units coded 0 in a panel running from 1 to 12", "never-treated units coded 1000", answer = "units treated before the panel begins, coded with a year that falls outside it"))`
::::

:::: {.reading}
::: {.titelreading}
<h2>Reading</h2>
:::
One paper covers this whole section and is written for users rather than econometricians: Jonathan Roth, Pedro Sant'Anna, Alyssa Bilinski and John Poe's synthesis of the recent literature, which ends with a practical checklist.[^did-synthesis] For the design at its most persuasive, read Card and Krueger's original minimum-wage study and then Dustmann and co-authors on the German minimum wage, which reaches a different conclusion about employment and a more interesting one about where workers ended up [@cunningham2021causal].[^did-dustmann]
::::

::: {.analysis-takeaway}
Difference-in-differences buys comparability out of timing. It tolerates any amount of pre-existing difference in levels and asks only that the two groups would have moved together — an assumption about a counterfactual, which the pre-period can make plausible but cannot establish, and which a pre-trend test checks with much less power than its p-value suggests.

Where treatment arrives in waves, the estimator matters as much as the design: the standard two-way fixed effects regression silently uses already-treated units as controls, and this section's simulated programme shows it converting a three-point gain into a precisely estimated zero. Estimate wave by wave and year by year, then choose the average yourself.

Section \@ref(discontinuities) narrows the comparison differently again: instead of a date, a threshold — and the same Startchancen programme supplies one, because the schools were selected by a published social index with a line drawn through it.
:::

### Readings {#did-readings -}

- Roth, J., Sant'Anna, P. H. C., Bilinski, A., & Poe, J. (2023). What's Trending in Difference-in-Differences? A Synthesis of the Recent Econometrics Literature. *Journal of Econometrics* 235(2), 2218-2244. The one-stop survey, with recommendations.
- Angrist, J. D., & Pischke, J.-S. (2009). *Mostly Harmless Econometrics*, chapter 5. The classical treatment, before the staggered-adoption literature.
- Cunningham, S. (2021). *Causal Inference: The Mixtape*, chapter 9 [@cunningham2021causal]. Free online, with the Bacon decomposition worked through by hand.
- Huntington-Klein, N. (2022). *The Effect: An Introduction to Research Design and Causality*. Chapman & Hall/CRC, chapters 17 and 18 — event studies and difference-in-differences. The most patient explanation of why the forbidden comparison is forbidden.
- Callaway, B., & Sant'Anna, P. H. C. `did` package vignettes. The reference implementation, written by the authors of the method.

[^did-startchancen]: Bund and Länder announced their agreement on 2 February 2024 and signed the *Bund-Länder-Vereinbarung* on 4 June 2024; the programme started on 1 August 2024 and runs to 2034. The federal government contributes up to €1 billion a year and the Länder match it, giving €20 billion over ten years for around 4,000 schools and well over a million pupils. The three pillars are an investment programme for learning environments, a discretionary *Chancenbudget* for school development, and staff for multi-professional teams. Note that the widely quoted 40 / 40 / 20 key — child poverty, migration background, and inverse economic strength — governs how the federal investment funds are distributed *across the sixteen Länder*, not which schools are selected; individual schools are chosen by each Land using its own social index, and those indices differ. <https://www.bmbfsfj.bund.de/>

[^did-iqb]: Shares failing the minimum standard for Germany as a whole, from the IQB *Bildungstrend* reports: reading and mathematics at the end of year 4 in 2021, reading at the end of year 9 in 2022, mathematics at the end of year 9 in 2024. Stanat, P., et al. (Eds.), *IQB-Bildungstrend*, Waxmann. The year-9 reports publish two thresholds — the minimum standard for the *erster Schulabschluss* and the one for the *mittlerer Schulabschluss* — and a figure quoted without saying which is meaningless. It is worth knowing that the programme's own target is stated for its own schools rather than for the national figure, so these numbers are the backdrop, not the baseline.

[^did-infas]: The evaluation was commissioned by the federal education ministry after an EU-wide tender and is led by the infas Institut für angewandte Sozialwissenschaft, with the Hector Institute (Tübingen), LIfBi, IEA Hamburg, the chair of empirical economics at Potsdam, Evaluation Office Caliendo & Partner and the BiB, with the IQB advising; the contract is worth close to €50 million. The published design is a repeated cross-section of year 4 and year 9 pupils at 300 randomly selected programme schools, with a baseline in early 2026 and a repeat in early 2029, using IQB *Bildungstrend* test items so that comparison values come from the national assessments; monitoring surveys go to every programme school roughly every eighteen months, and the project runs to 2030. Note what the public documentation does *not* say: it names no identification strategy at all. Reading the design as a difference-in-differences, with a threshold design available alongside it, is this book's inference from the data being collected — not a claim the consortium has made. A pleasing detail for readers of Section \@ref(matching-and-balancing): the Potsdam chair and the evaluation office belong to Marco Caliendo, whose implementation guide is the standard reference for propensity-score matching cited there. <https://evaluation-startchancen.de>

[^did-snow]: Snow, J., *On the Mode of Communication of Cholera*, 2nd edition, John Churchill, 1855. See the box in Section \@ref(experiments); the Lambeth company moved its intake upstream between the 1849 and 1854 epidemics while its competitor did not, which is a difference in differences in everything but name. Coleman's re-analysis is worth knowing about: once over-dispersion is allowed for, the effect is only modestly significant, which is a fair reminder that a famous design and a decisive test are two different things.

[^did-cardkrueger]: Card, D., & Krueger, A. B. (1994). Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania. *American Economic Review* 84(4), 772-793. New Jersey raised its minimum wage from $4.25 to $5.05 on 1 April 1992; the study surveyed 410 restaurants on both sides of the border. Full-time-equivalent employment per store went from 20.44 to 21.03 in New Jersey and from 23.33 to 21.17 in Pennsylvania, a relative gain of 2.76 employees, or 13%, with a t-statistic of 2.03.

[^did-milog]: The *Mindestlohngesetz* introduced a national minimum wage of €8.50 an hour on 1 January 2015. Two careful evaluations reaching different conclusions: Caliendo, M., Fedorets, A., Preuss, M., Schröder, C., & Wittbrodt, L. (2018), *The short-run employment effects of the German minimum wage reform*, *Labour Economics* 53, 46-62, <https://doi.org/10.1016/j.labeco.2018.07.002>, which finds moderate employment losses concentrated in marginal employment; and Bossler, M., & Gerner, H.-D. (2020), *Employment Effects of the New German Minimum Wage: Evidence from Establishment-Level Microdata*, *ILR Review* 73(5), 1070-1094, <https://doi.org/10.1177/0019793919889635>.

[^did-greenrev]: Gollin, D., Hansen, C. W., & Wingender, A. M. (2021). Two Blades of Grass: The Impact of the Green Revolution. *Journal of Political Economy* 129(8), 2344-2384, <https://doi.org/10.1086/714444>. Identification comes from the staggered arrival of high-yielding varieties across countries and crops; the headline counterfactual is that a ten-year delay would have left GDP per capita in the developing world about 17% lower in 2010. Norman Borlaug received the Nobel Peace Prize in 1970 for the plant breeding behind it.

[^did-rollout]: 2,125 schools took part in the first school year, 2024/25; the target of roughly 4,000 was reached by 2026. A programme that arrives in waves is not an inconvenience for the evaluator — it is the situation Sections \@ref(did-staggered) to \@ref(did-modern) are about.

[^did-athey]: Athey, S., & Imbens, G. W. (2006). Identification and Inference in Nonlinear Difference-in-Differences Models. *Econometrica* 74(2), 431-497, <https://doi.org/10.1111/j.1468-0262.2006.00668.x>.

[^did-ashenfelter]: Ashenfelter, O. (1978). Estimating the Effect of Training Programs on Earnings. *Review of Economics and Statistics* 60(1), 47-57, JSTOR 1924332. The name "Ashenfelter's dip" was attached later by the evaluation literature, not by Ashenfelter.

[^did-roth]: Roth, J. (2022). Pretest with Caution: Event-Study Estimates after Testing for Parallel Trends. *American Economic Review: Insights* 4(3), 305-322, <https://doi.org/10.1257/aeri.20210236>.

[^did-rambachan]: Rambachan, A., & Roth, J. (2023). A More Credible Approach to Parallel Trends. *Review of Economic Studies* 90(5), 2555-2591, <https://doi.org/10.1093/restud/rdad018>. Implemented in the `HonestDiD` package, on CRAN, with a development version on GitHub.

[^did-bdm]: Bertrand, M., Duflo, E., & Mullainathan, S. (2004). How Much Should We Trust Differences-in-Differences Estimates? *Quarterly Journal of Economics* 119(1), 249-275, <https://doi.org/10.1162/003355304772839588>. The 45% is an upper bound across their specifications: "we find an 'effect' significant at the 5 percent level for up to 45 percent of the placebo interventions".

[^did-cgm]: Cameron, A. C., Gelbach, J. B., & Miller, D. L. (2008). Bootstrap-Based Improvements for Inference with Clustered Errors. *Review of Economics and Statistics* 90(3), 414-427, <https://doi.org/10.1162/rest.90.3.414>.

[^did-bacon]: Goodman-Bacon, A. (2021). Difference-in-differences with variation in treatment timing. *Journal of Econometrics* 225(2), 254-277, <https://doi.org/10.1016/j.jeconom.2021.03.014>. The decomposition is implemented in the `bacondecomp` package by Evan Flack and Edward Jee.

[^did-chaisemartin]: de Chaisemartin, C., & D'Haultfœuille, X. (2020). Two-Way Fixed Effects Estimators with Heterogeneous Treatment Effects. *American Economic Review* 110(9), 2964-2996, <https://doi.org/10.1257/aer.20181169>.

[^did-cs]: Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-Differences with multiple time periods. *Journal of Econometrics* 225(2), 200-230, <https://doi.org/10.1016/j.jeconom.2020.12.001>. The `did` package aggregates group-time effects with `type = "simple"`, `"dynamic"`, `"group"` or `"calendar"`.

[^did-sunab]: Sun, L., & Abraham, S. (2021). Estimating dynamic treatment effects in event studies with heterogeneous treatment effects. *Journal of Econometrics* 225(2), 175-199, <https://doi.org/10.1016/j.jeconom.2020.09.006>. Implemented as `sunab()` in `fixest` by Laurent Bergé. A third robust estimator, imputation-based, is Borusyak, K., Jaravel, X., & Spiess, J. (2024), *Revisiting Event-Study Designs: Robust and Efficient Estimation*, *Review of Economic Studies* 91(6), 3253-3285, <https://doi.org/10.1093/restud/rdae007>.

[^did-abadie]: Abadie, A. (2005). Semiparametric Difference-in-Differences Estimators. *Review of Economic Studies* 72(1), 1-19, <https://doi.org/10.1111/0034-6527.00321>.

[^did-synthesis]: Roth, J., Sant'Anna, P. H. C., Bilinski, A., & Poe, J. (2023). What's trending in difference-in-differences? A synthesis of the recent econometrics literature. *Journal of Econometrics* 235(2), 2218-2244, <https://doi.org/10.1016/j.jeconom.2023.03.008>.

[^did-dustmann]: Dustmann, C., Lindner, A., Schönberg, U., Umkehrer, M., & vom Berge, P. (2022). Reallocation Effects of the Minimum Wage. *Quarterly Journal of Economics* 137(1), 267-328, <https://doi.org/10.1093/qje/qjab028>. The reform raised wages without lowering employment, and moved low-wage workers towards larger and more productive firms - a reallocation that accounts for up to 17% of the wage gain.
