## Common Tests Are Linear Models {#tests-as-models}

::: {.analysis-subtitle}
One framework behind the whole toolbox
:::

<small>

> "Teaching statistics as a zoo of named tests is like teaching zoology as a list of animal names."
>
> `r tufte::quote_footer('The complaint that motivated this section')`

</small>

Look back at what the Compare chapter asked you to learn. A one-sample t-test. A two-sample t-test, in a Student and a Welch flavour. A paired t-test. A correlation test. A one-way ANOVA. An ANCOVA. A Wilcoxon test, a Mann-Whitney test, a Kruskal-Wallis test. Nine names, nine sets of assumptions, nine entries in a decision tree that students memorise for an exam and forget by Christmas.

Now that you know what a linear model is, the honest summary can finally be given: **there are not nine tests. There is one model, fitted with different right-hand sides.** Every test in that list is `lm()` wearing a costume.

This observation is not new, but its clearest modern statement is a cheat sheet by Jonas Kristoffer Lindeløv, *Common statistical tests are linear models*, which lines up each classical test against its equivalent `lm()` formula.[^tm-lindeloev] This section works through that correspondence on our own data, so that you can check every claim rather than take it on faith.

### The one model {#tm-one-model}

Recall the notation. In R, `y ~ 1` means "predict `y` with an intercept and nothing else", and the intercept of a model with no predictors is simply the mean of `y`. Everything else is built by adding terms to the right of the tilde:

| Right-hand side | What it says | Classical name |
|:---|:---|:---|
| `y ~ 1` | one number predicts `y` | one-sample t-test |
| `(y2 - y1) ~ 1` | one number predicts the within-unit change | paired t-test |
| `y ~ 1 + x` | intercept plus a slope on continuous `x` | correlation / simple regression |
| `y ~ 1 + G` | one intercept per group (2 groups) | two-sample t-test |
| `y ~ 1 + G` | one intercept per group (3+ groups) | one-way ANOVA |
| `y ~ 1 + G + x` | group intercepts plus a slope | ANCOVA |
| `y ~ 1 + G * S` | group effects that differ by a second factor | two-way ANOVA |
| `rank(y) ~ ...` | the same models fitted to ranks | the non-parametric family |

Read the fourth and fifth rows together. The two-sample t-test and the one-way ANOVA are *the same formula*; the only difference is how many levels the grouping variable happens to have. The wall between "t-test" and "ANOVA" that every introductory course erects is a wall between two and three, and there is nothing there.

::: {.analysis-note}
One caveat before the demonstrations. The equivalences below are **exact** for the parametric tests and **approximate** for the rank-based ones — good approximations, and better the larger the sample, but approximations. Where the match is exact, R will agree with itself to fifteen decimal places. Where it is approximate, the p-values will be close rather than identical, and we say so.
:::

```{r tm-load, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

coursedata <- read.csv("data/Course/GF_AllTime.csv", sep = ";") %>%
  filter(Age < 100)
```

### One mean is an intercept {#tm-one-sample}

Section \@ref(cm-one-sample) tested whether students' self-rated R background differs from the scale midpoint of 3, and found $t = -20.70$. Here is the same question asked of a regression with no predictors at all. Subtract the benchmark from the outcome, then let the intercept do the work:

```{r tm-one-sample, echo=TRUE}
t.test(coursedata$Background.in.R, mu = 3)$statistic

lm(I(Background.in.R - 3) ~ 1, data = coursedata) %>% summary() %>% coef()
```

Identical to four decimal places, because they are the same computation. The intercept of `y ~ 1` is the mean; its standard error is $sd(y)/\sqrt{n}$; the ratio is Gosset's $t$. What the t-test calls "testing against $\mu_0$" the regression calls "is the intercept zero", and every regression table you will ever read tests exactly that for every coefficient it prints.

### A paired test is a one-sample test on differences {#tm-paired}

Section \@ref(cm-paired) already made this point without the vocabulary: pairing is subtraction. Build the differences and fit the emptiest model there is.

```{r tm-paired, echo=TRUE}
differences <- coursedata$Background.in.Statistics - coursedata$Background.in.R

t.test(coursedata$Background.in.Statistics,
       coursedata$Background.in.R, paired = TRUE)$statistic

lm(differences ~ 1) %>% summary() %>% coef()
```

Both give $t = 14.10$. The paired t-test was never a separate test — it is `y ~ 1` applied to a column you constructed.

### Two groups is a dummy variable {#tm-two-sample}

This is the equivalence that section \@ref(comparing-means) promised and deferred. Male students in this course are older than female students. The classical test and the regression:

```{r tm-two-sample, echo=TRUE}
t.test(Age ~ Gender, data = coursedata, var.equal = TRUE)$statistic

lm(Age ~ Gender, data = coursedata) %>% summary() %>% coef() %>% round(3)
```

Both report $|t| = 2.33$; the sign differs only because R's dummy coding takes `Female` as the reference and measures the *male* effect, whereas the t-test subtracts in the other order.

Read the coefficients as the chapter taught you to. The intercept, 24.500, is the mean age of the reference group — the women. The slope on `GenderMale`, 1.122, is *not* the mean age of men but the **difference** between the groups: men average $24.500 + 1.122 = 25.622$ years. A dummy variable turns a group label into a number, and once it is a number the regression machinery does not care that it only takes two values.

::: {.analysis-warning}
Note `var.equal = TRUE`. Ordinary least squares assumes one common error variance across the whole model, so it reproduces **Student's** t-test, not Welch's. If the groups genuinely differ in spread, the regression inherits the same problem the classical test had, and the fix is the same: model the variances, or use robust standard errors.

This is worth noticing rather than glossing over. "Everything is a linear model" does not mean everything is *safe* — it means the assumptions you were taught to check per test are, underneath, the same three assumptions checked once.
:::

### Correlation is a slope {#tm-correlation}

Section \@ref(relationships) tested whether age and total semesters move together. A correlation test and a regression of one on the other:

```{r tm-correlation, echo=TRUE}
cor.test(coursedata$Age, coursedata$Total.Semesters)$statistic

lm(Total.Semesters ~ Age, data = coursedata) %>% summary() %>% coef() %>% round(4)
```

Again identical: $t = 1.7264$, $p = 0.086$, a positive but unconvincing association. The correlation coefficient and the regression slope are different numbers — $r = 0.113$ is unit-free, the slope is in semesters per year — but they are the same *evidence*, because the slope is just the correlation rescaled by the two standard deviations. Testing "is $r$ zero" and "is the slope zero" cannot give different answers.

### Many groups is many dummies {#tm-anova}

The step from two groups to seven changes nothing structurally. Our students arrive across seven terms; do the cohorts differ in age?

```{r tm-anova, echo=TRUE}
summary(aov(Age ~ Term, data = coursedata))[[1]][1, 4]   # F from aov()

summary(lm(Age ~ Term, data = coursedata))$fstatistic[1] # F from lm()
```

$F = 2.15$ on both sides, $p = 0.049$ — a whisker under the conventional threshold, which is a fine reminder of the last box in section \@ref(cm-threshold). Behind the scenes `lm()` created six dummy variables for the seven terms, and the F-statistic asks whether all six slopes are zero at once. That is precisely the question ANOVA's F-test asks. The two functions print different tables because they were written for different audiences, not because they compute different things.

Adding a continuous variable to the same formula turns it into ANCOVA, which is why section \@ref(partitioning-variation) could hand the dog-size effects to a height covariate without changing tools. `aov(y ~ G + x)` and `lm(y ~ G + x)` are one model.

### The non-parametric family is the same models on ranks {#tm-ranks}

The tests recommended when data look non-normal — Mann-Whitney, Kruskal-Wallis, Wilcoxon — are usually presented as a separate methodology with its own decision tree. They are, to a very good approximation, the same linear models fitted to **ranks** instead of raw values.

```{r tm-ranks, echo=TRUE}
wilcox.test(Age ~ Gender, data = coursedata)$p.value            # Mann-Whitney

coef(summary(lm(rank(Age) ~ Gender, data = coursedata)))[2, 4]  # lm on ranks
```

```{r tm-ranks-table, echo=FALSE}
kw  <- kruskal.test(Age ~ Term, data = coursedata)$p.value
kwl <- anova(lm(rank(Age) ~ Term, data = coursedata))[1, 5]
mw  <- wilcox.test(Age ~ Gender, data = coursedata)$p.value
mwl <- coef(summary(lm(rank(Age) ~ Gender, data = coursedata)))[2, 4]

data.frame(
  `Classical test` = c("Mann-Whitney U (age by gender)", "Kruskal-Wallis (age by term)"),
  `Linear model on ranks` = c("lm(rank(Age) ~ Gender)", "lm(rank(Age) ~ Term)"),
  `p, classical` = round(c(mw, kw), 4),
  `p, linear model` = round(c(mwl, kwl), 4),
  check.names = FALSE) %>%
  knitr::kable(align = "llrr",
               caption = "Rank-based tests and their linear-model equivalents. Close, not identical - these are approximations that improve with sample size.")
```

Two lessons sit in that table. First, the approximation is good: 0.0043 against 0.0041, 0.0238 against 0.0219. Lindeløv's cheat sheet marks these as reliable for samples above roughly $n = 11$, and ours is twenty times that.

Second, and more interesting: the rank test on age by gender gives $p = 0.004$ where the ordinary t-test in section \@ref(cm-two-sample) gave $p = 0.021$. The rank test is the *more* convincing one here. Ranking pulled in the long right tail of mature students — the 42-year-old and the handful in their thirties — which had been inflating both group variances. When a rank test and a t-test disagree, the disagreement is information about the shape of your data, and worth a histogram before it is worth a decision.

::: {.analysis-takeaway}
The signed-rank family is where the approximation is loosest, and Lindeløv is explicit that the Wilcoxon signed-rank equivalence via `lm(signed_rank(y) ~ 1)` should be treated as an approximation for $n > 14$ rather than an identity. Use it to understand what the test is doing, not to replace the test in a published analysis.
:::

### Why this matters {#tm-why}

Three reasons, in rising order of importance.

**It halves what you have to remember.** A decision tree with nine leaves becomes one function and a formula syntax. When you meet a design this book never covered — three groups measured twice, a covariate plus an interaction, an ordinal predictor — you do not go looking for the name of the right test. You write the model.

**It makes the assumptions visible.** Every test in the list above inherits the same handful of assumptions from its parent model: the errors are independent, roughly constant in variance, roughly normal. Presented as nine separate tests, those become twenty-seven assumptions to memorise. Presented as one model, they are three things to check, and `plot(model)` checks them.

**It moves the conversation from testing to estimating.** This is the real prize. A t-test hands you a verdict; a linear model hands you a coefficient, a standard error and a confidence interval — how big, in what units, how precisely known. Section \@ref(cm-threshold) argued that significance is a threshold on evidence rather than a statement about importance. The linear-model framing makes that argument structural: once every test is a model, the *estimate* is what you naturally report, and the p-value becomes one column of the table rather than the point of it.

:::: {.dedicated}
::: {.titeldedicated}
<h2> Truly Dedicated: What the framework does not swallow </h2>
:::
Honesty requires the boundaries. "Everything is a linear model" is a claim about the tests in an introductory syllabus, not about all of statistics.

The **generalized** linear models of section \@ref(generalized-models) extend the idea rather than fit inside it: logistic and Poisson regression keep the linear predictor but attach a link function and a non-normal error distribution. The chi-square test of independence belongs there, as a log-linear model, which is why Lindeløv's sheet lists it under `glm(..., family = poisson)` rather than `lm()`.

**Exact** tests for tiny samples — Fisher's exact test, the permutation logic behind the lady tasting tea in section \@ref(partitioning-variation) — are not approximations of a linear model at all; they enumerate possibilities. And multilevel models, survival models, and the structural equation models of chapter \@ref(reveal) each add machinery that `lm()` does not have.

The right summary is therefore narrower than the slogan and more useful: *the classical tests you were taught as a list are special cases of one model, and the models that are not special cases are generalisations of the same idea.* Either way there is one framework, not a zoo.
::::

:::: {.challenge}
::: {.titelchallenge}
<h2> Your Turn </h2>
:::
1. You want to compare mean income across four regions. Which call does the job? `r mcq(c("aov(income ~ region) - lm() only works for two groups", answer = "lm(income ~ region) - the number of levels does not matter", "lm(income ~ 1) with the regions filtered one at a time"))`

2. `lm(y ~ G)` where `G` has exactly two levels is equivalent to which classical test? `r mcq(c("One-sample t-test", answer = "Two-sample t-test with var.equal = TRUE", "Welch's t-test", "Paired t-test"))`

3. In `lm(Age ~ Gender)` the intercept was 24.500 and the slope 1.122. What is the mean age of men? `r fitb(25.622, num = TRUE, tol = 0.01)`

4. `r torf(FALSE)` Because every test is a linear model, the assumptions of the linear model do not need to be checked for t-tests.

5. A colleague reports "we used a Mann-Whitney test because the data were skewed". Explain in one sentence what model they fitted, in the language of this section.
::::

[^tm-lindeloev]: Lindeløv, J. K., *Common statistical tests are linear models (or: how to teach stats)*, 2019, <https://lindeloev.github.io/tests-as-linear/>. The accompanying one-page cheat sheet tabulates each test, its built-in R function, the equivalent `lm()` call, and whether the correspondence is exact or approximate. The idea itself is older — it is the standard view in the general-linear-model literature — but Lindeløv's presentation is what made it teachable.
