# First Difference

## Application: Mincer earnings function

## Data Management (4 pt)

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
# Load data
library(haven)
library(tidyverse)
library(magrittr)

# Read
pl <- readRDS("C:/Users/Marco2020/Desktop/test/pl_reduced.rds")
pequiv <- readRDS("C:/Users/Marco2020/Desktop/test/pequiv_reduced.rds")

# Merge data
master <- merge(pl, pequiv, by = c("pid", "syear"))
```

All negative values represent missing values. Do a complete case analysis (n should be 338892). Drop all Stata labels with `remove_all_labels()` from `sjlabelled` package. 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
# Drop missings
master[master < 0] <- NA
master <- master[complete.cases(master), ]

# Drop labels 
library(sjlabelled)
master <- remove_all_labels(master)
```

Calculate each information in `master` per individual (group by `id`). *Rename* `plh0182` to `satis`, `d11109` to `years_educ` and `d11101` to `age`.

1. **Hourly wages** can be calculated as total annual income (`ijob1`) divided by total annual work hours (`e11101`). 
    + You may store log of hourly wage as a separate column/variable `log_inc_hourly`. 
    + Please filter for annual work hours larger than 0 (otherwise results might be `Inf`), before dividing. 
    + Please filter for hourly wages larger than 1, before taking the log.
2. **Years of schooling** are readily available in SOEP.
3. **Tenure** is not available in SOEP.
    + @reilich2011smoking suggests "Experience is the potential experience following Mincer. We calculate the age minus the years of schooling minus 6 and minus the number of years with the experience of being unemployed.".
    + Calculate `tenure` like this `tenure = age - years_educ - 6 - years_unemp`.
    + Calculate the number of years being unemployed `years_unemp` as the number of observations per individual minus the sum of all years that this individual was employed. `e11102` is the employment status (0 for unemployed, 1 for employed) in a particular year. 
    + Some people exhibit negative tenure for various reasons. Filter for positive values.

Store the prepared data as `mincer`. Summary statistics should roughly be equal to the following table:

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
# Data stuff
mincer <- master %>%
  rename(satis = plh0182,
         years_educ = d11109,
         age = d11101) %>%
  group_by(pid) %>%
  filter(e11101 > 0) %>%
  mutate(inc_hourly = ijob1/e11101) %>%
  filter(inc_hourly > 1) %>%
  mutate(log_inc_hourly = log(inc_hourly)) %>%
  mutate(years_obs = n()) %>%
  mutate(years_emp = sum(e11102)) %>%
  mutate(years_unemp = n() - sum(e11102)) %>%
  mutate(tenure = age - years_educ - 6 - years_unemp) %>%
  filter(tenure > 0) %>%
  arrange(pid)
  #select(c("pid", "syear", "years_educ", "tenure", "inc_hourly", "log_inc_hourly"))
```

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
mincer <- as.data.frame(mincer)
library(stargazer)
stargazer(mincer[,c(3:5)], type="text")
```

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
library(gtsummary)

mincer %>%
  select("years_educ", "tenure", "inc_hourly") %>%
  tbl_summary() %>%
  bold_labels() %>%
  add_n()
```

## Data Visualization (2 pt)

The graph shows the hourly wage (not logged) in relation to the work experience for 5 random people that are observed for at least 5 years. You can use the `sample_n_groups()` function from previous exercises. 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
sample_n_groups = function(grouped_df, size, replace = FALSE, weight=NULL) {
  grp_var <- grouped_df %>% 
    groups %>%
    unlist %>% 
    as.character
  random_grp <- grouped_df %>% 
    summarise() %>% 
    sample_n(size, replace, weight) %>% 
    mutate(unique_id = 1:NROW(.))
  grouped_df %>% 
    right_join(random_grp, by=grp_var) %>% 
    group_by_(grp_var) 
}
```

```{r warning=FALSE, message=FALSE, echo=TRUE}
library(ggplot2)
set.seed(44)

mincer %>% 
  group_by(pid) %>%
  nest() %>% 
  filter(purrr::map_lgl(data, ~ nrow(.x) >= 5)) %>%
  unnest() %>%
  sample_n_groups(5) %>% # select 3 pid for all time
  ggplot(aes(x = tenure, y = inc_hourly, col=factor(pid))) +  
  geom_line() + geom_point() + 
  labs(title="Income and Work Experience", 
       subtitle="Selected Sample of 5 People",
       x="Work Experience (Years)", y="Hourly Wage (Euro)", 
       colour="person") + # give some illustrative labels
       theme(legend.position = "none")
```

## Data Modeling (2 pt)

We select a subset of data for the years 2017 and 2018 (T=2). We calculate a pooling model, first difference and fixed effects. We use the `plm` package for all model setups. Estimate models without education and without the constant:

$$\ln w_{it} = \beta_1 \text{tenure}_{it} + \beta_2 \text{tenure}^2_{it}  $$

```{r warning=FALSE, message=FALSE, echo=TRUE}
mincer1718 <- mincer %>%
  filter(syear == 2017 | syear == 2018)

# OLS 
library(plm)
LM2 <- plm(log_inc_hourly ~ tenure + I(tenure^2) -1, data=mincer1718, 
          model="pooling", 
          index = c("pid","syear"))

# Fixed Effects
FE2 <- plm(log_inc_hourly ~ tenure + I(tenure^2) -1, data=mincer1718, 
          model="within", 
          index = c("pid","syear"))

# Fixed Effects
FD2 <- plm(log_inc_hourly ~ tenure + I(tenure^2) -1, data=mincer1718, 
          model="fd", 
          index = c("pid","syear"))
```

Create a **stargazer** table for model comparison, such as:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# Stargazer Comparison
library(stargazer)
stargazer(LM2, FD2, FE2, type="text",
          column.labels=c("Pooling", "FD", "FE"),
          #keep=c("inc"),
          dep.var.labels.include = FALSE, model.numbers = FALSE,
          omit.stat="f")
```

## Minimal Example 

Please create a **minimal working example** to illustrate the differences between FD and FE. How minimal can it be? 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
MWE <- data.frame(id = c(1,1,1,1,1,1,1,2,2,2,2,2,2,2),
                  time = seq(from=1, to=7, by=1),
                  married = c(0,0,1,1,0,1,1, 0,0,1,1,0,1,1),
                  happy = c(5,5,8,7,4,7,5, 7,7,9,8,5,8,6))
                  #married = c(0,0,1,1,0,1,1, 1,1,0,0,0,1,0),
                  #happy = c(6,5,8,7,4,7,5, 7,7,3,3,5,5,6))
                  #married = round(runif(14, min = 0, max = 1),0),
                  #happy = round(runif(14, min = 4, max = 10),0))
```

1. Create a dataset `MWE` for two people (`id` 1 and 2) with 7 observations (`time` periods from 1 to 7). 
2. Create two variables. A dummy called `married` (regressor) and a `happiness` outcome variable (natural numbers ranging from 0 to 10). You may  create the values of `married` and `happy` manually. 
3. Create a line plot for happiness over time colored by each individual. (In total, you should have two lines of different color.)
4. Mark the marriage status by any reasonable label (there are several options in `ggplot`). 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
library(ggplot2)
ggplot(MWE, aes(x = time, y = happy, col=factor(id))) +  
  geom_line(position = position_jitter(w = 0.2, h = 0.2)) + 
  geom_point(aes(shape=factor(married)), size=4, position = position_jitter(w = 0.2, h = 0.2)) + 
  labs(title="Marriage and Happiness", subtitle = "Jitter added to confuse the reader.",
       x="Time (Years)", y="Happiness", 
       colour="Person", shape="Marriage")
```

Calculate a FE and FD estimator for how the marriage dummy affects happiness. Play around with different values and see how the FE and FD estimators changes. Select values such that FD is reasonable higher than FE. The two models look like this:

```{r warning=FALSE, message=FALSE, echo=TRUE}
# Fixed Effects
FE_marry <- plm(happy ~ married -1, data=MWE, 
          model="within", 
          index = c("id","time"))

#FE_marry <- lm(happy ~ married + as.factor(id), data=MWE)

# First Difference
FD_marry <- plm(happy ~ married -1, data=MWE, 
          model="fd", 
          index = c("id","time"))

stargazer(FE_marry, FD_marry, type="text", 
          column.labels=c("FE", "FD"),
          keep=c("married"),
          dep.var.labels.include = FALSE, 
          model.numbers = FALSE)
```

From this table, it can be seen that the immediate impact getting married (2.83) is higher than the overall level difference of being married (1.75). 


