# Fixed effects

## Cross-sectional Data (2 pt)

Suppose you want to learn the **effect of price on the demand** for back massages. Read in the following data from four Midwest locations (call it `crosssection`) and print the table with the `gt` package. 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, fig.align='center'}
# Crosssection
crosssection <- data.frame(Location = c("Chicago", "Peoria", "Milwaukee", "Madison"),
                           Year = rep(2003, 4), 
                           Price = c(75, 50, 60, 55),
                           Quantity = c(2.0, 1.0, 1.5, 0.8))

library(ggplot2)
library(gt)
table_1 <-
  gt(data = crosssection) %>%
  tab_header(title = "Table 1: Cross-sectional Data") %>%
  cols_label(
    Location = md("**Location**"),
    Year = md("**Year**"),
    Price = md("**Price**"),
    Quantity = md("**Per Capita<br>Quantity**")
  )  

# Show the gt Table
table_1
```

Please create the plot (in Base R or `ggplot`).

```{r warning=FALSE, message=FALSE, echo=FALSE, eval=TRUE, fig.align='center'}
# Plot relationship
plot(crosssection$Price, crosssection$Quantity, 
     main="Relationship between Price and Demand in 2003",
     xlab="Price", ylab="Quantity")
abline(lm(Quantity ~ Price, data=crosssection))
```

## Panel Data (2 pt)

Read in additional data from Table 2 (call it `paneldata`). 

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, fig.align='center'}
library(ggplot2)
library(tidyverse)

# Panel Data
paneldata <- data.frame(Location = c("Chicago", "Chicago", "Peoria", "Peoria", "Milwaukee", "Milwaukee", "Madison", "Madison"),
                           Year = rep(2003:2004, 4), 
                           Price = c(75, 85, 50, 48, 60, 65, 55, 60),
                           Quantity = c(2.0, 1.8, 1.0, 1.1, 1.5, 1.4, 0.8, 0.7))

table_2 <-
  gt(data = paneldata) %>%
  tab_header(title = "Table 2: Panel Data") %>%
  cols_label(
    Location = md("**Location**"),
    Year = md("**Year**"),
    Price = md("**Price**"),
    Quantity = md("**Per Capita<br>Quantity**")
  )  
table_2
```

Please create the plot.

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, fig.align='center'}
ggplot(paneldata, aes(x=Price, y=Quantity)) + 
  geom_point() + 
  labs(title="Relationship between Price and Quantity",
       x="Price", y="Quantity") +
  geom_smooth(aes(colour = Location), method = "lm", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE)
```

## Fixed Effects Regression

### First Difference (2 pt)

Please add two columns, i.e. the change in price (name it `P_diff`) and the change in quantity (name it `Q_diff`) to your dataframe (1 pt).

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, results='asis', fig.align='center'}
table3 <- paneldata %>%
  group_by(Location) %>%
  mutate(P_diff = Price - lag(Price),
         Q_diff = Quantity - lag(Quantity))

#library(xtable) # may handle formulas in colnames
#print(xtable(table3)), type = "html")
#print(xtable(head(cars)), type = "html")
#print(xtable(table3), type = "html")

library(gt)
gt(data = table3, rowname_col = TRUE, groupname_col=FALSE) %>%
  tab_header(title = "Table 3: Data for Difference Equation Estimation.") %>%
  cols_label(
    Location = md("**Location**"),
    Year = md("**Year**"),
    Price = md("**Price**"),
    Quantity = md("**Per Capita<br>Quantity**"),
    P_diff = md("**Delta P**"),
    Q_diff = md("**Delta Q**")) 
```


<!-- ```{=html} -->
<!-- <style type="text/css"> -->
<!-- .tg  {border-collapse:collapse;border-spacing:0;} -->
<!-- .tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px; -->
<!--   overflow:hidden;padding:10px 5px;word-break:normal;} -->
<!-- .tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px; -->
<!--   font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;} -->
<!-- .tg .tg-1wig{font-weight:bold;text-align:left;vertical-align:top} -->
<!-- .tg .tg-0lax{text-align:left;vertical-align:top} -->
<!-- </style> -->
<!-- ``` -->

<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Location  | Year | Price | Per Capita Quantity  | $$       | $$       | -->
<!-- |           |      |       |                      | \Delta P | \Delta Q | -->
<!-- |           |      |       |                      | $$       | $$       | -->
<!-- +:==========+======+=======+======================+==========+==========+ -->
<!-- | Chicago   | 2003 | 75    | 2.0                  |          |          | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Chicago   | 2004 | 85    | 1.8                  | 10       | -0.2     | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Milwaukee | 2003 | 60    | 1.5                  |          |          | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Milwaukee | 2004 | 65    | 1.4                  | 5        | -0.1     | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Madison   | 2003 | 55    | 0.8                  |          |          | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Madison   | 2004 | 60    | 0.7                  | 5        | -0.1     | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Peoria    | 2003 | 50    | 1.0                  |          |          | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->
<!-- | Peoria    | 2004 | 48    | 1.1                  | -2       | 0.1      | -->
<!-- +-----------+------+-------+----------------------+----------+----------+ -->

<!-- : Table 3: Data for Difference Equation Estimation. -->

Run a first difference estimation by regressing change in quantity on change in price (1 pt).

### Within Estimator (2 pt)

Run a within fixed effects regression in two ways. 

- Function `plm()` from `plm` package.
- Replicate time demeaning programmatic in R. 
    + Calculate the mean of Q and P per individual (group). 
    + Subtract the individual mean from each observation. 
    + Regress these time-demeaned observation (with standard lm command).

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, results='asis', fig.align='center'}
table4 <- paneldata %>%
  group_by(Location) %>%
  mutate(P_diff = Price - lag(Price),
         Q_diff = Quantity - lag(Quantity),
         P_P_mean = Price - mean(Price),
         Q_Q_mean = Quantity - mean(Quantity))

library(gt)
gt(data = table4, rowname_col = TRUE, groupname_col=FALSE) %>%
  tab_header(title = "Table 4: Data for Within Time Demean Estimation.") %>%
  cols_label(
    Location = md("**Location**"),
    Year = md("**Year**"),
    Price = md("**Price**"),
    Quantity = md("**Per Capita<br>Quantity**"),
    P_diff = md("**Delta P**"),
    Q_diff = md("**Delta Q**"),
    P_P_mean = md("**P - mean(P)**"),
    Q_Q_mean = md("**Q - mean(Q)**")) 
```

### LSDV (1 pt)

Run a LSDV regression, i.e. include all individuals as dummy variables in your standard `lm` regression.

### Comparison (1 pt)

Create a model overview with `stargazer`. Start with a pooling model. You should have 5 models in total. You need to play around with stargazer options in order to make it good looking. Drop the constant in all your models. Drop `model.numbers` and `dep.var.labels.include` or the F statistic. You can relabel `column.labels` in stargazer as well. Your result should look like this:

```{r warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE}
paneldata <- data.frame(Location = c("Chicago", "Chicago", "Peoria", "Peoria", 
                                     "Milwaukee", "Milwaukee", "Madison", "Madison"),
                        Year = rep(2003:2004, 4),
                        Price = c(75, 85, 50, 48, 60, 65, 55, 60),
                        Quantity = c(2.0, 1.8, 1.0, 1.1, 1.5, 1.4, 0.8, 0.7))

library(sjmisc)
library(tidyverse)

paneldata <- paneldata %>%
  group_by(Location) %>%
  mutate(dP = last(Price) - first(Price),
         dQ = last(Quantity) - first(Quantity),
         dP=replace(dP, row_number()==1, NA),
         dQ=replace(dQ, row_number()==1, NA),
         mean_P = mean(Price),
         mean_Q = mean(Quantity),
         P_mean_P = Price - mean_P,
         Q_mean_Q = Quantity - mean_Q,
         Price_ave = Price - ave(Price, Location),
         Quantity_ave = Quantity - ave(Quantity, Location)) %>%
  mutate_at(.vars = c("Quantity","Price"),
            .funs = list("centered" = center)) %>%
  ungroup()

# for better comparison in stargazer
paneldata_demeaned <- with(paneldata,
                           data.frame(Quantity = Quantity - ave(Quantity, Location),
                                      Price = Price - ave(Price, Location)))

# Pooling
pooling <- lm(Quantity ~ Price - 1, data=paneldata)

# First Difference Estimation
FD <- lm(dQ ~ dP -1, data=paneldata)
names(FD$coefficients) <- c('Price')

# Time demean
TDM <- lm(Quantity ~ Price -1, data=paneldata_demeaned)

# Fixed Effects
library(plm)
FE <- plm(Quantity ~ Price -1, data=paneldata,
          model="within",
          index = c("Location","Year"))

# Least Squares Dummy Variable Estimation (Individual FE)
LSDV <- lm(Quantity ~ Price + Location -1, data=paneldata)
```

```{r warning=FALSE, message=FALSE, echo=TRUE}
# Stargazer Comparison
library(stargazer)
stargazer(pooling, FD, TDM, FE, LSDV, type="text", keep=c("Price"),
          column.labels=c("Pooling", "FD", "TD", "FE", "LSDV"),
          dep.var.labels.include = FALSE, model.numbers = FALSE,
          omit.stat="f")
```


